### 2022-06-21 paper/perfmod.tex
separate performance model from paper/gmgimp.tex in https://github.com/bueler/glaciers-gmg-implicit and start correspondence paper

\documentclass[twocolumn,letterpaper]{igs}
%\documentclass[twocolumn]{igs}  % A4 paper needs no option
%\documentclass[review,letterpaper]{igs}

\usepackage{verbatim,xspace,amsmath,amssymb,bm,multirow,dashbox}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{igsnatbib,lineno}

\usepackage[kw]{pseudo}
\pseudoset{left-margin=0mm,topsep=5mm,idfont=\texttt}

\usepackage{hyperref}
\hypersetup{pdfauthor={Ed Bueler},
            pdfcreator={pdflatex},
            colorlinks=true,
            citecolor=purple,
            linkcolor=red,
            urlcolor=blue,
            }

% math macros
\newcommand\bb{\mathbf{b}}
\newcommand\bc{\mathbf{c}}
\newcommand\bbf{\mathbf{f}}
\newcommand\bg{\mathbf{g}}
\newcommand\bn{\mathbf{n}}
\newcommand\bq{\mathbf{q}}
\newcommand\bu{\mathbf{u}}
\newcommand\bv{\mathbf{v}}
\newcommand\bx{\mathbf{x}}
\newcommand\by{\mathbf{y}}

\newcommand\bA{\mathbf{A}}
\newcommand\bF{\mathbf{F}}
\newcommand\bH{\mathbf{H}}
\newcommand\bN{\mathbf{N}}
\newcommand\bQ{\mathbf{Q}}
\newcommand\bU{\mathbf{U}}
\newcommand\bV{\mathbf{V}}
\newcommand\bW{\mathbf{W}}
\newcommand\bX{\mathbf{X}}

\newcommand\bzero{\bm{0}}

\newcommand{\Div}{\nabla\cdot}
\newcommand\eps{\epsilon}
\newcommand{\grad}{\nabla}
\newcommand{\ip}[2]{\ensuremath{\left<#1,#2\right>}}
\newcommand\lam{\lambda}
\newcommand\lap{\triangle}
\newcommand\RR{\mathbb{R}}
\newcommand{\half}{\tfrac{1}{2}}

\newcommand{\rhoi}{\rho_{\text{i}}}
\newcommand{\pp}{{\text{p}}}
\newcommand{\qq}{{\text{q}}}
\newcommand{\rr}{{\text{r}}}

\newcommand{\mR}{R^{\bm{\oplus}}}
\newcommand{\iR}{R^{\bullet}}

\newcommand{\sold}{s_{\text{o}}}
\newcommand{\told}{t_{\text{o}}}

\newcommand{\byold}{\by_{\text{o}}}

\newcommand{\onecol}[1]{\includegraphics[width=86mm]{#1}}
\newcommand{\onecolless}[1]{\includegraphics[width=80mm]{#1}}
\newcommand{\onecolthin}[1]{\includegraphics[width=75mm]{#1}}

%\newcommand{\twocol}[1]{\includegraphics[width=178mm]{#1}}
\newcommand{\twocol}[1]{\includegraphics[width=165mm]{#1}}
\newcommand{\twocolless}[1]{\includegraphics[width=140mm]{#1}}

\newcommand{\ds}{\displaystyle}

\begin{document}

\title{Correspondence: Performance modeling of glacier simulations}

%\abstract{\today. FIXME}

\author{Ed Bueler}

\affiliation{Department of Mathematics and Statistics, University of Alaska Fairbanks, USA \\
E-mail: \emph{\texttt{elbueler\@@alaska.edu}}}

%\keywords{}

\maketitle

\sectionsize

We expect that time-stepping glacier and ice-sheet numerical models become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.  This claim can be understood and assessed by a \emph{performance model} for the scheme, one which combines simplified computational work estimates for subproblems into an asymptotic (in the limit of highly-refined computational grids) description of of the work required for the overall simulation task.

Consider numerical simulations based on $n$ mesh nodes in the map-plane, i.e.~$n$ total horizontal grid points.  Such a simulation uses $O(n)$ surface elevation variables, the \emph{degrees of freedom} in our analysis.  Indeed it requires $O(n)$ memory to store the instantaneous model state, including geometry, velocity, and thermodynamical variables, once we assume that the vertical grid/mesh has an \emph{a priori} bounded vertical resolution, a common \citep[for example]{Brinkerhoffetal2017,Hoffmanetal2018}, though not universal \citep{IsaacStadlerGhattas2015}, property of such models.

The glacier simulation covers a model domain of horizontal span $L>0$.  For concreteness, in 1D the domain is an interval $[0,L]$, while in 2D it is $[0,L]^2$, or a subset thereof.  Let $h_x$ be the characteristic width of the grid cells, thus $h_x = O(L n^{-1})$ in 1D and $h_x = O(L n^{-1/2})$ in 2D.

We assume that the simulation is over a time interval $[0,T]$; $T>0$ is the \emph{duration} in model years.  Ignoring issues of stability, scientific application of glacier and ice sheet models climate interaction, especially through the SMB, which is dominated by the annual cycle.  Let $Q$ be the number of time steps per year needed to incorporate this climate interaction; typical values $Q=1,12,365 \,\text{a}^{-1}$ correspond to yearly, monthly, and daily climatic interation frequency.

Almost all current-technology glacier and ice-sheet models use explicit time-stepping which is only conditionally stable \citep[for example]{LeVeque2007}.  In that case we assume that for the resolutions used in scientific applications maintenance of model stability requires more than $Q$ time steps per year.

For an explicit SIA model the well-known stability restriction is $\Delta t < D h_x^2$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,Robinsonetal2022}.

For Stokes the question of explicit time-stepping stability is largely unexplored.  We propose a \emph{pessimistic} stability paradigm, namely the above SIA restriction $\Delta t < D h_x^2$, motivated by the small geometric aspect ratio analysis of the Stokes model leading to the SIA model \citep{GreveBlatter2009}.  Alternatively one might subscribe to an \emph{optimistic} paradigm supposing that stability is controlled by the advective restriction $\Delta t < u_0^{-1} h_x$ for some typical horizontal velocity scale $u_0$ in $\text{km}\,\text{a}^{-1}$.  So-called ``higher-order'' and ``hybrid'' are better-studied than numerical Stokes models, especially over a large range of horizontal resolutions.  Certain ``hybrid'' schemes simply use the SIA-type stability restriction \citep{Winkelmannetal2011}.  The optimistic concept is partially supported by a recent analysis of a certain higher-order scheme (``DIVA'') for which $\Delta t < \min\left\{C,u_0^{-1} h_x\right\}$ \citep[equations (52) and (56)]{Robinsonetal2022}, wherein $C>0$ is independent of $h_x$.  However, practical Greenland DIVA simulations suggest $\Delta t \sim O(h_x^{1.6})$ instead \citep[Figure 3(a)]{Robinsonetal2022}.

For fixed geometry, as in an explicit simulation, the momentum balance equations generally require nontrivial solutions as PDEs.  For the Stokes model we suppose one such solution requires $O(n^{1+\delta})$ work to, in particular, determine the surface velocity appearing in the SKE.  However, the value of $\delta \ge 0$ depends radically on velocity/pressure solver implementation.  Naive solution might involve $\delta=2$, for example if dense linear algebra is involved, but application of geometric multigrid methods by \cite{IsaacStadlerGhattas2015} suggest that $\delta \approx 0$ is perhaps attainable with an solver specifically adapted to ice sheets and a high-resolution simulation.  (In any case the constant $C$ in $(\text{work}) \sim C n^{1+\delta}$ is large.)  On the other hand, the SIA model velocity computation is a trivialization of the Stokes problem.  Each velocity unknown is computed in $O(1)$ work, thus the velocity solution is attained (from fixed geometry) in $O(n)$ work.

On the other hand, an unconditionally-stable implicit geometric-update scheme, e.g.~as demostrated in \citep{Bueler2016} and the current paper for the SIA, will take at most $Q$ steps per model year in order to follow the climate forcing.  For such an implicit scheme, our stability model is that there is no restriction, equivalently that $\Delta t < O(T)$ in the fine-resolution limit ($h_x \to 0$).  Let $M=QT$ is the total number of steps.  The steady-state SIA solver by \cite{Bueler2016} uses $M=1$ when $T=+\infty$, with $Q=0.01 \, \leftrightarrow \Delta t = 100 \,\text{a}$ as a solver-recovery strategy.

At each time step the geometry is updated, for example the old surface elevation values are replaced by new ones.  For an explicit stepping scheme the work in this update is small, specifically $O(n)$ because, once the ice velocity is computed from fixed geometry by the chosen momentum balance model, each grid point requires $O(1)$ work to update.  The nMCD method proposed in this paper is more expensive, and in fact our model for the performance of this process is that each simultaneous geometry/velocity update takes $O(n^{1+\eps})$ for $\eps > 0$ small, or optimistically $O(n \log n)$, but in any case with a large constant.  The \cite{Bueler2016} geometry-update method, which is fundamentally not a multigrid method, is $O(n^{1+\eps})$ with a larger value of $\eps$.

\begin{table}
\begin{center}
\begin{tabular}{c|l}
\emph{name} & \emph{meaning} \\ \hline
$D$ & typical geometric diffusivity of glacier (equation FIXME) \\
$\delta$ & one Stokes/HO velocity solution requires $O(n^{1+\delta})$ work\\
$\Delta t$ & length of time step, in model years \\
$h_x$ & characteristic width of grid cells, in km \\
$\eps$ & one geometry-update solution requires $O(n^{1+\eps})$ work \\
$L$ & width of simulation domain, in km \\
$n$ & degrees of freedom: number of horizontal grid points \\
$Q$ & fixed time steps per model year in implicit simulations \\
$T$ & simulation duration, in model years \\
\end{tabular}
\end{center}

\medskip
\caption{Notation for glacier model performance analysis.  (\emph{Subsection only.})}
\label{tab:performanceconstants}
\end{table}

Table \ref{tab:performanceconstants} shows our notation, coming from the above considerations.

How much work, then, is the entire simulation?  For explicit methods note $\Delta t = T/m$ for $m$ actual (i.e.~stability-limited) time steps.  Combining this with our observations about $h_x$, we have $m \gtrsim (DL^2)^{-1} T n^2$ in the 1D SIA and pessimistic-Stokes cases, and $m \gtrsim (DL^2)^{-1} T n$ in the 1D optimistic-Stokes case.  The power on $n$ is lower in the realistic 2D cases because $h_x = L/\sqrt{n}$, but of course the simulations themselves are much more expensive for a given resolution $h_x$.  The $m$ value is multiplied by the per-step expense to give a total work estimate.  The resulting scaling estimates, e.g.~$m O(n) = (DL^2)^{-1} T n^3$ for explicit SIA models in 1D, are shown in the ``expl.'' rows of Table \ref{tab:performancemodel}.

For implicit methods, however, $\Delta t$ does not scale with grid resolution; it has fixed value $\Delta t = T / M = 1/Q$ years.  Thus the work of the entire implicit SIA simulation is $M n^{1+\eps} = Q^{-1} n^{1+\eps}$.  For an implicit Stokes simulation the geometry/velocity solve is supposed to be, in the absense of constraining research, $O(n^{1+\delta+\eps})$, with corresponding total simulation work estimate; see Table \ref{tab:performancemodel}.

Noting that $n$ is the number of horizontal grid points, the work estimate does not depend on the dimension unless $\eps$ and/or $\delta$ depend on dimension; such a difference is indeed likely to be revealed by further research.  Furthermore, the estimate reported in Table \ref{tab:performancemodel} is only at leading order, and must be multiplied by a potentially very large scheme-dependent constant.  On the other hand, the promise of multigrid methods for the velocity and geometry-update problems is to make $\delta \approx 0$ and $\eps \approx 0$, respectively.

\setlength{\tabcolsep}{5pt}
\newcommand{\tabshow}[1]{{\small $\ds #1$}}
\renewcommand{\arraystretch}{1.5}
\begin{table}
\begin{center}
\begin{tabular}{lc|cc}
\emph{scheme} & & \multicolumn{2}{c}{\emph{work (flops)}} \\
                             & & \emph{1D hor.} & \emph{2D hor.} \\ \hline
expl.~SIA       & & {\Large $\strut$} \dbox{\tabshow{(DL^2)^{-1} T n^3}} & \tabshow{(DL^2)^{-1} T n^2} \\
\multirow{2}{*}{expl.~Stokes} & \emph{pessim.} & \tabshow{(DL^2)^{-1} T n^{3+\delta}} & \tabshow{(DL^2)^{-1} T n^{2+\delta}} \\
                              & \emph{optim.} & \tabshow{u_0 L^{-1} T n^{2+\delta}} & \tabshow{u_0 L^{-1} T n^{1.5+\delta}} \\ \hline
impl.~SIA & & \multicolumn{2}{c}{{\Large $\strut$} \fbox{\tabshow{Q T n^{1+\eps}}}} \\
impl.~Stokes & & \multicolumn{2}{c}{\tabshow{Q T n^{1+\eps+\delta}}}
\end{tabular}
\end{center}

\medskip
\caption{Estimated scaling of computational work for high-resolution ($n\to\infty$) time-stepping numerical glacier simulations, over duration $T$ and horizontal domain width $L$.  Our nMCD method is shown to reduce $\eps$ to near zero in the 1D implicit SIA case (boxed), significantly-improving previous performance (dashed box).  See Table \ref{tab:performanceconstants} and the text for notation.}
\label{tab:performancemodel}
\end{table}

%         References
\bibliography{perfmod}
\bibliographystyle{igs}

\end{document}

### 2022-06-22 paper/perfmod.tex
discard unlikely citations
make sure to cite recent work
two column tables; table 2 needs rebuild with more columns
fix tables and notation; add structure

@@
 
-\title{Correspondence: Performance modeling of glacier simulations}
+\title[Performance models of ice sheet simulations]{Letter: Performance models of ice sheet simulations}
 
-%\abstract{\today. FIXME}
+\abstract{\today. FIXME}
 
@@
 
-\affiliation{Department of Mathematics and Statistics, University of Alaska Fairbanks, USA \\
+\affiliation{Dept.~Mathematics and Statistics, University of Alaska Fairbanks, USA \\
 E-mail: \emph{\texttt{elbueler\@@alaska.edu}}}
@@
 
-We expect that time-stepping glacier and ice-sheet numerical models become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.
-This claim can be understood and assessed by a \emph{performance model} for the scheme, one which combines simplified computational work estimates for subproblems into an asymptotic (in the limit of highly-refined computational grids) description of of the work required for the overall simulation task.
+\subsection{Introduction}
+
+FIXME INCREASING UNDERSTANDING OF TIME STEP LIMITS \citep{Chengetal2017,Robinsonetal2022}; MORE THAN SIA CASE \citep[for comparison]{HindmarshPayne1996}, POSSIBILITY OF IMPLICIT SOLVERS \citep{Bueler2016,WirbelJarosch2020}, NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS
+
+FIXME We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.
+This hope can be understood and assessed by a \emph{performance model} for the scheme, one which combines computational work estimates for subproblems into an asymptotic, in the limit of highly-refined meshes, description of of the work required for a given simulation task.
+
+\subsection{Performance models}
 
 Consider numerical simulations based on $n$ mesh nodes in the map-plane, i.e.~$n$ total horizontal grid points.
-Such a simulation uses $O(n)$ surface elevation variables, the \emph{degrees of freedom} in our analysis.
-Indeed it requires $O(n)$ memory to store the instantaneous model state, including geometry, velocity, and thermodynamical variables, once we assume that the vertical grid/mesh has an \emph{a priori} bounded vertical resolution, a common \citep[for example]{Brinkerhoffetal2017,Hoffmanetal2018}, though not universal \citep{IsaacStadlerGhattas2015}, property of such models.
+(Table \ref{tab:notation} lists all of our notation.)  Such a simulation uses $O(n)$ surface elevation variables, the \emph{degrees of freedom} in our analysis.
+It requires $O(n)$ memory to store the instantaneous model state, including geometry, velocity, and thermodynamical variables, if we assume that the vertical grid/mesh has an \emph{a priori} bounded vertical resolution, a common \citep[for example]{Brinkerhoffetal2017,Hoffmanetal2018}, though not universal \citep{IsaacStadlerGhattas2015}, property of such models.
 
-The glacier simulation covers a model domain of horizontal span $L>0$.
-For concreteness, in 1D the domain is an interval $[0,L]$, while in 2D it is $[0,L]^2$, or a subset thereof.
-Let $h_x$ be the characteristic width of the grid cells, thus $h_x = O(L n^{-1})$ in 1D and $h_x = O(L n^{-1/2})$ in 2D.
-
-We assume that the simulation is over a time interval $[0,T]$; $T>0$ is the \emph{duration} in model years.
-Ignoring issues of stability, scientific application of glacier and ice sheet models climate interaction, especially through the SMB, which is dominated by the annual cycle.
-Let $Q$ be the number of time steps per year needed to incorporate this climate interaction; typical values $Q=1,12,365 \,\text{a}^{-1}$ correspond to yearly, monthly, and daily climatic interation frequency.
-
-Almost all current-technology glacier and ice-sheet models use explicit time-stepping which is only conditionally stable \citep[for example]{LeVeque2007}.
-In that case we assume that for the resolutions used in scientific applications maintenance of model stability requires more than $Q$ time steps per year.
-
-For an explicit SIA model the well-known stability restriction is $\Delta t < D h_x^2$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,Robinsonetal2022}.
-
-For Stokes the question of explicit time-stepping stability is largely unexplored.
-We propose a \emph{pessimistic} stability paradigm, namely the above SIA restriction $\Delta t < D h_x^2$, motivated by the small geometric aspect ratio analysis of the Stokes model leading to the SIA model \citep{GreveBlatter2009}.
-Alternatively one might subscribe to an \emph{optimistic} paradigm supposing that stability is controlled by the advective restriction $\Delta t < u_0^{-1} h_x$ for some typical horizontal velocity scale $u_0$ in $\text{km}\,\text{a}^{-1}$.
-So-called ``higher-order'' and ``hybrid'' are better-studied than numerical Stokes models, especially over a large range of horizontal resolutions.
-Certain ``hybrid'' schemes simply use the SIA-type stability restriction \citep{Winkelmannetal2011}.
-The optimistic concept is partially supported by a recent analysis of a certain higher-order scheme (``DIVA'') for which $\Delta t < \min\left\{C,u_0^{-1} h_x\right\}$ \citep[equations (52) and (56)]{Robinsonetal2022}, wherein $C>0$ is independent of $h_x$.
-However, practical Greenland DIVA simulations suggest $\Delta t \sim O(h_x^{1.6})$ instead \citep[Figure 3(a)]{Robinsonetal2022}.
-
-For fixed geometry, as in an explicit simulation, the momentum balance equations generally require nontrivial solutions as PDEs.
-For the Stokes model we suppose one such solution requires $O(n^{1+\delta})$ work to, in particular, determine the surface velocity appearing in the SKE.
-However, the value of $\delta \ge 0$ depends radically on velocity/pressure solver implementation.
-Naive solution might involve $\delta=2$, for example if dense linear algebra is involved, but application of geometric multigrid methods by \cite{IsaacStadlerGhattas2015} suggest that $\delta \approx 0$ is perhaps attainable with an solver specifically adapted to ice sheets and a high-resolution simulation.
-(In any case the constant $C$ in $(\text{work}) \sim C n^{1+\delta}$ is large.)  On the other hand, the SIA model velocity computation is a trivialization of the Stokes problem.
-Each velocity unknown is computed in $O(1)$ work, thus the velocity solution is attained (from fixed geometry) in $O(n)$ work.
-
-On the other hand, an unconditionally-stable implicit geometric-update scheme, e.g.~as demostrated in \citep{Bueler2016} and the current paper for the SIA, will take at most $Q$ steps per model year in order to follow the climate forcing.
-For such an implicit scheme, our stability model is that there is no restriction, equivalently that $\Delta t < O(T)$ in the fine-resolution limit ($h_x \to 0$).
-Let $M=QT$ is the total number of steps.
-The steady-state SIA solver by \cite{Bueler2016} uses $M=1$ when $T=+\infty$, with $Q=0.01 \, \leftrightarrow \Delta t = 100 \,\text{a}$ as a solver-recovery strategy.
-
-At each time step the geometry is updated, for example the old surface elevation values are replaced by new ones.
-For an explicit stepping scheme the work in this update is small, specifically $O(n)$ because, once the ice velocity is computed from fixed geometry by the chosen momentum balance model, each grid point requires $O(1)$ work to update.
-The nMCD method proposed in this paper is more expensive, and in fact our model for the performance of this process is that each simultaneous geometry/velocity update takes $O(n^{1+\eps})$ for $\eps > 0$ small, or optimistically $O(n \log n)$, but in any case with a large constant.
-The \cite{Bueler2016} geometry-update method, which is fundamentally not a multigrid method, is $O(n^{1+\eps})$ with a larger value of $\eps$.
-
-\begin{table}
+\begin{table*}[ht]
 \begin{center}
-\begin{tabular}{c|l}
-\emph{name} & \emph{meaning} \\ \hline
-$D$ & typical geometric diffusivity of glacier (equation FIXME) \\
-$\delta$ & one Stokes/HO velocity solution requires $O(n^{1+\delta})$ work\\
-$\Delta t$ & length of time step, in model years \\
-$h_x$ & characteristic width of grid cells, in km \\
+\begin{tabular}{cll}
+\emph{name} & \emph{meaning} & \emph{units} \\ \hline
+$D$         & representative geometric (shallow ice approximation) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
+$\delta$    & one velocity solution requires $O(n^{1+\delta})$ work\\
 $\eps$      & one geometry-update solution requires $O(n^{1+\eps})$ work \\
-$L$ & width of simulation domain, in km \\
+$L$         & width of simulation domain & km \\
+$m$         & total number of stability-limited time steps in actual simulation \\
+$m_c$       & total number of time steps needed to resolve modeled climate interactions ($m_c=qT$) \\
 $n$         & degrees of freedom: number of horizontal grid points \\
-$Q$ & fixed time steps per model year in implicit simulations \\
-$T$ & simulation duration, in model years \\
+$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
+$T$         & simulation duration in model years & a \\
+$\Delta t$  & length of time step in model years & a \\
+$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
+$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
 \end{tabular}
@@
 \medskip
-\caption{Notation for glacier model performance analysis.  (\emph{Subsection only.})}
-\label{tab:performanceconstants}
-\end{table}
+\caption{Notation for performance modeling.  Parameters $\delta,\eps,m, m_c,n$ are pure numbers.}
+\label{tab:notation}
+\end{table*}
 
-Table \ref{tab:performanceconstants} shows our notation, coming from the above considerations.
+The glacier simulation covers a model domain of horizontal span $L>0$.
+For concreteness, in 1D the domain is an interval $[0,L]$, while in 2D it is $[0,L]^2$, or a subset thereof.
+Let $\Delta x$ be the characteristic width (diameter) of map-plane mesh cells, thus $\Delta x = O(L n^{-1})$ in 1D and $\Delta x = O(L n^{-1/2})$ in 2D.
 
-How much work, then, is the entire simulation?
-For explicit methods note $\Delta t = T/m$ for $m$ actual (i.e.~stability-limited) time steps.
-Combining this with our observations about $h_x$, we have $m \gtrsim (DL^2)^{-1} T n^2$ in the 1D SIA and pessimistic-Stokes cases, and $m \gtrsim (DL^2)^{-1} T n$ in the 1D optimistic-Stokes case.
+We assume that the simulation is over a time interval $[0,T]$ where $T>0$ is the \emph{duration} in model years.
+Momentarily ignoring issues of stability, ice sheet models with climate interaction, especially SMB, the interaction is dominated by the annual cycle.
+Let $q$ be the number of time steps per year needed to incorporate this climate interaction.
+(Typical values $q=1,12,365 \,\text{a}^{-1}$ correspond to yearly, monthly, and daily climatic interation frequency.)  The number of time steps imposed by the need to resolve climatic interaction is $m_c=qT$.
+
+Almost all current-technology glacier and ice-sheet models use explicit time-stepping which is only conditionally stable \citep{LeVeque2007}.
+We will assume that for the resolutions used in scientific applications, maintenance of the stability of explicit time-stepping requires substantially more than $Q$ time steps per year.
+
+For an explicit SIA model the well-known stability restriction is $\Delta t < D \Delta x^2$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,HindmarshPayne1996}.
+
+For Stokes dynamics the question of explicit time-stepping stability is largely unexplored.
+We propose a \emph{pessimistic} explicit time-step stability paradigm, namely the above SIA restriction $\Delta t < D \Delta x^2$.
+This is motivated by the small geometric aspect ratio analysis of the Stokes model leading to the SIA model \citep{GreveBlatter2009}.
+Alternatively one might subscribe to an \emph{optimistic} paradigm supposing that stability is controlled by the advective restriction $\Delta t < U^{-1} \Delta x$ for some typical horizontal velocity scale $U$ in $\text{km}\,\text{a}^{-1}$.
+
+So-called ``higher-order'' and ``hybrid'' are perhaps better-studied than numerical Stokes models, especially over a large range of horizontal resolutions.
+Certain ``hybrid'' schemes simply use the SIA-type stability restriction \citep{Winkelmannetal2011}.
+The optimistic concept above is partially supported by a recent analysis of a certain higher-order scheme (``DIVA'') for which $\Delta t < \min\left\{C,U^{-1} h_x\right\}$ \citep[equations (52) and (56)]{Robinsonetal2022}, wherein $C>0$ is independent of $\Delta x$.
+However, practical Greenland DIVA simulations in the same work actually suggest $\Delta t \sim O(\Delta x^{1.6})$ instead \citep[Figure 3(a)]{Robinsonetal2022}.
+
+For fixed geometry, as in each time step of an explicit simulation, the momentum balance equations generally require nontrivial solutions as PDEs.
+For the Stokes model we suppose one such solution requires $O(n^{1+\delta})$ work.
+Note that such a scheme will determine the surface velocity appearing in the SKE.
+However, the value of $\delta \ge 0$ depends strongly on the numerical velocity/pressure solver implementation, that is, the numerics by which the Stokes momentum balance is actually solved.
+Naive solution might involve $\delta=2$, for example if dense linear algebra is involved, but application of geometric multigrid methods by \cite{IsaacStadlerGhattas2015} suggest that $\delta \approx 0$ is perhaps attainable with an solver specifically adapted to ice sheets and a high-resolution simulation.
+(In any case the constant $C$ in $(\text{work}) \sim C n^{1+\delta}$ is large.)
+
+On the other hand, the SIA model velocity computation is a trivialization of the Stokes problem.
+Each velocity unknown is computed in $O(1)$ work, thus the velocity solution is attained (from fixed geometry) in $O(n)$ work.
+
+On the other hand, an unconditionally-stable implicit geometric-update scheme, e.g.~as demonstrated in \citep{Bueler2016} for the SIA, will take at most $Q$ steps per model year in order to follow the climate forcing.
+For such an implicit scheme, our stability model is that there is no restriction in the fine-resolution limit ($\Delta x \to 0$).
+For example, the steady-state SIA solver in \cite{Bueler2016} uses $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence.
+
+At each time step of an ice sheet simulation the geometry is updated, for example the old surface elevation values are replaced by new ones.
+For an explicit stepping scheme the work in this update is small, specifically $O(n)$, because, once the ice velocity is computed from fixed geometry by the chosen momentum balance model, each grid point requires $O(1)$ work to update.
+The \cite{Bueler2016} geometry-update method, which is fundamentally not a multigrid method, is $O(n^{1+\eps})$ with a larger value of $\eps$.
+
+FIXME The nMCD method proposed in [Bueler 2022] is more expensive, and in fact our model for the performance of this process is that each simultaneous geometry/velocity update takes $O(n^{1+\eps})$ for $\eps > 0$ small, or optimistically $O(n \log n)$, but in any case with a large constant.
+
+How much work, then, is the entire simulation?  Table \ref{tab:performancemodel} summarizes our performance models.
+
+\newcommand{\oo}[1]{{\LARGE \strut} \displaystyle O\left(#1\right)}
+\setlength{\tabcolsep}{5pt}
+\renewcommand{\arraystretch}{1.5}
+\begin{table*}[ht]
+\begin{center}
+\begin{tabular}{lllll}
+\emph{time-stepping} & \emph{dynamics} & \emph{work in 1D} & \emph{work in 2D} & \emph{caveat} \\ \hline
+explicit & SIA    & $\oo{\frac{D}{L^2} n^3}$       & $\oo{\frac{D}{L^2} n^2}$ \\
+explicit & Stokes & $\oo{\frac{D}{L^2} n^{3+\delta}}$ & $\oo{\frac{D}{L^2} n^{2+\delta}}$ &  $D$ effects dominant \\
+explicit & Stokes & $\oo{\frac{U}{L} n^{2+\delta}}$    & $\oo{\frac{U}{L} n^{1.5+\delta}}$  & $U$ effects dominant \\
+implicit & SIA    & \multicolumn{2}{c}{$\oo{q\, n^{1+\eps}}$} \\
+implicit & Stokes & \multicolumn{2}{c}{$\oo{q\, n^{1+\eps+\delta}}$}
+\end{tabular}
+\end{center}
+
+\medskip
+\caption{Asymptotic scaling of computational work, in flops per model year, for time-stepping numerical ice sheet simulations over domain width $L$, as the degrees of freedom $n$ go to infinity (high horizontal resolution).
+See Table \ref{tab:notation} for remaining notation; \emph{1D} and \emph{2D} refer to flow-line and map-plane simulations, respectively.}
+\label{tab:performancemodel}
+\end{table*}
+
+For explicit methods note that $\Delta t = T/m$ for $m$ actual, stability-limited time steps in the simulation.
+Combining this with our observations about $\Delta x$, we have $m \gtrsim (DL^2)^{-1} T n^2$ in the 1D SIA and pessimistic-Stokes cases, and $m \gtrsim (DL^2)^{-1} T n$ in the 1D optimistic-Stokes case.
 The power on $n$ is lower in the realistic 2D cases because $h_x = L/\sqrt{n}$, but of course the simulations themselves are much more expensive for a given resolution $h_x$.
-The $m$ value is multiplied by the per-step expense to give a total work estimate.
-The resulting scaling estimates, e.g.~$m O(n) = (DL^2)^{-1} T n^3$ for explicit SIA models in 1D, are shown in the ``expl.'' rows of Table \ref{tab:performancemodel}.
 
-For implicit methods, however, $\Delta t$ does not scale with grid resolution; it has fixed value $\Delta t = T / M = 1/Q$ years.
-Thus the work of the entire implicit SIA simulation is $M n^{1+\eps} = Q^{-1} n^{1+\eps}$.
-For an implicit Stokes simulation the geometry/velocity solve is supposed to be, in the absense of constraining research, $O(n^{1+\delta+\eps})$, with corresponding total simulation work estimate; see Table \ref{tab:performancemodel}.
+This $m$ value is multiplied by the per-step expense to give a total work estimate.
+The resulting scaling estimates, e.g.~$m O(n) = (DL^2)^{-1} T n^3$ for explicit SIA models in 1D, are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.
+
+For implicit methods, however, $\Delta t$ does not scale with grid resolution; it has fixed value $\Delta t = T / m_c = 1/Q$ years.
+That is, $m=m_c$ because the number of time steps is determined only by the need to resolve climatic interactions.
+Thus the work of the entire implicit SIA simulation is $m_c n^{1+\eps} = q T n^{1+\eps}$.
+For an implicit Stokes simulation the geometry/velocity solve is assumed to be, in the absense of constraining research, $O(n^{1+\delta+\eps})$, with corresponding total simulation work estimate; see Table \ref{tab:performancemodel}.
 
@@
 
-\setlength{\tabcolsep}{5pt}
-\newcommand{\tabshow}[1]{{\small $\ds #1$}}
-\renewcommand{\arraystretch}{1.5}
-\begin{table}
-\begin{center}
-\begin{tabular}{lc|cc}
-\emph{scheme} & & \multicolumn{2}{c}{\emph{work (flops)}} \\
-                             & & \emph{1D hor.} & \emph{2D hor.} \\ \hline
-expl.~SIA       & & {\Large $\strut$} \dbox{\tabshow{(DL^2)^{-1} T n^3}} & \tabshow{(DL^2)^{-1} T n^2} \\
-\multirow{2}{*}{expl.~Stokes} & \emph{pessim.} & \tabshow{(DL^2)^{-1} T n^{3+\delta}} & \tabshow{(DL^2)^{-1} T n^{2+\delta}} \\
-                              & \emph{optim.} & \tabshow{u_0 L^{-1} T n^{2+\delta}} & \tabshow{u_0 L^{-1} T n^{1.5+\delta}} \\ \hline
-impl.~SIA & & \multicolumn{2}{c}{{\Large $\strut$} \fbox{\tabshow{Q T n^{1+\eps}}}} \\
-impl.~Stokes & & \multicolumn{2}{c}{\tabshow{Q T n^{1+\eps+\delta}}}
-\end{tabular}
-\end{center}
+\subsection{Discussion and Conclusion}
 
-\medskip
-\caption{Estimated scaling of computational work for high-resolution ($n\to\infty$) time-stepping numerical glacier simulations, over duration $T$ and horizontal domain width $L$.
-Our nMCD method is shown to reduce $\eps$ to near zero in the 1D implicit SIA case (boxed), significantly-improving previous performance (dashed box).
-See Table \ref{tab:performanceconstants} and the text for notation.}
-\label{tab:performancemodel}
-\end{table}
+FIXME also cite \citep{LofgrenAhlkronaHelanow2021}
 

### 2022-06-22 paper/perfmod.tex
update titles and tables

@@
 
-\title[Performance models of ice sheet simulations]{Letter: Performance models of ice sheet simulations}
+\title[Performance models for high-resolution ice sheet simulations]{Letter: Performance models for \\ high-resolution ice sheet simulations}
 
@@
 \begin{table*}[ht]
-\begin{center}
 \begin{tabular}{cll}
 \emph{name} & \emph{meaning} & \emph{units} \\ \hline
+$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
+$\beta$     & one implicit SIA geometry-update requires $O(n^{1+\beta})$ work \\
+$\gamma$    & one implicit Stokes geometry-update plus velocity solution requires $O(n^{1+\gamma})$ work \\
 $D$         & representative geometric (shallow ice approximation) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
-$\delta$    & one velocity solution requires $O(n^{1+\delta})$ work\\
-$\eps$      & one geometry-update solution requires $O(n^{1+\eps})$ work \\
 $L$         & width of simulation domain & km \\
-$m$         & total number of stability-limited time steps in actual simulation \\
-$m_c$       & total number of time steps needed to resolve modeled climate interactions ($m_c=qT$) \\
-$n$         & degrees of freedom: number of horizontal grid points \\
+$n$         & degrees of freedom: number of nodes in the horizontal (flowline or map-plane) mesh \\
 $q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
@@
 \end{tabular}
-\end{center}
-
-\medskip
-\caption{Notation for performance modeling.  Parameters $\delta,\eps,m, m_c,n$ are pure numbers.}
+\caption{Notation for performance modeling.  Parameters $\alpha,\beta,\gamma,n$ are pure numbers.}
 \label{tab:notation}
@@
 \renewcommand{\arraystretch}{1.5}
-\begin{table*}[ht]
-\begin{center}
-\begin{tabular}{lllll}
-\emph{time-stepping} & \emph{dynamics} & \emph{work in 1D} & \emph{work in 2D} & \emph{caveat} \\ \hline
-explicit & SIA    & $\oo{\frac{D}{L^2} n^3}$       & $\oo{\frac{D}{L^2} n^2}$ \\
-explicit & Stokes & $\oo{\frac{D}{L^2} n^{3+\delta}}$ & $\oo{\frac{D}{L^2} n^{2+\delta}}$ &  $D$ effects dominant \\
-explicit & Stokes & $\oo{\frac{U}{L} n^{2+\delta}}$    & $\oo{\frac{U}{L} n^{1.5+\delta}}$  & $U$ effects dominant \\
-implicit & SIA    & \multicolumn{2}{c}{$\oo{q\, n^{1+\eps}}$} \\
-implicit & Stokes & \multicolumn{2}{c}{$\oo{q\, n^{1+\eps+\delta}}$}
+\begin{table}[ht]
+\begin{tabular}{lll}
+\emph{time-stepping} & \emph{dynamics} & \emph{work (flops per model year)} \\ \hline
+explicit & SIA    & 1D:\quad $\oo{\frac{D}{L^2} n^3}$ \\
+         &        & 2D:\quad $\oo{\frac{D}{L^2} n^2}$ \\
+explicit & Stokes & 1D:\quad $\oo{\frac{D}{L^2} n^{3+\alpha} + \frac{U}{L} n^{2+\alpha}}$ \\
+         &        & 2D:\quad $\oo{\frac{D}{L^2} n^{2+\alpha} + \frac{U}{L} n^{1.5+\alpha}}$ \\
+implicit & SIA    & $\oo{q\, n^{1+\beta}}$ \\
+implicit & Stokes & $\oo{q\, n^{1+\gamma}}$
 \end{tabular}
-\end{center}
-
-\medskip
-\caption{Asymptotic scaling of computational work, in flops per model year, for time-stepping numerical ice sheet simulations over domain width $L$, as the degrees of freedom $n$ go to infinity (high horizontal resolution).
-See Table \ref{tab:notation} for remaining notation; \emph{1D} and \emph{2D} refer to flow-line and map-plane simulations, respectively.}
+\caption{Asymptotic scaling of computational work for time-stepping numerical ice sheet simulations as the degrees of freedom $n$ go to infinity (high horizontal resolution).
+Note 1D and 2D refer to flow-line and map-plane simulations, respectively.
+See Table \ref{tab:notation} for remaining notation.}
 \label{tab:performancemodel}
-\end{table*}
+\end{table}
 

### 2022-06-22 paper/perfmod.tex
sweep through to make more sensible ... now needs good writing

@@
 
+FIXME \citep{GreveBlatter2009,SchoofHewitt2013} for SIA and Stokes equations
+
 FIXME We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.
@@
 
-We assume that the simulation is over a time interval $[0,T]$ where $T>0$ is the \emph{duration} in model years.
-Momentarily ignoring issues of stability, ice sheet models with climate interaction, especially SMB, the interaction is dominated by the annual cycle.
+Ice sheet models involve climate interaction, especially SMB, and the interaction is dominated by the annual cycle.
 Let $q$ be the number of time steps per year needed to incorporate this climate interaction.
-(Typical values $q=1,12,365 \,\text{a}^{-1}$ correspond to yearly, monthly, and daily climatic interation frequency.)  The number of time steps imposed by the need to resolve climatic interaction is $m_c=qT$.
+Typical values $q=1,12,365 \,\text{a}^{-1}$ correspond to yearly, monthly, and daily climatic interaction frequency.
 
@@
 
-For an explicit SIA model the well-known stability restriction is $\Delta t < D \Delta x^2$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,HindmarshPayne1996}.
-
+For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,HindmarshPayne1996}.\footnote{Recall $\displaystyle D = \frac{2 A (\rho g)^n}{n+2} H^{n+2} |\grad s|^{n-1}$ where $A$ is ice softness, $\rho$ is ice density, $g$ is gravity, $H$ is ice thickness, and $s$ is ice surface elevation.}
 For Stokes dynamics the question of explicit time-stepping stability is largely unexplored.
-We propose a \emph{pessimistic} explicit time-step stability paradigm, namely the above SIA restriction $\Delta t < D \Delta x^2$.
+We propose a \emph{pessimistic} explicit time-step stability paradigm, namely the above SIA restriction $\Delta t < O(D^{-1} \Delta x^2)$.
 This is motivated by the small geometric aspect ratio analysis of the Stokes model leading to the SIA model \citep{GreveBlatter2009}.
-Alternatively one might subscribe to an \emph{optimistic} paradigm supposing that stability is controlled by the advective restriction $\Delta t < U^{-1} \Delta x$ for some typical horizontal velocity scale $U$ in $\text{km}\,\text{a}^{-1}$.
+Alternatively one might subscribe to an \emph{optimistic} paradigm supposing that stability is controlled by the advective restriction $\Delta t < O(U^{-1} \Delta x)$ for some typical horizontal velocity scale $U$ in $\text{km}\,\text{a}^{-1}$.
 
@@
 Certain ``hybrid'' schemes simply use the SIA-type stability restriction \citep{Winkelmannetal2011}.
-The optimistic concept above is partially supported by a recent analysis of a certain higher-order scheme (``DIVA'') for which $\Delta t < \min\left\{C,U^{-1} h_x\right\}$ \citep[equations (52) and (56)]{Robinsonetal2022}, wherein $C>0$ is independent of $\Delta x$.
-However, practical Greenland DIVA simulations in the same work actually suggest $\Delta t \sim O(\Delta x^{1.6})$ instead \citep[Figure 3(a)]{Robinsonetal2022}.
+The optimistic concept above is partially supported by a recent analysis of a certain higher-order scheme (``DIVA'') for which $\Delta t < \min\left\{C,O(U^{-1} \Delta x)\right\}$ \citep[equations (52) and (56)]{Robinsonetal2022}, wherein $C>0$ is independent of $\Delta x$.
+However, practical Greenland DIVA simulations in the same work actually suggest $\Delta t = O(\Delta x^{1.6})$ instead \citep[Figure 3(a)]{Robinsonetal2022}.
 
 For fixed geometry, as in each time step of an explicit simulation, the momentum balance equations generally require nontrivial solutions as PDEs.
-For the Stokes model we suppose one such solution requires $O(n^{1+\delta})$ work.
+For the Stokes model we suppose one such solution requires $O(n^{1+\alpha})$ work.
 Note that such a scheme will determine the surface velocity appearing in the SKE.
-However, the value of $\delta \ge 0$ depends strongly on the numerical velocity/pressure solver implementation, that is, the numerics by which the Stokes momentum balance is actually solved.
-Naive solution might involve $\delta=2$, for example if dense linear algebra is involved, but application of geometric multigrid methods by \cite{IsaacStadlerGhattas2015} suggest that $\delta \approx 0$ is perhaps attainable with an solver specifically adapted to ice sheets and a high-resolution simulation.
-(In any case the constant $C$ in $(\text{work}) \sim C n^{1+\delta}$ is large.)
+However, the value of $\alpha \ge 0$ depends strongly on the numerical velocity/pressure solver implementation, that is, the numerics by which the Stokes momentum balance is actually solved.
+Naive solution might involve $\alpha=2$, for example if dense linear algebra is involved, but application of geometric multigrid methods by \cite{IsaacStadlerGhattas2015}, with a solver specifically adapted to the geometry and flow law applicable to ice sheets, suggest that $\alpha \approx 0$ is potentially attainable.
+(Of course, the constant $C$ in $(\text{work}) \le C n^{1+\alpha}$ is large.)  Geometric multigrid methods have also been applied to the higher-order (hydrostatic) momentum balance equations \citep{BrownSmithAhmadia2013}, a shallow approximation which resolves both vertical shear stresses and membrane stresses, and in this case a velocity solution work estimate $O(n^{1+\alpha})$ with $\alpha \approx 0$ is a reasonable aspiration.
 
-On the other hand, the SIA model velocity computation is a trivialization of the Stokes problem.
+The SIA model velocity computation is a trivialization of the Stokes problem.
 Each velocity unknown is computed in $O(1)$ work, thus the velocity solution is attained (from fixed geometry) in $O(n)$ work.
+
+FIXME \citep{LofgrenAhlkronaHelanow2021} suggest how to reduce the constant in explicit methods for Stokes geometry updates
 
@@
 For an explicit stepping scheme the work in this update is small, specifically $O(n)$, because, once the ice velocity is computed from fixed geometry by the chosen momentum balance model, each grid point requires $O(1)$ work to update.
-The \cite{Bueler2016} geometry-update method, which is fundamentally not a multigrid method, is $O(n^{1+\eps})$ with a larger value of $\eps$.
 
-FIXME The nMCD method proposed in [Bueler 2022] is more expensive, and in fact our model for the performance of this process is that each simultaneous geometry/velocity update takes $O(n^{1+\eps})$ for $\eps > 0$ small, or optimistically $O(n \log n)$, but in any case with a large constant.
-
-How much work, then, is the entire simulation?  Table \ref{tab:performancemodel} summarizes our performance models.
+Table \ref{tab:performancemodel} summarizes our performance models.
 
@@
 \renewcommand{\arraystretch}{1.5}
-\begin{table}[ht]
+\begin{table*}[ht]
 \begin{tabular}{lll}
@@
          &        & 2D:\quad $\oo{\frac{D}{L^2} n^2}$ \\
-explicit & Stokes & 1D:\quad $\oo{\frac{D}{L^2} n^{3+\alpha} + \frac{U}{L} n^{2+\alpha}}$ \\
-         &        & 2D:\quad $\oo{\frac{D}{L^2} n^{2+\alpha} + \frac{U}{L} n^{1.5+\alpha}}$ \\
+explicit & Stokes & 1D:\quad $\oo{\frac{U}{L} n^{2+\alpha}}$ \quad $\left[\oo{\frac{D}{L^2} n^{3+\alpha}} \text{if pessimistic}\right]$ \\
+         &        & 2D:\quad $\oo{\frac{U}{L} n^{1.5+\alpha}}$ \quad $\left[\oo{\frac{D}{L^2} n^{2+\alpha}} \text{if pessimistic}\right]$ \\
 implicit & SIA    & $\oo{q\, n^{1+\beta}}$ \\
@@
 \end{tabular}
-\caption{Asymptotic scaling of computational work for time-stepping numerical ice sheet simulations as the degrees of freedom $n$ go to infinity (high horizontal resolution).
+\caption{Asymptotic scaling of computational work for time-stepping numerical ice sheet simulations as the degrees of freedom $n$ goes to infinity (high horizontal resolution).
 Note 1D and 2D refer to flow-line and map-plane simulations, respectively.
@@
 \label{tab:performancemodel}
-\end{table}
+\end{table*}
 
+We assume that the simulation is over a time interval $[0,T]$ where $T>0$ is the \emph{duration} in model years.
 For explicit methods note that $\Delta t = T/m$ for $m$ actual, stability-limited time steps in the simulation.
-Combining this with our observations about $\Delta x$, we have $m \gtrsim (DL^2)^{-1} T n^2$ in the 1D SIA and pessimistic-Stokes cases, and $m \gtrsim (DL^2)^{-1} T n$ in the 1D optimistic-Stokes case.
-The power on $n$ is lower in the realistic 2D cases because $h_x = L/\sqrt{n}$, but of course the simulations themselves are much more expensive for a given resolution $h_x$.
+That is, we may define $m=T/\Delta t$ as the number of actual time steps taken.
 
-This $m$ value is multiplied by the per-step expense to give a total work estimate.
-The resulting scaling estimates, e.g.~$m O(n) = (DL^2)^{-1} T n^3$ for explicit SIA models in 1D, are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.
+Combining our observations about the scaling of $\Delta x$ with $n$, for explicit schemes we have
+\begin{equation}
+\Delta t^{-1} = \frac{m}{T} > O\left(\frac{D n^2}{L^2}\right)
+\end{equation}
+in the 1D SIA and pessimistic-Stokes cases and
+\begin{equation}
+\Delta t^{-1} > O\left(\frac{D n}{L^2}\right)
+\end{equation}
+in the corresponding 2D cases.  (Recall that the power on $n$ is lower in 2D cases because $\Delta x = L/\sqrt{n}$.)  In the optimistic-Stokes case these estimates become
+\begin{equation}
+\Delta t^{-1} > O\left(\frac{U n}{L}\right), O\left(\frac{U \sqrt{n}}{L}\right)
+\end{equation}
+in 1D and 2D respectively.
 
-For implicit methods, however, $\Delta t$ does not scale with grid resolution; it has fixed value $\Delta t = T / m_c = 1/Q$ years.
-That is, $m=m_c$ because the number of time steps is determined only by the need to resolve climatic interactions.
-Thus the work of the entire implicit SIA simulation is $m_c n^{1+\eps} = q T n^{1+\eps}$.
-For an implicit Stokes simulation the geometry/velocity solve is assumed to be, in the absense of constraining research, $O(n^{1+\delta+\eps})$, with corresponding total simulation work estimate; see Table \ref{tab:performancemodel}.
+This estimate of $\Delta t^{-1}$ is multiplied by the per-step expense to give a work estimate for each model year in a simulation.
+The results are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.
+Note that 2D simulationsare much more expensive for a given resolution $\Delta x$ because $n$ is so much larger; $n = O(\Delta x^{-1})$ in 1D while $n = O(\Delta x^{-2})$ in 2D.
 
-Noting that $n$ is the number of horizontal grid points, the work estimate does not depend on the dimension unless $\eps$ and/or $\delta$ depend on dimension; such a difference is indeed likely to be revealed by further research.
-Furthermore, the estimate reported in Table \ref{tab:performancemodel} is only at leading order, and must be multiplied by a potentially very large scheme-dependent constant.
-On the other hand, the promise of multigrid methods for the velocity and geometry-update problems is to make $\delta \approx 0$ and $\eps \approx 0$, respectively.
+For unconditionally-stable implicit methods, however, $\Delta t$ does not need scale with grid resolution; it has fixed value $\Delta t = 1/q$ model years.
+That is, the time steps are determined only by the need to resolve climatic interactions.
+
+We suppose that the work per geometry-update solution of an implicit SIA simulation is $O(n^{1+\beta})$ for some $\beta>0$.
+The \cite{Bueler2016} geometry-update method is fundamentally not a multigrid method, and it has FIXME WHAT $\beta$?
+
+For an implicit Stokes simulation the geometry/velocity solve is assumed to be, in the absense of constraining research, $O(n^{1+\gamma})$ for some $\gamma>0$ to be determined.
+The nMCD method proposed in [Bueler 2022] is expensive but multigrid, so it may turn out that $\gamma$ is small.
+
+Noting that $n$ is the number of horizontal grid points, the work estimate for an implicit time step may not depend on the dimension, but in fact it is likely that $\alpha,\beta,\gamma$ each depend on dimension in some manner; this should be informed by further research.
+Furthermore, the estimate reported in Table \ref{tab:performancemodel} must be multiplied by a potentially very large scheme-dependent constant.
 
@@
 
-FIXME also cite \citep{LofgrenAhlkronaHelanow2021}
+Existing multigrid methods for the velocity solution process show $\alpha$ is close to zero.
+That is, the evidence shows that $\alpha$ is small for well-engineered Stokes \citep{IsaacStadlerGhattas2015} and higher-order \citep{BrownSmithAhmadia2013} solvers.
+
+The promise of multigrid methods for the implicit, coupled geometry-update plus velocity solution problems is to also make $\beta,\gamma$ smaller.
+However, this aspiration is long-term because work is barely started on it.
 

### 2022-06-22 paper/perfmod.tex
an abstract

@@
 
-\abstract{\today. FIXME}
+\abstract{Modern numerical ice sheet models, with horizontal resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various widely-discussed stress-balance approximations and boundary conditions.
+The practical performance of such numerical models is, however, often dominated by stability conditions for currently-standard explicit time-stepping.
+(Current schemes generally update ice thickness after a velocity solution.)  Such models take many short, stability-limited time steps.
+In this note, performance of numerical models is clarified by quantifying computational effort, in the asymptotic limit of many map-plane mesh nodes, in terms of major computational parameters.
+The potential of coming generations of Stokes dynamics and/or implicit time-stepping numerical models is assessed.}
 

### 2022-06-23 paper/perfmod.tex
recent ism refs, and start on intro

--- 2022-06-22 paper/perfmod.tex
@@
 
-\abstract{\today. FIXME}
+\abstract{Modern numerical ice sheet models, with horizontal resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various widely-discussed stress-balance approximations and boundary conditions.
+The practical performance of such numerical models is, however, often dominated by the stability restrictions of currently-standard explicit time-stepping.
+Such models take many short, stability-limited time steps because ice thickness is updated following each velocity solution.
+We clarify the performance of numerical model choices, in the high-resolution limit, by quantifying computational effort in terms of major parameters.
+The potential performance of coming generations of Stokes dynamics and/or implicit time-stepping numerical models is assessed.}
 
@@
 
-FIXME INCREASING UNDERSTANDING OF TIME STEP LIMITS \citep{Chengetal2017,Robinsonetal2022}; MORE THAN SIA CASE \citep[for comparison]{HindmarshPayne1996}, POSSIBILITY OF IMPLICIT SOLVERS \citep{Bueler2016,WirbelJarosch2020}, NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS
+Numerical ice sheet models with evolving ice geometry are now in common use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and quantification of long-term glacial erosion \citep{SeguinotDelaney2021}.
+It is generally accepted that horizontal mesh or grid cells must be smaller than about 10 km in order to generate valid results.
+Whether using local mesh refinement \citep[for example]{Hoffmanetal2018} or not, resolutions of a few kilometers to one kilometer \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} have been used at ice sheet scale.
+With available computational resources, sub-kilometer meshes will become increasingly common even for long-term, whole ice sheet scale studies.
 
-FIXME \citep{GreveBlatter2009,SchoofHewitt2013} for SIA and Stokes equations
+The current generation of ice sheet models use a variety of stress balances, from simple approximations like the shallow ice approximation (SIA), through ``hybrid'' \citep[for example]{Winkelmannetal2011} and higher-order approximations \citep{GreveBlatter2009}, to the non-shallow Stokes approximation.
+With very few exceptions, however, current models alternate between solving the stress balance for velocity, using geometry from the previous time step, and then updating the ice thickness and/or surface elevation.
+The latter geometry-update operation, using the mass continuity or surface kinematical equations \cite{GreveBlatter2009}, is where interaction between the ice sheet and the climate is (primarily) simulated, especially through surface mass balance (SMB) and calving processes.
+Such schemes are called \emph{explicit time-stepping} for the coupled mass and momentum system which describes ice dynamics.
 
-FIXME We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.
+For simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
+The stability limitations of SIA numerical models have been understood, at least empirically, for some time \citep{HindmarshPayne1996}.
+More recent studies have focussed on time-step limits applicable to models using hybrid, higher-order, and Stokes dynamics \citep{Chengetal2017,Robinsonetal2022} or on techniques to mitigate such limits \citep{LofgrenAhlkronaHelanow2021}.
+
+FIXME POSSIBILITY OF IMPLICIT SOLVERS \citep{Bueler2016,WirbelJarosch2020}
+
+FIXME see \citep{GreveBlatter2009,SchoofHewitt2013} for SIA and Stokes equations; CONFUSION FROM TWO FORMATS
+    $$\frac{\partial H}{\partial t} + \Div \left(U H\right) = a$$
+    $$\frac{\partial H}{\partial t} = \Div \left(D \grad s \right) + a$$
+where FIXME; Recall $\displaystyle D = \frac{2 A (\rho g)^n}{n+2} H^{n+2} |\grad s|^{n-1}$ where $A$ is ice softness, $\rho$ is ice density, $g$ is gravity, $H$ is ice thickness, and $s$ is ice surface elevation; TSUNAMI WAVES EXAMPLE OF FORMER NOT LATER; NOTHING LIKE TSUNMAMIS IN ICE BECAUSE ICE FLOWS DOWNHILL WITH RARE EXCEPTIONS
+
+FIXME SITUATION COMPLICATED AND PRONE TO MISUNDERSTANDING; NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS; We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.
 This hope can be understood and assessed by a \emph{performance model} for the scheme, one which combines computational work estimates for subproblems into an asymptotic, in the limit of highly-refined meshes, description of of the work required for a given simulation task.
@@
 
-For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,HindmarshPayne1996}.\footnote{Recall $\displaystyle D = \frac{2 A (\rho g)^n}{n+2} H^{n+2} |\grad s|^{n-1}$ where $A$ is ice softness, $\rho$ is ice density, $g$ is gravity, $H$ is ice thickness, and $s$ is ice surface elevation.}
+For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,HindmarshPayne1996}.
 For Stokes dynamics the question of explicit time-stepping stability is largely unexplored.

### 2022-06-23 paper/perfmod.tex
edits to mostly complete introduction

--- 2022-06-22 paper/perfmod.tex
@@
 
+\newcommand{\Divx}{\nabla_\bx \cdot}
+\newcommand{\gradx}{\nabla_\bx}
+
 \newcommand{\rhoi}{\rho_{\text{i}}}
@@
 
-\abstract{\today. FIXME}
+\abstract{High-resolution numerical ice sheet models, with horizontal resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.
+These models usually require many short, stability-limited time steps because ice thickness is updated following each velocity solution.
+Practical performance in the high-spatial-resolution limit is degraded by the stability conditions of such explicit time-stepping.
+In this short note, which considers the shallow ice approximation and Stokes models as end members, we attempt to clarify numerical model performance by quantifying the simulation cost per model year in terms of the number of degrees of freedom.
+The performance of coming generations of implicit time-stepping numerical models is then assessed.}
 
@@
 
-\subsection{Introduction}
+\section{Introduction}
 
-FIXME INCREASING UNDERSTANDING OF TIME STEP LIMITS \citep{Chengetal2017,Robinsonetal2022}; MORE THAN SIA CASE \citep[for comparison]{HindmarshPayne1996}, POSSIBILITY OF IMPLICIT SOLVERS \citep{Bueler2016,WirbelJarosch2020}, NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS
+Numerical ice sheet (glacier) models with evolving ice geometry are now in common use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and quantification of long-term glacial erosion \citep{SeguinotDelaney2021}, among other applications.
+It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results.
+Whether using local mesh refinement \citep[for example]{Hoffmanetal2018} or not, resolutions of a few kilometers to one kilometer \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used at ice sheet scale.
+With currently-available computational resources, sub-kilometer meshes should become common for long-duration, whole ice sheet studies.
 
-FIXME \citep{GreveBlatter2009,SchoofHewitt2013} for SIA and Stokes equations
+Current-generation ice sheet models use a variety of stress balances, from simple approximations like the shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, to the non-shallow Stokes approximation.
+With very few exceptions, however, current models alternate between solving the stress balance for velocity, over the geometry determined by the previous time step, and then updating the geometry, i.e.~the ice thickness and/or surface elevation.
+The geometry-update operation, using the mass continuity or surface kinematical equations \citep{GreveBlatter2009}, is where interaction between the ice sheet and the climate is (primarily) simulated, especially through surface mass balance, sub-shelf (basal) mass balance, and calving processes.
+Such schemes are using \emph{explicit} time-stepping for the coupled mass and momentum system which describes ice dynamics.
 
-FIXME We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.
+For simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
+The stability limitations of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the time-step limits of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening the steps \citep{LofgrenAhlkronaHelanow2021}.
+
+However, \emph{implicit} time-stepping should also be considered.
+Here the velocity and geometry (thickness or surface elevation) are updated simultaneously by solving a coupled mass-momentum problem.
+While implicit time-stepping is a routine solution process for certain simpler partial differential equations \citep[for example]{LeVeque2007}, for ice sheets the implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}.
+The problem is of free-boundary type in map-plane (horizontal) directions \citep{SchoofHewitt2013}, as well as in the easily-resolved vertical direction.
+An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.
+The same reference shows how the steady-state problem \citep{JouvetBueler2012} can often be solved; steady-state solutions are an implicit time step of infinite duration.
+Early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} illuminates some of the techniques, and difficulties, needed to make such a free-boundary problem work with a membrane-stress-resolving balance.
+
+This is the context in which we will relate time-stepping and stress-balance choices to overall computational effort.
+The reader is assumed to have access to the SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}, regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}, and familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009}.
+
+However, the manner of presenting the mass continuity equation should first be examined.
+For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, this equation says
+\begin{equation}
+\frac{\partial H}{\partial t} + \Divx \left(\bu H\right) = a, \label{eq:masscontinuity}
+\end{equation}
+where $\bx=(x,y)$ denotes horizontal coordinates.
+
+Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, regarding the growth of numerical instabilities.
+This is because ice flows dominantly downhill.
+Thus, as addressed by linearized analysis \citep{Chengetal2017,Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of advective and diffusive mechanisms.
+
+Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation for bed elevation $b(\bx)$.
+If the numerical thickness is perturbed then the velocity $\bu$ often responds by growing in the downhill direction $-\gradx s$, a direction often correlated to $-\gradx H$.
+In membrane-stress-resolving models this happens through the nonlocal solution of the stress balance; the driving stress is along $\gradx s$.
+Equation \eqref{eq:masscontinuity} has velocity $\bu$ which is really a non-local function ``$\bu(H,\grad_x s)$''; a solution of the stress balance equations is required to evaluate this function.
+
+Thus the mass continuity equation is substantially, though not wholely, diffusive.
+The diffusive nature dominates in the small-parameter limit which generates the SIA, in which a revised form of the mass continuity equation holds:
+\begin{equation}
+\frac{\partial H}{\partial t} = \Divx \left(D \gradx s \right) + a  \label{eq:siamasscontinuity}
+\end{equation}
+where
+\begin{equation}
+D = \Gamma H^{n+2} |\gradx s|^{n-1} \label{eq:siadiffusivity}
+\end{equation}
+is the nonlinear diffusivity.
+Here $\Gamma = 2 A (\rho g)^n/(n+2)$ is a positive constant in the isothermal case, with $A$ the ice softness, $\rho$ the ice density, $g$ gravity, and $n$ the Glen exponent in the flow law \citep{GreveBlatter2009}.
+
+Unlike equation \eqref{eq:masscontinuity}, equation \eqref{eq:siamasscontinuity} does not hold for all models, but the essentially-geometric diffusivity $D$ can be computed from equation \eqref{eq:siadiffusivity}, regardless.
+In any numerical model of grounded ice sheets, large values of $D$ will coincide with locations of unstable mode growth if explicit time-steps are chosen too large.
+
+FIXME no characteristics
+
+FIXME SITUATION COMPLICATED AND PRONE TO MISUNDERSTANDING; NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS; We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.
 This hope can be understood and assessed by a \emph{performance model} for the scheme, one which combines computational work estimates for subproblems into an asymptotic, in the limit of highly-refined meshes, description of of the work required for a given simulation task.
 
-\subsection{Performance models}
+\section{Performance models}
 
@@
 
-\subsection{Discussion and Conclusion}
+\section{Discussion and Conclusion}
 

### 2022-06-24 paper/perfmod.tex
progress on performance model text itself

@@
 Equation \eqref{eq:masscontinuity} has velocity $\bu$ which is really a non-local function ``$\bu(H,\grad_x s)$''; a solution of the stress balance equations is required to evaluate this function.
+Equivalently, ice sheet flow has no characteristic curves because the velocity $\bu$ actually depends indirectly on the gradient of thickness.
 
 Thus the mass continuity equation is substantially, though not wholely, diffusive.
-The diffusive nature dominates in the small-parameter limit which generates the SIA, in which a revised form of the mass continuity equation holds:
+The diffusive nature dominates in the small-parameter limit which generates the SIA, in which a revised mass continuity equation holds:
 \begin{equation}
-\frac{\partial H}{\partial t} = \Divx \left(D \gradx s \right) + a  \label{eq:siamasscontinuity}
+\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a, \,  \label{eq:siamasscontinuity}
 \end{equation}
-where
-\begin{equation}
-D = \Gamma H^{n+2} |\gradx s|^{n-1} \label{eq:siadiffusivity}
-\end{equation}
-is the nonlinear diffusivity.
-Here $\Gamma = 2 A (\rho g)^n/(n+2)$ is a positive constant in the isothermal case, with $A$ the ice softness, $\rho$ the ice density, $g$ gravity, and $n$ the Glen exponent in the flow law \citep{GreveBlatter2009}.
+where $d = C H^{n+2} |\gradx s|^{n-1}$ is the nonlinear diffusivity, $C = 2 A (\rho g)^n/(n+2)$ is a positive constant, $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $n$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.
 
-Unlike equation \eqref{eq:masscontinuity}, equation \eqref{eq:siamasscontinuity} does not hold for all models, but the essentially-geometric diffusivity $D$ can be computed from equation \eqref{eq:siadiffusivity}, regardless.
-In any numerical model of grounded ice sheets, large values of $D$ will coincide with locations of unstable mode growth if explicit time-steps are chosen too large.
+Unlike equation \eqref{eq:masscontinuity}, equation \eqref{eq:siamasscontinuity} does not hold for all models.
+Regardless, the same diffusivity $d$, an essentially geometric quantity, can be computed.
+For grounded ice sheets large values of $d$ will indicate locations of unstable mode growth if explicit time-steps are chosen too large.
 
-FIXME no characteristics
-
-FIXME SITUATION COMPLICATED AND PRONE TO MISUNDERSTANDING; NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS; We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.
-This hope can be understood and assessed by a \emph{performance model} for the scheme, one which combines computational work estimates for subproblems into an asymptotic, in the limit of highly-refined meshes, description of of the work required for a given simulation task.
+The present-day understanding of ice sheet model time stepping is complicated.
+While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping and advanced solver techniques like multigrid \citep{Briggsetal2000} are applied, this belief should be inspected quantitatively to the extent possible.
+A simplified performance model can expose the most important considerations and trade-offs.
 
@@
 
-Consider numerical simulations based on $n$ mesh nodes in the map-plane, i.e.~$n$ total horizontal grid points.
-(Table \ref{tab:notation} lists all of our notation.)  Such a simulation uses $O(n)$ surface elevation variables, the \emph{degrees of freedom} in our analysis.
-It requires $O(n)$ memory to store the instantaneous model state, including geometry, velocity, and thermodynamical variables, if we assume that the vertical grid/mesh has an \emph{a priori} bounded vertical resolution, a common \citep[for example]{Brinkerhoffetal2017,Hoffmanetal2018}, though not universal \citep{IsaacStadlerGhattas2015}, property of such models.
+We analyze simulation performance using $n$, the number of mesh/grid nodes in the horizontal map-plane.
+(Table \ref{tab:notation} lists our notation.)  A numerical ice sheet model uses $n$ ice thickness or surface elevation variables, the number of \emph{degrees of freedom}.
+It requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This bounded-vertical-nodes assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}, but it may hold only in an approximate sense \citep{IsaacStadlerGhattas2015}.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.
 
@@
 
-The glacier simulation covers a model domain of horizontal span $L>0$.
-For concreteness, in 1D the domain is an interval $[0,L]$, while in 2D it is $[0,L]^2$, or a subset thereof.
-Let $\Delta x$ be the characteristic width (diameter) of map-plane mesh cells, thus $\Delta x = O(L n^{-1})$ in 1D and $\Delta x = O(L n^{-1/2})$ in 2D.
+Ice sheet models also have $O(n)$ velocity variables.  However, these are not state variables because a very-viscous stress balance can recompute velocity as a function of (true) state variables.
 
-Ice sheet models involve climate interaction, especially SMB, and the interaction is dominated by the annual cycle.
-Let $q$ be the number of time steps per year needed to incorporate this climate interaction.
-Typical values $q=1,12,365 \,\text{a}^{-1}$ correspond to yearly, monthly, and daily climatic interaction frequency.
+The model domain is of (horizontal) width $L>0$.
+If $\Delta x$ is a representative diameter of the map-plane mesh cells then $\Delta x = O(L n^{-1/2})$; there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.
+In flow-line models, not our primary motivation but included for completeness, $\Delta x = O(L n^{-1})$.
 
-Almost all current-technology glacier and ice-sheet models use explicit time-stepping which is only conditionally stable \citep{LeVeque2007}.
-We will assume that for the resolutions used in scientific applications, maintenance of the stability of explicit time-stepping requires substantially more than $Q$ time steps per year.
+Ice sheet models involve climate interactions, especially via surface mass balance, which are dominated by an annual cycle and longer time-scale changes.
+Let $q$ be the number of ice-dynamical time steps per year needed to capture this climate interaction; typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly flow/climate interaction frequency, respectively.
+Energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated from an ice velocity solution.
 
-For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,HindmarshPayne1996}.
-For Stokes dynamics the question of explicit time-stepping stability is largely unexplored.
-We propose a \emph{pessimistic} explicit time-step stability paradigm, namely the above SIA restriction $\Delta t < O(D^{-1} \Delta x^2)$.
-This is motivated by the small geometric aspect ratio analysis of the Stokes model leading to the SIA model \citep{GreveBlatter2009}.
-Alternatively one might subscribe to an \emph{optimistic} paradigm supposing that stability is controlled by the advective restriction $\Delta t < O(U^{-1} \Delta x)$ for some typical horizontal velocity scale $U$ in $\text{km}\,\text{a}^{-1}$.
+Current-technology ice sheet models mostly use explicit time-stepping which is only conditionally stable \citep{LeVeque2007}.
+For the spatial resolutions used in present-day scientific applications, we observe that maintenance of explicit time-stepping stability requires time steps substantially shorter than $1/q$ model years.
+Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $1/q$ itself, reflecting the simulation purpose and not the maintenance of stability.
 
-So-called ``higher-order'' and ``hybrid'' are perhaps better-studied than numerical Stokes models, especially over a large range of horizontal resolutions.
-Certain ``hybrid'' schemes simply use the SIA-type stability restriction \citep{Winkelmannetal2011}.
-The optimistic concept above is partially supported by a recent analysis of a certain higher-order scheme (``DIVA'') for which $\Delta t < \min\left\{C,O(U^{-1} \Delta x)\right\}$ \citep[equations (52) and (56)]{Robinsonetal2022}, wherein $C>0$ is independent of $\Delta x$.
-However, practical Greenland DIVA simulations in the same work actually suggest $\Delta t = O(\Delta x^{1.6})$ instead \citep[Figure 3(a)]{Robinsonetal2022}.
+For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value from the diffusivity formula in equation \eqref{eq:siamasscontinuity}.
 
-For fixed geometry, as in each time step of an explicit simulation, the momentum balance equations generally require nontrivial solutions as PDEs.
-For the Stokes model we suppose one such solution requires $O(n^{1+\alpha})$ work.
-Note that such a scheme will determine the surface velocity appearing in the SKE.
-However, the value of $\alpha \ge 0$ depends strongly on the numerical velocity/pressure solver implementation, that is, the numerics by which the Stokes momentum balance is actually solved.
-Naive solution might involve $\alpha=2$, for example if dense linear algebra is involved, but application of geometric multigrid methods by \cite{IsaacStadlerGhattas2015}, with a solver specifically adapted to the geometry and flow law applicable to ice sheets, suggest that $\alpha \approx 0$ is potentially attainable.
-(Of course, the constant $C$ in $(\text{work}) \le C n^{1+\alpha}$ is large.)  Geometric multigrid methods have also been applied to the higher-order (hydrostatic) momentum balance equations \citep{BrownSmithAhmadia2013}, a shallow approximation which resolves both vertical shear stresses and membrane stresses, and in this case a velocity solution work estimate $O(n^{1+\alpha})$ with $\alpha \approx 0$ is a reasonable aspiration.
+For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense \citep[compare][]{Chengetal2017}.
+We propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, computed from simulated ice velocities, represents the \emph{optimistic} paradigm.
+The corresponding \emph{pessimistic} paradigm enforces $\Delta t < O(D^{-1} \Delta x^2)$, using diagnostic diffusivity values computed in the SIA manner.
 
-The SIA model velocity computation is a trivialization of the Stokes problem.
-Each velocity unknown is computed in $O(1)$ work, thus the velocity solution is attained (from fixed geometry) in $O(n)$ work.
+Explicit time-stepping with hybrid and higher-order schemes is better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.
+Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}.
+The optimistic time step is supported for a certain higher-order scheme \citep[``DIVA'';][equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
 
-FIXME \citep{LofgrenAhlkronaHelanow2021} suggest how to reduce the constant in explicit methods for Stokes geometry updates
+Regardless of time-step stability, solution of the momentum balance equations requires effort.
+In general this requires a nontrivial solution of spatial partial differential equations.
+For the Stokes model, in particular, we suppose one such solution requires $O(n^{1+\alpha})$ work, namely floating point operations in our simple performance model, where $\alpha\ge 0$ depends on the design of the stress balance solver.
+Naive solution using direct linear algebra might involve $\alpha=3$, or $\alpha \ge 2$ when exploiting sparsity.
+However, application of a multigrid method, such as the \cite{IsaacStadlerGhattas2015} solver specifically adapted to the geometry and flow law applicable to ice sheets, suggest that $\alpha$ can potentially be close to zero, with the important caveat that the constant implied in ``$O(n^{1+\alpha})$'' is very large.
+Multigrid methods have also been applied to higher-order stress balance equations \citep{BrownSmithAhmadia2013}; again $\alpha \approx 0$ is a credible aspiration.
+On the other hand, the SIA velocity computation is a trivialization of the Stokes problem; here each velocity is computed in $O(1)$ work, and the velocity solution in $O(n)$ work.
 
-On the other hand, an unconditionally-stable implicit geometric-update scheme, e.g.~as demonstrated in \citep{Bueler2016} for the SIA, will take at most $Q$ steps per model year in order to follow the climate forcing.
+The stress-balance scheme determines the velocity appearing in equation \eqref{eq:masscontinuity}.
+From the velocity field, regardless of the stress balance source, an explicit time-stepping scheme computes the updated ice thickness using $O(n)$ work.
+
+On the other hand, an unconditionally-stable implicit geometric-update scheme in the SIA, as demonstrated by \cite{Bueler2016}, FIXME $O(n^{1+\beta})$ work; for Stokes this is a different task of solving coupled equations and we merely write $O(n^{1+\gamma})$ for now
+
+FIXME $\Delta t^{-1}$ time steps per model year (if $\Delta t$ is in model years)
+
 For such an implicit scheme, our stability model is that there is no restriction in the fine-resolution limit ($\Delta x \to 0$).
@@
 \begin{table*}[ht]
-\begin{tabular}{lll}
-\emph{time-stepping} & \emph{dynamics} & \emph{work (flops per model year)} \\ \hline
-explicit & SIA    & 1D:\quad $\oo{\frac{D}{L^2} n^3}$ \\
-         &        & 2D:\quad $\oo{\frac{D}{L^2} n^2}$ \\
-explicit & Stokes & 1D:\quad $\oo{\frac{U}{L} n^{2+\alpha}}$ \quad $\left[\oo{\frac{D}{L^2} n^{3+\alpha}} \text{if pessimistic}\right]$ \\
-         &        & 2D:\quad $\oo{\frac{U}{L} n^{1.5+\alpha}}$ \quad $\left[\oo{\frac{D}{L^2} n^{2+\alpha}} \text{if pessimistic}\right]$ \\
+\begin{tabular}{llll}
+\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year \, $\left[\text{1D}\right]$} & \emph{pessimistic \, $\left[\text{1D}\right]$} \\ \hline
+explicit & SIA    & $\oo{\frac{D}{L^2} n^2}$ \, $\left[n^3\right]$ \\
+explicit & Stokes & $\oo{\frac{U}{L} n^{1.5+\alpha}}$ \, $\left[n^{2+\alpha}\right]$ & $\oo{\frac{D}{L^2} n^{2+\alpha}}$ \, $\left[n^{3+\alpha}\right]$ \\
 implicit & SIA    & $\oo{q\, n^{1+\beta}}$ \\
@@
 \end{tabular}
-\caption{Asymptotic scaling of computational work for time-stepping numerical ice sheet simulations as the degrees of freedom $n$ goes to infinity (high horizontal resolution).
-Note 1D and 2D refer to flow-line and map-plane simulations, respectively.
+\caption{Asymptotic scaling of computational work, floating point operations per model year, for time-stepping numerical ice sheet simulations, as the degrees of freedom $n$ goes to infinity.
+Note 1D refers to flow-line geometry; corrected power on $n$ for 1D is in square brackets.
 See Table \ref{tab:notation} for remaining notation.}

### 2022-06-24 paper/perfmod.tex
sweep through first three pages

@@
 
-\title[Performance models for high-resolution ice sheet simulations]{Letter: Performance models for \\ high-resolution ice sheet simulations}
+\title[Performance analysis of high-resolution ice sheet simulations]{Letter: Performance analysis of \\ high-resolution ice sheet simulations}
 
-\abstract{High-resolution numerical ice sheet models, with horizontal resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.
-These models usually require many short, stability-limited time steps because ice thickness is updated following each velocity solution.
-Practical performance in the high-spatial-resolution limit is degraded by the stability conditions of such explicit time-stepping.
-In this short note, which considers the shallow ice approximation and Stokes models as end members, we attempt to clarify numerical model performance by quantifying the simulation cost per model year in terms of the number of degrees of freedom.
+\abstract{High spatial resolution numerical ice sheet models, with horizontal mesh/grid resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.
+These models usually require many time steps, usually shorter than climate-coupling time scales, because ice thickness is updated following each velocity solution.
+That is, practical performance in the high-resolution limit is degraded by the stability conditions imposed by such explicit time-stepping.
+In this short note, regarding the shallow ice approximation and Stokes models as stress balance end members, we attempt to clarify numerical model performance by quantifying the simulation cost per model year in terms of the mesh resolution and number of degrees of freedom.
 The performance of coming generations of implicit time-stepping numerical models is then assessed.}
@@
 
-Numerical ice sheet (glacier) models with evolving ice geometry are now in common use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and quantification of long-term glacial erosion \citep{SeguinotDelaney2021}, among other applications.
-It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results.
+Numerical ice sheet (glacier) models with evolving ice geometry are now in routine use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and quantification of long-term glacial erosion \citep{SeguinotDelaney2021}, among other applications.
+It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows suggest the need for finer resolution.
 Whether using local mesh refinement \citep[for example]{Hoffmanetal2018} or not, resolutions of a few kilometers to one kilometer \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used at ice sheet scale.
-With currently-available computational resources, sub-kilometer meshes should become common for long-duration, whole ice sheet studies.
 
-Current-generation ice sheet models use a variety of stress balances, from simple approximations like the shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, to the non-shallow Stokes approximation.
-With very few exceptions, however, current models alternate between solving the stress balance for velocity, over the geometry determined by the previous time step, and then updating the geometry, i.e.~the ice thickness and/or surface elevation.
-The geometry-update operation, using the mass continuity or surface kinematical equations \citep{GreveBlatter2009}, is where interaction between the ice sheet and the climate is (primarily) simulated, especially through surface mass balance, sub-shelf (basal) mass balance, and calving processes.
-Such schemes are using \emph{explicit} time-stepping for the coupled mass and momentum system which describes ice dynamics.
+Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, to the non-shallow and non-hydrostatic Stokes approximation.
+With very few exceptions, however, current models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the last-computed velocity field.
+Therefore ice thickness and surface elevation, which are functionally-equivalent variables for this purpose, are updated after the velocity is fixed.
+This geometry-update operation, using the mass continuity or surface kinematical equations \citep{GreveBlatter2009}, enables interaction between the ice sheet and the surrounding climate.
+Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes in particular.
 
-For simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
+These standard schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use the velocity computed from the geometry at the previous time step \citep[for example]{Chengetal2017}.}  Now, at least for simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
 The stability limitations of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the time-step limits of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening the steps \citep{LofgrenAhlkronaHelanow2021}.
 
-However, \emph{implicit} time-stepping should also be considered.
-Here the velocity and geometry (thickness or surface elevation) are updated simultaneously by solving a coupled mass-momentum problem.
-While implicit time-stepping is a routine solution process for certain simpler partial differential equations \citep[for example]{LeVeque2007}, for ice sheets the implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}.
-The problem is of free-boundary type in map-plane (horizontal) directions \citep{SchoofHewitt2013}, as well as in the easily-resolved vertical direction.
-An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.
-The same reference shows how the steady-state problem \citep{JouvetBueler2012} can often be solved; steady-state solutions are an implicit time step of infinite duration.
-Early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} illuminates some of the techniques, and difficulties, needed to make such a free-boundary problem work with a membrane-stress-resolving balance.
+However, actual \emph{implicit} time-stepping should also be considered.
+Here the velocity and geometry are updated simultaneously by solving a coupled mass-momentum problem.
+While implicit time-stepping always requires the solution of systems of equations at each step \citep{LeVeque2007}, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}.
+The problem is of free-boundary type, both in map-plane (horizontal) \citep{SchoofHewitt2013} and (easily-resolved) vertical directions.
 
-This is the context in which we will relate time-stepping and stress-balance choices to overall computational effort.
-The reader is assumed to have access to the SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}, regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}, and familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009}.
+A successful implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.
+The same reference shows how the steady-state problem \citep[see][for example]{JouvetBueler2012} can often be solved; observe that steady-state solutions correspond to an implicit time step of infinite duration.
+The corresponding problem for the Stokes equations has not, to the author's knowledge, yet been attempted.
+However, important early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} to solve for a free boundary illuminates some of the techniques and difficulties needed to make such a strategy work for a membrane-stress-resolving balance.
 
-However, the manner of presenting the mass continuity equation should first be examined.
+This is the complicated context in which the current note relates time-stepping and stress-balance choices to computational effort.
+While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping is applied, along with advanced solver techniques like multigrid \citep{Briggsetal2000}, this belief should be inspected quantitatively to the extent possible.
+A simplified performance (computational effort) analysis can expose the most important considerations and trade-offs.
+
+\section{Mass continuity equation}
+
+The reader is assumed to have access to the SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.
+These (continuum) models are regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}.
+
+Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.
+However, before presenting a performance analysis, the form of the mass continuity equation must be examined.
 For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, this equation says
@@
 
-Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, regarding the growth of numerical instabilities.
+Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.
 This is because ice flows dominantly downhill.
+Indeed, ice sheet flow has no characteristic curves, as would a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.
 Thus, as addressed by linearized analysis \citep{Chengetal2017,Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of advective and diffusive mechanisms.
@@
 Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation for bed elevation $b(\bx)$.
-If the numerical thickness is perturbed then the velocity $\bu$ often responds by growing in the downhill direction $-\gradx s$, a direction often correlated to $-\gradx H$.
-In membrane-stress-resolving models this happens through the nonlocal solution of the stress balance; the driving stress is along $\gradx s$.
-Equation \eqref{eq:masscontinuity} has velocity $\bu$ which is really a non-local function ``$\bu(H,\grad_x s)$''; a solution of the stress balance equations is required to evaluate this function.
-Equivalently, ice sheet flow has no characteristic curves because the velocity $\bu$ actually depends indirectly on the gradient of thickness.
+If a numerical thickness perturbation grows then the velocity $\bu$ often responds by increasing in a direction close to downhill ($-\gradx s$), a direction often correlated to $-\gradx H$.
+In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, but the gravitational driving stress is always along $\gradx s$.
+Equation \eqref{eq:masscontinuity} therefore has velocity $\bu$ which is really a non-local function ``$\bu(H,\grad_x s)$'', though a stress balance solution is required to evaluate this function.
+A numerical instability occurs when the ice thickness under/over-shoots its correct (i.e.~continuum) value because the numerical velocity response is too strong.
 
-Thus the mass continuity equation is substantially, though not wholely, diffusive.
-The diffusive nature dominates in the small-parameter limit which generates the SIA, in which a revised mass continuity equation holds:
+\newcommand{\nn}{\text{n}}
+The mass continuity equation is not wholely diffusive for Stokes dynamics, but the diffusive nature dominates in the small-parameter limit which generates the SIA.
+In the isothermal case we see the following, revised mass continuity equation:
 \begin{equation}
-\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a, \,  \label{eq:siamasscontinuity}
+\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
 \end{equation}
-where $d = C H^{n+2} |\gradx s|^{n-1}$ is the nonlinear diffusivity, $C = 2 A (\rho g)^n/(n+2)$ is a positive constant, $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $n$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.
+Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity and $C$ is a positive constant.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ where $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.
+Typically $\nn\approx 3$.}
 
-Unlike equation \eqref{eq:masscontinuity}, equation \eqref{eq:siamasscontinuity} does not hold for all models.
-Regardless, the same diffusivity $d$, an essentially geometric quantity, can be computed.
-For grounded ice sheets large values of $d$ will indicate locations of unstable mode growth if explicit time-steps are chosen too large.
-
-The present-day understanding of ice sheet model time stepping is complicated.
-While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping and advanced solver techniques like multigrid \citep{Briggsetal2000} are applied, this belief should be inspected quantitatively to the extent possible.
-A simplified performance model can expose the most important considerations and trade-offs.
-
-\section{Performance models}
-
-We analyze simulation performance using $n$, the number of mesh/grid nodes in the horizontal map-plane.
-(Table \ref{tab:notation} lists our notation.)  A numerical ice sheet model uses $n$ ice thickness or surface elevation variables, the number of \emph{degrees of freedom}.
-It requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This bounded-vertical-nodes assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}, but it may hold only in an approximate sense \citep{IsaacStadlerGhattas2015}.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.
+Equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics.
+However, the same diffusivity $d$, an essentially geometric quantity, can be computed.
+For grounded ice sheets one observes that large values of $d$ often indicate locations of unstable mode growth if explicit time-steps are chosen too large.
 
 \begin{table*}[ht]
+{\normalsize
 \begin{tabular}{cll}
@@
 \end{tabular}
+}
 \caption{Notation for performance modeling.  Parameters $\alpha,\beta,\gamma,n$ are pure numbers.}
@@
 
-Ice sheet models also have $O(n)$ velocity variables.  However, these are not state variables because a very-viscous stress balance can recompute velocity as a function of (true) state variables.
+\section{Performance analysis}
 
-The model domain is of (horizontal) width $L>0$.
-If $\Delta x$ is a representative diameter of the map-plane mesh cells then $\Delta x = O(L n^{-1/2})$; there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.
-In flow-line models, not our primary motivation but included for completeness, $\Delta x = O(L n^{-1})$.
+We parameterize simulation performance using either $\Delta x$, the representative horizontal mesh/grid cell diameter, i.e.~the resolution, or $n$, the number of nodes or vertices in the horizontal mesh or grid.
+The model domain is of (horizontal) width $L$ so $\Delta x = O(L n^{-1/2})$; there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L n^{-1})$, but our analysis focusses on spatially-3D models with map-plane horizontal meshes or grids.}  Table \ref{tab:notation} lists our notation.
 
-Ice sheet models involve climate interactions, especially via surface mass balance, which are dominated by an annual cycle and longer time-scale changes.
-Let $q$ be the number of ice-dynamical time steps per year needed to capture this climate interaction; typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly flow/climate interaction frequency, respectively.
-Energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated from an ice velocity solution.
+A standard numerical ice sheet model uses $n$ ice thickness or surface elevation variables, the number of \emph{degrees of freedom}, and it requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This bounded-vertical-nodes assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}, but it may hold only in an approximate sense \citep{IsaacStadlerGhattas2015}.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.
+Ice sheet models also have $O(n)$ velocity variables, but these are not state variables because a very-viscous stress balance can recompute velocity as a function of (true) state variables.
 
-Current-technology ice sheet models mostly use explicit time-stepping which is only conditionally stable \citep{LeVeque2007}.
+Ice sheet models involve climate interactions (coupling), especially via surface mass balance, on time scales which are dominated by an annual cycle and certain longer scales.
+We define $q$ as the number of ice-dynamical time steps per year needed to capture this coupling.
+Some typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly coupling frequency, respectively.
+Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated from an ice velocity solution and the latest surface mass balance values.
+
+As already noted, current-technology ice sheet models use explicit time-stepping which is only conditionally stable \citep{LeVeque2007}.
 For the spatial resolutions used in present-day scientific applications, we observe that maintenance of explicit time-stepping stability requires time steps substantially shorter than $1/q$ model years.
-Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $1/q$ itself, reflecting the simulation purpose and not the maintenance of stability.
-
 For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value from the diffusivity formula in equation \eqref{eq:siamasscontinuity}.
 
-For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense \citep[compare][]{Chengetal2017}.
-We propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, computed from simulated ice velocities, represents the \emph{optimistic} paradigm.
+For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but we propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, computed from simulated ice velocities, represents the \emph{optimistic} paradigm.
 The corresponding \emph{pessimistic} paradigm enforces $\Delta t < O(D^{-1} \Delta x^2)$, using diagnostic diffusivity values computed in the SIA manner.
@@
 Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}.
-The optimistic time step is supported for a certain higher-order scheme \citep[``DIVA'';][equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
+The optimistic time step is supported for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
 
-Regardless of time-step stability, solution of the momentum balance equations requires effort.
-In general this requires a nontrivial solution of spatial partial differential equations.
+Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $1/q$ itself.
+This restriction reflects the simulation purpose and not the maintenance of stability.
+Thus, for an implicit scheme the frequency of climate coupling determines the time-stepping simulation cost, through the need to solve the coupled mass/momentum equations with frequency $q$.
+
+FIXME WORK FROM HERE
+
+Regardless of time-stepping design, solution of the stress balance equations requires effort.
+In general this requires a nontrivial solution of (spatial) partial differential equations.
 For the Stokes model, in particular, we suppose one such solution requires $O(n^{1+\alpha})$ work, namely floating point operations in our simple performance model, where $\alpha\ge 0$ depends on the design of the stress balance solver.
@@
 \begin{table*}[ht]
+{\normalsize
 \begin{tabular}{llll}
-\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year \, $\left[\text{1D}\right]$} & \emph{pessimistic \, $\left[\text{1D}\right]$} \\ \hline
-explicit & SIA    & $\oo{\frac{D}{L^2} n^2}$ \, $\left[n^3\right]$ \\
-explicit & Stokes & $\oo{\frac{U}{L} n^{1.5+\alpha}}$ \, $\left[n^{2+\alpha}\right]$ & $\oo{\frac{D}{L^2} n^{2+\alpha}}$ \, $\left[n^{3+\alpha}\right]$ \\
-implicit & SIA    & $\oo{q\, n^{1+\beta}}$ \\
-implicit & Stokes & $\oo{q\, n^{1+\gamma}}$
+\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{pessimistic} \\ \hline
+explicit & SIA    & $\oo{\frac{D L^2}{\Delta x^4}} = \oo{\frac{D\, n^2}{L^2}}${\Huge \strut} \\
+explicit & Stokes & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U n^{1.5+\alpha}}{L}}${\Huge \strut}\phantom{x} & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,n^{2+\alpha}}{L^2}}$ \\
+implicit & SIA    & $\oo{\frac{q L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, n^{1+\beta}}${\Huge \strut} \\
+implicit & Stokes & $\oo{\frac{q L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, n^{1+\gamma}}${\Huge \strut}
 \end{tabular}
-\caption{Asymptotic scaling of computational work, floating point operations per model year, for time-stepping numerical ice sheet simulations, as the degrees of freedom $n$ goes to infinity.
-Note 1D refers to flow-line geometry; corrected power on $n$ for 1D is in square brackets.
-See Table \ref{tab:notation} for remaining notation.}
+}
+\caption{FIXME CHECK FOR ERRORS BEFORE INTERPRETATION.
+Asymptotic scaling of computational work, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $n\to\infty$.
+See Table \ref{tab:notation} for notation.}
 \label{tab:performancemodel}

### 2022-06-25 paper/perfmod.tex
better writing and table, using both dx and n

@@
 In this short note, regarding the shallow ice approximation and Stokes models as stress balance end members, we attempt to clarify numerical model performance by quantifying the simulation cost per model year in terms of the mesh resolution and number of degrees of freedom.
-The performance of coming generations of implicit time-stepping numerical models is then assessed.}
+The performance of coming generations of implicit time-stepping numerical models is then assessed.
+Our main result, Table \ref{tab:performancemodel}, emphasizes the key role played by solver performance in Stokes models.}
 
@@
 $\beta$     & one implicit SIA geometry-update requires $O(n^{1+\beta})$ work \\
-$\gamma$    & one implicit Stokes geometry-update plus velocity solution requires $O(n^{1+\gamma})$ work \\
-$D$         & representative geometric (shallow ice approximation) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
+$\gamma$    & one implicit Stokes geometry/velocity solution requires $O(n^{1+\gamma})$ work \\
+$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
 $L$         & width of simulation domain & km \\
-$n$         & degrees of freedom: number of nodes in the horizontal (flowline or map-plane) mesh \\
+$n$         & degrees of freedom: number of nodes in the horizontal mesh \\
 $q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
-$T$         & simulation duration in model years & a \\
 $\Delta t$  & length of time step in model years & a \\
@@
 }
-\caption{Notation for performance modeling.  Parameters $\alpha,\beta,\gamma,n$ are pure numbers.}
+\caption{Parameters for performance analysis; $\alpha,\beta,\gamma,n$ are pure numbers.}
 \label{tab:notation}
@@
 We parameterize simulation performance using either $\Delta x$, the representative horizontal mesh/grid cell diameter, i.e.~the resolution, or $n$, the number of nodes or vertices in the horizontal mesh or grid.
-The model domain is of (horizontal) width $L$ so $\Delta x = O(L n^{-1/2})$; there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L n^{-1})$, but our analysis focusses on spatially-3D models with map-plane horizontal meshes or grids.}  Table \ref{tab:notation} lists our notation.
+The model domain is of (horizontal) width $L$ so $\Delta x = O(L n^{-1/2})$; there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L n^{-1})$, but our analysis focusses on spatially-3D models with map-plane horizontal meshes or grids.}  Table \ref{tab:notation} lists the parameters used in our analysis.
+Big-O notation applies only in the high resolution limit where $\Delta x \to 0$ and $n\to \infty$.
 
-A standard numerical ice sheet model uses $n$ ice thickness or surface elevation variables, the number of \emph{degrees of freedom}, and it requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This bounded-vertical-nodes assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}, but it may hold only in an approximate sense \citep{IsaacStadlerGhattas2015}.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.
+A standard numerical ice sheet model uses $n$ ice thickness or surface elevation variables, the number of \emph{degrees of freedom}, and it requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This bounded-vertical-nodes assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.
+However, the \cite{IsaacStadlerGhattas2015} Stokes and \cite{BrownSmithAhmadia2013} higher-order multigrid solvers use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.
 Ice sheet models also have $O(n)$ velocity variables, but these are not state variables because a very-viscous stress balance can recompute velocity as a function of (true) state variables.
@@
 
-FIXME WORK FROM HERE
+Regardless of time-stepping design, for each explicit time step the velocity (or velocity/pressure) solution of the stress balance equations requires effort, namely the solution of (spatial) partial differential equations.
+For the Stokes model we suppose one such solution requires $O(n^{1+\alpha})$ work, which we describe the the number floating point operations in our simple performance analysis.
 
-Regardless of time-stepping design, solution of the stress balance equations requires effort.
-In general this requires a nontrivial solution of (spatial) partial differential equations.
-For the Stokes model, in particular, we suppose one such solution requires $O(n^{1+\alpha})$ work, namely floating point operations in our simple performance model, where $\alpha\ge 0$ depends on the design of the stress balance solver.
-Naive solution using direct linear algebra might involve $\alpha=3$, or $\alpha \ge 2$ when exploiting sparsity.
-However, application of a multigrid method, such as the \cite{IsaacStadlerGhattas2015} solver specifically adapted to the geometry and flow law applicable to ice sheets, suggest that $\alpha$ can potentially be close to zero, with the important caveat that the constant implied in ``$O(n^{1+\alpha})$'' is very large.
-Multigrid methods have also been applied to higher-order stress balance equations \citep{BrownSmithAhmadia2013}; again $\alpha \approx 0$ is a credible aspiration.
-On the other hand, the SIA velocity computation is a trivialization of the Stokes problem; here each velocity is computed in $O(1)$ work, and the velocity solution in $O(n)$ work.
+Of course, the power $\alpha$ depends on the design of the stress balance solver.
+A direct linear algebra Stokes solver implementation might yield $\alpha=3$, or $\alpha \approx 2$ when exploiting sparsity, but a multigrid method, such as the algebraic multigrid \citep{Trottenbergetal2001} solver by \cite{IsaacStadlerGhattas2015}, reduces $\alpha$.
+Their Antarctic ice sheet results show Newton-Krylov iterations \citep{Bueler2021} growing only slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
+% from Table 8.1:
+% [7*66 10*75 11*90] = [462 750 990]
+%>> polyfit(log([1 8 64]),log([462 750 990]),1)
+%ans = 0.18326   6.17004
+A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ also close to zero for certain simplified geometries.\footnote{The \cite{IsaacStadlerGhattas2015} and \cite{BrownSmithAhmadia2013} results also show good parallel scaling, something not considered in our simplified analysis.}  At the even easier end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula, $O(1)$ work; a velocity solution requires only $O(n)$ work ($\alpha=0$).
 
-The stress-balance scheme determines the velocity appearing in equation \eqref{eq:masscontinuity}.
-From the velocity field, regardless of the stress balance source, an explicit time-stepping scheme computes the updated ice thickness using $O(n)$ work.
+When analyzing solver scaling in this simple manner, one must be aware that the constant in $O(n^{1+\alpha})$ can be very large.  Furthermore it will vary greatly based on solver design.
 
-On the other hand, an unconditionally-stable implicit geometric-update scheme in the SIA, as demonstrated by \cite{Bueler2016}, FIXME $O(n^{1+\beta})$ work; for Stokes this is a different task of solving coupled equations and we merely write $O(n^{1+\gamma})$ for now
+Regardless of the stress balance source, an explicit time-stepping scheme applies the mass continuity equation to update the ice thickness using $O(n)$ work.
+That is, once the velocity is computed for the geometry from the previous time step, old thickness values are replaced by new ones by a formula needing only $O(1)$ work per mesh node.
 
-FIXME $\Delta t^{-1}$ time steps per model year (if $\Delta t$ is in model years)
-
-For such an implicit scheme, our stability model is that there is no restriction in the fine-resolution limit ($\Delta x \to 0$).
-For example, the steady-state SIA solver in \cite{Bueler2016} uses $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence.
-
-At each time step of an ice sheet simulation the geometry is updated, for example the old surface elevation values are replaced by new ones.
-For an explicit stepping scheme the work in this update is small, specifically $O(n)$, because, once the ice velocity is computed from fixed geometry by the chosen momentum balance model, each grid point requires $O(1)$ work to update.
-
-Table \ref{tab:performancemodel} summarizes our performance models.
-
-\newcommand{\oo}[1]{{\LARGE \strut} \displaystyle O\left(#1\right)}
+\newcommand{\oo}[1]{\displaystyle O\left(#1\right)}
 \setlength{\tabcolsep}{5pt}
@@
 \emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{pessimistic} \\ \hline
-explicit & SIA    & $\oo{\frac{D L^2}{\Delta x^4}} = \oo{\frac{D\, n^2}{L^2}}${\Huge \strut} \\
+explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, n^2}{L^2}}${\Huge \strut} \\
 explicit & Stokes & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U n^{1.5+\alpha}}{L}}${\Huge \strut}\phantom{x} & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,n^{2+\alpha}}{L^2}}$ \\
-implicit & SIA    & $\oo{\frac{q L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, n^{1+\beta}}${\Huge \strut} \\
-implicit & Stokes & $\oo{\frac{q L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, n^{1+\gamma}}${\Huge \strut}
+implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, n^{1+\beta}}${\Huge \strut} \\
+implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, n^{1+\gamma}}${\Huge \strut}
 \end{tabular}
 }
-\caption{FIXME CHECK FOR ERRORS BEFORE INTERPRETATION.
-Asymptotic scaling of computational work, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $n\to\infty$.
+\caption{Asymptotic scaling of computational work, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $n\to\infty$.
 See Table \ref{tab:notation} for notation.}
@@
 
-We assume that the simulation is over a time interval $[0,T]$ where $T>0$ is the \emph{duration} in model years.
-For explicit methods note that $\Delta t = T/m$ for $m$ actual, stability-limited time steps in the simulation.
-That is, we may define $m=T/\Delta t$ as the number of actual time steps taken.
+Now suppose that a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.
+Recalling the scaling of $\Delta x$ with $n$, namely that $\Delta x = O(L n^{-1/2})$, equivalently $n=O(L^2 \Delta x^{-2})$, for explicit schemes we may write the stability restriction as a required number of time-steps per model year in order to maintain stability, namely
+% $\Delta t < O(D^{-1} \Delta x^2)$
+\begin{equation}
+\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D n}{L^2}} \label{eq:explicitsiarequired}
+\end{equation}
+in the SIA and pessimistic-Stokes cases.  In the optimistic-Stokes case the estimate becomes
+% $\Delta t < O(U^{-1} \Delta x)$
+\begin{equation}
+\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U n^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
+\end{equation}
 
-Combining our observations about the scaling of $\Delta x$ with $n$, for explicit schemes we have
-\begin{equation}
-\Delta t^{-1} = \frac{m}{T} > O\left(\frac{D n^2}{L^2}\right)
-\end{equation}
-in the 1D SIA and pessimistic-Stokes cases and
-\begin{equation}
-\Delta t^{-1} > O\left(\frac{D n}{L^2}\right)
-\end{equation}
-in the corresponding 2D cases.  (Recall that the power on $n$ is lower in 2D cases because $\Delta x = L/\sqrt{n}$.)  In the optimistic-Stokes case these estimates become
-\begin{equation}
-\Delta t^{-1} > O\left(\frac{U n}{L}\right), O\left(\frac{U \sqrt{n}}{L}\right)
-\end{equation}
-in 1D and 2D respectively.
+The number of time steps per model year is multiplied by the per-step computational cost to give a work estimate for each model year in a simulation.
+The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.
 
-This estimate of $\Delta t^{-1}$ is multiplied by the per-step expense to give a work estimate for each model year in a simulation.
-The results are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.
-Note that 2D simulationsare much more expensive for a given resolution $\Delta x$ because $n$ is so much larger; $n = O(\Delta x^{-1})$ in 1D while $n = O(\Delta x^{-2})$ in 2D.
-
-For unconditionally-stable implicit methods, however, $\Delta t$ does not need scale with grid resolution; it has fixed value $\Delta t = 1/q$ model years.
-That is, the time steps are determined only by the need to resolve climatic interactions.
-
-We suppose that the work per geometry-update solution of an implicit SIA simulation is $O(n^{1+\beta})$ for some $\beta>0$.
-The \cite{Bueler2016} geometry-update method is fundamentally not a multigrid method, and it has FIXME WHAT $\beta$?
+For unconditionally-stable implicit methods, however, $\Delta t$ does not scale with grid resolution; it has fixed value $\Delta t = 1/q$ determined by the need to resolve climatic interactions.
+On the other hand, the per-step expense is much greater because nontrivial equations, indeed a free-boundary problem, must be solved.
+FIXME for the unconditionally-stable implicit geometric-update scheme in the SIA, as demonstrated by \cite{Bueler2016}, $O(n^{1+\beta})$ work; it is not multigrid; indeed the steady-state SIA solver in \cite{Bueler2016} allows $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence
 
@@
 
-Noting that $n$ is the number of horizontal grid points, the work estimate for an implicit time step may not depend on the dimension, but in fact it is likely that $\alpha,\beta,\gamma$ each depend on dimension in some manner; this should be informed by further research.
-Furthermore, the estimate reported in Table \ref{tab:performancemodel} must be multiplied by a potentially very large scheme-dependent constant.
+FIXME emphasize the estimate reported in Table \ref{tab:performancemodel} must be multiplied by a potentially very large scheme-dependent constant.
 
@@
 
-Existing multigrid methods for the velocity solution process show $\alpha$ is close to zero.
-That is, the evidence shows that $\alpha$ is small for well-engineered Stokes \citep{IsaacStadlerGhattas2015} and higher-order \citep{BrownSmithAhmadia2013} solvers.
+FIXME from Table \ref{tab:performancemodel} we see for explicit SIA that work scales as $\Delta x^{-4}$ for a given problem with fixed $D$ and $L$; because work per step scales as $\Delta x^{-2}$ but number of steps also scales as $\Delta x^{-2}$; explicit time-stepping Stokes is even worse under the pessimistic model; if solver only achieves $\alpha=2$ then work scales like horrific $\Delta x^{-8}$ (or $\Delta x^{-7}$ if optimistic about time-stepping stability); even for explicit time stepping the implementation burden of multigrid should pay off
 
-The promise of multigrid methods for the implicit, coupled geometry-update plus velocity solution problems is to also make $\beta,\gamma$ smaller.
+FIXME one can also fix resolution (fix $\Delta x$) and look at the expense of larger ice sheets ($L\to \infty$) to see why Stokes is difficult: scales as $L^{2+2\alpha}$ for optimistic and pessimistic cases; again want $\alpha$ is small
+
+FIXME The real promise of multigrid methods for the implicit, coupled geometry-update plus velocity solution problems is to make $\beta,\gamma$ seriously smaller.
+Specifically $\gamma < 1$ might be regarded as the goal of the next few decades of work on Stokes models.
 However, this aspiration is long-term because work is barely started on it.

### 2022-06-25 paper/perfmod.tex
quick look at Isaac et al 2015 and Bueler 2016 scaling

@@
 Their Antarctic ice sheet results show Newton-Krylov iterations \citep{Bueler2021} growing only slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
-% from Table 8.1:
-% [7*66 10*75 11*90] = [462 750 990]
-%>> polyfit(log([1 8 64]),log([462 750 990]),1)
-%ans = 0.18326   6.17004
+% see data/scalingisaac.m
 A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ also close to zero for certain simplified geometries.\footnote{The \cite{IsaacStadlerGhattas2015} and \cite{BrownSmithAhmadia2013} results also show good parallel scaling, something not considered in our simplified analysis.}  At the even easier end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula, $O(1)$ work; a velocity solution requires only $O(n)$ work ($\alpha=0$).
@@
 Recalling the scaling of $\Delta x$ with $n$, namely that $\Delta x = O(L n^{-1/2})$, equivalently $n=O(L^2 \Delta x^{-2})$, for explicit schemes we may write the stability restriction as a required number of time-steps per model year in order to maintain stability, namely
-% $\Delta t < O(D^{-1} \Delta x^2)$
 \begin{equation}
@@
 in the SIA and pessimistic-Stokes cases.  In the optimistic-Stokes case the estimate becomes
-% $\Delta t < O(U^{-1} \Delta x)$
 \begin{equation}
@@
 On the other hand, the per-step expense is much greater because nontrivial equations, indeed a free-boundary problem, must be solved.
-FIXME for the unconditionally-stable implicit geometric-update scheme in the SIA, as demonstrated by \cite{Bueler2016}, $O(n^{1+\beta})$ work; it is not multigrid; indeed the steady-state SIA solver in \cite{Bueler2016} allows $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence
+FIXME for the unconditionally-stable implicit geometric-update scheme in the SIA $O(n^{1+\beta})$ work; \cite{Bueler2016} simplified dome geometry case has $\beta=1.8$, which is not adequate; it is not multigrid
+% see data/bueler.perf and data/scalingbueler.m
+the steady-state SIA solver in \cite{Bueler2016} allows $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence
 

### 2022-06-25 paper/data/scalingbueler.m

@@ -1,0 +1,9 @@
+% generate serial scaling for bueler 2016
+% LU solve of Newton step equations
+% see sia-fve/petsc/verif/domeperf.sh
+% see generated file bueler.perf
+
+n = [18 36 72 144 288].^2;
+flops = [1.776e7 1.749e8 2.374e9 2.608e10 2.927e11]
+polyfit(log(n),log(flops),1)
+%ans = 1.76186   6.45957

### 2022-06-25 paper/data/scalingisaac.m

@@ -1,0 +1,6 @@
+% from Table 8.1 in issac et al 2015
+its = [7*66 10*75 11*90]
+polyfit(log([1 8 64]),log(its),1)
+%ans = 0.18326   6.17004
+polyfit(log([7e6 51e6 383e6]),log(its),1)
+%ans = 0.19030   3.17188

### 2022-06-25 paper/perfmod.tex
edits first two sections

@@
 \abstract{High spatial resolution numerical ice sheet models, with horizontal mesh/grid resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.
-These models usually require many time steps, usually shorter than climate-coupling time scales, because ice thickness is updated following each velocity solution.
-That is, practical performance in the high-resolution limit is degraded by the stability conditions imposed by such explicit time-stepping.
-In this short note, regarding the shallow ice approximation and Stokes models as stress balance end members, we attempt to clarify numerical model performance by quantifying the simulation cost per model year in terms of the mesh resolution and number of degrees of freedom.
-The performance of coming generations of implicit time-stepping numerical models is then assessed.
-Our main result, Table \ref{tab:performancemodel}, emphasizes the key role played by solver performance in Stokes models.}
+These models usually require many time steps, shorter than climate-coupling time scales, because ice thickness is updated following each velocity solution.
+High-resolution performance is degraded by the stability restrictions of such explicit time-stepping.
+This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, attempts to clarify numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.
+The prospective performance of implicit time-stepping numerical models is assessed.
+The main result, Table \ref{tab:performancemodel}, emphasizes the key role played by the performance of stress-balance and implicit-step solvers.}
 
@@
 
-Numerical ice sheet (glacier) models with evolving ice geometry are now in routine use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and quantification of long-term glacial erosion \citep{SeguinotDelaney2021}, among other applications.
-It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows suggest the need for finer resolution.
+Numerical ice sheet (glacier) models with evolving ice geometry are now in routine use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.
+It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows need finer resolution.
 Whether using local mesh refinement \citep[for example]{Hoffmanetal2018} or not, resolutions of a few kilometers to one kilometer \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used at ice sheet scale.
@@
 Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, to the non-shallow and non-hydrostatic Stokes approximation.
-With very few exceptions, however, current models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the last-computed velocity field.
-Therefore ice thickness and surface elevation, which are functionally-equivalent variables for this purpose, are updated after the velocity is fixed.
-This geometry-update operation, using the mass continuity or surface kinematical equations \citep{GreveBlatter2009}, enables interaction between the ice sheet and the surrounding climate.
+With very few exceptions, however, current models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the just-computed velocity field.
+Therefore ice thickness and surface elevation, functionally-equivalent variables for this purpose, are updated after velocity is fixed.
+Interaction between the ice sheet and the surrounding climate occurs during the geometry-update operation via the mass continuity or surface kinematical equations \citep{GreveBlatter2009}.
 Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes in particular.
 
-These standard schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use the velocity computed from the geometry at the previous time step \citep[for example]{Chengetal2017}.}  Now, at least for simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
-The stability limitations of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the time-step limits of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening the steps \citep{LofgrenAhlkronaHelanow2021}.
+Thus these standard schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use the velocity computed from the geometry at the previous time step \citep[for example]{Chengetal2017}.}  At least for simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}, and stability conditions of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the time-step limits of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.
 
-However, actual \emph{implicit} time-stepping should also be considered.
-Here the velocity and geometry are updated simultaneously by solving a coupled mass-momentum problem.
+However, actual \emph{implicit} time-stepping \citep{Bueler2021,LeVeque2007} should also be considered.
+Here the velocity and geometry are updated by simultaneously solving mass and momentum conservation equations.
 While implicit time-stepping always requires the solution of systems of equations at each step \citep{LeVeque2007}, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}.
@@
 
-A successful implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.
-The same reference shows how the steady-state problem \citep[see][for example]{JouvetBueler2012} can often be solved; observe that steady-state solutions correspond to an implicit time step of infinite duration.
-The corresponding problem for the Stokes equations has not, to the author's knowledge, yet been attempted.
+An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}, and the same reference shows how steady-state problem \citep[compare][]{JouvetBueler2012} can often be solved, demonstrating unconditional stability.
+(Observe that steady-state solutions correspond to an implicit time step of infinite duration.)  The corresponding problem for the Stokes equations has not, to the author's knowledge, yet been attempted.
 However, important early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} to solve for a free boundary illuminates some of the techniques and difficulties needed to make such a strategy work for a membrane-stress-resolving balance.
 
-This is the complicated context in which the current note relates time-stepping and stress-balance choices to computational effort.
+This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.
+A simplified performance (computational work) analysis can expose the most important considerations and trade-offs.
 While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping is applied, along with advanced solver techniques like multigrid \citep{Briggsetal2000}, this belief should be inspected quantitatively to the extent possible.
-A simplified performance (computational effort) analysis can expose the most important considerations and trade-offs.
 
@@
 
-The reader is assumed to have access to the SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.
+The reader is assumed to have access to the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.
 These (continuum) models are regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}.
@@
 However, before presenting a performance analysis, the form of the mass continuity equation must be examined.
-For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, this equation says
+For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged horizontal velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, this equation says
 \begin{equation}
@@
 This is because ice flows dominantly downhill.
-Indeed, ice sheet flow has no characteristic curves, as would a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.
-Thus, as addressed by linearized analysis \citep{Chengetal2017,Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of advective and diffusive mechanisms.
+Indeed, ice sheet flow has no characteristic curves, as would \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.
+Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.
 
-Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation for bed elevation $b(\bx)$.
-If a numerical thickness perturbation grows then the velocity $\bu$ often responds by increasing in a direction close to downhill ($-\gradx s$), a direction often correlated to $-\gradx H$.
-In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, but the gravitational driving stress is always along $\gradx s$.
-Equation \eqref{eq:masscontinuity} therefore has velocity $\bu$ which is really a non-local function ``$\bu(H,\grad_x s)$'', though a stress balance solution is required to evaluate this function.
-A numerical instability occurs when the ice thickness under/over-shoots its correct (i.e.~continuum) value because the numerical velocity response is too strong.
+Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.
+Under a numerical thickness perturbation the velocity $\bu$ often responds by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.
+In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational driving stress source term is along $\gradx s$.
+In fact equation \eqref{eq:masscontinuity} has velocity $\bu$ which is a non-local function $\bu(H,\grad_x s)$, though a stress balance solution is required to evaluate it.
+A numerical instability occurs when the ice thickness under/over-shoots its correct value because the numerical velocity response from evaluating this non-local function is too strong.
 
 \newcommand{\nn}{\text{n}}
-The mass continuity equation is not wholely diffusive for Stokes dynamics, but the diffusive nature dominates in the small-parameter limit which generates the SIA.
-In the isothermal case we see the following, revised mass continuity equation:
+Neither is the mass continuity equation wholely diffusive for Stokes dynamics.
+A diffusive description is valid, however, in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}.
+In the isothermal case we see a revised mass continuity equation:
 \begin{equation}
@@
 Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity and $C$ is a positive constant.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ where $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.
-Typically $\nn\approx 3$.}
-
-Equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics.
-However, the same diffusivity $d$, an essentially geometric quantity, can be computed.
-For grounded ice sheets one observes that large values of $d$ often indicate locations of unstable mode growth if explicit time-steps are chosen too large.
+Typically $\nn\approx 3$.}  While equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed.
+For grounded ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.
 

### 2022-06-26 paper/data/scaling.m
show scaling graphically in one code

@@ -1,0 +1,34 @@
+% generate serial scaling for bueler 2016 and isaac et al 2015
+% bueler:  LU solve of Newton step equations; see sia-fve/petsc/verif/domeperf.sh; see generated file bueler.perf
+% isaac: AMG-Krylov-Newton Stokes solve; see table 8.1 in paper
+
+nb = [18 36 72 144 288].^2;
+flopsb = [1.776e7 1.749e8 2.374e9 2.608e10 2.927e11];
+pb = polyfit(log(nb),log(flopsb),1);
+%ans = 1.76186   6.45957
+
+ni = [7e6 51e6 383e6];
+itsi = [7*66 10*75 11*90];
+%polyfit(log([1 8 64]),log(its),1) --> ans = 0.18326   6.17004
+pi = polyfit(log(ni),log(itsi),1);
+%ans = 0.19030   3.17188
+
+subplot(2,1,1)
+loglog(nb, flopsb, 'bo')
+xlabel('degrees of freedom')
+ylabel('flops')
+title('bueler')
+hold on
+loglog(nb, exp(pb(1) * log(nb) + pb(2)), 'b--')
+hold off
+legend(sprintf('O(n^{%.2d})',pb(1)))
+
+subplot(2,1,2)
+loglog(ni, itsi, 'ro')
+xlabel('degrees of freedom')
+ylabel('krylov iterations')
+title('isaac')
+hold on
+loglog(ni, exp(pi(1) * log(ni) + pi(2)), 'r--')
+hold off
+legend(sprintf('O(n^{%.2d})',pi(1)))

### 2022-06-26 paper/data/scalingbueler.m

@@ -1,9 +1,0 @@
-% generate serial scaling for bueler 2016
-% LU solve of Newton step equations
-% see sia-fve/petsc/verif/domeperf.sh
-% see generated file bueler.perf
-
-n = [18 36 72 144 288].^2;
-flops = [1.776e7 1.749e8 2.374e9 2.608e10 2.927e11]
-polyfit(log(n),log(flops),1)
-%ans = 1.76186   6.45957

### 2022-06-26 paper/data/scalingisaac.m

@@ -1,6 +1,0 @@
-% from Table 8.1 in issac et al 2015
-its = [7*66 10*75 11*90]
-polyfit(log([1 8 64]),log(its),1)
-%ans = 0.18326   6.17004
-polyfit(log([7e6 51e6 383e6]),log(its),1)
-%ans = 0.19030   3.17188

### 2022-06-26 paper/data/scaling.m
minor: clarify scaling

@@ -17,18 +17,18 @@ subplot(2,1,1)
 loglog(nb, flopsb, 'bo')
 xlabel('degrees of freedom')
 ylabel('flops')
-title('bueler')
+title('Bueler')
 hold on
 loglog(nb, exp(pb(1) * log(nb) + pb(2)), 'b--')
 hold off
-legend(sprintf('O(n^{%.2d})',pb(1)))
+legend(sprintf('O(n^{%.3d})',pb(1)))
 
 subplot(2,1,2)
 loglog(ni, itsi, 'ro')
 xlabel('degrees of freedom')
-ylabel('krylov iterations')
-title('isaac')
+ylabel('total krylov iterations')
+title('Isaac')
 hold on
 loglog(ni, exp(pi(1) * log(ni) + pi(2)), 'r--')
 hold off
-legend(sprintf('O(n^{%.2d})',pi(1)))
+legend(sprintf('O(n^{%.3d})',pi(1)))

### 2022-06-26 paper/perfmod.tex
edits

@@
 $\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
-$\beta$     & one implicit SIA geometry-update requires $O(n^{1+\beta})$ work \\
-$\gamma$    & one implicit Stokes geometry/velocity solution requires $O(n^{1+\gamma})$ work \\
+$\beta$     & one implicit SIA geometry-update solution requires $O(n^{1+\beta})$ work \\
+$\gamma$    & one implicit Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
 $D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
@@
 
-We parameterize simulation performance using either $\Delta x$, the representative horizontal mesh/grid cell diameter, i.e.~the resolution, or $n$, the number of nodes or vertices in the horizontal mesh or grid.
-The model domain is of (horizontal) width $L$ so $\Delta x = O(L n^{-1/2})$; there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L n^{-1})$, but our analysis focusses on spatially-3D models with map-plane horizontal meshes or grids.}  Table \ref{tab:notation} lists the parameters used in our analysis.
-Big-O notation applies only in the high resolution limit where $\Delta x \to 0$ and $n\to \infty$.
+Table \ref{tab:notation} lists the parameters used in our simulation performance  analysis.
+Our primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $n$, the number of nodes (vertices) in the horizontal mesh.
+\emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $n\to\infty$ limits, and big-O notation is only used in this limit.
+If the model domain is of width $L$ then these parameters are related by
+\begin{equation}
+\Delta x = O\left(\frac{L}{\sqrt{n}}\right) \quad \leftrightarrow \quad n = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
+\end{equation}
+That is, for a given value of $L$ there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L n^{-1})$.
+However, our analysis addresses spatially-3D models with map-plane horizontal meshes.}
 
-A standard numerical ice sheet model uses $n$ ice thickness or surface elevation variables, the number of \emph{degrees of freedom}, and it requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This bounded-vertical-nodes assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.
+A numerical ice sheet model uses $n$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.
+It requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.
 However, the \cite{IsaacStadlerGhattas2015} Stokes and \cite{BrownSmithAhmadia2013} higher-order multigrid solvers use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.
-Ice sheet models also have $O(n)$ velocity variables, but these are not state variables because a very-viscous stress balance can recompute velocity as a function of (true) state variables.
+Ice sheet models also have $O(n)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.
 

### 2022-06-26 paper/perfmod.tex
replace n->m to avoid Glen clash

@@
 $L$         & width of simulation domain & km \\
-$n$         & degrees of freedom: number of nodes in the horizontal mesh \\
+$m$         & degrees of freedom: number of nodes in the horizontal mesh \\
 $q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
@@
 Table \ref{tab:notation} lists the parameters used in our simulation performance  analysis.
-Our primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $n$, the number of nodes (vertices) in the horizontal mesh.
-\emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $n\to\infty$ limits, and big-O notation is only used in this limit.
+Our primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.
+\emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits, and big-O notation is only used in this limit.
 If the model domain is of width $L$ then these parameters are related by
 \begin{equation}
-\Delta x = O\left(\frac{L}{\sqrt{n}}\right) \quad \leftrightarrow \quad n = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
+\Delta x = O\left(\frac{L}{\sqrt{m}}\right) \quad \leftrightarrow \quad m = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
 \end{equation}
-That is, for a given value of $L$ there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L n^{-1})$.
+That is, for a given value of $L$ there are $O(\sqrt{m})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$.
 However, our analysis addresses spatially-3D models with map-plane horizontal meshes.}
 
-A numerical ice sheet model uses $n$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.
-It requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.
-However, the \cite{IsaacStadlerGhattas2015} Stokes and \cite{BrownSmithAhmadia2013} higher-order multigrid solvers use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.
-Ice sheet models also have $O(n)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.
+A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.
+It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.
+However, the \cite{IsaacStadlerGhattas2015} Stokes and \cite{BrownSmithAhmadia2013} higher-order multigrid solvers use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.
+Ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.
 
@@
 Regardless of time-stepping design, for each explicit time step the velocity (or velocity/pressure) solution of the stress balance equations requires effort, namely the solution of (spatial) partial differential equations.
-For the Stokes model we suppose one such solution requires $O(n^{1+\alpha})$ work, which we describe the the number floating point operations in our simple performance analysis.
+For the Stokes model we suppose one such solution requires $O(m^{1+\alpha})$ work, which we describe the the number floating point operations in our simple performance analysis.
 
@@
 % see data/scalingisaac.m
-A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ also close to zero for certain simplified geometries.\footnote{The \cite{IsaacStadlerGhattas2015} and \cite{BrownSmithAhmadia2013} results also show good parallel scaling, something not considered in our simplified analysis.}  At the even easier end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula, $O(1)$ work; a velocity solution requires only $O(n)$ work ($\alpha=0$).
+A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ also close to zero for certain simplified geometries.\footnote{The \cite{IsaacStadlerGhattas2015} and \cite{BrownSmithAhmadia2013} results also show good parallel scaling, something not considered in our simplified analysis.}  At the even easier end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula, $O(1)$ work; a velocity solution requires only $O(m)$ work ($\alpha=0$).
 
-When analyzing solver scaling in this simple manner, one must be aware that the constant in $O(n^{1+\alpha})$ can be very large.  Furthermore it will vary greatly based on solver design.
+When analyzing solver scaling in this simple manner, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large.  Furthermore it will vary greatly based on solver design.
 
-Regardless of the stress balance source, an explicit time-stepping scheme applies the mass continuity equation to update the ice thickness using $O(n)$ work.
+Regardless of the stress balance source, an explicit time-stepping scheme applies the mass continuity equation to update the ice thickness using $O(m)$ work.
 That is, once the velocity is computed for the geometry from the previous time step, old thickness values are replaced by new ones by a formula needing only $O(1)$ work per mesh node.
@@
 \emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{pessimistic} \\ \hline
-explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, n^2}{L^2}}${\Huge \strut} \\
-explicit & Stokes & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U n^{1.5+\alpha}}{L}}${\Huge \strut}\phantom{x} & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,n^{2+\alpha}}{L^2}}$ \\
-implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, n^{1+\beta}}${\Huge \strut} \\
-implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, n^{1+\gamma}}${\Huge \strut}
+explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}${\Huge \strut} \\
+explicit & Stokes & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}${\Huge \strut}\phantom{x} & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
+implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}${\Huge \strut} \\
+implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}${\Huge \strut}
 \end{tabular}
 }
-\caption{Asymptotic scaling of computational work, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $n\to\infty$.
+\caption{Asymptotic scaling of computational work, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$.
 See Table \ref{tab:notation} for notation.}
@@
 Now suppose that a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.
-Recalling the scaling of $\Delta x$ with $n$, namely that $\Delta x = O(L n^{-1/2})$, equivalently $n=O(L^2 \Delta x^{-2})$, for explicit schemes we may write the stability restriction as a required number of time-steps per model year in order to maintain stability, namely
+Recalling the scaling of $\Delta x$ with $m$, namely that $\Delta x = O(L m^{-1/2})$, equivalently $m=O(L^2 \Delta x^{-2})$, for explicit schemes we may write the stability restriction as a required number of time-steps per model year in order to maintain stability, namely
 \begin{equation}
-\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D n}{L^2}} \label{eq:explicitsiarequired}
+\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}} \label{eq:explicitsiarequired}
 \end{equation}
@@
 \begin{equation}
-\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U n^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
+\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
 \end{equation}
@@
 On the other hand, the per-step expense is much greater because nontrivial equations, indeed a free-boundary problem, must be solved.
-FIXME for the unconditionally-stable implicit geometric-update scheme in the SIA $O(n^{1+\beta})$ work; \cite{Bueler2016} simplified dome geometry case has $\beta=1.8$, which is not adequate; it is not multigrid
+FIXME for the unconditionally-stable implicit geometric-update scheme in the SIA $O(m^{1+\beta})$ work; \cite{Bueler2016} simplified dome geometry case has $\beta=1.8$, which is not adequate; it is not multigrid
 % see data/bueler.perf and data/scalingbueler.m
@@
 
-For an implicit Stokes simulation the geometry/velocity solve is assumed to be, in the absense of constraining research, $O(n^{1+\gamma})$ for some $\gamma>0$ to be determined.
+For an implicit Stokes simulation the geometry/velocity solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma>0$ to be determined.
 The nMCD method proposed in [Bueler 2022] is expensive but multigrid, so it may turn out that $\gamma$ is small.

### 2022-06-26 paper/perfmod.tex
the first three pages are as solid as they are likely to be

@@
 
-Ice sheet models involve climate interactions (coupling), especially via surface mass balance, on time scales which are dominated by an annual cycle and certain longer scales.
-We define $q$ as the number of ice-dynamical time steps per year needed to capture this coupling.
-Some typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly coupling frequency, respectively.
-Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated from an ice velocity solution and the latest surface mass balance values.
+Ice sheet models resolve climate interactions (coupling), especially via surface mass balance, on time scales which are dominated by an annual cycle and certain longer scales.
+Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.
+Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly coupling frequency, respectively.
+Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated using velocity, i.e.~via solution of the mass continuity or surface kinematical equation.
 
-As already noted, current-technology ice sheet models use explicit time-stepping which is only conditionally stable \citep{LeVeque2007}.
-For the spatial resolutions used in present-day scientific applications, we observe that maintenance of explicit time-stepping stability requires time steps substantially shorter than $1/q$ model years.
-For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value from the diffusivity formula in equation \eqref{eq:siamasscontinuity}.
+As already noted, current-technology ice sheet models use explicit time-stepping which is only conditionally stable.
+For the spatial resolutions used in present-day scientific applications, we observe that maintenance of explicit time-stepping stability requires steps substantially shorter than $1/q$ model years.
 
-For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but we propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, computed from simulated ice velocities, represents the \emph{optimistic} paradigm.
-The corresponding \emph{pessimistic} paradigm enforces $\Delta t < O(D^{-1} \Delta x^2)$, using diagnostic diffusivity values computed in the SIA manner.
+For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value (from equation \eqref{eq:siamasscontinuity}).
+For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but we propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, or computed from modeled ice velocities, represents the \emph{optimistic} paradigm.
+The corresponding \emph{pessimistic} paradigm enforces $\Delta t < O(D^{-1} \Delta x^2)$, using (otherwise diagnostic) diffusivity values computed in the SIA manner.
 
@@
 Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}.
-The optimistic time step is supported for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
+The optimistic time step is supported for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
 
@@
 This restriction reflects the simulation purpose and not the maintenance of stability.
-Thus, for an implicit scheme the frequency of climate coupling determines the time-stepping simulation cost, through the need to solve the coupled mass/momentum equations with frequency $q$.
+Indeed, for an implicit scheme the frequency of climate coupling determines the time-stepping simulation cost, according to the solution cost of coupled mass and momentum equations.
 
-Regardless of time-stepping design, for each explicit time step the velocity (or velocity/pressure) solution of the stress balance equations requires effort, namely the solution of (spatial) partial differential equations.
-For the Stokes model we suppose one such solution requires $O(m^{1+\alpha})$ work, which we describe the the number floating point operations in our simple performance analysis.
+For each explicit time step the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.
+For the Stokes model we suppose one such solution requires $O(m^{1+\alpha})$ floating point operations.\footnote{Many considerations are suppresed in our simple analysis, including memory latency, memory bandwidth, and process/thread/GPU parallelism.}
 
-Of course, the power $\alpha$ depends on the design of the stress balance solver.
+The power $\alpha$ depends on the design of the stress balance solver.
 A direct linear algebra Stokes solver implementation might yield $\alpha=3$, or $\alpha \approx 2$ when exploiting sparsity, but a multigrid method, such as the algebraic multigrid \citep{Trottenbergetal2001} solver by \cite{IsaacStadlerGhattas2015}, reduces $\alpha$.

### 2022-06-26 paper/perfmod.tex
apparently completed performance analysis section

@@
 $\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
-$\beta$     & one implicit SIA geometry-update solution requires $O(n^{1+\beta})$ work \\
-$\gamma$    & one implicit Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
+$\beta$     & one implicit SIA geometry-update (and velocity) solution requires $O(n^{1+\beta})$ work \\
+$\gamma$    & one implicit, coupled Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
 $D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
@@
 A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.
-It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.
-However, the \cite{IsaacStadlerGhattas2015} Stokes and \cite{BrownSmithAhmadia2013} higher-order multigrid solvers use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.
+It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012}.
+However, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015} models use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.
 Ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.
@@
 For each explicit time step the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.
-For the Stokes model we suppose one such solution requires $O(m^{1+\alpha})$ floating point operations.\footnote{Many considerations are suppresed in our simple analysis, including memory latency, memory bandwidth, and process/thread/GPU parallelism.}
+We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.
+A Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited.
+By contrast, a multigrid method \citep{Trottenbergetal2001} will greatly reduce $\alpha$ if correctly applied.
+Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the number of preconditioned Newton-Krylov iterations \citep{Bueler2021} grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
+% see data/scaling.m
+The algebraic-multigrid based solver in \cite{Lengetal2012} may have similar algorithmic scaling, but reported results do not constrain $\alpha$.
+A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.\footnote{The \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Lengetal2012} results all show good parallel scaling.}  Note that $\alpha=0$ describes an \emph{optimal} solver  \citep{Bueler2021} in the language of scientific computation.
 
-The power $\alpha$ depends on the design of the stress balance solver.
-A direct linear algebra Stokes solver implementation might yield $\alpha=3$, or $\alpha \approx 2$ when exploiting sparsity, but a multigrid method, such as the algebraic multigrid \citep{Trottenbergetal2001} solver by \cite{IsaacStadlerGhattas2015}, reduces $\alpha$.
-Their Antarctic ice sheet results show Newton-Krylov iterations \citep{Bueler2021} growing only slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
-% see data/scalingisaac.m
-A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ also close to zero for certain simplified geometries.\footnote{The \cite{IsaacStadlerGhattas2015} and \cite{BrownSmithAhmadia2013} results also show good parallel scaling, something not considered in our simplified analysis.}  At the even easier end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula, $O(1)$ work; a velocity solution requires only $O(m)$ work ($\alpha=0$).
+At the easiest end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula.
+Therefore, for the SIA a velocity solution requires only $O(m)$ work: $\alpha=0$.
 
-When analyzing solver scaling in this simple manner, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large.  Furthermore it will vary greatly based on solver design.
+When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and will vary depending on solver design.
+In fact, many considerations are suppressed in our simple flops-based analysis of algorithmic scaling, and actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.
 
-Regardless of the stress balance source, an explicit time-stepping scheme applies the mass continuity equation to update the ice thickness using $O(m)$ work.
-That is, once the velocity is computed for the geometry from the previous time step, old thickness values are replaced by new ones by a formula needing only $O(1)$ work per mesh node.
+Regardless of the stress balance source, an explicit time-stepping scheme then applies the mass continuity equation to update the ice thickness using $O(m)$ work.
+That is, once the velocity is computed for the geometry from the previous time step, an explicit scheme replaces old thickness values by new ones using a pointwise formula.\footnote{Additional computation needed to remesh the updated geometry, a strongly design-dependent cost, is omitted here.}
 
@@
 \begin{tabular}{llll}
-\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{pessimistic} \\ \hline
+\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{[pessimistic stability]} \\ \hline
 explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}${\Huge \strut} \\
@@
 }
-\caption{Asymptotic scaling of computational work, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$.
+\caption{Asymptotic estimates of algorithmic scaling, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$.
 See Table \ref{tab:notation} for notation.}
@@
 
-Now suppose that a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.
-Recalling the scaling of $\Delta x$ with $m$, namely that $\Delta x = O(L m^{-1/2})$, equivalently $m=O(L^2 \Delta x^{-2})$, for explicit schemes we may write the stability restriction as a required number of time-steps per model year in order to maintain stability, namely
+Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.
+Recalling the scaling of $\Delta x$ with $m$, in \eqref{eq:paramrelation}, we now write time step restrictions as required numbers of time-steps per model year in order to maintain stability, as a function of horizontal resolution $\Delta x$ or degrees of freedom $m$.
+In the explicit SIA and pessimistic-Stokes cases we have
 \begin{equation}
-\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}} \label{eq:explicitsiarequired}
+\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
 \end{equation}
-in the SIA and pessimistic-Stokes cases.  In the optimistic-Stokes case the estimate becomes
+The explicit, optimistic-Stokes estimate becomes
 \begin{equation}
@@
 \end{equation}
-
-The number of time steps per model year is multiplied by the per-step computational cost to give a work estimate for each model year in a simulation.
+The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.
 The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.
 
-For unconditionally-stable implicit methods, however, $\Delta t$ does not scale with grid resolution; it has fixed value $\Delta t = 1/q$ determined by the need to resolve climatic interactions.
+Unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve climatic interactions.
 On the other hand, the per-step expense is much greater because nontrivial equations, indeed a free-boundary problem, must be solved.
-FIXME for the unconditionally-stable implicit geometric-update scheme in the SIA $O(m^{1+\beta})$ work; \cite{Bueler2016} simplified dome geometry case has $\beta=1.8$, which is not adequate; it is not multigrid
+We parameterize the flops of such coupled velocity and geometry-update solutions as $O(m^{1+\beta})$, with the understanding that $\beta \ge \alpha$ for the $\alpha$ value associated to the corresponding fixed-geometry velocity solution.
+
+To our knowledge there is only one implemented, unconditionally-stable, fully-implicit geometric-update solver, namely the SIA scheme in \cite{Bueler2016}.
+For simplified dome geometry this method, based on Newton steps solved via direct linear algebra, shows $\beta=1.8$.
 % see data/bueler.perf and data/scalingbueler.m
-the steady-state SIA solver in \cite{Bueler2016} allows $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence
+The same work shows how a steady-state ($\Delta t=\infty$) SIA solution is practical for many geometries, but in any case convergence is robust for time steps of years to centuries on kilometer-scale grids, even using realistic, thus irregular, bed topographies.
 
-For an implicit Stokes simulation the geometry/velocity solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma>0$ to be determined.
-The nMCD method proposed in [Bueler 2022] is expensive but multigrid, so it may turn out that $\gamma$ is small.
+The hybrid \cite{Brinkerhoffetal2017} flowline solver is implicit but apparently subject to a time-stepping condition when the margin moves, and its algorithmic scaling is unknown.
 
-FIXME emphasize the estimate reported in Table \ref{tab:performancemodel} must be multiplied by a potentially very large scheme-dependent constant.
+For an implicit Stokes simulation, a prospective coupled velocity and geometry-update solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma>0$ to be determined.
+Again $\beta \ge \alpha$ is automatic.
+It is not clear if $\gamma\ge \beta$ in general, which might be the expectation, but to the author's knowledge there are no implemented cases to measure.
+% The nMCD method proposed in [Bueler 2022] is expensive but multigrid, so it may turn out that $\gamma$ is small.
+
+These comments complete Table \ref{tab:performancemodel}.
+Again one must be aware that all asymptotic performance estimates in the Table involve a large, scheme-dependent constant, with this especially true of implicit schemes.
 
@@
 
-FIXME The real promise of multigrid methods for the implicit, coupled geometry-update plus velocity solution problems is to make $\beta,\gamma$ seriously smaller.
-Specifically $\gamma < 1$ might be regarded as the goal of the next few decades of work on Stokes models.
+FIXME The real promise of multigrid methods for the implicit, coupled geometry-update plus velocity solution problems; large constants are acceptable if $\beta,\gamma$ are less than 2, for example.
+Specifically $\gamma < 1$ might be regarded as the goal of the next decades of research on numerical ice sheet models using Stokes dynamics.
 However, this aspiration is long-term because work is barely started on it.

### 2022-06-28 paper/perfmod.tex
attempt a final paragraph

@@
 
-FIXME The real promise of multigrid methods for the implicit, coupled geometry-update plus velocity solution problems; large constants are acceptable if $\beta,\gamma$ are less than 2, for example.
-Specifically $\gamma < 1$ might be regarded as the goal of the next decades of research on numerical ice sheet models using Stokes dynamics.
-However, this aspiration is long-term because work is barely started on it.
+The real promise of near-optimal methods is revealed when implicit, coupled mass/momentum solvers are considered.
+That is, if a multigrid method can be designed which yields $\beta\approx 0$ for SIA models or $\gamma\approx 0$ for Stokes models then numerical ice sheet modeling tools will be qualitatively faster for many tasks.
+Large constants would be acceptable if $\beta,\gamma$ are small.
+Specifically, and more modestly, a $\gamma < 1$ implicit, essentially unconditionally-stable, Stokes time-stepping method might be regarded as an appropriate goal for the next decades of research on numerical ice sheet models.
+Such a numerical ice sheet model would scale with resolution as well as current explicit SIA models, at $O(\Delta x^{-4})$ rate or better, and yet avoid shallow approximations.
+In fact the same applies to the construction of capable higher-order or other membrane-stress resolving solvers.
+These aspirations are admittedly long-term because, to the author's knowledge, no progress has yet been made.
 

### 2022-06-28 paper/perfmod.tex
do not forget tuminaro

--- 2022-06-26 paper/perfmod.tex
@@
 A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.
-It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012}.
-However, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015} models use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.
+It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.
+The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.
 Ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.
+
+The assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012,
+Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.
+For example, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016} solvers use 3D refinement, though with various modifications regarding vertical refinement.
+This and other mesh/solver details cannot be pursued here.
 
@@
 A Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited.
+
 By contrast, a multigrid method \citep{Trottenbergetal2001} will greatly reduce $\alpha$ if correctly applied.
-Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the number of preconditioned Newton-Krylov iterations \citep{Bueler2021} grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
+Antarctic ice sheet results by \cite[Table 8.1]{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$.
+\cite[Table 7.5]{Tuminaroetal2016} reports total algebraic-multigrid-preconditioned iterations growing as $\alpha \approx 0.05$ in a higher-order velocity solver for the Greenland ice sheet.
+(Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.)
 % see data/scaling.m
-The algebraic-multigrid based solver in \cite{Lengetal2012} may have similar algorithmic scaling, but reported results do not constrain $\alpha$.
-A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.\footnote{The \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Lengetal2012} results all show good parallel scaling.}  Note that $\alpha=0$ describes an \emph{optimal} solver  \citep{Bueler2021} in the language of scientific computation.
+An algebraic-multigrid based solver by \cite{Lengetal2012} may have similar algorithmic scaling, but reported results do not constrain $\alpha$.
+A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.\footnote{The \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Lengetal2012,Tumarinoetal2016} results all show good parallel scaling.}  Note that $\alpha=0$ describes an \emph{optimal} solver  \citep{Bueler2021} in the language of scientific computation.
 
@@
 
-FIXME The real promise of multigrid methods for the implicit, coupled geometry-update plus velocity solution problems; large constants are acceptable if $\beta,\gamma$ are less than 2, for example.
-Specifically $\gamma < 1$ might be regarded as the goal of the next decades of research on numerical ice sheet models using Stokes dynamics.
-However, this aspiration is long-term because work is barely started on it.
+The real promise of near-optimal methods is revealed when implicit solvers are considered, ones which simultanously satisfy the mass and momentum equations at each time step.
+If a multigrid method can be designed which yields $\gamma\approx 0$ for a Stokes model, for example, then numerical ice sheet modeling tools will be qualitatively faster for many tasks, and a large constant would be acceptable.
+Specifically, and more modestly, a $\gamma < 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method is a goal for the next decades of research on numerical ice sheet models.
+Such a tool would scale at $O(\Delta x^{-3})$, better than current explicit SIA models, avoiding shallow approximations while computing time steps with a frequency determined only by the scientific need to resolve climate coupling.
+The same goal applies to higher-order, or other membrane-stress-resolving, solvers.
+These aspirations are admittedly long-term because, to the author's knowledge, no progress has yet been made on such implicit designs.
 

### 2022-06-28 paper/data/scaling.m

@@ -1,6 +1,7 @@
-% generate serial scaling for bueler 2016 and isaac et al 2015
-% bueler:  LU solve of Newton step equations; see sia-fve/petsc/verif/domeperf.sh; see generated file bueler.perf
-% isaac: AMG-Krylov-Newton Stokes solve; see table 8.1 in paper
+% generate serial scaling for bueler 2016 and isaac et al 2015 and tumarino et al 2016
+% bueler:  LU solve of Newton step equations for dome; see sia-fve/petsc/verif/domeperf.sh; see generated file bueler.perf
+% isaac: AMG-Krylov-Newton Stokes solve for antarctica; see table 8.1 in paper
+% tumarino: MDSC–AMG-Krylov-Newton higher-order solve for greenland; see table 7.5 in paper; area of Greenland taken from Searise data: 300x560 5km grid
 
 nb = [18 36 72 144 288].^2;
 flopsb = [1.776e7 1.749e8 2.374e9 2.608e10 2.927e11];
@@ -8,12 +9,16 @@ pb = polyfit(log(nb),log(flopsb),1);
 %ans = 1.76186   6.45957
 
 ni = [7e6 51e6 383e6];
-itsi = [7*66 10*75 11*90];
-%polyfit(log([1 8 64]),log(its),1) --> ans = 0.18326   6.17004
+itsi = [7*66 10*75 11*90]; % total Krylov iterations over Newton steps
 pi = polyfit(log(ni),log(itsi),1);
 %ans = 0.19030   3.17188
 
-subplot(2,1,1)
+nt = ((300*560*5^2) / (8^2)) * [1 4 16 64 256];
+itst = [36*13.3 34*12.7 35*12.3 36*13.3 37*17.4]; % total Krylov iterations over Newton steps
+pt = polyfit(log(nt),log(itst),1);
+%ans = 0.05017   5.49297
+
+subplot(3,1,1)
 loglog(nb, flopsb, 'bo')
 xlabel('degrees of freedom')
 ylabel('flops')
@@ -23,7 +28,7 @@ loglog(nb, exp(pb(1) * log(nb) + pb(2)), 'b--')
 hold off
 legend(sprintf('O(n^{%.3d})',pb(1)))
 
-subplot(2,1,2)
+subplot(3,1,2)
 loglog(ni, itsi, 'ro')
 xlabel('degrees of freedom')
[…]

### 2022-06-28 paper/perfmod.tex
Add an ISSM ref.  People are really cagey about their time steps.

@@
 For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value (from equation \eqref{eq:siamasscontinuity}).
-For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but we propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, or computed from modeled ice velocities, represents the \emph{optimistic} paradigm.
+For Stokes or other membrane-stress-resolving dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, or computed from modeled ice velocities \citep[for example]{Fischleretal2022}, represents the \emph{optimistic} paradigm.
 The corresponding \emph{pessimistic} paradigm enforces $\Delta t < O(D^{-1} \Delta x^2)$, using (otherwise diagnostic) diffusivity values computed in the SIA manner.
@@
 Explicit time-stepping with hybrid and higher-order schemes is better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.
-Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}.
-The optimistic time step is supported for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
+Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}, but some models apparently require the user to choose a time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.
+The optimistic time step is supported for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)), and thus some other manner of determining the time step.
 

### 2022-06-28 paper/perfmod.tex
ready for final draft of discussion and conclusions

--- 2022-06-28 paper/perfmod.tex
@@
 
-\abstract{High spatial resolution numerical ice sheet models, with horizontal mesh/grid resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.
-These models usually require many time steps, shorter than climate-coupling time scales, because ice thickness is updated following each velocity solution.
-High-resolution performance is degraded by the stability restrictions of such explicit time-stepping.
+\abstract{Numerical ice sheet models compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.
+At high spatial resolution, with horizontal mesh/grid resolutions of a few kilometers or smaller, these models usually require time steps shorter than climate-coupling time scales because these models update ice thickness after each velocity solution.
+Although precise time-step conditions are often lacking, high-resolution performance is degraded by the stability restrictions of such explicit time-stepping.
 This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, attempts to clarify numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.
-The prospective performance of implicit time-stepping numerical models is assessed.
-The main result, Table \ref{tab:performancemodel}, emphasizes the key role played by the performance of stress-balance and implicit-step solvers.}
+The performance of current-generation explicit time-stepping models is assessed, and then the prospective performance of implicit schemes.
+The main result, Table \ref{tab:performancemodel}, highlights the key role which will be played by the algorithmic scaling of stress-balance and implicit-step solvers.}
 
@@
 Numerical ice sheet (glacier) models with evolving ice geometry are now in routine use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.
-It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows need finer resolution.
-Whether using local mesh refinement \citep[for example]{Hoffmanetal2018} or not, resolutions of a few kilometers to one kilometer \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used at ice sheet scale.
+It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows need yet finer resolution.
+Whether using local mesh refinement \citep[for example]{Fischleretal2022,Hoffmanetal2018} or not, resolutions of two or one kilometers \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used for science at ice sheet scale.
 
-Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, to the non-shallow and non-hydrostatic Stokes approximation.
+Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, up to the non-shallow and non-hydrostatic Stokes approximation.
 With very few exceptions, however, current models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the just-computed velocity field.
 Therefore ice thickness and surface elevation, functionally-equivalent variables for this purpose, are updated after velocity is fixed.
-Interaction between the ice sheet and the surrounding climate occurs during the geometry-update operation via the mass continuity or surface kinematical equations \citep{GreveBlatter2009}.
+Note that the interaction between the ice sheet and the surrounding climate occurs during this geometry-update operation, via the mass continuity or surface kinematical equations \citep{GreveBlatter2009}.
 Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes in particular.
 
-Thus these standard schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use the velocity computed from the geometry at the previous time step \citep[for example]{Chengetal2017}.}  At least for simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}, and stability conditions of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the time-step limits of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.
+These current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use the velocity computed from the geometry at the previous time step \citep[for example]{Chengetal2017}.}  At least for simpler, textbook partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}, and stability conditions of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the explicit time-stepping conditions of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.
 
 However, actual \emph{implicit} time-stepping \citep{Bueler2021,LeVeque2007} should also be considered.
-Here the velocity and geometry are updated by simultaneously solving mass and momentum conservation equations.
-While implicit time-stepping always requires the solution of systems of equations at each step \citep{LeVeque2007}, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}.
+Here the velocity and geometry are updated simultaneously by solving coupled mass and momentum conservation equations.
+While implicit time-stepping always requires the solution of systems of equations at each step, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}, namely the 3D extend of the ice (once the coupled solution converges).
 The problem is of free-boundary type, both in map-plane (horizontal) \citep{SchoofHewitt2013} and (easily-resolved) vertical directions.
 
-An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}, and the same reference shows how steady-state problem \citep[compare][]{JouvetBueler2012} can often be solved, demonstrating unconditional stability.
+An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.
+The same reference shows how the steady-state problem \citep[compare][]{JouvetBueler2012} can often be solved, demonstrating unconditional stability.
 (Observe that steady-state solutions correspond to an implicit time step of infinite duration.)  The corresponding problem for the Stokes equations has not, to the author's knowledge, yet been attempted.
-However, important early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} to solve for a free boundary illuminates some of the techniques and difficulties needed to make such a strategy work for a membrane-stress-resolving balance.
+However, important early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} to solve a free boundary problem illuminates some of the techniques and difficulties needed to make such a strategy work for a membrane-stress-resolving balance.
 
 This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.
-A simplified performance (computational work) analysis can expose the most important considerations and trade-offs.
-While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping is applied, along with advanced solver techniques like multigrid \citep{Briggsetal2000}, this belief should be inspected quantitatively to the extent possible.
+Our simplified performance analysis can expose the most important considerations and trade-offs.
+While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping is applied, along with advanced solver techniques like multigrid \citep{Briggsetal2000}, this belief should be inspected in advance, and quantitatively to the extent possible.
 
@@
 
-The reader is assumed to have access to the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.
+Reader familiarity with the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013} is assumed.
 These (continuum) models are regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}.
@@
 Indeed, ice sheet flow has no characteristic curves, as would \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.
+
 Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.
-
 Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.
-Under a numerical thickness perturbation the velocity $\bu$ often responds by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.
-In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational driving stress source term is along $\gradx s$.
+Under a numerical thickness perturbation the velocity $\bu$ will often respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.
+In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term is along $\gradx s$.
 In fact equation \eqref{eq:masscontinuity} has velocity $\bu$ which is a non-local function $\bu(H,\grad_x s)$, though a stress balance solution is required to evaluate it.
-A numerical instability occurs when the ice thickness under/over-shoots its correct value because the numerical velocity response from evaluating this non-local function is too strong.
+A numerical instability occurs when the ice thickness under/over-shoots its correct value because the numerical velocity response from evaluating this non-local function is too strong for the given time step.
 
 \newcommand{\nn}{\text{n}}
-Neither is the mass continuity equation wholely diffusive for Stokes dynamics.
-A diffusive description is valid, however, in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}.
-In the isothermal case we see a revised mass continuity equation:
+Neither is the mass continuity equation wholely diffusive for stress-complete Stokes dynamics.
+A diffusive description is valid, however, in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}, and we see a revised mass continuity equation:
 \begin{equation}
@@
 \end{equation}
-Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity and $C$ is a positive constant.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ where $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.
+Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity, in the isothermal case, in which $C$ is a positive constant.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ where $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.
 Typically $\nn\approx 3$.}  While equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed.
-For grounded ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.
+For grounded ice sheets one observes that large values of $d$ tend to indicate locations of unstable mode growth if explicit time-steps are chosen too large.
 
@@
 $D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
-$L$         & width of simulation domain & km \\
+$L$         & width of map-plane domain & km \\
 $m$         & degrees of freedom: number of nodes in the horizontal mesh \\
@@
 
-Table \ref{tab:notation} lists the parameters used in our simulation performance  analysis.
-Our primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.
+Table \ref{tab:notation} lists the parameters used in our performance analysis.
+The primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.
 \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits, and big-O notation is only used in this limit.
-If the model domain is of width $L$ then these parameters are related by
+If the map-plane model domain is of width $L$ then these parameters are related by
 \begin{equation}
-\Delta x = O\left(\frac{L}{\sqrt{m}}\right) \quad \leftrightarrow \quad m = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
+\Delta x = O\left(\frac{L}{\sqrt{m}}\right) \quad \text{and} \quad m = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
 \end{equation}
@@
 A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.
-It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.
+It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the mesh/grid has an \emph{a priori} bounded resolution in the vertical direction.
 The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.
-Ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.
+Such ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance computes velocity as a function of the true state variables.
 
-The assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012,
+Our assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012,
 Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.
 For example, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016} solvers use 3D refinement, though with various modifications regarding vertical refinement.
-This and other mesh/solver details cannot be pursued here.
+This and many other mesh/solver details cannot be pursued here.
 
-Ice sheet models resolve climate interactions (coupling), especially via surface mass balance, on time scales which are dominated by an annual cycle and certain longer scales.
+Ice sheet models resolve climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle and certain longer scales.
 Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.
-Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly coupling frequency, respectively.
-Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated using velocity, i.e.~via solution of the mass continuity or surface kinematical equation.
+Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.
+Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated using ice velocity, i.e.~via solution of the mass continuity or surface kinematical equation.
 
@@
 
-For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value (from equation \eqref{eq:siamasscontinuity}).
-For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but we propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, or computed from modeled ice velocities, represents the \emph{optimistic} paradigm.
-The corresponding \emph{pessimistic} paradigm enforces $\Delta t < O(D^{-1} \Delta x^2)$, using (otherwise diagnostic) diffusivity values computed in the SIA manner.
+For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value (equation \eqref{eq:siamasscontinuity}).
+For Stokes or other membrane-stress-resolving dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, represents the \emph{optimistic} paradigm.
+The corresponding \emph{pessimistic} paradigm requires $\Delta t < O(D^{-1} \Delta x^2)$, using a representative diffusivity value $D$ computed in the SIA manner.
 
-Explicit time-stepping with hybrid and higher-order schemes is better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.
-Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}.
-The optimistic time step is supported for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
+Explicit time-stepping with hybrid and higher-order schemes is somewhat better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.
+Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}, but other models apparently require the user to choose a time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.
+The optimistic time step is supported in theory for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)), and thus some other manner of determining the time step.
 
-Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $1/q$ itself.
-This restriction reflects the simulation purpose and not the maintenance of stability.
-Indeed, for an implicit scheme the frequency of climate coupling determines the time-stepping simulation cost, according to the solution cost of coupled mass and momentum equations.
+Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$.
+This restriction reflects the simulation purpose, not maintenance of stability.
+For an implicit scheme the frequency of climate coupling determines the total simulation cost, according to the (large) solution cost, at each time step, of coupled mass and momentum equations.
 
-For each explicit time step the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.
+For each explicit time step of a model using a membrane-stress-resolving balance the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.
 We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.
-A Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited.
+For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited.
 
-By contrast, a multigrid method \citep{Trottenbergetal2001} will greatly reduce $\alpha$ if correctly applied.
-Antarctic ice sheet results by \cite[Table 8.1]{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$.
-\cite[Table 7.5]{Tuminaroetal2016} reports total algebraic-multigrid-preconditioned iterations growing as $\alpha \approx 0.05$ in a higher-order velocity solver for the Greenland ice sheet.
-(Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.)
-% see data/scaling.m
-An algebraic-multigrid based solver by \cite{Lengetal2012} may have similar algorithmic scaling, but reported results do not constrain $\alpha$.
-A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.\footnote{The \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Lengetal2012,Tumarinoetal2016} results all show good parallel scaling.}  Note that $\alpha=0$ describes an \emph{optimal} solver  \citep{Bueler2021} in the language of scientific computation.
+% see data/scaling.m for results in next paragraphs
 
-At the easiest end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula.
-Therefore, for the SIA a velocity solution requires only $O(m)$ work: $\alpha=0$.
+By contrast, a multigrid method \citep{Trottenbergetal2001} can greatly reduce $\alpha$.
+For example, Antarctic ice sheet results by \cite[Table 8.1]{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$.\footnote{The \cite{Lengetal2012} algebraic-multigrid solver may have similar scaling, but reported results do not constrain $\alpha$.}  For higher-order stress balances, \cite{Tuminaroetal2016} report total algebraic-multigrid-preconditioned iterations following $\alpha \approx 0.05$ for the Greenland ice sheet,\footnote{See Table 7.5.
+Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.} and a geometric multigrid method by \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.
 
-When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and will vary depending on solver design.
-In fact, many considerations are suppressed in our simple flops-based analysis of algorithmic scaling, and actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.
+Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling \citep{Bueler2021}.
+At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula.
+Thus an SIA velocity solution requires optimal $O(m)$, $\alpha=0$, flops.
 
-Regardless of the stress balance source, an explicit time-stepping scheme then applies the mass continuity equation to update the ice thickness using $O(m)$ work.
-That is, once the velocity is computed for the geometry from the previous time step, an explicit scheme replaces old thickness values by new ones using a pointwise formula.\footnote{Additional computation needed to remesh the updated geometry, a strongly design-dependent cost, is omitted here.}
+When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, strongly depending on solver design.
+Furthermore, many considerations are suppressed in a flops-based analysis of algorithmic scaling.
+Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.
+For example, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results all show good parallel scaling, something we do not address.
+
+Regardless of the stress balance, an explicit time-stepping scheme then applies the mass continuity equation to update the ice thickness using $O(m)$ work.
+That is, once the velocity is computed for the previous time-step's geometry, an explicit scheme replaces old thickness values by new ones using a pointwise formula.
+Any additional computation needed to remesh the updated geometry, a design-dependent cost, is omitted here.
 
@@
 
+How many flops are needed to simulate one model year?
 Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.
-Recalling the scaling of $\Delta x$ with $m$, in \eqref{eq:paramrelation}, we now write time step restrictions as required numbers of time-steps per model year in order to maintain stability, as a function of horizontal resolution $\Delta x$ or degrees of freedom $m$.
-In the explicit SIA and pessimistic-Stokes cases we have
+We may write time-step restrictions as required numbers of time-steps per model year to maintain stability.
+That is, $\Delta t^{-1}$ is bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.
+Recalling scaling \eqref{eq:paramrelation} of $\Delta x$ with $m$, in the explicit SIA and pessimistic-Stokes cases we have
 \begin{equation}
@@
 
-Unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve climatic interactions.
-On the other hand, the per-step expense is much greater because nontrivial equations, indeed a free-boundary problem, must be solved.
-We parameterize the flops of such coupled velocity and geometry-update solutions as $O(m^{1+\beta})$, with the understanding that $\beta \ge \alpha$ for the $\alpha$ value associated to the corresponding fixed-geometry velocity solution.
+As already explained, unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve climatic interactions.
+On the other hand, the per-step expense is much greater because nontrivial coupled equations, indeed a free-boundary problem, must be solved for the velocity and geometry update simultaneously.
+We parameterize the flops of such coupled solutions as $O(m^{1+\beta})$, with the understanding that $\beta \ge \alpha$, for $\alpha$ associated to a corresponding velocity-only solution, and that a large constant is present.
 
-To our knowledge there is only one implemented, unconditionally-stable, fully-implicit geometric-update solver, namely the SIA scheme in \cite{Bueler2016}.
-For simplified dome geometry this method, based on Newton steps solved via direct linear algebra, shows $\beta=1.8$.
-% see data/bueler.perf and data/scalingbueler.m
-The same work shows how a steady-state ($\Delta t=\infty$) SIA solution is practical for many geometries, but in any case convergence is robust for time steps of years to centuries on kilometer-scale grids, even using realistic, thus irregular, bed topographies.
+% regarding next paragraph, see data/bueler.perf and data/scaling.m
 
-The hybrid \cite{Brinkerhoffetal2017} flowline solver is implicit but apparently subject to a time-stepping condition when the margin moves, and its algorithmic scaling is unknown.
+It appears there is only one implemented, unconditionally-stable, fully-implicit geometric-update solver, namely the SIA scheme in \cite{Bueler2016}.\footnote{The hybrid \cite{Brinkerhoffetal2017} flowline solver is partly implicit but apparently subject to a time-stepping condition when the margin moves.}  For simplified dome geometry this method, based on Newton steps solved via direct linear algebra, and on ice margin determination on a single grid, shows $\beta=1.8$.
+The same work shows how a steady-state ($\Delta t=\infty$) SIA solution is practical for many geometries.
+Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, even using realistic, thus irregular, bed topographies.
 
-For an implicit Stokes simulation, a prospective coupled velocity and geometry-update solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma>0$ to be determined.
-Again $\beta \ge \alpha$ is automatic.
-It is not clear if $\gamma\ge \beta$ in general, which might be the expectation, but to the author's knowledge there are no implemented cases to measure.
-% The nMCD method proposed in [Bueler 2022] is expensive but multigrid, so it may turn out that $\gamma$ is small.
+% regarding the nMCD method proposed in [Bueler unpublished], which is expensive but multigrid, it may turn out that $\gamma$ is small
 
-These comments complete Table \ref{tab:performancemodel}.
-Again one must be aware that all asymptotic performance estimates in the Table involve a large, scheme-dependent constant, with this especially true of implicit schemes.
+For implicit Stokes time-stepping, a prospective, coupled, and free-boundary velocity and geometry-update solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.
+One may suppose $\gamma\ge \beta$, but there are no implemented cases to measure.
+
+These comments complete Table \ref{tab:performancemodel}.  All estimates involve a scheme-dependent constant, with especially large values for the implicit schemes.
 
@@
 
-FIXME from Table \ref{tab:performancemodel} we see for explicit SIA that work scales as $\Delta x^{-4}$ for a given problem with fixed $D$ and $L$; because work per step scales as $\Delta x^{-2}$ but number of steps also scales as $\Delta x^{-2}$; explicit time-stepping Stokes is even worse under the pessimistic model; if solver only achieves $\alpha=2$ then work scales like horrific $\Delta x^{-8}$ (or $\Delta x^{-7}$ if optimistic about time-stepping stability); even for explicit time stepping the implementation burden of multigrid should pay off
-
-FIXME one can also fix resolution (fix $\Delta x$) and look at the expense of larger ice sheets ($L\to \infty$) to see why Stokes is difficult: scales as $L^{2+2\alpha}$ for optimistic and pessimistic cases; again want $\alpha$ is small
+FIXME
 

### 2022-06-28 paper/perfmod.tex
write delicate text to clarify the Jouvet and Brinkerhoff approaches

@@
 
-Our assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012,
+Our assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,BrinkerhoffJohnson2015,Hoffmanetal2018,Lengetal2012,
 Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.
@@
 Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}, but other models apparently require the user to choose a time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.
-The optimistic time step is supported in theory for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)), and thus some other manner of determining the time step.
+The optimistic time step is supported in theory for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
 
@@
 
-Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling \citep{Bueler2021}.
+Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling (or solver complexity \citep{Bueler2021}).
 At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula.
@@
 
-It appears there is only one implemented, unconditionally-stable, fully-implicit geometric-update solver, namely the SIA scheme in \cite{Bueler2016}.\footnote{The hybrid \cite{Brinkerhoffetal2017} flowline solver is partly implicit but apparently subject to a time-stepping condition when the margin moves.}  For simplified dome geometry this method, based on Newton steps solved via direct linear algebra, and on ice margin determination on a single grid, shows $\beta=1.8$.
-The same work shows how a steady-state ($\Delta t=\infty$) SIA solution is practical for many geometries.
-Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, even using realistic, thus irregular, bed topographies.
+The only implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.
+For simplified dome geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=1.8$.
+(An earlier non-Newton implementation by \cite{JouvetBueler2012} seems not to scale any better.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.
+Practical steady-state ($\Delta t=\infty$) solutions at ice sheet scale are also demonstrated by \cite{JouvetBueler2012,Bueler2016}.
 
-% regarding the nMCD method proposed in [Bueler unpublished], which is expensive but multigrid, it may turn out that $\gamma$ is small
+The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are also solved implicitly.
+\cite{JouvetGraeser2013} use a multigrid method, but their published, simplified-geometry results do not constrain $\beta$.
+Overall time-stepping in these hybrids is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects the thickness, so the SIA-portion solvers are apparently only tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$.
+% equation (20) in JouvetGraeser2013
+% equation (A21) in BrinkerhoffJohnson2015
 
@@
 
-These comments complete Table \ref{tab:performancemodel}.  All estimates involve a scheme-dependent constant, with especially large values for the implicit schemes.
+% regarding the nMCD method proposed in [Bueler unpublished], which is expensive but multigrid, it may turn out that $\gamma$ is small
+
+These comments complete Table \ref{tab:performancemodel}.  All estimates involve a scheme-dependent constant, one which must be especially large for the implicit schemes.
 

### 2022-06-29 paper/perfmod.tex
good to post at arxiv

@@
-\documentclass[twocolumn,letterpaper]{igs}
-%\documentclass[twocolumn]{igs}  % A4 paper needs no option
+%\documentclass[twocolumn,letterpaper]{igs}
+\documentclass[twocolumn]{igs}  % A4 paper needs no option
 %\documentclass[review,letterpaper]{igs}
@@
 
-\title[Performance analysis of high-resolution ice sheet simulations]{Letter: Performance analysis of \\ high-resolution ice sheet simulations}
+\title[Performance analysis of high-resolution ice sheet simulations]{Performance analysis of high-resolution \\ ice sheet simulations}
 
 \abstract{Numerical ice sheet models compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.
-At high spatial resolution, with horizontal mesh/grid resolutions of a few kilometers or smaller, these models usually require time steps shorter than climate-coupling time scales because these models update ice thickness after each velocity solution.
-Although precise time-step conditions are often lacking, high-resolution performance is degraded by the stability restrictions of such explicit time-stepping.
+At high spatial resolution, with horizontal mesh/grid resolutions of a few kilometers or smaller, these models usually require time steps shorter than climate-coupling time scales because they update ice thickness after each velocity solution.
+High-resolution performance is degraded by the stability restrictions of such explicit time-stepping.
 This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, attempts to clarify numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.
-The performance of current-generation explicit time-stepping models is assessed, and then the prospective performance of implicit schemes.
-The main result, Table \ref{tab:performancemodel}, highlights the key role which will be played by the algorithmic scaling of stress-balance and implicit-step solvers.}
+The performance of current-generation explicit time-stepping models is assessed, and then compared to the prospective performance of implicit schemes.
+The main result, Table \ref{tab:performancemodel}, highlights the key roles played by the algorithmic scaling of stress-balance and implicit-step solvers.}
 
@@
 
-Numerical ice sheet (glacier) models with evolving ice geometry are now in routine use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.
+Numerical ice sheet (glacier) models with evolving ice geometry are now routinely used for addressing scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.
 It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows need yet finer resolution.
@@
 Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, up to the non-shallow and non-hydrostatic Stokes approximation.
-With very few exceptions, however, current models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the just-computed velocity field.
+With very few exceptions, however, current time-stepping models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the just-computed velocity field.
 Therefore ice thickness and surface elevation, functionally-equivalent variables for this purpose, are updated after velocity is fixed.
@@
 
-These current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use the velocity computed from the geometry at the previous time step \citep[for example]{Chengetal2017}.}  At least for simpler, textbook partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}, and stability conditions of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the explicit time-stepping conditions of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.
+These current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use velocity computed from the geometry at the start of the time step \citep[for example]{Chengetal2017}.}  At least for simpler, textbook partial differential equation problems, the conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
+Stability conditions of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.
 
-However, actual \emph{implicit} time-stepping \citep{Bueler2021,LeVeque2007} should also be considered.
+However, actual \emph{implicit} time-stepping \citep{LeVeque2007} should also be considered.
 Here the velocity and geometry are updated simultaneously by solving coupled mass and momentum conservation equations.
-While implicit time-stepping always requires the solution of systems of equations at each step, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}, namely the 3D extend of the ice (once the coupled solution converges).
-The problem is of free-boundary type, both in map-plane (horizontal) \citep{SchoofHewitt2013} and (easily-resolved) vertical directions.
+While implicit time-stepping always requires the solution of systems of equations at each step, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}, namely the 3D extent of the ice once the coupled solution has converged.
+The problem is of free-boundary type, in map-plane (horizontal) directions \citep{SchoofHewitt2013}, and also in easily-resolved vertical directions.
 
-An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.
-The same reference shows how the steady-state problem \citep[compare][]{JouvetBueler2012} can often be solved, demonstrating unconditional stability.
-(Observe that steady-state solutions correspond to an implicit time step of infinite duration.)  The corresponding problem for the Stokes equations has not, to the author's knowledge, yet been attempted.
+An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016,JouvetGraeser2013}.
+\cite{Bueler2016} also shows how the steady-state problem \citep{JouvetBueler2012} can often be solved, demonstrating unconditional stability.
+(Observe that steady-state equations correspond to an implicit time step of infinite duration.)
+
+The corresponding problem for the Stokes equations has not, to the author's knowledge, yet been attempted.
 However, important early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} to solve a free boundary problem illuminates some of the techniques and difficulties needed to make such a strategy work for a membrane-stress-resolving balance.
@@
 Our simplified performance analysis can expose the most important considerations and trade-offs.
-While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping is applied, along with advanced solver techniques like multigrid \citep{Briggsetal2000}, this belief should be inspected in advance, and quantitatively to the extent possible.
+While the author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, these beliefs should be inspected quantitatively to the extent possible.
 
@@
 
-Reader familiarity with the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013} is assumed.
+Reader familiarity is assumed with the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.
 These (continuum) models are regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}.
+Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.
+However, before analyzing performance, the form of the mass continuity equation must be examined.
 
-Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.
-However, before presenting a performance analysis, the form of the mass continuity equation must be examined.
 For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged horizontal velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, this equation says
@@
 In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term is along $\gradx s$.
-In fact equation \eqref{eq:masscontinuity} has velocity $\bu$ which is a non-local function $\bu(H,\grad_x s)$, though a stress balance solution is required to evaluate it.
+Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\grad_x s)$ which is a non-local function of geometry, and a stress balance solution is required to evaluate it.
 A numerical instability occurs when the ice thickness under/over-shoots its correct value because the numerical velocity response from evaluating this non-local function is too strong for the given time step.
@@
 \newcommand{\nn}{\text{n}}
-Neither is the mass continuity equation wholely diffusive for stress-complete Stokes dynamics.
-A diffusive description is valid, however, in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}, and we see a revised mass continuity equation:
+A diffusive description of the mass continuity equation is valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
 \begin{equation}
@@
 \end{equation}
-Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity, in the isothermal case, in which $C$ is a positive constant.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ where $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.
-Typically $\nn\approx 3$.}  While equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed.
-For grounded ice sheets one observes that large values of $d$ tend to indicate locations of unstable mode growth if explicit time-steps are chosen too large.
+Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed.
+Generically, across stress balance choices, for grounded ice sheets one observes that large values of $d$ tend to indicate locations of unstable mode growth if explicit time-steps are chosen too large.
 
@@
 }
-\caption{Parameters for performance analysis; $\alpha,\beta,\gamma,n$ are pure numbers.}
+\caption{Parameters for performance analysis.  Note $\alpha$, $\beta$, $\gamma$, and $m$ are pure numbers.}
 \label{tab:notation}
@@
 \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits, and big-O notation is only used in this limit.
-If the map-plane model domain is of width $L$ then these parameters are related by
+
+If the map-plane model domain is of width $L$ then these primary parameters are related by
 \begin{equation}
@@
 \end{equation}
-That is, for a given value of $L$ there are $O(\sqrt{m})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$.
+For a given domain width $L$ there are $O(\sqrt{m})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$.
 However, our analysis addresses spatially-3D models with map-plane horizontal meshes.}
@@
 A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.
-It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the mesh/grid has an \emph{a priori} bounded resolution in the vertical direction.
-The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.
-Such ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance computes velocity as a function of the true state variables.
+Storing these model state variables, plus the thermodynamical state, requires $O(m)$ memory if the mesh/grid has \emph{a priori} bounded resolution in the vertical direction.
+The amount of fast memory needed by the simulation is also $O(m)$ if prior states are discarded or transferred to storage.
+Such ice sheet models also have $O(m)$ velocity variables, but these are not state variables.
+That is, a very-viscous stress balance computes velocity as a function of the true state variables.
 
@@
 Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.
-For example, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016} solvers use 3D refinement, though with various modifications regarding vertical refinement.
-This and many other mesh/solver details cannot be pursued here.
+Some solvers use 3D refinement \citep{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016}, with various distinctions between how horizontal and vertical meshing is handled.
+This and many other details cannot be pursued here.
 
-Ice sheet models resolve climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle and certain longer scales.
+Ice sheet models resolve climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and longer scales.
 Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.
 Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.
-Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated using ice velocity, i.e.~via solution of the mass continuity or surface kinematical equation.
+Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ describes the frequency on which ice geometry is updated using ice velocity, i.e.~via solution of the mass continuity or surface kinematical equation.
 
-As already noted, current-technology ice sheet models use explicit time-stepping which is only conditionally stable.
+Current-technology ice sheet models use explicit time-stepping which is only conditionally stable.
 For the spatial resolutions used in present-day scientific applications, we observe that maintenance of explicit time-stepping stability requires steps substantially shorter than $1/q$ model years.
 
-For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value (equation \eqref{eq:siamasscontinuity}).
+For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value, i.e.~of the $d$ in equation \eqref{eq:siamasscontinuity}.
 For Stokes or other membrane-stress-resolving dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, represents the \emph{optimistic} paradigm.
@@
 Explicit time-stepping with hybrid and higher-order schemes is somewhat better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.
-Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}, but other models apparently require the user to choose a time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.
-The optimistic time step is supported in theory for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
+Some hybrid schemes apply the pessimistic paradigm as an adaptive restriction \citep{Winkelmannetal2011}.
+Other models apparently require the user to choose a fixed time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.
+The optimistic paradigm is supported in theory for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
 
@@
 We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.
-For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited.
+For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited \citep{Bueler2021}.
 
@@
 By contrast, a multigrid method \citep{Trottenbergetal2001} can greatly reduce $\alpha$.
-For example, Antarctic ice sheet results by \cite[Table 8.1]{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$.\footnote{The \cite{Lengetal2012} algebraic-multigrid solver may have similar scaling, but reported results do not constrain $\alpha$.}  For higher-order stress balances, \cite{Tuminaroetal2016} report total algebraic-multigrid-preconditioned iterations following $\alpha \approx 0.05$ for the Greenland ice sheet,\footnote{See Table 7.5.
+For example, Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[Table 8.1]{IsaacStadlerGhattas2015}.
+The \cite{Lengetal2012} algebraic-multigrid Stokes solver may have similar scaling, but reported results do not constrain $\alpha$.
+For a higher-order stress balance on the Greenland ice sheet, \cite{Tuminaroetal2016} report total algebraic-multigrid-preconditioned iterations suggesting $\alpha \approx 0.05$,\footnote{See Table 7.5.
 Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.} and a geometric multigrid method by \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.
 
-Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling (or solver complexity \citep{Bueler2021}).
-At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula.
-Thus an SIA velocity solution requires optimal $O(m)$, $\alpha=0$, flops.
+Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.
+At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula; an SIA velocity solution requires optimal $O(m)$ flops.
 
 When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, strongly depending on solver design.
-Furthermore, many considerations are suppressed in a flops-based analysis of algorithmic scaling.
+Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling.
 Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.
-For example, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results all show good parallel scaling, something we do not address.
+For example, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good parallel scaling, something we do not address.
 
@@
 
-How many flops are needed to simulate one model year?
+Now, how many flops are needed to simulate one model year?
 Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.
-We may write time-step restrictions as required numbers of time-steps per model year to maintain stability.
+We may write stability restrictions as required numbers of time-steps per model year.
 That is, $\Delta t^{-1}$ is bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.
@@
 As already explained, unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve climatic interactions.
-On the other hand, the per-step expense is much greater because nontrivial coupled equations, indeed a free-boundary problem, must be solved for the velocity and geometry update simultaneously.
-We parameterize the flops of such coupled solutions as $O(m^{1+\beta})$, with the understanding that $\beta \ge \alpha$, for $\alpha$ associated to a corresponding velocity-only solution, and that a large constant is present.
+On the other hand, the per-step expense is much greater because nontrivial coupled equations, indeed a free-boundary problem, must be solved for the velocity and geometry-update simultaneously.
+For SIA models we parameterize the flops of such coupled solutions as $O(m^{1+\beta})$ with $\beta \ge 0$.
+A large constant is assumed to be present.
 
@@
 For simplified dome geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=1.8$.
-(An earlier non-Newton implementation by \cite{JouvetBueler2012} seems not to scale any better.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.
-Practical steady-state ($\Delta t=\infty$) solutions at ice sheet scale are also demonstrated by \cite{JouvetBueler2012,Bueler2016}.
+(An earlier non-Newton implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.
+Practical steady-state ($\Delta t=\infty$) solutions at ice sheet scale are also demonstrated by \cite{JouvetBueler2012} and \cite{Bueler2016}.
 
@@
 \cite{JouvetGraeser2013} use a multigrid method, but their published, simplified-geometry results do not constrain $\beta$.
-Overall time-stepping in these hybrids is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects the thickness, so the SIA-portion solvers are apparently only tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$.
+Overall time-stepping in these hybrids is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects the thickness, so the SIA-portion solvers are apparently only tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$ with $U$ scaling according to the sliding.
 % equation (20) in JouvetGraeser2013
@@
 For implicit Stokes time-stepping, a prospective, coupled, and free-boundary velocity and geometry-update solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.
-One may suppose $\gamma\ge \beta$, but there are no implemented cases to measure.
+One might also suppose $\gamma\ge \beta$, but in any case there are no implemented cases to measure.
+These comments complete Table \ref{tab:performancemodel}, in which all estimates involve a scheme-dependent constant, something which is especially large for the implicit schemes.
 
@@
 
-These comments complete Table \ref{tab:performancemodel}.  All estimates involve a scheme-dependent constant, one which must be especially large for the implicit schemes.
-
 \section{Discussion and Conclusion}
 
-FIXME
+From Table \ref{tab:performancemodel} we first observe a known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that effort scales as $O(\Delta x^{-4})$.
+Recall that this follows because $\Delta t < O(\Delta x^2)$, and because the expense of one geometry-update operation is $O(m) = O(\Delta x^{-2})$.
+Spatial mesh refinement by a factor of two therefore imposes an impressive 16-times increase in effort.
 
-The real promise of near-optimal methods is revealed when implicit solvers are considered, ones which simultanously satisfy the mass and momentum equations at each time step.
-If a multigrid method can be designed which yields $\gamma\approx 0$ for a Stokes model, for example, then numerical ice sheet modeling tools will be qualitatively faster for many tasks, and a large constant would be acceptable.
-Specifically, and more modestly, a $\gamma < 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method is a goal for the next decades of research on numerical ice sheet models.
-Such a tool would scale at $O(\Delta x^{-3})$, better than current explicit SIA models, avoiding shallow approximations while computing time steps with a frequency determined only by the scientific need to resolve climate coupling.
-The same goal applies to higher-order, or other membrane-stress-resolving, solvers.
-These aspirations are admittedly long-term because, to the author's knowledge, no progress has yet been made on such implicit designs.
+The \cite{Bueler2016} implicit SIA solver is really no better, however.
+Although long time steps can be taken by this implicit solver, the cost of $\beta=1.8$ scaling is $O(\Delta x^{-5.6})$ because $2+2\beta=5.6$.
+However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably superior performance.
+
+Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of solvers completely dominates.
+In particular, an $\alpha=2$ explicit Stokes model, e.g.~one using direct linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-8})$ under a pessimistic stability condition.
+Optimistic stability yields still-horrific $O(\Delta x^{-7})$.
+These simple observations emphasize the key role of multigrid-based stress balance solvers.
+That is, even retaining explicit time-stepping, nearly-optimal ($\alpha \approx 0$) scaling of computation effort in Stokes and higher-order solvers will be necessary for routine application on high-resolution meshes.
+
+On the other hand, suppose resolution ($\Delta x$) is fixed.
+The Table also shows why algorithmic scaling remains important in the large-domain $L\to\infty$ limit when applying Stokes or higher-order dynamics.
+An $\alpha=2$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.
+By contrast, the computational work of a nearly-optimal solver will be proportional to ice sheet area $L^2$.
+
+The promise of nearly-optimal solvers is profoundly revealed when implicit geometry updates are considered.
+Future methods which simultanously satisfy the mass and momentum equations at each time step, and which do work essentially proportional to the number of degrees of freedom, have the greatest promise.
+A $\gamma \approx 0$ implicit Stokes method would be hugely more capable for many tasks.
+Specifically, and more achievably, a $\gamma \le 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method, presumably based on multigrid solution of the free-boundary problem for the coupled mass and momentum equations, is an appropriate goal for decades of research on numerical ice sheet models.
+The computation cost would scale at $O(\Delta x^{-3})$, better than explicit SIA models, and the method would only update ice geometry when determined by the scientific need to resolve climate coupling, while avoiding shallow approximations.
+The same goal makes sense for all membrane-stress-resolving solvers.
+No apparent technical progress has yet been made on such an implicit Stokes design, and so these aspirations are decidedly long-term.
+However, the above discussion suggests why measured values for $\alpha,\beta,\gamma$, or equivalent algorithmic scaling measures, are important performance metrics to report when describing new ice sheet solvers.
 

### 2022-07-04 paper/data/scaling.m
I screwed up beta for the Bueler (2016) solver (flops per model year should have beta=0.8)

@@ -1,4 +1,7 @@
 % generate serial scaling for bueler 2016 and isaac et al 2015 and tumarino et al 2016
+
+% note on iterations: if iterations scale as O(m^alpha) then flops scale as O(m^{1+alpha}); this is under assumption that each iteration costs O(m) flops
+
 % bueler:  LU solve of Newton step equations for dome; see sia-fve/petsc/verif/domeperf.sh; see generated file bueler.perf
 % isaac: AMG-Krylov-Newton Stokes solve for antarctica; see table 8.1 in paper
 % tumarino: MDSC–AMG-Krylov-Newton higher-order solve for greenland; see table 7.5 in paper; area of Greenland taken from Searise data: 300x560 5km grid
@@ -6,44 +9,44 @@
 nb = [18 36 72 144 288].^2;
 flopsb = [1.776e7 1.749e8 2.374e9 2.608e10 2.927e11];
 pb = polyfit(log(nb),log(flopsb),1);
-%ans = 1.76186   6.45957
+%ans = 1.76186   6.45957   ==>  beta = 0.76186
 
 ni = [7e6 51e6 383e6];
 itsi = [7*66 10*75 11*90]; % total Krylov iterations over Newton steps
 pi = polyfit(log(ni),log(itsi),1);
-%ans = 0.19030   3.17188
+%ans = 0.19030   3.17188   ==>  alpha = 0.19030
 
 nt = ((300*560*5^2) / (8^2)) * [1 4 16 64 256];
 itst = [36*13.3 34*12.7 35*12.3 36*13.3 37*17.4]; % total Krylov iterations over Newton steps
 pt = polyfit(log(nt),log(itst),1);
-%ans = 0.05017   5.49297
+%ans = 0.05017   5.49297   ==>  alpha = 0.05017
 
 subplot(3,1,1)
 loglog(nb, flopsb, 'bo')
-xlabel('degrees of freedom')
+xlabel('degrees of freedom m')
 ylabel('flops')
 title('Bueler')
 hold on
 loglog(nb, exp(pb(1) * log(nb) + pb(2)), 'b--')
 hold off
-legend(sprintf('O(n^{%.3d})',pb(1)))
+legend(sprintf('O(m^{%.3d})',pb(1)))
 
 subplot(3,1,2)
 loglog(ni, itsi, 'ro')
-xlabel('degrees of freedom')
[…]

### 2022-07-04 paper/perfmod.tex
this version posted to arxiv ... email friends once processed
posted at arxiv (https://arxiv.org/abs/2206.14352) and submitted to J Glaciol
I submitted it again in the review format
correct errors (alpha=1 not 2 for sparsity-exploiting direct LA; beta=0.8 not beta=1.8 for Bueler 2016 effort)

@@
-%\documentclass[twocolumn,letterpaper]{igs}
-\documentclass[twocolumn]{igs}  % A4 paper needs no option
+\documentclass[twocolumn,letterpaper]{igs}
+%\documentclass[twocolumn]{igs}  % A4 paper needs no option
 %\documentclass[review,letterpaper]{igs}
@@
 
-%\keywords{}
+%\keywords{glacier flow,glacier modelling,ice-sheet modelling}
+% suggested reviewers:  Dan Goldberg, William Lipscomb, Toby Isaac, Josefin Ahlkrona, Alex Jarosch, Doug Brinkerhoff
 
@@
 This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.
-Our simplified performance analysis can expose the most important considerations and trade-offs.
-While the author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, these beliefs should be inspected quantitatively to the extent possible.
+Our simplified performance analysis exposes the most important considerations and trade-offs.
+While the author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, these beliefs should be assessed quantitatively to the extent possible.
 
@@
 We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.
-For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited \citep{Bueler2021}.
+For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 1$ if sparsity is exploited or $\alpha \approx 2$ if not \citep{Bueler2021}.
 
@@
 Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.
-At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula; an SIA velocity solution requires optimal $O(m)$ flops.
+At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula; an SIA velocity solution therefore requires optimal $O(m)$ flops.
 
@@
 Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.
-For example, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good parallel scaling, something we do not address.
+For example, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good parallel scaling, something we do not address here.
 
@@
 Now, how many flops are needed to simulate one model year?
-Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.
-We may write stability restrictions as required numbers of time-steps per model year.
+Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ steps per model year.
+We may write stability restrictions as required numbers of steps per model year.
 That is, $\Delta t^{-1}$ is bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.
@@
 The only implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.
-For simplified dome geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=1.8$.
+For simplified dome geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.
 (An earlier non-Newton implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.
@@
 \cite{JouvetGraeser2013} use a multigrid method, but their published, simplified-geometry results do not constrain $\beta$.
-Overall time-stepping in these hybrids is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects the thickness, so the SIA-portion solvers are apparently only tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$ with $U$ scaling according to the sliding.
+Overall time-stepping in these hybrids is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects the thickness, so the SIA-portion solvers are apparently only tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$ for $U$ which scales according to the sliding.
 % equation (20) in JouvetGraeser2013
@@
 
-For implicit Stokes time-stepping, a prospective, coupled, and free-boundary velocity and geometry-update solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.
+For implicit Stokes time-stepping, a prospective, coupled, and free-boundary velocity and geometry-update solve is assumed to be, in the absence of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.
 One might also suppose $\gamma\ge \beta$, but in any case there are no implemented cases to measure.
@@
 
-From Table \ref{tab:performancemodel} we first observe a known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that effort scales as $O(\Delta x^{-4})$.
+From Table \ref{tab:performancemodel} we first observe a known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that effort, here flops per model year, scales as $O(\Delta x^{-4})$.
 Recall that this follows because $\Delta t < O(\Delta x^2)$, and because the expense of one geometry-update operation is $O(m) = O(\Delta x^{-2})$.
@@
 
-The \cite{Bueler2016} implicit SIA solver is really no better, however.
-Although long time steps can be taken by this implicit solver, the cost of $\beta=1.8$ scaling is $O(\Delta x^{-5.6})$ because $2+2\beta=5.6$.
-However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably superior performance.
+The \cite{Bueler2016} implicit SIA solver is not enough better, however.
+Although long time steps can be taken by this implicit solver, the $\beta=0.8$ scaling gives $O(\Delta x^{-3.6})$ effort because $2+2\beta=3.6$.
+However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably superior performance compared to an explicit, $O(\Delta x^{-4})$ SIA scheme.
 
 Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of solvers completely dominates.
-In particular, an $\alpha=2$ explicit Stokes model, e.g.~one using direct linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-8})$ under a pessimistic stability condition.
-Optimistic stability yields still-horrific $O(\Delta x^{-7})$.
+In particular, an $\alpha=1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition.
+Optimistic stability yields still-bad $O(\Delta x^{-5})$.
 These simple observations emphasize the key role of multigrid-based stress balance solvers.
@@
 The Table also shows why algorithmic scaling remains important in the large-domain $L\to\infty$ limit when applying Stokes or higher-order dynamics.
-An $\alpha=2$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.
+An $\alpha=1$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.
 By contrast, the computational work of a nearly-optimal solver will be proportional to ice sheet area $L^2$.
@@
 A $\gamma \approx 0$ implicit Stokes method would be hugely more capable for many tasks.
-Specifically, and more achievably, a $\gamma \le 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method, presumably based on multigrid solution of the free-boundary problem for the coupled mass and momentum equations, is an appropriate goal for decades of research on numerical ice sheet models.
+Specifically, and more achievably, a $\gamma < 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method, presumably based on multigrid solution of the free-boundary problem for the coupled mass and momentum equations, is an appropriate goal for coming decades of research on numerical ice sheet models.
 The computation cost would scale at $O(\Delta x^{-3})$, better than explicit SIA models, and the method would only update ice geometry when determined by the scientific need to resolve climate coupling, while avoiding shallow approximations.
@@
 However, the above discussion suggests why measured values for $\alpha,\beta,\gamma$, or equivalent algorithmic scaling measures, are important performance metrics to report when describing new ice sheet solvers.
+
+\subsubsection{Note added 4 July 2022.}  Errors in the first posted version (\url{https://arxiv.org/abs/2206.14352}) are corrected here.
 

### 2022-09-25 paper/reviews/response.tex
start a response doc; reviewer 1 comments handled

\documentclass[letterpaper,final,12pt,reqno]{amsart}

\usepackage[total={6.3in,9.2in},top=1.1in,left=1.1in]{geometry}

\usepackage{times,bm,bbm,empheq,fancyvrb,graphicx,amsthm,amssymb}
\usepackage[dvipsnames]{xcolor}
\usepackage{longtable}
\usepackage{booktabs}

\usepackage{tabto}
\TabPositions{1.5cm}

\usepackage{float}

% hyperref should be the last package we load
\usepackage[pdftex,
colorlinks=true,
plainpages=false, % only if colorlinks=true
linkcolor=blue,   % ...
citecolor=Red,    % ...
urlcolor=black    % ...
]{hyperref}

\renewcommand{\baselinestretch}{1.05}

\allowdisplaybreaks[1]  % allow display breaks in align environments, if they avoid major underfull

\newcommand{\eps}{\epsilon}

\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}

\newcommand{\grad}{\nabla}
\newcommand{\Div}{\nabla\cdot}
\newcommand{\trace}{\operatorname{tr}}

\newcommand{\hbn}{\hat{\mathbf{n}}}

\newcommand{\bb}{\mathbf{b}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bbf}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bz}{\mathbf{z}}

\newcommand{\bF}{\mathbf{F}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bX}{\mathbf{X}}

\newcommand{\bxi}{\bm{\xi}}
\newcommand{\bzero}{\bm{0}}

\newcommand{\cK}{\mathcal{K}}
\newcommand{\cV}{\mathcal{V}}

\newcommand{\rhoi}{\rho_{\text{i}}}

\newcommand{\ip}[2]{\left<#1,#2\right>}

\newcommand{\maxR}{R^{\bm{\oplus}}}
\newcommand{\minR}{R^{\bm{\ominus}}}
\newcommand{\iR}{R^{\bullet}}

\newcommand{\nn}{{\text{n}}}
\newcommand{\pp}{{\text{p}}}
\newcommand{\qq}{{\text{q}}}
\newcommand{\rr}{{\text{r}}}

\newcommand{\supp}{\operatorname{supp}}
\newcommand{\Span}{\operatorname{span}}

\newenvironment{review}%
{\bigskip \par \begin{quote} \selectfont \sl}%
{\end{quote}}

\begin{document}
\title{Response to reviews of \emph{Performance analysis of high-resolution ice sheet simulations}}

\author{Ed Bueler}

\date{\today}

\maketitle

%\tableofcontents

\thispagestyle{empty}
%\bigskip

The thoughtful comments of the two reviewers are much appreciated.

\section{Responses to Reviewer 1}

\begin{review}
Let me congratulate you to this much needed performance analysis of ice sheet models, which is presented clearly and in detail in the manuscript at hand. I have only a few very minor comments to add.

As you include glaciers in your introduction (L\# 21) you could reference Clarke et al (Clarke, G. K., Jarosch, A. H., Anslow, F. S., Radić, V., \& Menounos, B. (2015). Projected deglaciation of western Canada in the twenty-first century. Nature Geoscience, 8(5), 372-377) around L\#30 as an example of a large, region scale glacier study using an explicit time stepping scheme on sub km grid sizes.
\end{review}

\noindent Thank you!  This is a good example to add, and I have done so.

\begin{review}
More style comments:

In lines \# 82 and on page 8 just befor equation (4), maybe use the word "equation" alongside the reference to an equation, as you do e.g. in L\#95 and other places, just for consistency.
\end{review}

\noindent Done.

\begin{review}
In lines \# 67, 150, 173, and maybe more places you use plural such as "our", "we", which strikes me as odd as you did the work on your own. But maybe that is only me as a non native English speaker.
\end{review}

\noindent I believe this is standard-enough usage, although I am happy to follow the Editor's guidance if it differs.  Speaking more precisely, I am trying to stick to ``inclusive we'' usage; see

\medskip
{\footnotesize \href{https://oxfordediting.com/to-we-or-not-to-we-the-first-person-in-academic-writing/}{\texttt{oxfordediting.com/to-we-or-not-to-we-the-first-person-in-academic-writing}}}

\medskip
\noindent Therefore I altered the text, away from ``we'' or ``our'', in lines 67, 113, 127, 173, and 207.  However, I kept ``us'', in the inclusive phrases ``let us assume'' and ``we see'', in lines 150, 186, and 217.

\begin{review}
In line \# 143, maybe add again the Robinson et al reference into the brackets to Figure 3(a) so that it is clear that you reference their figure.
\end{review}

\noindent Yes.  Done.

\begin{review}
Anyhow just a few comments from my side. Else the paper is, as said, crystal clear and well presented and very relevant for directing the model development efforts in a performance aware direction.
\end{review}

Thanks.

\section{Responses to Reviewer 2}

\end{document}

### 2022-09-26 paper/reviews/response.tex
start on reviewer 2 ... this will be work

@@
 
+\begin{review}
+\textbf{Summary}
+
+In this short manuscript, the author provides a prospectus on the benefits of adopting implicit methods for the time-discretization of the equations of ice sheet motion, and in particular how such methods may improve model efficiency relative to the use of explicit methods (which represent the current dominant, though not universal, paradigm).
+Beginning with some background on the relationship between ice sheet velocity (which is diagnostic) and geometry (which is prognostic) under different simplifications to the stress balance, the manuscript then goes through various scaling arguments that establish the essential amount of work needed to evolve an ice sheet model for a given length of time.
+The author re-establishes the classical result of fourth-power in spatial resolution effort for the shallow ice approximation, but presents the surprising result of (pessimistic) sixth-power growth for Stokes’ equations (which is quite oppressive), then shows that these resolution-dependent challenges can be significantly ameliorated with implicit schemes, which have the potential to realistically reduce computational cost to cubic in resolution (at least asymptotically).
+
+Overall, the manuscript is a worthwhile contribution that will help to guide future model development and to justify why such developments are worth the effort.
+It is, in a sense, an opinion piece, so there is little to criticize with respect to scientific merit.
+I do have a limited set of comments that might help clarify points and answer questions that some readers may have.
+\end{review}
+
+\noindent This summary of my intent is quite accurate, and I appreciate the reviewer's effort to improve my manuscript.
+
+\begin{review}
+\textbf{Title} I don't find this title to be a particularly helpful characterization of the work.
+Maybe something more specific e.g.
+'Asymptotic analysis of implicit versus explicit methods in ice sheet models' would be better.
+\end{review}
+
+\noindent Some thought on a revised title is worthwhile, so this comment is welcomed.
+Unfortunately ``Asymptotic analysis of implicit versus explicit methods in ice sheet models'' would not clarify which is the asymptotic limit.
+(It is not aspect ratio $H/L \to 0$, nor convergence $\Delta x \to 0$, nor approach to steady state $t\to \infty$, but rather number of degrees of freedom.)  Furthermore, my intent is to address \emph{performance}, as a broader issue which relates to multiple modeling choices, specifically mesh resolution, time-stepping type (explicit vs.~implicit), solver design (multigrid versus not), and even operational ice-climate coupling frequency, all over long-duration runs.
+
+\medskip \noindent In other words, I don't see how to write a title of acceptable length without requiring the interested reader to (at least) dive into the abstract for clarification of some brief phrase such as ``performance analysis''.
+(Of course, a shorter title will, itself, increase the number of such interested readers.)  A possible title is ``Performance analysis of numerical ice sheet simulations in the high-resolution limit'', and this captures the correct asymptotic, but I find the current title to be essentially as clear, and clearly briefer.
+Does the Editor have a title suggestion?
+
+\begin{review}
+\textbf{L25} 'it is generally accepted...'  I am not sure that I agree that this is generally accepted.
+There remain important works in Antarctica where models are run at 20km resolution or more and I’m not sure that they are ‘invalid’.
+It is worth perhaps being more specific about the resolution requirements of specific tasks, e.g.
+capturing detailed perturbations to the grounding line or calving front, the influence of steep subglacial topography, etc.
+\end{review}
+
+\noindent I stand by the assertion, but this reasonable challenge has caused me to add some words to clarify.  The sentence in question now says:
+
+\begin{quote}  In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows need yet finer resolution.\end{quote}
+
+\noindent A few points about this sentence will, I think, be understood by any reader with minimal exposure to ice sheet modeling:
+\begin{itemize}
+\item This is a necessarily-imprecise description of practical resolution needs for ``resolving'' physical mechanisms understood to be important by a community of scientists.
+This is not a mathematically or statistically based assertion.
+\item The added phrase ''ice streams as fluid features'' refers to model results which are necessarily more than one grid cell wide.
+As fluids are locally infinite-dimensional, you need a few cells, more than one or two across for ice streams, to capture a ``fluid feature''.
+\item The added word ``it is \underline{now} generally accepted'' clearly refers to expectations in the 2020s.
+There were indeed ``important'' 20km results before now, but I would disagree with any assertion in a present-day journal submission that, for example, flux across grounding lines in Antarctica is ``validly modeled'' at 20 km resolution; that would be a consequence of not trying very hard, or choosing a deficient model.
+\item 10 km resolution is itself not enough for narrower features than ice streams.
+\end{itemize}
+With that background, I am happy if readers conclude, for example, that the author thinks that 20km or 40km Antarctic model results are no longer ``valid'' in a scientific sense.
+That is what I think, and I was under the impression that this was a generally-accepted judgement.
+
+\begin{review}
+\textbf{L34}  While explicit time stepping is indeed the paradigm, the ‘very few exceptions’ ought to be referenced.
+While I am sure there are more, here are three papers that use a coupled and implicit geometry-non-SIA-velocity solution scheme: Gudmundsson (2013); Brinkerhoff et al.
+(2017); Shapero et al.
+(2021)
+\end{review}
+
+\noindent FIXME This reviewer comment drives me to make the most significant alterations to the paper.
+The reviewer, and I suspect many readers, have never imagined an unconditionally stable implicit ice sheet model, so they don't recognize the only semi-implicit nature of various assertions of implicitness in the literature.
+The SIA solver in Bueler (2016) is actually implicit, and demonstrably unconditionally stable on flat beds, in the sense that one can take a single time step of arbitrary length to construct the geometry of an ice sheet from knowledge only of the surface mass balance (accumulation/ablation function) and the bed.
+This means a free-boundary problem must be solved during the time step to determine the map-plane region occupied by the ice.
+Otherwise, in the cited models wherein an ``implicit'' claim is made, in fact a major aspect of the geometry at the current time step, specifically the ice-covered domain of the map-plane, is \emph{fixed} during the time step.
+Only the vertical extent of the ice is updated implicitly.
+
+\begin{review}
+\end{review}
+
+\noindent x
+\begin{review}
+\end{review}
+
+\noindent x
+\begin{review}
+\end{review}
+
+\noindent x
+\begin{review}
+\end{review}
+
+\noindent x
+\begin{review}
+\end{review}
+
+\noindent x
+\begin{review}
+\end{review}
+
+\noindent x
 \end{document}

### 2022-09-26 paper/reviews/response.tex
clarify for myself how to proceed

@@
 
+FIXME owe the reader a definition of ``explicit'', ``semi-implicit'', and ``implicit'' as I use them; put this at the end of the (new) ``Coupled geometry-velocity time-stepping'' section; ``explicit if all aspects of ice geometry are fixed during the velocity solve, and then that velocity is used (without further alteration) to compute an updated velocity''; ``semi-implicit if any major aspect of the ice geometry is fixed during the (partly-coupled) geometry-velocity solve'' (and the ice-covered map-plane region certainly counts as a major aspect); ``implicit if all aspects of ice geometry and velocity are solved for simultaneously'' (and this is necessarily a free-boundary problem in horizontal and vertical directions; quote ``weakly recognized'' comment from Schoof \& Hewitt
+
 \begin{review}

### 2022-09-28 paper/reviews/response.tex
a bit of work, but no completion

@@
 
-\noindent FIXME This reviewer comment drives me to make the most significant alterations to the paper.
-The reviewer, and I suspect many readers, have never imagined an unconditionally stable implicit ice sheet model, so they don't recognize the only semi-implicit nature of various assertions of implicitness in the literature.
-The SIA solver in Bueler (2016) is actually implicit, and demonstrably unconditionally stable on flat beds, in the sense that one can take a single time step of arbitrary length to construct the geometry of an ice sheet from knowledge only of the surface mass balance (accumulation/ablation function) and the bed.
+\noindent This reviewer comment drives me to make the most significant addition to the paper.
+The reviewer, and I suspect many readers, have never really imagined an unconditionally-stable and fully-implicit ice sheet model, so they don't recognize the only semi-implicit nature of various assertions of implicitness in the literature.
+The only such solver is the SIA implementation in Bueler (2016), though in fact the idea has existed as a theoretical matter since Calvo et al (2003).
+
+My addition is the following definitions-of-terms inserted at the end of the second section, which has been re-titled from ``Mass continuity equation'' to ``Coupled geometry-velocity modeling'':
+
+\begin{quote}
+Precise definitions of the terms \emph{explicit} and \emph{implicit}, for coupled geometry-velocity updates, equivalently for coupled mass-momentum conservation time steps, will be needed for understanding the performance analysis which follows.
+As defined earlier, an \emph{explicit} scheme for the coupled problem alternates between solving the stress balance for velocity, and then holding that velocity fixed while solving mass continuity or surface kinematical equation for geometry (thickness or surface elevation), and then accepting both fields as the scheme proceeds to the next time step.
+At the other extreme a (\emph{fully}) \emph{implicit} scheme
+\end{quote}
+
+FIXME returning to the reviewer's specific comment and citations, ...
+
+ is actually implicit, and demonstrably unconditionally stable on flat beds, in the sense that one can take a single time step of arbitrary length to construct the geometry of an ice sheet from knowledge only of the surface mass balance (accumulation/ablation function) and the bed.
 This means a free-boundary problem must be solved during the time step to determine the map-plane region occupied by the ice.

### 2022-10-06 paper/perfmod.tex
again put in review format
edits through big addition

@@
-\documentclass[twocolumn,letterpaper]{igs}
+%\documentclass[twocolumn,letterpaper]{igs}
 %\documentclass[twocolumn]{igs}  % A4 paper needs no option
-%\documentclass[review,letterpaper]{igs}
+\documentclass[review,letterpaper]{igs}
 
@@
 High-resolution performance is degraded by the stability restrictions of such explicit time-stepping.
-This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, attempts to clarify numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.
+This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, clarifies the scaling of numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.
 The performance of current-generation explicit time-stepping models is assessed, and then compared to the prospective performance of implicit schemes.
-The main result, Table \ref{tab:performancemodel}, highlights the key roles played by the algorithmic scaling of stress-balance and implicit-step solvers.}
+The main result, Table \ref{tab:performancemodel}, highlights the key roles played by the algorithmic scaling of stress-balance and coupled, implicit-step solvers.}
 
@@
 
-Numerical ice sheet (glacier) models with evolving ice geometry are now routinely used for addressing scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.
-It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows need yet finer resolution.
-Whether using local mesh refinement \citep[for example]{Fischleretal2022,Hoffmanetal2018} or not, resolutions of two or one kilometers \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used for science at ice sheet scale.
+Numerical ice sheet (glacier) models with evolving ice geometry are now routinely used for addressing scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenlandic \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.
+In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows need yet finer resolution.
+Whether using local mesh refinement \citep[for example]{Fischleretal2022,Hoffmanetal2018} or not, resolutions of two or one kilometers \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019,Clarkeetal2015} are increasingly used for science at ice sheet scale.
 
-Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, up to the non-shallow and non-hydrostatic Stokes approximation.
+Current-generation ice sheet models apply a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, up to the non-shallow and non-hydrostatic Stokes approximation.
 With very few exceptions, however, current time-stepping models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the just-computed velocity field.
-Therefore ice thickness and surface elevation, functionally-equivalent variables for this purpose, are updated after velocity is fixed.
-Note that the interaction between the ice sheet and the surrounding climate occurs during this geometry-update operation, via the mass continuity or surface kinematical equations \citep{GreveBlatter2009}.
-Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes in particular.
+Therefore ice thickness and surface elevation, equivalent geometry variables for most modeling purposes, are updated after velocity is fixed.
+The interaction between the ice sheet and the surrounding climate occurs during this geometry-update operation, via the mass continuity or surface kinematical equations \citep{GreveBlatter2009}.
+Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes, in particular.
 
-These current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use velocity computed from the geometry at the start of the time step \citep[for example]{Chengetal2017}.}  At least for simpler, textbook partial differential equation problems, the conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
-Stability conditions of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.
+In other words, these current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system describing the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use velocity computed from the geometry at the start of the time step \citep[for example]{Chengetal2017}.}  However, at least for simpler, textbook partial differential equation problems, the limited and conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
+Stability conditions of explicit SIA models appeared some time ago \citep[e.g.][]{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.
 
@@
 Here the velocity and geometry are updated simultaneously by solving coupled mass and momentum conservation equations.
-While implicit time-stepping always requires the solution of systems of equations at each step, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}, namely the 3D extent of the ice once the coupled solution has converged.
-The problem is of free-boundary type, in map-plane (horizontal) directions \citep{SchoofHewitt2013}, and also in easily-resolved vertical directions.
+While such implicit time-stepping requires the solution of systems of equations at each step, for glaciers and ice sheets an implicit step must also \emph{simultaneously} compute the velocity field and the \emph{domain on which the velocity is defined}, namely the 3D extent of the ice once the coupled solution has converged.
+The problem is of free-boundary type, especially in map-plane (horizontal) directions \citep{SchoofHewitt2013}, but also in the (easily-resolved) vertical direction.
 
-An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016,JouvetGraeser2013}.
-\cite{Bueler2016} also shows how the steady-state problem \citep{JouvetBueler2012} can often be solved, demonstrating unconditional stability.
+An implicit strategy has been demonstrated at high (900m) resolution for the Greenlandic ice sheet in the simplest frozen-base, isothermal SIA case \citep{Bueler2016,JouvetGraeser2013}.
+That work also shows how the steady-state SIA problem \citep[c.f.][]{JouvetBueler2012} can be solved, demonstrating unconditional stability.
 (Observe that steady-state equations correspond to an implicit time step of infinite duration.)
@@
 This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.
-Our simplified performance analysis exposes the most important considerations and trade-offs.
-While the author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, these beliefs should be assessed quantitatively to the extent possible.
+The simplified performance analysis here exposes the most important considerations and trade-offs.
+While this author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, such beliefs should be assessed quantitatively to the extent possible.
 
-\section{Mass continuity equation}
+\section{Coupled geometry-velocity modeling}
 
 Reader familiarity is assumed with the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.
-These (continuum) models are regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}.
+These (continuum) models are regarded here as end members of current-usage stress-balance approximations.
 Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.
-However, before analyzing performance, the form of the mass continuity equation must be examined.
+However, before analyzing the performance consequences of numerical modeling choices, the form of the mass continuity equation will be examined, and then the terms ``explicit'' and ``implicit'' will be carefully defined.
 
-For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged horizontal velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, this equation says
+For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged horizontal velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, the mass continuity equation says
 \begin{equation}
@@
 \end{equation}
-where $\bx=(x,y)$ denotes horizontal coordinates.
-
+where $\bx=(x,y)$ denotes the horizontal coordinates.
 Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.
 This is because ice flows dominantly downhill.
-Indeed, ice sheet flow has no characteristic curves, as would \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.
+Indeed, ice sheet flow has no characteristic curves, as would Equation \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.
 
@@
 Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.
-Under a numerical thickness perturbation the velocity $\bu$ will often respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.
-In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term is along $\gradx s$.
-Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\grad_x s)$ which is a non-local function of geometry, and a stress balance solution is required to evaluate it.
-A numerical instability occurs when the ice thickness under/over-shoots its correct value because the numerical velocity response from evaluating this non-local function is too strong for the given time step.
+A numerical thickness perturbation will often cause the velocity $\bu$ to respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.
+In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term is along the surface gradient $\gradx s$.
+One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\grad_x s)$, a non-local function of geometry; a stress balance solution is required to evaluate this function.
+A numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step.
 
@@
 \end{equation}
-Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed.
+Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.
 Generically, across stress balance choices, for grounded ice sheets one observes that large values of $d$ tend to indicate locations of unstable mode growth if explicit time-steps are chosen too large.
+
+FIXME big addition
+
+\section{Performance analysis}
+
+Table \ref{tab:notation} lists the parameters used in our performance analysis.
+The primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.
+\emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits, and big-O notation is only used in this limit.
 
@@
 \end{table*}
-
-\section{Performance analysis}
-
-Table \ref{tab:notation} lists the parameters used in our performance analysis.
-The primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.
-\emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits, and big-O notation is only used in this limit.
 

### 2022-10-06 paper/perfmod.tex
make the major addition

@@
 Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.
-However, before analyzing the performance consequences of numerical modeling choices, the form of the mass continuity equation will be examined, and then the terms ``explicit'' and ``implicit'' will be carefully defined.
+However, before analyzing the performance consequences of numerical modeling choices, the form of the mass continuity equation is examined, and then the terms ``explicit'' and ``implicit'' are carefully defined.
 
@@
 A numerical thickness perturbation will often cause the velocity $\bu$ to respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.
-In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term is along the surface gradient $\gradx s$.
-One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\grad_x s)$, a non-local function of geometry; a stress balance solution is required to evaluate this function.
+In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term effectively acts along the surface gradient $\gradx s$ because of the ice-domain geometry.
+One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\gradx s)$, a non-local function of geometry; a stress balance solution is required to evaluate this function.
 A numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step.
@@
 \newcommand{\nn}{\text{n}}
-A diffusive description of the mass continuity equation is valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
+A diffusive description of the mass continuity equation is also valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
 \begin{equation}
@@
 \end{equation}
-Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.
-Generically, across stress balance choices, for grounded ice sheets one observes that large values of $d$ tend to indicate locations of unstable mode growth if explicit time-steps are chosen too large.
+Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While Equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.
+Generically across stress balance choices, for grounded glaciers and ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.
 
-FIXME big addition
+However, either statement \eqref{eq:masscontinuity} or \eqref{eq:siamasscontinuity} of the mass continuity equation is fundamentally incomplete without modeling the changing (map-plane) boundary position of the glacier or ice sheet.
+The problem of determining the updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).
+The mathematical role of this constraint has been understood for some time in the SIA model \citep{Calvoetal2002,JouvetBueler2012,SchoofHewitt2013}, but its role as the source of the margin location is universal across fluid-layer problems with signed source terms \citep{Bueler2021conservation}.
+
+It is in this context that we can distinguish time-stepping types in coupled geometry-velocity, i.e.~mass and momentum conserving, glacier and ice sheet models.
+A coupled model computes (\emph{fully}) \emph{implicit} time steps if no significant aspects of the geometry or velocity are held fixed during the (coupled) solution.
+That is, in an implicit scheme a coupled and well-posed model for the updated ice surface elevation (or thickness) and velocity fields is solved.
+A scheme which does not allow the map-plane ice-covered region to change during a time step cannot respond physically (conservatively) to a change in climate inputs; such a scheme is not actually implicit for the coupled problem and it cannot validly compute long-duration time steps.
+Thus we say that a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during its coupled geometry-velocity solution.
+A scheme is only semi-implicit if it computes margin advance or retreat after the velocity field is updated.
+Finally, a scheme is \emph{explicit} for the coupled problem, and uses a time-step \emph{splitting} scheme \citep{HundsdorferVerwer2003}, if the ice geometry is held fixed during the velocity solution, and then this (accepted) velocity solution is used to update the geometry, e.g.~through an advection step for Equation \eqref{eq:masscontinuity}.
+
+Thus a key point about time-stepping schemes for the coupled geometry and velocity problem for glaciers and ice sheets is that implicit schemes are not merely solving fixed-domain systems of coupled PDEs.
+The mathematical problem for the coupled and implicit time-step includes the condition which controls the free boundary, namely the inequality constraint of nonnegative thickness \citep{Bueler2021conservation}.
+Implicit schemes which solve these free-boundary problems at each time step can, at high spatial resolution, transcend stability limitations on time step duration.
+They can do time-stepping at a rate controlled by climatic-input time scales.
 

### 2022-10-06 paper/perfmod.tex
ongoing edits

@@
 \end{equation}
-Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While Equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.
+Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, for the isothermal case where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}, one has $C = 2 A (\rho g)^\nn/(\nn+2)$.}  While Equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.
 Generically across stress balance choices, for grounded glaciers and ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.
@@
 
-Table \ref{tab:notation} lists the parameters used in our performance analysis.
-The primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.
-\emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits, and big-O notation is only used in this limit.
+Table \ref{tab:notation} lists the parameters used in the numerical model performance analysis of this section.
+The primary parameters are $\Delta x$, which is a representative value for the horizontal mesh (grid) cell diameter, and $m$, which is the number of nodes (vertices) in the horizontal mesh.
+\emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits; asymptotic and big-O notation is used only in this limit.
 
@@
 \emph{name} & \emph{meaning} & \emph{units} \\ \hline
-$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
-$\beta$     & one implicit SIA geometry-update (and velocity) solution requires $O(n^{1+\beta})$ work \\
-$\gamma$    & one implicit, coupled Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
+$\alpha$    & one fixed-geometry Stokes velocity solution is $O(n^{1+\alpha})${\large \strut} work\\
+$\beta$     & one implicit SIA geometry (and velocity) solution is $O(n^{1+\beta})$ work \\
+$\gamma$    & one implicit, coupled Stokes geometry and velocity solution is $O(n^{1+\gamma})$ work \\
 $D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
@@
 }
-\caption{Parameters for performance analysis.  Note $\alpha$, $\beta$, $\gamma$, and $m$ are pure numbers.}
+\caption{Parameters for performance analysis.  Note $\alpha$, $\beta$, $\gamma$, and $m$ are pure, unit-less numbers.}
 \label{tab:notation}
@@
 
-If the map-plane model domain is of width $L$ then these primary parameters are related by
+Assuming the map-plane extent of the model domain is 2D and of width $L$, these primary parameters are related by
 \begin{equation}
-\Delta x = O\left(\frac{L}{\sqrt{m}}\right) \quad \text{and} \quad m = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
+\Delta x \sim \frac{L}{m^{1/2}} \quad \text{and} \quad m \sim \frac{L^2}{\Delta x^2}.  \label{eq:paramrelation}
 \end{equation}
-For a given domain width $L$ there are $O(\sqrt{m})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$.
-However, our analysis addresses spatially-3D models with map-plane horizontal meshes.}
+That is, for fixed domain width $L$ there are $O(m^{1/2})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$, but this paper addresses 3D models with map-plane (2D) horizontal meshes.}
 
-A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.
+A numerical glacier or ice sheet model uses $m$ ice thickness or surface elevation variables, i.e.~one \emph{degree of freedom} per mesh node, including at ice-free nodes where the thickness has value zero.
 Storing these model state variables, plus the thermodynamical state, requires $O(m)$ memory if the mesh/grid has \emph{a priori} bounded resolution in the vertical direction.
-The amount of fast memory needed by the simulation is also $O(m)$ if prior states are discarded or transferred to storage.
-Such ice sheet models also have $O(m)$ velocity variables, but these are not state variables.
-That is, a very-viscous stress balance computes velocity as a function of the true state variables.
+The amount of computer memory needed by the simulation is also $O(m)$, assuming prior states are discarded.
 
-Our assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,BrinkerhoffJohnson2015,Hoffmanetal2018,Lengetal2012,
+Such models also have $O(m)$ velocity variables, but note that these are not state variables.
+That is, a very-viscous stress balance computes velocity as a function of the true state variables, especially ice thickness or surface elevation.
+
+The above assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,BrinkerhoffJohnson2015,Hoffmanetal2018,Lengetal2012,
 Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.
@@
 
-Ice sheet models resolve climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and longer scales.
+Glacier and ice sheet models necessarily resolve and integrate ice-climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and on longer scales.
 Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.
 Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.
-Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ describes the frequency on which ice geometry is updated using ice velocity, i.e.~via solution of the mass continuity or surface kinematical equation.
+Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ describes the \emph{coupling} frequency on which ice geometry needs to be updated via solution of the mass continuity or surface kinematical equation.
 
-Current-technology ice sheet models use explicit time-stepping which is only conditionally stable.
-For the spatial resolutions used in present-day scientific applications, we observe that maintenance of explicit time-stepping stability requires steps substantially shorter than $1/q$ model years.
+Current-technology glacier and ice sheet models use explicit time-stepping which is only conditionally stable.
+For the spatial resolutions used in present-day scientific applications, maintenance of explicit time-stepping stability requires steps substantially shorter than $1/q$ model years.
+That is, the performance of current-generation models is limited by numerical analysis choices, and not by scientific needs.
 
-For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value, i.e.~of the $d$ in equation \eqref{eq:siamasscontinuity}.
-For Stokes or other membrane-stress-resolving dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, represents the \emph{optimistic} paradigm.
+For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996}, where $D$ is a representative diffusivity value, i.e.~of $d$ in equation \eqref{eq:siamasscontinuity}.
+For Stokes or other membrane-stress-resolving dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, might be said to represent the \emph{optimistic} paradigm.
 The corresponding \emph{pessimistic} paradigm requires $\Delta t < O(D^{-1} \Delta x^2)$, using a representative diffusivity value $D$ computed in the SIA manner.
@@
 Explicit time-stepping with hybrid and higher-order schemes is somewhat better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.
-Some hybrid schemes apply the pessimistic paradigm as an adaptive restriction \citep{Winkelmannetal2011}.
+Some hybrid schemes apply the pessimistic paradigm as an adaptive restriction \citep{Winkelmannetal2011} across all spatial scales.
 Other models apparently require the user to choose a fixed time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.
-The optimistic paradigm is supported in theory for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).
+The optimistic paradigm has theoretical support for a certain higher-order DIVA scheme \citep[Equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t \approx O(\Delta x^{1.6})$ \citep[Figure 3(a)]{Robinsonetal2022}.
+An intermediate power, a restriction $\Delta t = O(\Delta x^\omega)$ for some fuzzy value $1.5<\alpha<2$, also aligns with this author's experience.
 
-Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$.
-This restriction reflects the simulation purpose, not maintenance of stability.
-For an implicit scheme the frequency of climate coupling determines the total simulation cost, according to the (large) solution cost, at each time step, of coupled mass and momentum equations.
-
-For each explicit time step of a model using a membrane-stress-resolving balance the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.
-We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.
-For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 1$ if sparsity is exploited or $\alpha \approx 2$ if not \citep{Bueler2021}.
+Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$, but this restriction reflects the simulation purpose, not maintenance of stability.
+For an implicit scheme the desired frequency of ice-climate interaction determines the total simulation cost only according to the (large) solution cost, at each time step, of solving coupled mass and momentum equations, including the surface kinematical or mass continuity equation.
 
@@
 
-By contrast, a multigrid method \citep{Trottenbergetal2001} can greatly reduce $\alpha$.
+For each explicit time step of a model which applies a membrane-stress-resolving balance, the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.
+Let us assume that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.
+For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 1$ if sparsity is exploited or $\alpha \approx 2$ if not \citep{Bueler2021}.
+However, a multigrid method \citep{Trottenbergetal2001} can greatly reduce $\alpha$, and the ideal value $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.
 For example, Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[Table 8.1]{IsaacStadlerGhattas2015}.
 The \cite{Lengetal2012} algebraic-multigrid Stokes solver may have similar scaling, but reported results do not constrain $\alpha$.
-For a higher-order stress balance on the Greenland ice sheet, \cite{Tuminaroetal2016} report total algebraic-multigrid-preconditioned iterations suggesting $\alpha \approx 0.05$,\footnote{See Table 7.5.
+For a higher-order stress balance on the Greenlandic ice sheet, \cite{Tuminaroetal2016} report total algebraic-multigrid-preconditioned iterations suggesting $\alpha \approx 0.05$,\footnote{See \citep[Table 7.5]{Tuminaroetal2016}.
 Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.} and a geometric multigrid method by \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.
+Observe that the SIA velocity computation, a trivialization of the Stokes problem in which velocity is computed by a pointwise formula, requires optimal $O(m)$ flops.
 
-Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.
-At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula; an SIA velocity solution therefore requires optimal $O(m)$ flops.
-
-When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, strongly depending on solver design.
+When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and it depends strongly on solver design.
 Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling.
 Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.
-For example, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good parallel scaling, something we do not address here.
+Also, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good \emph{parallel} scaling, something not addressed here.
 
-Regardless of the stress balance, an explicit time-stepping scheme then applies the mass continuity equation to update the ice thickness using $O(m)$ work.
-That is, once the velocity is computed for the previous time-step's geometry, an explicit scheme replaces old thickness values by new ones using a pointwise formula.
-Any additional computation needed to remesh the updated geometry, a design-dependent cost, is omitted here.
+Regardless of the stress balance, an explicit time-stepping scheme will first compute velocity from current geometry and then apply the mass continuity equation to update the ice thickness using $O(m)$ work.
+That is, once the velocity is computed from the last-known geometry, an explicit scheme replaces old thickness values by new ones using a pointwise formula.\footnote{Any additional computation needed to remesh the updated geometry, a highly design-dependent cost, is omitted here.}
 
@@
 Now, how many flops are needed to simulate one model year?
-Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ steps per model year.
-We may write stability restrictions as required numbers of steps per model year.
-That is, $\Delta t^{-1}$ is bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.
-Recalling scaling \eqref{eq:paramrelation} of $\Delta x$ with $m$, in the explicit SIA and pessimistic-Stokes cases we have
+Suppose a numerical model takes time steps of $\Delta t$ model years duration, equivalently $\Delta t^{-1}$ steps per model year.
+In these terms one may write stability restrictions as required numbers of steps per model year.
+That is, stability requires that $\Delta t^{-1}$ be bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.
+Recalling the scaling in Equation \eqref{eq:paramrelation}, of $\Delta x$ with $m$, the explicit SIA and pessimistic-Stokes cases have
 \begin{equation}
@@
 \end{equation}
+
 The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.
@@
 
-As already explained, unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve climatic interactions.
-On the other hand, the per-step expense is much greater because nontrivial coupled equations, indeed a free-boundary problem, must be solved for the velocity and geometry-update simultaneously.
-For SIA models we parameterize the flops of such coupled solutions as $O(m^{1+\beta})$ with $\beta \ge 0$.
-A large constant is assumed to be present.
+As already explained, unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve ice-climate interactions.
+On the other hand, the per-step expense is much greater because nontrivial coupled equations, and a free-boundary problem, must be solved for the velocity and geometry-update simultaneously.
+For SIA models, let us assume that the flops of such coupled solutions scale as $O(m^{1+\beta})$ with $\beta \ge 0$; a large constant is assumed.
 
@@
 
-The only implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.
-For simplified dome geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.
+The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.
+For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.
 (An earlier non-Newton implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.

### 2022-10-07 paper/perfmod.tex
it is hard work to describe what Ua does

@@
 A scheme which does not allow the map-plane ice-covered region to change during a time step cannot respond physically (conservatively) to a change in climate inputs; such a scheme is not actually implicit for the coupled problem and it cannot validly compute long-duration time steps.
-Thus we say that a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during its coupled geometry-velocity solution.
-A scheme is only semi-implicit if it computes margin advance or retreat after the velocity field is updated.
+Thus we say that a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during its coupled geometry-velocity solution, or if the velocity is held fixed during a time step as the geometry is updated.
+A scheme which computes margin advance or retreat after the velocity field is updated is only semi-implicit, for example.
 Finally, a scheme is \emph{explicit} for the coupled problem, and uses a time-step \emph{splitting} scheme \citep{HundsdorferVerwer2003}, if the ice geometry is held fixed during the velocity solution, and then this (accepted) velocity solution is used to update the geometry, e.g.~through an advection step for Equation \eqref{eq:masscontinuity}.
@@
 For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.
-(An earlier non-Newton implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.
-Practical steady-state ($\Delta t=\infty$) solutions at ice sheet scale are also demonstrated by \cite{JouvetBueler2012} and \cite{Bueler2016}.
+(An earlier steady, Picard iteration implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is also robust for implicit time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, irregular, bed topographies.
 
-The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are also solved implicitly.
-\cite{JouvetGraeser2013} use a multigrid method, but their published, simplified-geometry results do not constrain $\beta$.
-Overall time-stepping in these hybrids is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects the thickness, so the SIA-portion solvers are apparently only tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$ for $U$ which scales according to the sliding.
+Certain recent numerical models are semi-implicit in innovative ways.
+The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are solved implicitly, with nonnegativity of thickness in their SIA portions enforced as part of a free-boundary solution.
+However, the time-stepping in these models is only semi-implicit (i.e.~time split) because the membrane-stress-resolving portion of the velocity explicitly advects thickness.
+The SIA time-step solvers in these schemes have apparently only been tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$, for $U$ derived from sliding speeds.
+Also, \cite{JouvetGraeser2013} use multigrid methods, but their published results do not constrain $\beta$.
 % equation (20) in JouvetGraeser2013
 % equation (A21) in BrinkerhoffJohnson2015
+In a different design, the semi-implicit scheme in the open-source \'Ua numerical model \citep{Gudmundsson2021manual} implicitly solves the mass continuity equation, using an active set method, but while holding the membrane-stress-resolved velocity fixed.
+Note that the formally third-order time-stepping in this model \citep{Gudmundsson2013}, while desirable for marine ice sheets, cannot achieve that accuracy when the nonnegativity-of-thickness constraint is actively determining the free boundary, e.g.~at retreating margins.
+% equations (1), (26) in Gudmundson2013
+% see semi-implicit comments in commit 9081fdbe3cf250d9cfd66504bf03134043ba2547
+% https://github.com/GHilmarG/UaSource/blob/master/NexthTG3in2D.m: "Third order impicit Taylor-Galerkin (implict with respect to h, not u and v)"
+% https://github.com/GHilmarG/UaSource/blob/master/SSS2dPrognostic.m: "semi-implicit forward integration. implicit with respect to thickness (h) explicit with respect to velocity"
+% and "error('Semi-implicit time integration only implemented for SSTREAM and not SSHEET or Hybrid formulations \n')"
 
-For implicit Stokes time-stepping, a prospective, coupled, and free-boundary velocity and geometry-update solve is assumed to be, in the absence of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.
-One might also suppose $\gamma\ge \beta$, but in any case there are no implemented cases to measure.
-These comments complete Table \ref{tab:performancemodel}, in which all estimates involve a scheme-dependent constant, something which is especially large for the implicit schemes.
-
-% regarding the nMCD method proposed in [Bueler unpublished], which is expensive but multigrid, it may turn out that $\gamma$ is small
+For implicit Stokes time-stepping, a coupled and free-boundary velocity and geometry-update solve is assumed to be, in the absence of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.
+One might also suppose $\gamma\ge \beta$, but there are no implemented cases to measure.
+These comments complete Table \ref{tab:performancemodel}, in which all implicit scheme estimates involve an especially-large scheme-dependent constant.
 
@@
 
-From Table \ref{tab:performancemodel} we first observe a known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that effort, here flops per model year, scales as $O(\Delta x^{-4})$.
+From Table \ref{tab:performancemodel} one first observes a known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that computational effort, here flops per model year, scales as $O(\Delta x^{-4})$.
 Recall that this follows because $\Delta t < O(\Delta x^2)$, and because the expense of one geometry-update operation is $O(m) = O(\Delta x^{-2})$.
@@
 The \cite{Bueler2016} implicit SIA solver is not enough better, however.
-Although long time steps can be taken by this implicit solver, the $\beta=0.8$ scaling gives $O(\Delta x^{-3.6})$ effort because $2+2\beta=3.6$.
-However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably superior performance compared to an explicit, $O(\Delta x^{-4})$ SIA scheme.
+Although long time steps can be taken by this implicit solver, the $\beta=0.8$ scaling gives $O(\Delta x^{-3.6})$ effort.
+However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably superior performance compared to an explicit SIA scheme.
 
-Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of solvers completely dominates.
-In particular, an $\alpha=1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition.
-Optimistic stability yields still-bad $O(\Delta x^{-5})$.
-These simple observations emphasize the key role of multigrid-based stress balance solvers.
+Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of solvers dominates performance concerns.
+In particular, an $\alpha=1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition, and optimistic stability yields a still-bad $O(\Delta x^{-5})$ rate.
+These simple observations emphasize the key role which will be played by multigrid-based stress balance solvers.
 That is, even retaining explicit time-stepping, nearly-optimal ($\alpha \approx 0$) scaling of computation effort in Stokes and higher-order solvers will be necessary for routine application on high-resolution meshes.
@@
 The Table also shows why algorithmic scaling remains important in the large-domain $L\to\infty$ limit when applying Stokes or higher-order dynamics.
-An $\alpha=1$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.
-By contrast, the computational work of a nearly-optimal solver will be proportional to ice sheet area $L^2$.
+The computational work of an optimal solver will be proportional to ice sheet area $L^2$.
+However, an $\alpha=1$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.
 
-The promise of nearly-optimal solvers is profoundly revealed when implicit geometry updates are considered.
-Future methods which simultanously satisfy the mass and momentum equations at each time step, and which do work essentially proportional to the number of degrees of freedom, have the greatest promise.
-A $\gamma \approx 0$ implicit Stokes method would be hugely more capable for many tasks.
-Specifically, and more achievably, a $\gamma < 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method, presumably based on multigrid solution of the free-boundary problem for the coupled mass and momentum equations, is an appropriate goal for coming decades of research on numerical ice sheet models.
-The computation cost would scale at $O(\Delta x^{-3})$, better than explicit SIA models, and the method would only update ice geometry when determined by the scientific need to resolve climate coupling, while avoiding shallow approximations.
-The same goal makes sense for all membrane-stress-resolving solvers.
-No apparent technical progress has yet been made on such an implicit Stokes design, and so these aspirations are decidedly long-term.
+The promise of nearly-optimal solvers is further revealed when implicit, coupled geometry-velocity updates are considered.
+If they become possible, methods which simultanously satisfy the mass and momentum equations at each time step, by doing work (essentially) proportional to the number of degrees of freedom have the greatest promise.
+A $\gamma \approx 0$ implicit Stokes method would be capable of many new tasks.
+More achievably, a $\gamma < 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method, presumably based on multigrid solution of the free-boundary problem for the coupled mass and momentum equations, is an appropriate goal for coming decades of research on numerical ice sheet models.
+The computation cost would scale at $O(\Delta x^{-3})$, better than explicit SIA models, and the method would only update ice geometry when needed by ice-climate coupling, while avoiding shallow approximations.
+(The same goal makes sense for any membrane-stress-resolving solver.)  No apparent technical progress has yet been made on such an implicit Stokes design, and these aspirations are decidedly long-term.
 However, the above discussion suggests why measured values for $\alpha,\beta,\gamma$, or equivalent algorithmic scaling measures, are important performance metrics to report when describing new ice sheet solvers.
-
-\subsubsection{Note added 4 July 2022.}  Errors in the first posted version (\url{https://arxiv.org/abs/2206.14352}) are corrected here.
 

### 2022-10-08 paper/perfmod.tex
these edits complete my revised version, but I need to finish the responses document

@@
 
-In other words, these current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system describing the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use velocity computed from the geometry at the start of the time step \citep[for example]{Chengetal2017}.}  However, at least for simpler, textbook partial differential equation problems, the limited and conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
+In other words, these current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system describing the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which fix the velocity following the computation which updates the geometry \citep[for example]{Chengetal2017}.}  However, at least for simpler, textbook partial differential equation problems, the limited and conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
 Stability conditions of explicit SIA models appeared some time ago \citep[e.g.][]{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.
@@
 Here the velocity and geometry are updated simultaneously by solving coupled mass and momentum conservation equations.
-While such implicit time-stepping requires the solution of systems of equations at each step, for glaciers and ice sheets an implicit step must also \emph{simultaneously} compute the velocity field and the \emph{domain on which the velocity is defined}, namely the 3D extent of the ice once the coupled solution has converged.
+While such implicit time-stepping requires the solution of systems of equations at each step, for glaciers and ice sheets an implicit step must simultaneously compute the velocity field and the \emph{domain on which the velocity is defined}, namely the 3D extent of the ice once the coupled solution has converged.
 The problem is of free-boundary type, especially in map-plane (horizontal) directions \citep{SchoofHewitt2013}, but also in the (easily-resolved) vertical direction.
 
-An implicit strategy has been demonstrated at high (900m) resolution for the Greenlandic ice sheet in the simplest frozen-base, isothermal SIA case \citep{Bueler2016,JouvetGraeser2013}.
+An implicit strategy has been demonstrated at high (900m) resolution for the Greenlandic ice sheet in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.
 That work also shows how the steady-state SIA problem \citep[c.f.][]{JouvetBueler2012} can be solved, demonstrating unconditional stability.
@@
 The simplified performance analysis here exposes the most important considerations and trade-offs.
-While this author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, such beliefs should be assessed quantitatively to the extent possible.
+While this author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will lead to better modeling, such beliefs should be assessed quantitatively to the extent possible.
 
@@
 
-Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.
+Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of  advective and diffusive mechanisms.
 Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.
@@
 However, either statement \eqref{eq:masscontinuity} or \eqref{eq:siamasscontinuity} of the mass continuity equation is fundamentally incomplete without modeling the changing (map-plane) boundary position of the glacier or ice sheet.
-The problem of determining the updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).
-The mathematical role of this constraint has been understood for some time in the SIA model \citep{Calvoetal2002,JouvetBueler2012,SchoofHewitt2013}, but its role as the source of the margin location is universal across fluid-layer problems with signed source terms \citep{Bueler2021conservation}.
+The problem of determining this updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).
+The mathematical role of this constraint has been understood for some time in the SIA model \citep{Calvoetal2002,JouvetBueler2012,SchoofHewitt2013}, but its role as the determinant of margin location is universal across fluid-layer problems with signed source terms \citep{Bueler2021conservation}.
 
-It is in this context that we can distinguish time-stepping types in coupled geometry-velocity, i.e.~mass and momentum conserving, glacier and ice sheet models.
+In this context one can distinguish time-stepping types in coupled geometry-velocity, i.e.~mass and momentum conserving, glacier and ice sheet models.
 A coupled model computes (\emph{fully}) \emph{implicit} time steps if no significant aspects of the geometry or velocity are held fixed during the (coupled) solution.
-That is, in an implicit scheme a coupled and well-posed model for the updated ice surface elevation (or thickness) and velocity fields is solved.
-A scheme which does not allow the map-plane ice-covered region to change during a time step cannot respond physically (conservatively) to a change in climate inputs; such a scheme is not actually implicit for the coupled problem and it cannot validly compute long-duration time steps.
-Thus we say that a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during its coupled geometry-velocity solution, or if the velocity is held fixed during a time step as the geometry is updated.
-A scheme which computes margin advance or retreat after the velocity field is updated is only semi-implicit, for example.
-Finally, a scheme is \emph{explicit} for the coupled problem, and uses a time-step \emph{splitting} scheme \citep{HundsdorferVerwer2003}, if the ice geometry is held fixed during the velocity solution, and then this (accepted) velocity solution is used to update the geometry, e.g.~through an advection step for Equation \eqref{eq:masscontinuity}.
+That is, in an implicit scheme a coupled and well-posed model for the ice surface elevation (or thickness) and velocity field updates is solved.
+A scheme which does not allow the map-plane ice-covered region to change during the time step cannot respond physically (conservatively with respect to mass and momentum conservation) to a change in climate inputs.
+A scheme which is implicit in the sense here can be unconditionally stable; it can stably compute time steps of arbitrary duration causing nontrivial changes in margin position.
+Next, a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during the geometry-velocity solution, for instance if the velocity is held fixed during a time step as the geometry is updated (perhaps ``implicitly'' on its own).
+Thus a scheme which computes margin advance or retreat only after the velocity field update is accepted is only semi-implicit, for example.
+Finally, a scheme is \emph{explicit} for the coupled problem if the ice geometry is held fixed during the velocity solution, and then this (accepted and now fixed) velocity solution is used to update the geometry through an explicit step for Equation \eqref{eq:masscontinuity}.
 
-Thus a key point about time-stepping schemes for the coupled geometry and velocity problem for glaciers and ice sheets is that implicit schemes are not merely solving fixed-domain systems of coupled PDEs.
-The mathematical problem for the coupled and implicit time-step includes the condition which controls the free boundary, namely the inequality constraint of nonnegative thickness \citep{Bueler2021conservation}.
+Thus a key point about time-stepping schemes for the coupled geometry and velocity problem for glaciers and ice sheets is that implicit schemes are not solving fixed-domain systems of coupled PDEs.
+The mathematical problem for the coupled and implicit time-step includes the condition which controls the moving and free boundary, namely the inequality constraint of nonnegative thickness.
 Implicit schemes which solve these free-boundary problems at each time step can, at high spatial resolution, transcend stability limitations on time step duration.
-They can do time-stepping at a rate controlled by climatic-input time scales.
+They, and only they, can do time-stepping at a rate controlled only by ice-climate interaction time scales.
 
@@
 
-Table \ref{tab:notation} lists the parameters used in the numerical model performance analysis of this section.
-The primary parameters are $\Delta x$, which is a representative value for the horizontal mesh (grid) cell diameter, and $m$, which is the number of nodes (vertices) in the horizontal mesh.
+Table \ref{tab:notation} lists the parameters used in the numerical model performance analysis which follows.
+The primary parameters are $\Delta x$, which is a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.
 \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits; asymptotic and big-O notation is used only in this limit.
@@
 $m$         & degrees of freedom: number of nodes in the horizontal mesh \\
-$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
-$\Delta t$  & length of time step in model years & a \\
+$q$         & time steps per model year needed to resolve ice-climate interactions & $\text{a}^{-1}$ \\
+$\Delta t$  & length of time step & a \\
 $U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
@@
 \end{equation}
-That is, for fixed domain width $L$ there are $O(m^{1/2})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$, but this paper addresses 3D models with map-plane (2D) horizontal meshes.}
+Specifically, for fixed domain width $L$ there are $O(m^{1/2})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$, but this paper addresses 3D models with map-plane (2D) horizontal meshes.}
 
@@
 Such models also have $O(m)$ velocity variables, but note that these are not state variables.
-That is, a very-viscous stress balance computes velocity as a function of the true state variables, especially ice thickness or surface elevation.
+That is, a very-viscous stress balance computes velocity as a function of the true state variables, such as ice thickness and temperature.
 
@@
 Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.
-Some solvers use 3D refinement \citep{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016}, with various distinctions between how horizontal and vertical meshing is handled.
-This and many other details cannot be pursued here.
+Some solvers use 3D refinement \citep{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016}, with various distinctions between how horizontal and vertical meshing is handled, but this and many other details cannot be pursued here.
 
-Glacier and ice sheet models necessarily resolve and integrate ice-climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and on longer scales.
+Glacier and ice sheet models resolve and integrate ice-climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and on longer scales.
 Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.
 Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.
-Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ describes the \emph{coupling} frequency on which ice geometry needs to be updated via solution of the mass continuity or surface kinematical equation.
+Note that energy balance and degree-day schemes for computing surface mass balance \citep{GreveBlatter2009} generally have much shorter time scales, but here $q$ describes the \emph{coupling} frequency on which ice geometry needs to be updated via solution of the mass continuity or surface kinematical equation.
 
@@
 For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996}, where $D$ is a representative diffusivity value, i.e.~of $d$ in equation \eqref{eq:siamasscontinuity}.
-For Stokes or other membrane-stress-resolving dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, might be said to represent the \emph{optimistic} paradigm.
-The corresponding \emph{pessimistic} paradigm requires $\Delta t < O(D^{-1} \Delta x^2)$, using a representative diffusivity value $D$ computed in the SIA manner.
+For Stokes models, or other membrane-stress-resolving dynamics, the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, might be said to represent the \emph{optimistic} paradigm.
+The corresponding \emph{pessimistic} paradigm for Stokes models says $\Delta t < O(D^{-1} \Delta x^2)$, using a representative diffusivity value $D$ computed in the SIA manner.
 
@@
 The optimistic paradigm has theoretical support for a certain higher-order DIVA scheme \citep[Equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t \approx O(\Delta x^{1.6})$ \citep[Figure 3(a)]{Robinsonetal2022}.
-An intermediate power, a restriction $\Delta t = O(\Delta x^\omega)$ for some fuzzy value $1.5<\alpha<2$, also aligns with this author's experience.
+An intermediate power, a restriction $\Delta t = O(\Delta x^\omega)$ for some fuzzy value $1.5<\omega<2$, also aligns with this author's experience.
 
 Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$, but this restriction reflects the simulation purpose, not maintenance of stability.
-For an implicit scheme the desired frequency of ice-climate interaction determines the total simulation cost only according to the (large) solution cost, at each time step, of solving coupled mass and momentum equations, including the surface kinematical or mass continuity equation.
+For an implicit scheme the desired frequency of ice-climate interaction will, however, determine the total simulation cost according to the (large) solution cost, at each time step, of solving coupled mass and momentum equations, including the surface kinematical or mass continuity equation.
 
@@
 However, a multigrid method \citep{Trottenbergetal2001} can greatly reduce $\alpha$, and the ideal value $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.
-For example, Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[Table 8.1]{IsaacStadlerGhattas2015}.
+For example, Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[Table 8.1]{IsaacStadlerGhattas2015}.
 The \cite{Lengetal2012} algebraic-multigrid Stokes solver may have similar scaling, but reported results do not constrain $\alpha$.
@@
 When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and it depends strongly on solver design.
-Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling.
-Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.
-Also, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good \emph{parallel} scaling, something not addressed here.
+Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling, as actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.
+Many models \citep[e.g.][]{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} also show good \emph{parallel} scaling, something not addressed here.
 
 Regardless of the stress balance, an explicit time-stepping scheme will first compute velocity from current geometry and then apply the mass continuity equation to update the ice thickness using $O(m)$ work.
-That is, once the velocity is computed from the last-known geometry, an explicit scheme replaces old thickness values by new ones using a pointwise formula.\footnote{Any additional computation needed to remesh the updated geometry, a highly design-dependent cost, is omitted here.}
+That is, once the velocity is computed from the last-known geometry, let us assume that an explicit scheme replaces old thickness values by new ones using an optimal pointwise formula.\footnote{Any additional computation needed to remesh the updated geometry, a highly design-dependent cost, is omitted here.}
 
@@
 Suppose a numerical model takes time steps of $\Delta t$ model years duration, equivalently $\Delta t^{-1}$ steps per model year.
-In these terms one may write stability restrictions as required numbers of steps per model year.
+In these terms stability restrictions give required numbers of steps per model year.
 That is, stability requires that $\Delta t^{-1}$ be bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.
-Recalling the scaling in Equation \eqref{eq:paramrelation}, of $\Delta x$ with $m$, the explicit SIA and pessimistic-Stokes cases have
+Recalling the scaling in Equation \eqref{eq:paramrelation}, the explicit SIA and pessimistic-Stokes cases have
 \begin{equation}
@@
 \end{equation}
-
-The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.
+The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$ for Stokes models, to give a work estimate for each model year in a simulation.
 The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.
 
-As already explained, unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve ice-climate interactions.
-On the other hand, the per-step expense is much greater because nontrivial coupled equations, and a free-boundary problem, must be solved for the velocity and geometry-update simultaneously.
-For SIA models, let us assume that the flops of such coupled solutions scale as $O(m^{1+\beta})$ with $\beta \ge 0$; a large constant is assumed.
+As already explained, unconditionally-stable implicit methods may have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve ice-climate interactions.
+On the other hand, the per-step expense is much greater because nontrivial coupled equations, and a free-boundary problem, must be solved simultaneously for the updated velocity and geometry.
 
 % regarding next paragraph, see data/bueler.perf and data/scaling.m
-
+For SIA models, let us assume that the flops of such coupled solutions scale as $O(m^{1+\beta})$ with $\beta \ge 0$; a large constant is assumed.
 The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.
 For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.
-(An earlier steady, Picard iteration implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is also robust for implicit time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, irregular, bed topographies.
+(An earlier steady, Picard iteration implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is also robust for implicit time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic and irregular bed topographies.
 
@@
 The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are solved implicitly, with nonnegativity of thickness in their SIA portions enforced as part of a free-boundary solution.
-However, the time-stepping in these models is only semi-implicit (i.e.~time split) because the membrane-stress-resolving portion of the velocity explicitly advects thickness.
+However, the time-stepping in these models is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects thickness.
 The SIA time-step solvers in these schemes have apparently only been tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$, for $U$ derived from sliding speeds.
-Also, \cite{JouvetGraeser2013} use multigrid methods, but their published results do not constrain $\beta$.
+(Also, \cite{JouvetGraeser2013} use multigrid methods, but their published results do not constrain the power $\beta$.)
 % equation (20) in JouvetGraeser2013
 % equation (A21) in BrinkerhoffJohnson2015
-In a different design, the semi-implicit scheme in the open-source \'Ua numerical model \citep{Gudmundsson2021manual} implicitly solves the mass continuity equation, using an active set method, but while holding the membrane-stress-resolved velocity fixed.
-Note that the formally third-order time-stepping in this model \citep{Gudmundsson2013}, while desirable for marine ice sheets, cannot achieve that accuracy when the nonnegativity-of-thickness constraint is actively determining the free boundary, e.g.~at retreating margins.
+In a different design, the semi-implicit scheme in the open-source \'Ua numerical model \citep{Gudmundsson2021manual} implicitly solves mass continuity Equation \eqref{eq:masscontinuity} using an active set method, but while holding the membrane-stress-resolved velocity fixed.
+Note that the formally third-order time-stepping in this model \citep{Gudmundsson2013}, while surely desirable for marine ice sheets, cannot achieve that accuracy when the nonnegativity-of-thickness constraint is actively determining the free boundary, e.g.~at retreating margins.
 % equations (1), (26) in Gudmundson2013
@@
 One might also suppose $\gamma\ge \beta$, but there are no implemented cases to measure.
-These comments complete Table \ref{tab:performancemodel}, in which all implicit scheme estimates involve an especially-large scheme-dependent constant.
+These comments complete Table \ref{tab:performancemodel}.
+Note that all implicit scheme estimates involve an especially-large scheme-dependent constant.
 
@@
 
-From Table \ref{tab:performancemodel} one first observes a known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that computational effort, here flops per model year, scales as $O(\Delta x^{-4})$.
-Recall that this follows because $\Delta t < O(\Delta x^2)$, and because the expense of one geometry-update operation is $O(m) = O(\Delta x^{-2})$.
+From Table \ref{tab:performancemodel} one first observes a well-known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that computational effort, here flops per model year, scales as $O(\Delta x^{-4})$.
+This follows because $\Delta t < O(\Delta x^2)$, and because the expense of one geometry-update operation is $O(m) = O(\Delta x^{-2})$.
 Spatial mesh refinement by a factor of two therefore imposes an impressive 16-times increase in effort.
@@
 Although long time steps can be taken by this implicit solver, the $\beta=0.8$ scaling gives $O(\Delta x^{-3.6})$ effort.
-However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably superior performance compared to an explicit SIA scheme.
+However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably-superior $O(\Delta x^{-3})$ performance or better.
 
-Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of solvers dominates performance concerns.
-In particular, an $\alpha=1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition, and optimistic stability yields a still-bad $O(\Delta x^{-5})$ rate.
+Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of membrane-stress-resolving solvers dominates performance concerns.
+In particular, an $\alpha=1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition.
+Optimistic stability yields a still-bad $O(\Delta x^{-5})$ rate.
 These simple observations emphasize the key role which will be played by multigrid-based stress balance solvers.
@@
 
-The promise of nearly-optimal solvers is further revealed when implicit, coupled geometry-velocity updates are considered.
-If they become possible, methods which simultanously satisfy the mass and momentum equations at each time step, by doing work (essentially) proportional to the number of degrees of freedom have the greatest promise.
+The promise of nearly-optimal solvers is truly revealed, however, when implicit, coupled geometry-velocity updates are considered.
+If they become possible, methods which simultanously satisfy the mass and momentum equations at each time step, doing work (essentially) proportional to the number of degrees of freedom, have great promise.
 A $\gamma \approx 0$ implicit Stokes method would be capable of many new tasks.
@@
 The computation cost would scale at $O(\Delta x^{-3})$, better than explicit SIA models, and the method would only update ice geometry when needed by ice-climate coupling, while avoiding shallow approximations.
-(The same goal makes sense for any membrane-stress-resolving solver.)  No apparent technical progress has yet been made on such an implicit Stokes design, and these aspirations are decidedly long-term.
+(The same goal makes sense for any membrane-stress-resolving solver.)  Little technical progress has yet been made on such a coupled, fully-implicit, and scalable Stokes design \cite[but see][]{WirbelJarosch2020}, so these aspirations are decidedly long-term.
 However, the above discussion suggests why measured values for $\alpha,\beta,\gamma$, or equivalent algorithmic scaling measures, are important performance metrics to report when describing new ice sheet solvers.

### 2022-10-09 paper/reviews/response.tex
reviewers handled

--- 2022-09-26 paper/reviews/response.tex
@@
 The thoughtful comments of the two reviewers are much appreciated.
+Below I both respond to their concerns---in block italics---and, in the important cases, quote the new or rewritten content of the paper which addresses these concerns.
 
@@
 Let me congratulate you to this much needed performance analysis of ice sheet models, which is presented clearly and in detail in the manuscript at hand. I have only a few very minor comments to add.
+\end{review}
 
+\noindent Thank you!
+
+\begin{review}
 As you include glaciers in your introduction (L\# 21) you could reference Clarke et al (Clarke, G.
@@
 
-\noindent Thank you!  This is a good example to add, and I have done so.
+\noindent This is a good example to add, and I have done so.
 
@@
 
-In lines \# 82 and on page 8 just befor equation (4), maybe use the word "equation" alongside the reference to an equation, as you do e.g. in L\#95 and other places, just for consistency.
+In lines \# 82 and on page 8 just before equation (4), maybe use the word "equation" alongside the reference to an equation, as you do e.g. in L\#95 and other places, just for consistency.
 \end{review}
@@
 
-Thanks.
+\noindent Thanks.
 
@@
 \noindent Some thought on a revised title is worthwhile, so this comment is welcomed.
+
 Unfortunately ``Asymptotic analysis of implicit versus explicit methods in ice sheet models'' would not clarify which is the asymptotic limit.
-(It is not aspect ratio $H/L \to 0$, nor convergence $\Delta x \to 0$, nor approach to steady state $t\to \infty$, but rather number of degrees of freedom.)  Furthermore, my intent is to address \emph{performance}, as a broader issue which relates to multiple modeling choices, specifically mesh resolution, time-stepping type (explicit vs.~implicit), solver design (multigrid versus not), and even operational ice-climate coupling frequency, all over long-duration runs.
+(It is not aspect ratio $H/L \to 0$, nor approach to steady state $t\to \infty$, nor even convergence under the high-resolution $\Delta x \to 0$, but rather performance in the number of degrees of freedom limit.)  My intent is to address \emph{performance}, as a broader issue which relates to multiple modeling choices, specifically mesh resolution, time-stepping type (explicit vs.~implicit), solver design (multigrid versus not), and even operational ice-climate coupling frequency, all over long-duration runs.
 
-\medskip \noindent In other words, I don't see how to write a title of acceptable length without requiring the interested reader to (at least) dive into the abstract for clarification of some brief phrase such as ``performance analysis''.
-(Of course, a shorter title will, itself, increase the number of such interested readers.)  A possible title is ``Performance analysis of numerical ice sheet simulations in the high-resolution limit'', and this captures the correct asymptotic, but I find the current title to be essentially as clear, and clearly briefer.
+In other words, I don't see how to write a title of acceptable length without requiring the interested reader to (at least) dive into the abstract for clarification of some brief phrase such as ``performance analysis''.
+(A shorter title may itself increase the number of such interested readers.)  A possible title is ``Performance analysis of numerical ice sheet simulations in the high-resolution limit'', and this identifies the asymptotic, but I find the current title to be essentially as clear, and clearly briefer.
 Does the Editor have a title suggestion?
@@
 
-\begin{quote}  In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows need yet finer resolution.\end{quote}
+\begin{quote}``In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows need yet finer resolution.''\end{quote}
 
-\noindent A few points about this sentence will, I think, be understood by any reader with minimal exposure to ice sheet modeling:
+\noindent A few points about this new sentence will, I think, be understood by any reader with minimal exposure to ice sheet modeling:
 \begin{itemize}
 \item This is a necessarily-imprecise description of practical resolution needs for ``resolving'' physical mechanisms understood to be important by a community of scientists.
-This is not a mathematically or statistically based assertion.
-\item The added phrase ''ice streams as fluid features'' refers to model results which are necessarily more than one grid cell wide.
-As fluids are locally infinite-dimensional, you need a few cells, more than one or two across for ice streams, to capture a ``fluid feature''.
+This is not a mathematically- or statistically-based assertion.
+\item The added phrase ''ice streams as fluid features'' refers to aspects of model results which are necessarily more than one grid cell wide.
+That is, as fluids are locally infinite-dimensional, you need a few cells, more than one or two across for ice streams, to capture a ``fluid feature''.
+\item 10 km resolution is itself not enough for yet narrower features.
 \item The added word ``it is \underline{now} generally accepted'' clearly refers to expectations in the 2020s.
-There were indeed ``important'' 20km results before now, but I would disagree with any assertion in a present-day journal submission that, for example, flux across grounding lines in Antarctica is ``validly modeled'' at 20 km resolution; that would be a consequence of not trying very hard, or choosing a deficient model.
-\item 10 km resolution is itself not enough for narrower features than ice streams.
+There were indeed ``important'' 20km results before now, but I would disagree with any assertion in a present-day journal submission that, for example, flux across grounding lines in Antarctica is ``validly modeled'' at 20 km resolution; that would now be a consequence of not trying very hard, or choosing a deficient model.
 \end{itemize}
@@
 
-\noindent FIXME This reviewer comment drives me to make the most significant alterations to the paper.
-The reviewer, and I suspect many readers, have never imagined an unconditionally stable implicit ice sheet model, so they don't recognize the only semi-implicit nature of various assertions of implicitness in the literature.
-The SIA solver in Bueler (2016) is actually implicit, and demonstrably unconditionally stable on flat beds, in the sense that one can take a single time step of arbitrary length to construct the geometry of an ice sheet from knowledge only of the surface mass balance (accumulation/ablation function) and the bed.
-This means a free-boundary problem must be solved during the time step to determine the map-plane region occupied by the ice.
-Otherwise, in the cited models wherein an ``implicit'' claim is made, in fact a major aspect of the geometry at the current time step, specifically the ice-covered domain of the map-plane, is \emph{fixed} during the time step.
-Only the vertical extent of the ice is updated implicitly.
+\noindent This reviewer comment drives me to make the most significant addition to the paper.
+The reviewer, and I suspect many readers, may never have fully imagined an unconditionally-stable and implicit ice sheet model, so they don't recognize the only semi-implicit nature of various assertions of implicitness in the literature.
+The only such demonstrated solver is the SIA implementation in Bueler (2016), though in fact the idea has existed as a theoretical matter since Calvo et al (2002).
+
+That is, I recognize that I must work much harder to clearly define my terms.
+My major addition, below, is in the second section, which has also been re-titled from ``Mass continuity equation'' to ``Coupled geometry-velocity modeling''.
+The following three-paragraph addition ends the section with precise definitions of ``implicit'' and ``explicit'' as these terms would be applied to solving coupled mass-momentum time steps, that is, in a context in which we seek the property of unconditional stability by choosing the scheme construction called implicitness.
+
+\begin{quote}
+``However, either statement (1) or (2) of the mass continuity equation is fundamentally incomplete without modeling the evolving (map-plane) boundary position of the glacier or ice sheet.
+The problem of determining this updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).
+The mathematical role of this constraint has been understood for some time in the SIA model (Calvo and others, 2002; Jouvet and Bueler, 2012; Schoof and Hewitt, 2013), but its role as the determinant of margin location is universal across fluid-layer problems with signed source terms (Bueler, 2021a).
+
+``In this context one can distinguish time-stepping types in coupled geometry-velocity, i.e.~mass and momentum conserving, glacier and ice sheet models.
+A coupled model computes (\emph{fully}) \emph{implicit} time steps if no significant aspects of the geometry or velocity are held fixed during the (coupled) solution.
+That is, in an implicit scheme a coupled and well-posed model for the ice surface elevation (or thickness) and velocity field updates is solved.
+A scheme which does not allow the map-plane ice-covered region to change during the time step cannot respond physically (i.e.~conservatively with respect to mass and momentum conservation) to a change in climate inputs.
+A scheme which is implicit in the sense here can be unconditionally stable; it can stably compute time steps of arbitrary duration causing nontrivial changes in margin position.
+Next, a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during the geometry-velocity solution, or, for instance, if the velocity is held fixed during a time step as the geometry is updated (though perhaps ``implicitly'' on its own).
+Thus a scheme which computes margin advance or retreat only after the velocity field update is accepted is only semi-implicit, for example.
+Finally, a scheme is (\emph{fully}) \emph{explicit} for the coupled problem if the ice geometry is held fixed during the velocity solution, and then this (accepted and now fixed) velocity solution is used to update the geometry through an explicit step for Equation (1).
+
+``Thus a key point about time-stepping schemes for the coupled geometry and velocity problem for glaciers and ice sheets is that implicit schemes are not solving fixed-domain systems of coupled PDEs.
+The mathematical problem for the coupled and implicit time-step includes the condition which controls the moving and free boundary, namely the inequality constraint of nonnegative thickness.
+Implicit schemes which solve these free-boundary problems at each time step can, at high spatial resolution, transcend stability limitations on time step duration.
+They, and only they, can do time-stepping at a rate controlled only by ice-climate interaction time scales.''
+\end{quote}
+
+With these clarified definitions, let me return to the reviewer's specific comment and citations.
+Two of the three cited works are now addressed in the following substantially-rewritten paragraph at the end of the third section.
+The semi-implicit structure of the Brinkerhoff \& Johnson (2015) model is clear from their published text, and likewise the Jouvet \& Graeser (2013) model which the reviewer does not mention, but to understand the time-stepping structure of the open source \'Ua model (Gudmundsson, 2013; Gudmundsson, 2021) I looked in the source code and saw clear comments about its structural semi-implicitness.\footnote{See the relevant files \url{https://github.com/GHilmarG/UaSource/blob/master/SSS2dPrognostic.m} and \url{https://github.com/GHilmarG/UaSource/blob/master/NexthTG3in2D.m}.
+I examined commit \texttt{9081fdbe3cf250d9cfd66504bf03134043ba2547}.}  I believe I have characterized the (innovative) semi-implicitness of these models correctly.
+
+\begin{quote}
+``Certain recent numerical models are semi-implicit in innovative ways.
+The SIA portions of hybrid time-stepping schemes by Jouvet \& Graeser (2013) and Brinkerhoff \& Johnson (2015) are solved implicitly, with nonnegativity of thickness in their SIA portions enforced as part of a free-boundary solution.
+However, the time-stepping in these models is only semi-implicit because the membrane-stress-resolving portion of the velocity is held fixed as the ice thickness is advected.
+The SIA time-step solvers in these schemes have apparently only been tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$, for $U$ derived from sliding speeds.
+(Also, Jouvet \& Graeser (2013) use multigrid methods, but their published results do not constrain the power $\beta$.)  The semi-implicit scheme in the open-source \'Ua numerical marine ice sheet model (Gudmundsson, 2013; Gudmundsson, 2021) implicitly solves mass continuity free-boundary problem for Equation (1) using an active set method, but also while holding the (membrane-stress-resolved) velocity fixed.''
+\end{quote}
+
+Regarding (Shapero et al., 2021)\footnote{Shapero, D.
+R., Badgeley, J.
+A., Hoffman, A.
+O., \& Joughin, I.
+R.
+(2021).
+\emph{icepack: A new glacier flow modeling package in Python, version 1.0}.
+Geoscientific Model Development, 14(7), 4593-4616.}, I am reasonably familiar with the model and the paper, and the authors make no claim that the time stepping has any implicitness, so I have not cited it in the paper.
+I am not sure what the reviewer has in mind.
 
 \begin{review}
+\textbf{L57--60}  `Unconditional stability' is a useful shorthand here, and I'm alright with its use, but it also deserves a bit of an asterisk (stated here or elsewhere) that there is a finite radius of convergence that when combined with a practical initialization procedure can still yield some kind of time-step restriction.
+This type of thing was seen, for example, with too-rough topography in Bueler (2016).
 \end{review}
 
-\noindent x
+\noindent This interesting comment is in cognitive dissonance with the previous comment!
+Together, roughly, the reviewer's last two comments say: ``We accept other literature making an implicit claim about their schemes, even if this is not true, and when there is not even a hope of unconditional stability in such designs, but for a design where unconditional stability has been demonstrated in some cases, we must complain that it has not also been demonstrated in difficult high-resolution, realistic-data cases.''
+
+In any case the mildly-rewritten relevant text in the paper makes clear that the Bueler (2016) implicit-step solver is capable of unconditional stability because infinite time steps are computable for simplified geometry.
+Furthermore, (Bueler, 2016) demonstrates long implicit time steps for realistic, high resolution data.
+
+\begin{quote}
+``The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.
+For simplified (dome) geometry the scheme in Bueler (2016), based on Newton steps solved via direct linear algebra and single-grid ice margin determination, directly computes steady states with scaling $\beta=0.8$.
+(An earlier steady, Picard iteration implementation by Jouvet \& Bueler (2012) scales worse.)  Convergence is also robust for implicit time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic and irregular bed topographies.''
+\end{quote}
+
+The reviewer's belief that the ``asterisk'' is because of ``a finite radius of convergence that when combined with a practical initialization procedure can still yield some kind of time-step restriction'' is not accurate, nor stated or implied by (Bueler, 2016).
+Rather, even at resolutions well under 5km, down to 600m, for the entire Greenlandic ice sheet, the solver converges with regularization of the nonlinear diffusivity.
+The problem is neither a ``finite radius of convergence'' nor ``a practical initialization procedure.''  Bueler (2016) actually says (page 238) that ``highly-resolved bed, which causes large and irregular values of [diffusivity and bed-slope pseudo-velocity] limits the success of our combined continuation scheme and Newton iteration.''  Direct computation of steady states, thereby demonstrating unconditional stability, is fine until you resolve the bed topography far beyond the reviewer's ``valid'' 20 km level, and the breakdown is in the Newton iteration.\footnote{Such breakdowns of Newton iterations typically relate to the conditioning of Jacobians, and the issue might even go away with use of 128 bit reals.
+In any case it has nothing to do with the reviewer's description.}
+
 \begin{review}
+\dots In addition, there are other considerations that might yield only conditional stability: complicated boundary conditions like calving fluxes (and especially knowing where to impose them) come to mind.
 \end{review}
 
-\noindent x
+\noindent This comment is completely correct, but the paper is fully aware of it and addresses it.
+Namely, calving is evidently a matter of ice-climate interaction.
+Clearly such ``complicated boundary conditions'' will impose time-stepping constraints, and these are handled by the parameter $q$, which the paper describes as the ``number of ice-dynamical time steps per year needed to capture'' ice-climate interaction.\footnote{Other than blanket statements that ``ice sheet modeling is complicated,'' or similar, I don't know how to address the reality that looms over the paper, namely that the full, operational boundary value problems in ice sheet modeling are messy.}
+
 \begin{review}
+\textbf{L83} It might be useful to state here that velocity depends also on the gradient of the bed elevation.
 \end{review}
 
-\noindent x
+\noindent The velocity depends on everything in the stress balance boundary-value problem.
+However, the text is describing instabilities of the mass continuity equation (1), and why an advection view of that equation is insufficient.
+The relevant dependence on the surface slope, thus the slope of the (advected) thickness, is therefore emphasized.
+Given that the mass continuity equation is usually stated in the glaciers literature with \emph{all} dependencies of the velocity unmentioned, cluttering the paragraph with this particular dependence is a peculiar ask.
+The text is left as is.
+
 \begin{review}
+\textbf{L92--94} This sentence is a bit mysterious. Is this trying to describe the instability caused by violating the CFL or a different type of wiggle?
 \end{review}
 
-\noindent x
+\noindent The point about any instability, under the view contained in the standard and very-basic definition of \emph{absolute stability} of ODE systems, therefore of course including the method-of-lines semidiscretizations of all PDEs, and therefore all stability violations for discretized PDEs, is that large time steps cause overshoots.
+The author is not invoking some weird concept here; the standard absolute stability view allows consideration of CFL violations, but also all other time-dependent PDE situations.
+
+Thus the text is only modified to add a citation to the standard Leveque (2007) reference on ODE and PDE numerical methods.
+I don't know what ``a different type of wiggle'' would mean, and I would not want to pursue this for the reader, nor do I want to limit the consideration only to CFL violations, which would assume the (wrong) advection-only view of (1).
+
 \begin{review}
+\textbf{L101} I'm not sure what ‘primary’ means here. Perhaps something like ‘The parameters that we consider model efficiency with respect to are ...’.
 \end{review}
 
-\noindent x
+\noindent The text says:
+
+\begin{quote}
+``Table 1 lists the parameters used in the numerical model performance analysis which follows.
+The primary parameters are $\Delta x$, which is a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.''
+\end{quote}
+
+Any reasonable reader should clearly see that all of the parameters in Table 1 are relevant to the performance analysis, but that attention should focus first on $\Delta x$ and $m$?
+
 \begin{review}
+\textbf{L111--112} Perhaps clarify that these are diagnostic (and thus not `state variables') because the time-derivative in the more general Navier-Stokes equation has been set to zero due to the very-viscous situation (or the low Reynolds number).
 \end{review}
 
-\noindent x
+\noindent I try to avoid use of ``prognostic''/``diagnostic'' language.
+It does not have meaning in the mathematics of the systems under consideration.
+It certainly ossifies traditional distinctions between variables, equations, and discretizations as they are used in the climate-modeling world, but (young?) readers who have not been previously programmed by the traditions of that narrow world cannot look up the meaning of these terms in the broader (e.g.~applied mathematics, computational fluid dynamics, or physics) literature.
+By contrast, the meaning of ``state variables'' is fundamental to the consideration of all dynamical systems.
+
+However, the reviewer is right to suggest a reminder to the reader about the lack of a time-derivative in the Stokes model.\footnote{That is, as one way to explain why velocities are not state variables.
+The reviewer and I would probably both want to avoid a treatise on differential-algebraic equations here.}  The relevant paragraph now says:
+
+\begin{quote}
+``Such models also have $O(m)$ velocity variables, but note that these are not state variables, essentially because the Stokes model lacks time derivatives.
+That is, a very-viscous stress balance computes velocity as a function of the true state variables, such as ice thickness and temperature or enthalpy.''
+\end{quote}
+
+\begin{review}
+\textbf{L133--134} I like casting things as optimistic and pessimistic, but I think that a more technically descriptive phrasing might be ``advective stabililty condition'' and ``diffusive stability condition.''
+\end{review}
+
+\noindent Fair enough, but the two conditions are clearly ordered with respect to performance consequences, and it is that ordering that I want to emphasize.
+As the ``optimistic'' and ``pessimistic'' usage is clearly attached to the advective/diffusive stability distinction, respectively, when the usage is introduced,\footnote{The introducing text says ``an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, might be said to represent the \emph{optimistic} paradigm.
+The corresponding \emph{pessimistic} paradigm for Stokes models says $\Delta t < O(D^{-1} \Delta x^2)$, using a representative diffusivity value $D$ computed in the SIA manner.''.} I think I will stick to my branding scheme.
+
+\begin{review}
+\textbf{L212} `Not enough better'.  Not enough better for what?
+\end{review}
+
+\noindent Good point.  The text now says ``The scaling of the Bueler (2016) implicit SIA solver is not enough better than such an explicit scheme, however.''
+
+\begin{review}
+\textbf{L225} `The Table' $\to$ `Table 2'.
+\end{review}
+
+\noindent Yup.
+
 \end{document}

### 2022-10-09 paper/perfmod.tex
almost done

@@
 That is, in an implicit scheme a coupled and well-posed model for the ice surface elevation (or thickness) and velocity field updates is solved.
-A scheme which does not allow the map-plane ice-covered region to change during the time step cannot respond physically (conservatively with respect to mass and momentum conservation) to a change in climate inputs.
+A scheme which does not allow the map-plane ice-covered region to change during the time step cannot respond physically (i.e.~conservatively with respect to mass and momentum conservation) to a change in climate inputs.
 A scheme which is implicit in the sense here can be unconditionally stable; it can stably compute time steps of arbitrary duration causing nontrivial changes in margin position.
-Next, a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during the geometry-velocity solution, for instance if the velocity is held fixed during a time step as the geometry is updated (perhaps ``implicitly'' on its own).
+Next, a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during the geometry-velocity solution, or, for instance, if the velocity is held fixed during a time step as the geometry is updated (though perhaps ``implicitly'' on its own).
 Thus a scheme which computes margin advance or retreat only after the velocity field update is accepted is only semi-implicit, for example.
-Finally, a scheme is \emph{explicit} for the coupled problem if the ice geometry is held fixed during the velocity solution, and then this (accepted and now fixed) velocity solution is used to update the geometry through an explicit step for Equation \eqref{eq:masscontinuity}.
+Finally, a scheme is (\emph{fully}) \emph{explicit} for the coupled problem if the ice geometry is held fixed during the velocity solution, and then this (accepted and now fixed) velocity solution is used to update the geometry through an explicit step for Equation \eqref{eq:masscontinuity}.
 
@@
 
-Such models also have $O(m)$ velocity variables, but note that these are not state variables.
-That is, a very-viscous stress balance computes velocity as a function of the true state variables, such as ice thickness and temperature.
+Such models also have $O(m)$ velocity variables, but note that these are not state variables, essentially because the Stokes model lacks time derivatives.
+That is, a very-viscous stress balance computes velocity as a function of the true state variables, such as ice thickness and temperature or enthalpy.
 
@@
 The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.
-For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.
+For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, directly computes steady states with scaling $\beta=0.8$.
 (An earlier steady, Picard iteration implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is also robust for implicit time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic and irregular bed topographies.
@@
 The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are solved implicitly, with nonnegativity of thickness in their SIA portions enforced as part of a free-boundary solution.
-However, the time-stepping in these models is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects thickness.
+However, the time-stepping in these models is only semi-implicit because the membrane-stress-resolving portion of the velocity is held fixed as the ice thickness is advected.
 The SIA time-step solvers in these schemes have apparently only been tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$, for $U$ derived from sliding speeds.
@@
 % equation (A21) in BrinkerhoffJohnson2015
-In a different design, the semi-implicit scheme in the open-source \'Ua numerical model \citep{Gudmundsson2021manual} implicitly solves mass continuity Equation \eqref{eq:masscontinuity} using an active set method, but while holding the membrane-stress-resolved velocity fixed.
-Note that the formally third-order time-stepping in this model \citep{Gudmundsson2013}, while surely desirable for marine ice sheets, cannot achieve that accuracy when the nonnegativity-of-thickness constraint is actively determining the free boundary, e.g.~at retreating margins.
+The semi-implicit scheme in the open-source \'Ua numerical marine ice sheet model \citep{Gudmundsson2013,Gudmundsson2021manual} implicitly solves mass continuity free-boundary problem for Equation \eqref{eq:masscontinuity} using an active set method, but also while holding the (membrane-stress-resolved) velocity fixed.
 % equations (1), (26) in Gudmundson2013
@@
 
-The \cite{Bueler2016} implicit SIA solver is not enough better, however.
+The scaling of the \cite{Bueler2016} implicit SIA solver is not enough better than such an explicit scheme, however.
 Although long time steps can be taken by this implicit solver, the $\beta=0.8$ scaling gives $O(\Delta x^{-3.6})$ effort.

### 2022-10-09 paper/reviews/response.tex
ready to ship out

@@
 Below I both respond to their concerns---in block italics---and, in the important cases, quote the new or rewritten content of the paper which addresses these concerns.
+The aggregate responses to these comments have made the paper about one page longer, i.e.~when printed in the final two-column format.
+
+\section{Responses to Scientific Editor Ralf Greve}
+
+\begin{review}
+As for the first paragraph of the introduction and the comment by Reviewer 2 on L25, it may also be worth mentioning the findings by Gladstone et al.
+(2017, TC, doi: 10.5194/tc-11-319-2017).
+They showed that the resolution requirement can be alleviated substantially by choosing proper representations for basal sliding and melting that ensure continuity across the grounding line.
+\end{review}
+
+\noindent Yes, that makes sense.  This idea and citation is integrated into my response to Reviewer 2 on L25.
 
@@
 
-\noindent I stand by the assertion, but this reasonable challenge has caused me to add some words to clarify.  The sentence in question now says:
+\noindent I stand by the assertion, but this reasonable challenge has caused me to add some words to clarify.
+The sentence in question is now expanded into two sentences which explain some of the needs for resolution:
 
-\begin{quote}``In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows need yet finer resolution.''\end{quote}
+\begin{quote}``In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows and alpine topography need yet finer resolution.
+Resolving the physics of marine ice sheets also benefits from fine resolution, although careful choice of basal parameterizations will reduce resolution dependence (Gladstone and others, 2017).''\end{quote}
 
-\noindent A few points about this new sentence will, I think, be understood by any reader with minimal exposure to ice sheet modeling:
+\noindent A few points about the first of these new sentences will, I think, be understood by any reader with minimal exposure to ice sheet modeling:
 \begin{itemize}
@@
 That is, as fluids are locally infinite-dimensional, you need a few cells, more than one or two across for ice streams, to capture a ``fluid feature''.
-\item 10 km resolution is itself not enough for yet narrower features.
+\item 10 km resolution is itself not enough for yet narrower or smaller features.
 \item The added word ``it is \underline{now} generally accepted'' clearly refers to expectations in the 2020s.
-There were indeed ``important'' 20km results before now, but I would disagree with any assertion in a present-day journal submission that, for example, flux across grounding lines in Antarctica is ``validly modeled'' at 20 km resolution; that would now be a consequence of not trying very hard, or choosing a deficient model.
+There were indeed ``important'' 20km results before now, but I would disagree with any assertion in a present-day journal submission that, for example, outlet glaciers in Antarctica or Greenland are ``validly modeled'' at 20 km resolution; that would now be a consequence of not trying very hard, or choosing a deficient model.
 \end{itemize}
-With that background, I am happy if readers conclude, for example, that the author thinks that 20km or 40km Antarctic model results are no longer ``valid'' in a scientific sense.
-That is what I think, and I was under the impression that this was a generally-accepted judgement.
+With that background, I am happy if readers conclude, for example, that the author thinks that 20km or 40km ice sheet model results are no longer ``valid'' in a scientific sense.
+That is what I think, and I was under the impression that this was an acceptable judgement call in the community.
 

### 2022-10-09 paper/perfmod.tex
ready to ship out

@@
 
-\abstract{Numerical ice sheet models compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.
+\abstract{Numerical glacier and ice sheet models compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.
 At high spatial resolution, with horizontal mesh/grid resolutions of a few kilometers or smaller, these models usually require time steps shorter than climate-coupling time scales because they update ice thickness after each velocity solution.
@@
 Numerical ice sheet (glacier) models with evolving ice geometry are now routinely used for addressing scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenlandic \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.
-In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows need yet finer resolution.
-Whether using local mesh refinement \citep[for example]{Fischleretal2022,Hoffmanetal2018} or not, resolutions of two or one kilometers \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019,Clarkeetal2015} are increasingly used for science at ice sheet scale.
+In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows and alpine topography need yet finer resolution.
+Resolving the physics of marine ice sheets also benefits from fine resolution, although careful choice of basal parameterizations will reduce resolution dependence \citep{Gladstoneetal2017}.
+Thus, whether using local mesh refinement \citep[for example]{Fischleretal2022,Hoffmanetal2018} or not, resolutions of two or one kilometers \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019,Clarkeetal2015} are increasingly used for science at ice sheet scale.
 
@@
 \begin{equation}
-\frac{\partial H}{\partial t} + \Divx \left(\bu H\right) = a, \label{eq:masscontinuity}
+\frac{\partial H}{\partial t} + \Divx \left(\bu H\right) = a. \label{eq:masscontinuity}
 \end{equation}
-where $\bx=(x,y)$ denotes the horizontal coordinates.
-Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.
+(Here $\bx=(x,y)$ denotes the horizontal coordinates.)  Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.
 This is because ice flows dominantly downhill.
@@
 In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term effectively acts along the surface gradient $\gradx s$ because of the ice-domain geometry.
-One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\gradx s)$, a non-local function of geometry; a stress balance solution is required to evaluate this function.
-A numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step.
+One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\gradx s)$, a non-local function of geometry, but a stress balance solution is required to evaluate this function.
+In any case, a numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step \citep{LeVeque2007}.
 
@@
 
-However, either statement \eqref{eq:masscontinuity} or \eqref{eq:siamasscontinuity} of the mass continuity equation is fundamentally incomplete without modeling the changing (map-plane) boundary position of the glacier or ice sheet.
+However, either statement \eqref{eq:masscontinuity} or \eqref{eq:siamasscontinuity} of the mass continuity equation is fundamentally incomplete without modeling the evolving (map-plane) boundary position of the glacier or ice sheet.
 The problem of determining this updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).
@@
 
-Assuming the map-plane extent of the model domain is 2D and of width $L$, these primary parameters are related by
+Assuming the map-plane extent of the model domain is 2D\footnote{In flow-line models $\Delta x \sim L m^{-1}$, but this paper addresses 3D models with map-plane (2D) horizontal meshes.} and of width $L$, these primary parameters are related by
 \begin{equation}
@@
 \end{equation}
-Specifically, for fixed domain width $L$ there are $O(m^{1/2})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$, but this paper addresses 3D models with map-plane (2D) horizontal meshes.}
+Specifically, for fixed domain width $L$ there are $O(m^{1/2})$ mesh cells in each horizontal dimension.
 
@@
 Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.
-Note that energy balance and degree-day schemes for computing surface mass balance \citep{GreveBlatter2009} generally have much shorter time scales, but here $q$ describes the \emph{coupling} frequency on which ice geometry needs to be updated via solution of the mass continuity or surface kinematical equation.
+Note that energy balance and degree-day schemes for computing surface mass balance \citep{Hock2005} generally have much shorter time scales, but here $q$ describes the \emph{coupling} frequency on which ice geometry needs to be updated via solution of the mass continuity or surface kinematical equation.
 
@@
 Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$, but this restriction reflects the simulation purpose, not maintenance of stability.
-For an implicit scheme the desired frequency of ice-climate interaction will, however, determine the total simulation cost according to the (large) solution cost, at each time step, of solving coupled mass and momentum equations, including the surface kinematical or mass continuity equation.
+For an implicit scheme the desired frequency of ice-climate interaction will, however, determine the total simulation cost according to the (large) solution cost, at each time step, of solving a coupled mass and momentum free-boundary problem, including the surface kinematical or mass continuity equation.
 
@@
 
-When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and it depends strongly on solver design.
+When analyzing solver scaling in our simplified form, one must be aware that the constant in ``$O(m^{1+\alpha})$'' above can be very large, and it depends strongly on solver design.
 Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling, as actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.
-Many models \citep[e.g.][]{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} also show good \emph{parallel} scaling, something not addressed here.
+Indeed, many models \citep[e.g.][]{BrownSmithAhmadia2013, Fischleretal2022, IsaacStadlerGhattas2015, Lengetal2012, Tuminaroetal2016} also show good \emph{parallel} scaling, something not addressed here.
 
@@
 \end{equation}
-The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$ for Stokes models, to give a work estimate for each model year in a simulation.
+The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$ for Stokes models and $O(m)$ for SIA models, to give a work estimate for each model year in a simulation.
 The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.
@@
 For SIA models, let us assume that the flops of such coupled solutions scale as $O(m^{1+\beta})$ with $\beta \ge 0$; a large constant is assumed.
-The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.
+The only currently implemented, unconditionally-stable, fully-implicit time-step solvers use the SIA stress balance.
 For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, directly computes steady states with scaling $\beta=0.8$.
@@
 Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of membrane-stress-resolving solvers dominates performance concerns.
-In particular, an $\alpha=1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition.
+In particular, an $\alpha \approx 1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition.
 Optimistic stability yields a still-bad $O(\Delta x^{-5})$ rate.
@@
 The Table also shows why algorithmic scaling remains important in the large-domain $L\to\infty$ limit when applying Stokes or higher-order dynamics.
-The computational work of an optimal solver will be proportional to ice sheet area $L^2$.
-However, an $\alpha=1$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.
+The computational work of a nearly-optimal $\alpha \approx 0$ solver will be proportional to ice sheet area $L^2$.
+However, an $\alpha \approx 1$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha \approx 4$) power of $L$.
 
@@
 
+\subsection{Acknowledgements}  Comments by Jed Brown, editor Ralf Greve, referee Alex Jarosch, and especially an anonymous referee have helped to focus and improve this work.
+However, opinions expressed here, and any remaining errors, are entirely those of the author.
+
 %         References

### 2022-11-06 paper/perfmod.tex
in form for sending preprints to people
remove from bibliography the references I did not use; move unpublished Ua manual to text

@@
-%\documentclass[twocolumn,letterpaper]{igs}
-%\documentclass[twocolumn]{igs}  % A4 paper needs no option
-\documentclass[review,letterpaper]{igs}
+%\documentclass[twocolumn]{igs}
+\documentclass[review]{igs}
+%\documentclass[review,letterpaper]{igs}
 
@@
 \usepackage{igsnatbib,lineno}
-
-\usepackage[kw]{pseudo}
-\pseudoset{left-margin=0mm,topsep=5mm,idfont=\texttt}
 
@@
 In other words, these current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system describing the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which fix the velocity following the computation which updates the geometry \citep[for example]{Chengetal2017}.}  However, at least for simpler, textbook partial differential equation problems, the limited and conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.
-Stability conditions of explicit SIA models appeared some time ago \citep[e.g.][]{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.
+Stability conditions of explicit SIA models appeared some time ago \citep[e.g.][]{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2022}.
 
@@
 % equation (A21) in BrinkerhoffJohnson2015
-The semi-implicit scheme in the open-source \'Ua numerical marine ice sheet model \citep{Gudmundsson2013,Gudmundsson2021manual} implicitly solves mass continuity free-boundary problem for Equation \eqref{eq:masscontinuity} using an active set method, but also while holding the (membrane-stress-resolved) velocity fixed.
+The semi-implicit scheme in the open-source \'Ua numerical marine ice sheet model \citep[see also G.~H.~Gudmundsson, \emph{{\'U}a {C}ompendium}, \href{https://github.com/GHilmarG/UaSource}{\texttt{github.com/GHilmarG/UaSource}}]{Gudmundsson2013} implicitly solves mass continuity free-boundary problem for Equation \eqref{eq:masscontinuity} using an active set method, but also while holding the (membrane-stress-resolved) velocity fixed.
 % equations (1), (26) in Gudmundson2013

### 2022-11-06 paper/perfmod.tex
add doi to all articles, and spell out full journal titles

@@
 The problem of determining this updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).
-The mathematical role of this constraint has been understood for some time in the SIA model \citep{Calvoetal2002,JouvetBueler2012,SchoofHewitt2013}, but its role as the determinant of margin location is universal across fluid-layer problems with signed source terms \citep{Bueler2021conservation}.
+The mathematical role of this constraint has been understood for some time in the SIA model \citep{Calvoetal2003,JouvetBueler2012,SchoofHewitt2013}, but its role as the determinant of margin location is universal across fluid-layer problems with signed source terms \citep{Bueler2021conservation}.
 

### 2022-11-06 paper/perfmod.tex
clean .tex for submission

@@
 
-\usepackage{verbatim,xspace,amsmath,amssymb,bm,multirow,dashbox}
-\usepackage{tikz}
-\usetikzlibrary{arrows}
-\usepackage{igsnatbib,lineno}
+\usepackage{amsmath}
+\usepackage{igsnatbib}
+\usepackage{lineno}
 
@@
             colorlinks=true,
-            citecolor=purple,
+            citecolor=blue,
             linkcolor=red,
@@
 % math macros
-\newcommand\bb{\mathbf{b}}
-\newcommand\bc{\mathbf{c}}
-\newcommand\bbf{\mathbf{f}}
-\newcommand\bg{\mathbf{g}}
-\newcommand\bn{\mathbf{n}}
-\newcommand\bq{\mathbf{q}}
 \newcommand\bu{\mathbf{u}}
-\newcommand\bv{\mathbf{v}}
 \newcommand\bx{\mathbf{x}}
-\newcommand\by{\mathbf{y}}
-
-\newcommand\bA{\mathbf{A}}
-\newcommand\bF{\mathbf{F}}
-\newcommand\bH{\mathbf{H}}
-\newcommand\bN{\mathbf{N}}
-\newcommand\bQ{\mathbf{Q}}
-\newcommand\bU{\mathbf{U}}
-\newcommand\bV{\mathbf{V}}
-\newcommand\bW{\mathbf{W}}
-\newcommand\bX{\mathbf{X}}
-
-\newcommand\bzero{\bm{0}}
-
-\newcommand{\Div}{\nabla\cdot}
-\newcommand\eps{\epsilon}
-\newcommand{\grad}{\nabla}
-\newcommand{\ip}[2]{\ensuremath{\left<#1,#2\right>}}
-\newcommand\lam{\lambda}
-\newcommand\lap{\triangle}
-\newcommand\RR{\mathbb{R}}
-\newcommand{\half}{\tfrac{1}{2}}
-
 \newcommand{\Divx}{\nabla_\bx \cdot}
 \newcommand{\gradx}{\nabla_\bx}
-
-\newcommand{\rhoi}{\rho_{\text{i}}}
-\newcommand{\pp}{{\text{p}}}
-\newcommand{\qq}{{\text{q}}}
-\newcommand{\rr}{{\text{r}}}
-
-\newcommand{\mR}{R^{\bm{\oplus}}}
-\newcommand{\iR}{R^{\bullet}}
-
-\newcommand{\sold}{s_{\text{o}}}
-\newcommand{\told}{t_{\text{o}}}
-
-\newcommand{\byold}{\by_{\text{o}}}
-
-\newcommand{\onecol}[1]{\includegraphics[width=86mm]{#1}}
-\newcommand{\onecolless}[1]{\includegraphics[width=80mm]{#1}}
-\newcommand{\onecolthin}[1]{\includegraphics[width=75mm]{#1}}
-
-%\newcommand{\twocol}[1]{\includegraphics[width=178mm]{#1}}
-\newcommand{\twocol}[1]{\includegraphics[width=165mm]{#1}}
-\newcommand{\twocolless}[1]{\includegraphics[width=140mm]{#1}}
-
-\newcommand{\ds}{\displaystyle}
+\newcommand{\nn}{\text{n}}
 
@@
 E-mail: \emph{\texttt{elbueler\@@alaska.edu}}}
-
-%\keywords{glacier flow,glacier modelling,ice-sheet modelling}
-% suggested reviewers:  Dan Goldberg, William Lipscomb, Toby Isaac, Josefin Ahlkrona, Alex Jarosch, Doug Brinkerhoff
 
@@
 
-\newcommand{\nn}{\text{n}}
 A diffusive description of the mass continuity equation is also valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
@@
 %         References
-\bibliography{perfmod}
-\bibliographystyle{igs}
+%\bibliography{perfmod}
+%\bibliographystyle{igs}
+
+\begin{thebibliography}{34}
+\providecommand{\natexlab}[1]{#1}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi: #1}\else
+  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+\bibitem[\protect\citename{Aschwanden and others, }2019]{Aschwandenetal2019}
+Aschwanden A, Fahnestock MA, Truffer M, Brinkerhoff DJ, Hock R, Khroulev C,
+  Mottram R and Khan SA (2019) Contribution of the {Greenland Ice Sheet} to sea
+  level over the next millennium. \emph{Science Advances}, \textbf{5}(6) (\doi
+  {10.1126/sciadv.aav9396})
+
+\bibitem[\protect\citename{Briggs and others, }2000]{Briggsetal2000}
+Briggs W, Henson VE and McCormick S (2000) \emph{A {M}ultigrid {T}utorial}.
+  SIAM Press, Philadelphia, 2nd edition
+
+\bibitem[\protect\citename{Brinkerhoff and Johnson,
+  }2015]{BrinkerhoffJohnson2015}
+Brinkerhoff DJ and Johnson JV (2015) Dynamics of thermally induced ice streams
+  simulated with a higher-order flow model. \emph{Journal of Geophysical
+  Research: Earth Surface}, \textbf{120}(9), 1743--1770 (\doi
+  {10.1002/2015JF003499})
+
+\bibitem[\protect\citename{Brown and others, }2013]{BrownSmithAhmadia2013}
+Brown J, Smith B and Ahmadia A (2013) Achieving textbook multigrid efficiency
+  for hydrostatic ice sheet flow. \emph{SIAM Journal on Scientific Computing},
+  \textbf{35}(2), 359--375 (\doi {10.1137/110834512})
+
+\bibitem[\protect\citename{Bueler, }2016]{Bueler2016}
+Bueler E (2016) Stable finite volume element schemes for the shallow ice
+  approximation. \emph{Journal of Glaciology}, \textbf{62}(232), 230--242 (\doi
+  {10.1017/jog.2015.3})
+
+\bibitem[\protect\citename{Bueler, }2021{\natexlab{a}}]{Bueler2021conservation}
+Bueler E (2021{\natexlab{a}}) Conservation laws for free-boundary fluid layers.
+  \emph{SIAM Journal on Applied Mathematics}, \textbf{81}(5), 2007--2032 (\doi
+  {10.1137/20M135217X})
+
+\bibitem[\protect\citename{Bueler, }2021{\natexlab{b}}]{Bueler2021}
+Bueler E (2021{\natexlab{b}}) \emph{{PETSc} for {P}artial {D}ifferential
+  {E}quations: {N}umerical {S}olutions in {C} and {P}ython}. SIAM Press,
+  Philadelphia
+
+\bibitem[\protect\citename{Bueler and others, }2005]{Bueleretal2005}
+Bueler E, Lingle CS, Kallen-Brown JA, Covey DN and Bowman LN (2005) Exact
+  solutions and verification of numerical models for isothermal ice sheets.
+  \emph{Journal of Glaciology}, \textbf{51}(173), 291--306 (\doi
+  {10.3189/172756505781829449})
+
+\bibitem[\protect\citename{Calvo and others, }2003]{Calvoetal2003}
+Calvo N and others (2003) On a doubly nonlinear parabolic obstacle problem
+  modelling ice sheet dynamics. \emph{SIAM Journal on Applied Mathematics},
+  \textbf{63}(2), 683--707 (\doi {10.1137/S0036139901385345})
+
+\bibitem[\protect\citename{Cheng and others, }2017]{Chengetal2017}
+Cheng G, L{\"o}tstedt P and von Sydow L (2017) Accurate and stable time
+  stepping in ice sheet modeling. \emph{Journal of Computational Physics},
+  \textbf{329}, 29--47 (\doi {10.1016/j.jcp.2016.10.060})
+
+\bibitem[\protect\citename{Clarke and others, }2015]{Clarkeetal2015}
+Clarke GKC, Jarosch AH, Anslow FS, Radić V and Menounos B (2015) Projected
+  deglaciation of western {C}anada in the twenty-first century. \emph{Nature
+  Geoscience}, \textbf{8}, 372--377 (\doi {10.1038/ngeo2407})
+
+\bibitem[\protect\citename{Fischler and others, }2022]{Fischleretal2022}
+Fischler Y, R\"uckamp M, Bischof C, Aizinger V, Morlighem M and Humbert A
+  (2022) A scalability study of the {Ice-sheet and Sea-level System Model}
+  ({ISSM}, version 4.18). \emph{Geoscientific Model Development},
+  \textbf{15}(9), 3753--3771 (\doi {10.5194/gmd-15-3753-2022})
+
+\bibitem[\protect\citename{Gladstone and others, }2017]{Gladstoneetal2017}
+Gladstone RM, Warner RC, Galton-Fenzi BK, Gagliardini O, Zwinger T and Greve R
+  (2017) Marine ice sheet model performance depends on basal sliding physics
+  and sub-shelf melting. \emph{The Cryosphere}, \textbf{11}(1), 319--329 (\doi
+  {10.5194/tc-11-319-2017})
+
+\bibitem[\protect\citename{Goelzer and others, }2020]{Goelzeretal2020}
+Goelzer H, Nowicki S and others (2020) The future sea-level contribution of the
+  {G}reenland ice sheet: a multi-model ensemble study of {ISMIP6}. \emph{The
+  Cryosphere}, \textbf{14}(9), 3071--3096 (\doi {10.5194/tc-14-3071-2020})
+
+\bibitem[\protect\citename{Greve and Blatter, }2009]{GreveBlatter2009}
+Greve R and Blatter H (2009) \emph{Dynamics of {I}ce {S}heets and {G}laciers}.
+  Advances in Geophysical and Environmental Mechanics and Mathematics,
+  Springer, Berlin
+
+\bibitem[\protect\citename{Gudmundsson, }2013]{Gudmundsson2013}
+Gudmundsson GH (2013) Ice-shelf buttressing and the stability of marine ice
+  sheets. \emph{The Cryosphere}, \textbf{7}(2), 647--655 (\doi
+  {10.5194/tc-7-647-2013})
+
+\bibitem[\protect\citename{Hindmarsh and Payne, }1996]{HindmarshPayne1996}
+Hindmarsh RCA and Payne AJ (1996) Time-step limits for stable solutions of the
+  ice-sheet equation. \emph{Annals of Glaciology}, \textbf{23}, 74--85 (\doi
+  {10.3189/S0260305500013288})
+
+\bibitem[\protect\citename{Hock, }2005]{Hock2005}
+Hock R (2005) Glacier melt: a review of processes and their modelling.
+  \emph{Progress in Physical Geography}, \textbf{29}(3), 362--391 (\doi
+  {10.1191/0309133305pp453ra})
+
+\bibitem[\protect\citename{Hoffman and others, }2018]{Hoffmanetal2018}
+Hoffman MJ, Perego M, Price SF, Lipscomb WH, Zhang T, Jacobsen D, Tezaur I,
+  Salinger AG, Tuminaro R and Bertagna L (2018) {MPAS-Albany Land Ice (MALI)}:
+  a variable-resolution ice sheet model for {E}arth system modeling using
+  {V}oronoi grids. \emph{Geoscientific Model Development}, \textbf{11}(9),
+  3747--3780 (\doi {10.5194/gmd-11-3747-2018})
+
+\bibitem[\protect\citename{Isaac and others, }2015]{IsaacStadlerGhattas2015}
+Isaac T, Stadler G and Ghattas O (2015) Solution of nonlinear {S}tokes
+  equations discretized by high-order finite elements on nonconforming and
+  anisotropic meshes, with application to ice sheet dynamics. \emph{SIAM
+  Journal on Scientific Computing}, \textbf{37}(6), B804--B833 (\doi
+  {10.1137/140974407})
+
+\bibitem[\protect\citename{Jouvet and Bueler, }2012]{JouvetBueler2012}
+Jouvet G and Bueler E (2012) Steady, shallow ice sheets as obstacle problems:
+  well-posedness and finite element approximation. \emph{SIAM Journal on
+  Applied Mathematics}, \textbf{72}(4), 1292--1314 (\doi {10.1137/110856654})
+
+\bibitem[\protect\citename{Jouvet and Gr{\"a}ser, }2013]{JouvetGraeser2013}
+Jouvet G and Gr{\"a}ser C (2013) An adaptive {N}ewton multigrid method for a
+  model of marine ice sheets. \emph{Journal of Computational Physics},
+  \textbf{252}, 419--437 (\doi {10.1016/j.jcp.2013.06.032})
+
+\bibitem[\protect\citename{Leng and others, }2012]{Lengetal2012}
+Leng W, Ju L, Gunzburger M, Price S and Ringler T (2012) A parallel high-order
+  accurate finite element nonlinear {S}tokes ice sheet model and benchmark
+  experiments. \emph{Journal of Geophysical Research: Earth Surface},
+  \textbf{117}(F1) (\doi {10.1029/2011JF001962})
+
+\bibitem[\protect\citename{LeVeque, }2007]{LeVeque2007}
+LeVeque RJ (2007) \emph{{Finite Difference Methods for Ordinary and Partial
+  Differential Equations: Steady-State and Time-Dependent Problems}}. SIAM
+  Press, Philadelphia
+
+\bibitem[\protect\citename{L{\"o}fgren and others,
+  }2022]{LofgrenAhlkronaHelanow2022}
+L{\"o}fgren A, Ahlkrona J and Helanow C (2022) Increasing stable time-step
+  sizes of the free-surface problem arising in ice-sheet simulations.
+  \emph{Journal of Computational Physics: X}, \textbf{16}(100114) (\doi
+  {10.1016/j.jcpx.2022.100114})
+
+\bibitem[\protect\citename{Robinson and others, }2022]{Robinsonetal2022}
+Robinson A, Goldberg D and Lipscomb WH (2022) A comparison of the stability and
+  performance of depth-integrated ice-dynamics solvers. \emph{The Cryosphere},
+  \textbf{16}(2), 689--709 (\doi {10.5194/tc-16-689-2022})
+
+\bibitem[\protect\citename{Schoof and Hewitt, }2013]{SchoofHewitt2013}
+Schoof C and Hewitt IJ (2013) Ice-sheet dynamics. \emph{Annual Review of Fluid
+  Mechanics}, \textbf{45}, 217--239 (\doi {10.1002/jgrf.20146})
+
+\bibitem[\protect\citename{Seguinot and Delaney, }2021]{SeguinotDelaney2021}
+Seguinot J and Delaney I (2021) Last-glacial-cycle glacier erosion potential in
+  the {A}lps. \emph{Earth Surface Dynamics}, \textbf{9}(4), 923--935 (\doi
+  {10.5194/esurf-9-923-2021})
+
+\bibitem[\protect\citename{Seroussi and others, }2020]{Seroussietal2020}
+Seroussi H, Nowicki S and others (2020) {ISMIP6 Antarctica}: a multi-model
+  ensemble of the {A}ntarctic ice sheet evolution over the 21st century.
+  \emph{The Cryosphere}, \textbf{14}(9), 3033--3070 (\doi
+  {10.5194/tc-14-3033-2020})
+
+\bibitem[\protect\citename{Trottenberg and others, }2001]{Trottenbergetal2001}
+Trottenberg U, Oosterlee CW and Schuller A (2001) \emph{Multigrid}. Elsevier,
+  Oxford
+
+\bibitem[\protect\citename{Tuminaro and others, }2016]{Tuminaroetal2016}
+Tuminaro R, Perego M, Tezaur I, Salinger A and Price S (2016) A matrix
+  dependent/algebraic multigrid approach for extruded meshes with applications
+  to ice sheet modeling. \emph{SIAM Journal on Scientific Computing},
+  \textbf{38}(5), C504--C532 (\doi {10.1137/15M1040839})
+
+\bibitem[\protect\citename{Weber and others, }2021]{Weberetal2021}
+Weber ME, Golledge NR, Fogwill CJ, Turney CSM and Thomas ZA (2021)
+  Decadal-scale onset and termination of {A}ntarctic ice-mass loss during the
+  last deglaciation. \emph{Nature Communications}, \textbf{12}(6683) (\doi
+  {10.1038/s41467-021-27053-6})
+
+\bibitem[\protect\citename{Winkelmann and others, }2011]{Winkelmannetal2011}
+Winkelmann R, Martin MA, Haseloff M, Albrecht T, Bueler E, Khroulev C and
+  Levermann A (2011) {The Potsdam Parallel Ice Sheet Model (PISM-PIK)--Part 1:
+  Model description}. \emph{The Cryosphere}, \textbf{5}(3), 715--726 (\doi
+  {10.5194/tc-5-715-2011})
+
+\bibitem[\protect\citename{Wirbel and Jarosch, }2020]{WirbelJarosch2020}
+Wirbel A and Jarosch AH (2020) Inequality-constrained free-surface evolution in
+  a full {S}tokes ice flow model (\textit{evolve\_glacier v1.1}).
+  \emph{Geoscientific Model Development}, \textbf{13}(12), 6425--6445 (\doi
+  {10.5194/gmd-13-6425-2020})
+
+\end{thebibliography}
 

### 2022-12-12 talk/slides.tex
slides stub

\documentclass[svgnames,
               hyperref={colorlinks,citecolor=DeepPink4,linkcolor=FireBrick,urlcolor=Maroon},
               usepdftitle=false]  % see \hypersetup{} below
               {beamer}

\mode<presentation>{
  \usetheme{Madrid}
  \usecolortheme{seagull}
  \setbeamercovered{transparent}
  \setbeamerfont{frametitle}{size=\large}
}

\setbeamercolor*{block title}{bg=red!10}
\setbeamercolor*{block body}{bg=red!5}

%\usepackage[svgnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
    pdftitle = {Online optimization},
    pdfauthor = {Ed Bueler},
    pdfsubject = {},
    pdfkeywords = {}
}

\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage[T1]{fontenc}
% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.

\usepackage{empheq,bm}
\usepackage{xspace}
\usepackage{fancyvrb}

\usepackage{tikz}
\usetikzlibrary{shapes,arrows.meta,decorations.markings,decorations.pathreplacing,fadings,positioning}

\usepackage[kw]{pseudo}
\pseudoset{left-margin=15mm,topsep=5mm,idfont=\texttt,st-left=,st-right=}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:
%\beamerdefaultoverlayspecification{<+->}

\newcommand{\ba}{\mathbf{a}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bc}{\mathbf{c}}
\newcommand{\bbf}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bw}{\mathbf{w}}

\newcommand{\bF}{\mathbf{F}}
\newcommand{\bG}{\mathbf{G}}
\newcommand{\bQ}{\mathbf{Q}}

\newcommand{\grad}{\nabla}
\newcommand{\Div}{\nabla\cdot}

\newcommand{\argmin}{\operatorname{argmin}}

\newcommand{\CC}{\mathbb{C}}
\newcommand{\EE}{\mathbb{E}}
\newcommand{\RR}{\mathbb{R}}

\newcommand{\ddt}[1]{\ensuremath{\frac{\partial #1}{\partial t}}}
\newcommand{\ddx}[1]{\ensuremath{\frac{\partial #1}{\partial x}}}
\newcommand{\Matlab}{\textsc{Matlab}\xspace}
\newcommand{\Octave}{\textsc{Octave}\xspace}
\newcommand{\eps}{\epsilon}

\newcommand{\ip}[2]{\left<#1,#2\right>}

\newcommand{\xiphalf}{{x_{i+\frac{1}{2}}}}
\newcommand{\ximhalf}{{x_{i-\frac{1}{2}}}}
\newcommand{\Fiphalf}{{F_{i+\frac{1}{2}}}}
\newcommand{\Fimhalf}{{F_{i-\frac{1}{2}}}}
\newcommand{\Fiphalfn}{{F^n_{i+\frac{1}{2}}}}
\newcommand{\Fimhalfn}{{F^n_{i-\frac{1}{2}}}}

\newcommand{\trefcolumn}[1]{\begin{bmatrix} \phantom{x} \\ #1 \\ \phantom{x} \end{bmatrix}}
\newcommand{\trefmatrixtwo}[2]{\left[\begin{array}{c|c|c} & & \\ #1 & \dots & #2 \\ & & \end{array}\right]}
\newcommand{\trefmatrixthree}[3]{\left[\begin{array}{c|c|c|c} & & & \\ #1 & #2 & \dots & #3 \\ & & & \end{array}\right]}
\newcommand{\trefmatrixgroups}[4]{\left[\begin{array}{c|c|c|c|c|c} & & & & & \\ #1 & \dots & #2 & #3 & \dots & #4 \\ & & & & & \end{array}\right]}

\newcommand{\blocktwo}[4]{\left[\begin{array}{c|c} #1 & #2 \\ \hline #3 & #4 \end{array}\right]}

\newcommand{\bqed}{{\color{blue}\qed}}
\newcommand{\ds}{\displaystyle}

\newcommand\mynum[1]{{\renewcommand{\insertenumlabel}{#1}%
      \usebeamertemplate{enumerate item} \,}}

\title[Making ice sheet models scale properly]{Making ice sheet models \\ scale properly}

%\subtitle{\emph{x}}

\author{Ed Bueler}

\institute[UAF]{University of Alaska Fairbanks}

\date[]{February 2023}

%\titlegraphic{\begin{picture}(0,0)
%    \put(0,180){\makebox(0,0)[rt]{\includegraphics[width=4cm]{figs/software.png}}}
%  \end{picture}
%}

%% to start section counter at 0 see
%% https://tex.stackexchange.com/questions/170222/change-the-numbering-in-beamers-table-of-content

\begin{document}
\beamertemplatenavigationsymbolsempty

\begin{frame}
  \maketitle
\end{frame}

\begin{frame}{Outline}
  \tableofcontents[hideallsubsections]
\end{frame}

\section{a section}

\begin{frame}{\alert{summary}}

\begin{itemize}
\item x
\item y
\end{itemize}
\end{frame}

\begin{frame}{references}

\begin{itemize}
\footnotesize
\item E.~Hazan, A.~Agarwal, \& S.~Kale (2007).  \href{https://link.springer.com/content/pdf/10.1007/s10994-007-5016-8.pdf}{\emph{Logarithmic regret algorithms for online convex optimization.}} Machine Learning, 69(2), 169-192
    \begin{itemize}
    \scriptsize
    \item[$-$] $O(\log j)$ regret bounds for positive definite Hessians, and for Newton algorithms
    \end{itemize}
\item D.~P.~Kingma \& J.~Ba (2014). \href{https://arxiv.org/abs/1412.6980}{\emph{Adam: A method for stochastic optimization}}, preprint arXiv:1412.6980.
    \begin{itemize}
    \scriptsize
    \item[$-$] claims $O(\sqrt{j})$ regret bound
    \end{itemize}
\item M.~Zinkevich (2003). \href{https://www.aaai.org/Papers/ICML/2003/ICML03-120.pdf}{\emph{Online convex programming and generalized infinitesimal gradient ascent}}, Proceedings of the 20th International Conference on Machine Learning, 928-936
    \begin{itemize}
    \scriptsize
    \item[$-$] introduced regret
    \item[$-$] $O(\sqrt{j})$ regret bound of OGD
    \end{itemize}
\end{itemize}
\end{frame}

\end{document}

### 2022-12-12 talk/slides.tex
start thinking about content

@@
 
-\section{a section}
+\section{introduction}
+
+\begin{frame}{what is an ice sheet model?}
+
+\begin{itemize}
+\item x
+\item y
+\end{itemize}
+\end{frame}
 
@@
 \item y
+\end{itemize}
+\end{frame}
+
+\section{a simplified performance analysis}
+
+\begin{frame}{\alert{summary}}
+
+\begin{itemize}
+\item x
+\item y
+\end{itemize}
+\end{frame}
+
+\newcommand{\oo}[1]{\displaystyle O\left(#1\right)}
+
+\begin{frame}{complexity table}
+
+%\setlength{\tabcolsep}{5pt}
+%\renewcommand{\arraystretch}{1.5}
+
+\begin{tabular}{llll}
+\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{[pessimistic stability]} \\ \hline
+explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
+explicit & Stokes & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
+implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$ \\
+implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}$
+\end{tabular}
+
+\begin{itemize}
+\item Asymptotic estimates of algorithmic scaling, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$
 \end{itemize}

### 2023-01-01 talk/slides.tex
start

@@
 \hypersetup{
-    pdftitle = {Online optimization},
+    pdftitle = {Making ice sheet models scale properly},
     pdfauthor = {Ed Bueler},
@@
 
-\section{introduction}
+\section{what is an ice sheet model}
+
+\begin{frame}{what is an ice sheet?}
+
+\begin{itemize}
+\item before numerical models \dots \emph{what} are we modeling?
+\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width ratio
+\item<5> it is \emph{not} sea ice
+\end{itemize}
+
+\bigskip
+\begin{minipage}[t][40cm][t]{\textwidth}
+\begin{center}
+\includegraphics<1>[height=0.64\textheight]{images/ant-pittard2021.png}
+\only<1>{\par {\scriptsize Antarctic ice sheet (Pittard et al 2021)}}
+\only<2>{\vspace{15mm}}
+\includegraphics<2>[width=\textwidth]{images/ant-schoofhewitt2013.png}
+\only<2>{\vspace{10mm}}
+\only<2>{\par {\scriptsize note vertical exageration and rough bed (Schoof \& Hewitt 2013)}}
+\includegraphics<3>[height=0.64\textheight]{images/alps-seguinot2018.png}
+\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum (Seguinot et al 2018)}}
+\includegraphics<4>[height=0.64\textheight]{images/british-clark2022.png}
+\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum (Clark et al 2022)}}
+\includegraphics<5>[height=0.70\textheight]{images/not-sea-ice.png}
+\end{center}
+\end{minipage}
+\end{frame}
 

### 2023-01-02 talk/slides.tex
progress

@@
 
+\titlegraphic{\hfill \includegraphics[width=0.2\textwidth]{images/uafbw.png}}
+
 %% to start section counter at 0 see
@@
 
-\section{what is an ice sheet model}
+\section{what is an ice sheet model?}
 
@@
 
+\begin{columns}
+\begin{column}{0.63\textwidth}
 \begin{itemize}
-\item x
-\item y
+\item ice is modeled as a non-Newtonian viscous fluid at large scale
+    \begin{itemize}
+    \item[$\circ$] more on that later \dots
     \end{itemize}
+\item ice sheets evolve their geometry and velocity \emph{in contact with climate}
+    \begin{itemize}
+    \item[$\circ$] snowfall
+    \item[$\circ$] surface melt
+    \item[$\circ$] subglacial melt
+    \item[$\circ$] melt and calving of parts which float in ocean (\emph{ice shelves})
+    \end{itemize}
+\end{itemize}
+\end{column}
+\begin{column}{0.4\textwidth}
+\includegraphics[width=\textwidth]{images/coverfig.png}
+\end{column}
+\end{columns}
 \end{frame}
 
-\begin{frame}{\alert{summary}}
+\begin{frame}{what is an ice sheet model?}
+
+\begin{columns}
+\begin{column}{0.63\textwidth}
+\begin{itemize}
+\item for purposes of clarity and analysis, let us simplify
+\item inputs into an ice sheet model:
+\item outputs from an ice sheet model:
+\end{itemize}
+\end{column}
+\begin{column}{0.4\textwidth}
+\includegraphics[width=\textwidth]{images/coverfig.png}
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}{z}
 
@@
 \section{a simplified performance analysis}
+
+\section{more on numerical stability}
+
+\section{}
 

### 2023-01-02 talk/slides.tex
progress on slides

--- 2023-01-01 talk/slides.tex
@@
 
-\newcommand{\ba}{\mathbf{a}}
+\newcommand{\eps}{\epsilon}
+\newcommand{\RR}{\mathbb{R}}
+
+\newcommand{\grad}{\nabla}
+\newcommand{\Div}{\nabla\cdot}
+\newcommand{\trace}{\operatorname{tr}}
+
+\newcommand{\hbn}{\hat{\mathbf{n}}}
+
 \newcommand{\bb}{\mathbf{b}}
-\newcommand{\bc}{\mathbf{c}}
+\newcommand{\be}{\mathbf{e}}
 \newcommand{\bbf}{\mathbf{f}}
@@
 \newcommand{\bn}{\mathbf{n}}
-\newcommand{\bq}{\mathbf{q}}
 \newcommand{\br}{\mathbf{r}}
+\newcommand{\bu}{\mathbf{u}}
+\newcommand{\bv}{\mathbf{v}}
+\newcommand{\bw}{\mathbf{w}}
 \newcommand{\bx}{\mathbf{x}}
-\newcommand{\by}{\mathbf{y}}
-\newcommand{\bv}{\mathbf{v}}
-\newcommand{\bu}{\mathbf{u}}
-\newcommand{\bw}{\mathbf{w}}
 
 \newcommand{\bF}{\mathbf{F}}
-\newcommand{\bG}{\mathbf{G}}
-\newcommand{\bQ}{\mathbf{Q}}
+\newcommand{\bV}{\mathbf{V}}
+\newcommand{\bX}{\mathbf{X}}
 
-\newcommand{\grad}{\nabla}
-\newcommand{\Div}{\nabla\cdot}
+\newcommand{\bxi}{\bm{\xi}}
 
-\newcommand{\argmin}{\operatorname{argmin}}
+\newcommand{\bzero}{\bm{0}}
 
-\newcommand{\CC}{\mathbb{C}}
-\newcommand{\EE}{\mathbb{E}}
-\newcommand{\RR}{\mathbb{R}}
+\newcommand{\rhoi}{\rho_{\text{i}}}
 
-\newcommand{\ddt}[1]{\ensuremath{\frac{\partial #1}{\partial t}}}
-\newcommand{\ddx}[1]{\ensuremath{\frac{\partial #1}{\partial x}}}
-\newcommand{\Matlab}{\textsc{Matlab}\xspace}
-\newcommand{\Octave}{\textsc{Octave}\xspace}
-\newcommand{\eps}{\epsilon}
+\newcommand{\ip}[2]{\left(#1,#2\right)}
 
-\newcommand{\ip}[2]{\left<#1,#2\right>}
+\newcommand{\mR}{R^{\bm{\oplus}}}
+\newcommand{\iR}{R^{\bullet}}
 
-\newcommand{\xiphalf}{{x_{i+\frac{1}{2}}}}
-\newcommand{\ximhalf}{{x_{i-\frac{1}{2}}}}
-\newcommand{\Fiphalf}{{F_{i+\frac{1}{2}}}}
-\newcommand{\Fimhalf}{{F_{i-\frac{1}{2}}}}
-\newcommand{\Fiphalfn}{{F^n_{i+\frac{1}{2}}}}
-\newcommand{\Fimhalfn}{{F^n_{i-\frac{1}{2}}}}
+\newcommand{\nn}{{\text{n}}}
+\newcommand{\pp}{{\text{p}}}
+\newcommand{\qq}{{\text{q}}}
+\newcommand{\rr}{{\text{r}}}
 
-\newcommand{\trefcolumn}[1]{\begin{bmatrix} \phantom{x} \\ #1 \\ \phantom{x} \end{bmatrix}}
-\newcommand{\trefmatrixtwo}[2]{\left[\begin{array}{c|c|c} & & \\ #1 & \dots & #2 \\ & & \end{array}\right]}
-\newcommand{\trefmatrixthree}[3]{\left[\begin{array}{c|c|c|c} & & & \\ #1 & #2 & \dots & #3 \\ & & & \end{array}\right]}
-\newcommand{\trefmatrixgroups}[4]{\left[\begin{array}{c|c|c|c|c|c} & & & & & \\ #1 & \dots & #2 & #3 & \dots & #4 \\ & & & & & \end{array}\right]}
-
-\newcommand{\blocktwo}[4]{\left[\begin{array}{c|c} #1 & #2 \\ \hline #3 & #4 \end{array}\right]}
-
-\newcommand{\bqed}{{\color{blue}\qed}}
-\newcommand{\ds}{\displaystyle}
-
-\newcommand\mynum[1]{{\renewcommand{\insertenumlabel}{#1}%
-      \usebeamertemplate{enumerate item} \,}}
+\newcommand{\bus}{\bu|_s}
+\newcommand{\oo}[1]{\displaystyle O\left(#1\right)}
+\newcommand{\sold}{s_{\text{o}}}
 
@@
 
+%\begin{frame}
+%  \maketitle
+%\end{frame}
+
+{
+  \usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/gray-british-clark2022.png}}
   \begin{frame}
-  \maketitle
+    \titlepage
   \end{frame}
+}
 
@@
 \begin{itemize}
-\item before numerical models \dots \emph{what} are we modeling?
+\item before numerical modeling \dots \emph{what} are we modeling?
 \item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width ratio
-\item<5> it is \emph{not} sea ice
 \end{itemize}
@@
 \bigskip
-\begin{minipage}[t][40cm][t]{\textwidth}
+\begin{minipage}[t][60cm][t]{\textwidth}
 \begin{center}
-\includegraphics<1>[height=0.64\textheight]{images/ant-pittard2021.png}
-\only<1>{\par {\scriptsize Antarctic ice sheet (Pittard et al 2021)}}
-\only<2>{\vspace{15mm}}
+\includegraphics<1>[height=0.69\textheight]{images/ant-pittard2021.png}
+\only<1>{\par {\scriptsize Antarctic ice sheet}} % (Pittard et al 2021)
+\only<2>{\vspace{17mm}}
 \includegraphics<2>[width=\textwidth]{images/ant-schoofhewitt2013.png}
-\only<2>{\vspace{10mm}}
-\only<2>{\par {\scriptsize note vertical exageration and rough bed (Schoof \& Hewitt 2013)}}
-\includegraphics<3>[height=0.64\textheight]{images/alps-seguinot2018.png}
+\only<2>{\vspace{13mm}}
+\only<2>{\par {\scriptsize note vertical exaggeration and rough bed (Schoof \& Hewitt 2013)}}
+\includegraphics<3>[height=0.69\textheight]{images/alps-seguinot2018.png}
 \only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum (Seguinot et al 2018)}}
-\includegraphics<4>[height=0.64\textheight]{images/british-clark2022.png}
+\includegraphics<4>[height=0.69\textheight]{images/british-clark2022.png}
 \only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum (Clark et al 2022)}}
-\includegraphics<5>[height=0.70\textheight]{images/not-sea-ice.png}
+\includegraphics<5>[height=0.69\textheight]{images/not-sea-ice.png}
+\only<5>{\par {\scriptsize but an ice sheet is \emph{not} sea ice!}}
 \end{center}
@@
 
+\begin{frame}{basic facts}
+
+\begin{columns}
+\begin{column}{0.6\textwidth}
+\begin{itemize}
+\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
+    \begin{itemize}
+    \item[$\circ$] more on that soon!
+    \end{itemize}
+\item ice sheet geometry and velocity evolve \emph{in contact with the climate}
+    \begin{itemize}
+    \item[$\circ$] snowfall
+    \item[$\circ$] surface melt
+    \item[$\circ$] subglacial melt
+    \item[$\circ$] melt and calving when floating in the ocean (\emph{ice shelves})
+    \end{itemize}
+\end{itemize}
+\end{column}
+\begin{column}{0.42\textwidth}
+\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}
+\end{column}
+\end{columns}
+\end{frame}
+
 \begin{frame}{what is an ice sheet model?}
 
+\begin{columns}
+\begin{column}{0.6\textwidth}
 \begin{itemize}
-\item x
-\item y
+\item for simplicity/clarity of the upcoming performance analysis, I will be ignoring quite a bit of detail!
+\item \alert{ignoring}:
+    \begin{itemize}
+    \item[$\circ$] floating ice
+    \item[$\circ$] subglacial hydrology
+    \item[$\circ$] ice temperature
+    \item[$\circ$] fracture processes (calving, crevasses)
+    \end{itemize}
+\item now, a simple definition of \emph{ice sheet model}
+\end{itemize}
+\end{column}
+\begin{column}{0.42\textwidth}
+\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}{what is an ice sheet model?}
+
+\begin{columns}
+\begin{column}{0.6\textwidth}
+\begin{itemize}
+\begin{definition}
+an \alert{ice sheet model} is a map
+\end{definition} 
+\item two inputs:
+    \begin{itemize}
+    \item[$\circ$] \emph{surface mass balance}
+$$a(t,x) = \begin{pmatrix}
+\text{balance of solid precipitation} \\
+\text{\, vs melt \& runoff at surfaces}
+\end{pmatrix}$$
+
+\vspace{-3mm}
+        \begin{itemize}
+        \item[{\scriptsize $\bullet$}] units of mass flux:\, $\text{kg}\, \text{m}^{-2} \text{s}^{-1}$
+        \end{itemize}
+
+    \item[$\circ$] \emph{bed elevation} $b(x)$ [m]
+    \end{itemize}
+\item one output:
+    \begin{itemize}
+    \item[$\circ$] \emph{upper surface elevation} $s(t,x)$ [m]
+    \end{itemize}
+\end{itemize}
+\end{column}
+\begin{column}{0.42\textwidth}
+\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}
+
+\vspace{2mm}
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}{conservation}
+
+\begin{itemize}
+\item but this is a mathematics talk in fact
+\item FIXME conservation laws; conserve momentum standard; conserve mass means both conserve mass in interior (incompressible) and at surface (SKE)
 \end{itemize}
@@
 
-\begin{frame}{\alert{summary}}
+\begin{frame}{strong form of ice sheet model}
 
 \begin{itemize}
-\item x
-\item y
+\item FIXME RETURN TIME DERIV
+\item nonlinear complementarity problem coupled to Stokes:
+\begin{align*}
+s - b &\ge 0 && \text{on $\Omega$} \\
+s - \Delta t\, \bu|_s \cdot \bn_s - (\sold + \Delta t\, a )&\ge 0 && \text{''} \\
+(s - b) (s - \Delta t\, \bu|_s \cdot \bn_s - (\sold + \Delta t\, a)) &= 0 && \text{''} \\
+- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \\
+\nabla \cdot \bu &= 0 && \text{''} \\
+\bu &= \bzero && \text{on $\Gamma_0$} \\
+\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$}
+\end{align*}
+
+    \begin{itemize}
+    \item with regularized Glen-law effective viscosity ($\text{p}=\frac{1}{\nn}+1$):
+      $$\nu_\eps(D\bu) = \frac{\Gamma}{2} \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
+    \end{itemize}
 \end{itemize}
@@
 
+\section{time-stepping and stability}
+
 \section{a simplified performance analysis}
-
-\begin{frame}{\alert{summary}}
-
-\begin{itemize}
-\item x
-\item y
-\end{itemize}
-\end{frame}
-
-\newcommand{\oo}[1]{\displaystyle O\left(#1\right)}
 
@@
 
+\section{Stokes solver scaling}
+
+\section{scalable implicit steps?}
+
+\begin{frame}{\alert{summary}}
+
+\begin{itemize}
+\item x
+\item y
+\end{itemize}
+\end{frame}
+
 \begin{frame}{references}
@@
 \footnotesize
-\item E.~Hazan, A.~Agarwal, \& S.~Kale (2007).
-\href{https://link.springer.com/content/pdf/10.1007/s10994-007-5016-8.pdf}{\emph{Logarithmic regret algorithms for online convex optimization.}} Machine Learning, 69(2), 169-192
-    \begin{itemize}
-    \scriptsize
-    \item[$-$] $O(\log j)$ regret bounds for positive definite Hessians, and for Newton algorithms
-    \end{itemize}
-\item D.~P.~Kingma \& J.~Ba (2014). \href{https://arxiv.org/abs/1412.6980}{\emph{Adam: A method for stochastic optimization}}, preprint arXiv:1412.6980.
-    \begin{itemize}
-    \scriptsize
-    \item[$-$] claims $O(\sqrt{j})$ regret bound
-    \end{itemize}
-\item M.~Zinkevich (2003).
-\href{https://www.aaai.org/Papers/ICML/2003/ICML03-120.pdf}{\emph{Online convex programming and generalized infinitesimal gradient ascent}}, Proceedings of the 20th International Conference on Machine Learning, 928-936
-    \begin{itemize}
-    \scriptsize
-    \item[$-$] introduced regret
-    \item[$-$] $O(\sqrt{j})$ regret bound of OGD
-    \end{itemize}
+\item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
+\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
+\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
+\item x
 \end{itemize}

### 2023-01-03 talk/slides.tex
progress slides

@@
 
+\newcommand{\btau}{\bm{\tau}}
 \newcommand{\bxi}{\bm{\xi}}
@@
 
-\titlegraphic{\hfill \includegraphics[width=0.2\textwidth]{images/uafbw.png}}
+\titlegraphic{\hfill \includegraphics[width=0.15\textwidth]{images/uafbw.png}}
 
@@
 \begin{column}{0.42\textwidth}
-\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}
+\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
 \end{column}
@@
 \begin{itemize}
-\item for simplicity/clarity of the upcoming performance analysis, I will be ignoring quite a bit of detail!
+\item for simplicity/clarity of the upcoming performance analysis, I will be ignoring quite a bit of physics!
 \item \alert{ignoring}:
@@
     \end{itemize}
-\item now, a simple definition of \emph{ice sheet model}
+\item now, a simple definition of ``ice sheet model'' \dots
 \end{itemize}
@@
 \begin{definition}
-an \alert{ice sheet model} is a map
+an \alert{ice sheet model} is a \underline{map} which simulates an ice sheet in a climate
 \end{definition} 
@@
 
-    \item[$\circ$] \emph{bed elevation} $b(x)$ [m]
+    \item[$\circ$] \emph{bed elevation} $b(x)$
     \end{itemize}
-\item one output:
+\item two outputs:
     \begin{itemize}
-    \item[$\circ$] \emph{upper surface elevation} $s(t,x)$ [m]
+    \item[$\circ$] \emph{upper surface elevation} $s(t,x)$
+    \item[$\circ$] \emph{ice velocity} $\bu(t,x)$
     \end{itemize}
+\item {\small $\begin{pmatrix} \text{climate} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{velocity} \end{pmatrix}$}
 \end{itemize}
@@
 \begin{column}{0.42\textwidth}
-\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}
+\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}
 
@@
 \begin{itemize}
-\item but this is a mathematics talk in fact
+\item as this is a mathematics talk, more or less, some notation \dots
 \item FIXME conservation laws; conserve momentum standard; conserve mass means both conserve mass in interior (incompressible) and at surface (SKE)
@@
 \begin{itemize}
-\item FIXME RETURN TIME DERIV
 \item nonlinear complementarity problem coupled to Stokes:
 \begin{align*}
-s - b &\ge 0 && \text{on $\Omega$} \\
-s - \Delta t\, \bu|_s \cdot \bn_s - (\sold + \Delta t\, a )&\ge 0 && \text{''} \\
-(s - b) (s - \Delta t\, \bu|_s \cdot \bn_s - (\sold + \Delta t\, a)) &= 0 && \text{''} \\
-- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \\
+s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
+\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
+(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
+- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda(t) \subset \RR^3$} \\
 \nabla \cdot \bu &= 0 && \text{''} \\
-\bu &= \bzero && \text{on $\Gamma_0$} \\
-\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$}
+\bu\cdot \bn = 0 \quad \& \quad \btau_b - \bbf(\bu) &= \bzero && \text{on $\Gamma_b(t) \subset \partial \Lambda(t)$} \\
+\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t) \subset \partial \Lambda(t)$}
 \end{align*}
@@
     \begin{itemize}
-    \item with regularized Glen-law effective viscosity ($\text{p}=\frac{1}{\nn}+1$):
-      $$\nu_\eps(D\bu) = \frac{\Gamma}{2} \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
+    \item viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
     \end{itemize}
@@
 
-\section{scalable implicit steps?}
+\section{scalable implicit steps by multigrid?}
 
@@
 \begin{itemize}
-\item x
-\item y
+\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
+   \begin{itemize}
+   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
+   \end{itemize}
+\item current-technology ice sheet models mostly use \alert{explicit time stepping}, \alert{non-optimal stress-balance solvers}, and shallow assumptions
+\item \alert{coming soon} from current research:
+   \begin{itemize}
+   \item[$\circ$] scalable Stokes solvers (Isaac et al.~2015)
+   \item[$\circ$] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
+   \item[$\circ$] ML emulators (Jouvet et al.~2021)
+   \end{itemize}
+\item scalable solvers for the implicit-step, coupled problem for ice sheet geometry requires advances in \alert{multilevel solutions of non-local variational inequalities}
 \end{itemize}
@@
 \item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
-\item x
+\item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015).
+\emph{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics}, SIAM J.~Sci.~Comput.~37 (6), B804--B833, \href{https://doi.org/10.1137/140974407}{10.1137/140974407}
+\item G.~Jouvet, G.~Cordonnier, B.~Kim, M.~L\"uthi, A.~Vieli, A.~Aschwanden (2021).
+\emph{Deep learning speeds up ice flow modelling by several orders of magnitude}, J.~Glaciol.~\href{https://doi.org/10.1017/jog.2021.120}{10.1017/jog.2021.120}
+\item A.~L{\"o}fgren, J.
+Ahlkrona, \& C.
+Helanow (2022).
+\emph{Increasing stable time-step sizes of the free-surface problem arising in ice-sheet simulations}, J.~Comput.~Phys.~X 16, \href{https://doi.org/10.1016/j.jcpx.2022.100114}{10.1016/j.jcpx.2022.100114}
 \end{itemize}

### 2023-01-05 talk/slides.tex
progress finally

@@
 
-\usepackage{empheq,bm}
-\usepackage{xspace}
-\usepackage{fancyvrb}
+\usepackage{empheq,bm,xspace,fancyvrb,soul}
 
@@
     \item[$\circ$] fracture processes (calving, crevasses)
+    \item[$\circ$] solid earth deformation
     \end{itemize}
@@
     \item[$\circ$] \emph{surface mass balance}
-$$a(t,x) = \begin{pmatrix}
-\text{balance of solid precipitation} \\
-\text{\, vs melt \& runoff at surfaces}
+$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
+\text{balance of precipitation vs} \\
+\text{melt \& runoff from surfaces}
 \end{pmatrix}$$
@@
 
-    \item[$\circ$] \emph{bed elevation} $b(x)$
+    \item[$\circ$] \emph{bed elevation} $b(x,y)$
     \end{itemize}
@@
     \begin{itemize}
-    \item[$\circ$] \emph{upper surface elevation} $s(t,x)$
-    \item[$\circ$] \emph{ice velocity} $\bu(t,x)$
+    \item[$\circ$] \emph{upper surface elevation} $s(t,x,y)$
+    \item[$\circ$] \emph{ice velocity} $\bu(t,x,y,z)$
     \end{itemize}
@@
 
+\begin{frame}{notation}
+
+\begin{itemize}
+%\item but this is a mathematics talk \dots notation! equations!
+\item the data $a(t,x,y)$, $b(x,y)$ is defined on fixed time and space domains:
+	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
+
+\begin{center}
+\includegraphics<1>[width=0.55\textheight]{images/domain-data.png}
+\includegraphics<2>[width=0.6\textheight]{images/domain-velocity.png}
+\end{center}
+
+\medskip
+\item<2> the solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
+    \begin{itemize}
+    \item[$\circ$] $s=b$ where there is no ice
+    \end{itemize}
+
+\item<2> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \Omega$ on which the velocity $\bu(t,x,y,z)$ is defined:
+    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$
+
+\vspace{-2mm}
+    \begin{itemize}
+    \item[$\circ$] $\Gamma_s(t),\Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
+    \end{itemize}
+\end{itemize}
+\end{frame}
+
 \begin{frame}{conservation}
@@
 \begin{itemize}
-\item as this is a mathematics talk, more or less, some notation \dots
-\item FIXME conservation laws; conserve momentum standard; conserve mass means both conserve mass in interior (incompressible) and at surface (SKE)
+\item ice sheet evolution is a \alert{free-boundary conservation} problem
+    \begin{itemize}
+    \item[$\circ$] mass conservation
+    \item[$\circ$] momentum conservation
+    \item[$\circ$] energy conservation (\emph{ignored here})
+    \end{itemize}
+\item key idea: conservation of mass means both conserve mass in interior (incompressible) and at surface (SKE)
 \end{itemize}

### 2023-01-06 talk/slides.tex
many edits

@@
 \item before numerical modeling \dots \emph{what} are we modeling?
-\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width ratio
+\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width
 \end{itemize}
@@
 \includegraphics<5>[height=0.69\textheight]{images/not-sea-ice.png}
-\only<5>{\par {\scriptsize but an ice sheet is \emph{not} sea ice!}}
+\only<5>{\par {\scriptsize \emph{not} sea ice!}}
 \end{center}
@@
 
-\begin{frame}{basic facts}
+\begin{frame}{basic facts about glaciers}
 
@@
 \begin{itemize}
-\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
-    \begin{itemize}
-    \item[$\circ$] more on that soon!
-    \end{itemize}
-\item ice sheet geometry and velocity evolve \emph{in contact with the climate}
+\item the geometry and velocity evolve \emph{in contact with its climate},
     \begin{itemize}
@@
     \item[$\circ$] subglacial melt
-    \item[$\circ$] melt and calving when floating in the ocean (\emph{ice shelves})
+    \item[$\circ$] melt when floating (\emph{ice shelf})
+    \item[$\circ$] calving into the ocean
+    \end{itemize}
+\item and the \emph{topography} on which it sits
+\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
+    \begin{itemize}
+    \item[$\circ$] more on that soon
     \end{itemize}
@@
 
-\begin{frame}{what is an ice sheet model?}
+\begin{frame}{simplifications}
 
@@
 \begin{itemize}
-\item for simplicity/clarity of the upcoming performance analysis, I will be ignoring quite a bit of physics!
+\item for simplicity/clarity of the upcoming model performance analysis, I will ignore quite a bit of physics
 \item \alert{ignoring}:
@@
     \end{itemize}
-\item now, a simple definition of ``ice sheet model'' \dots
+
+\medskip
+\item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})} for a model which does \emph{not} make these simplifications}
+\item<2> {\footnotesize next, a simple definition of ``ice sheet model''}
 \end{itemize}
@@
     \end{itemize}
-\item {\small $\begin{pmatrix} \text{climate} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{velocity} \end{pmatrix}$}
+
+\bigskip
+\item<2-> {\small map: \quad $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$}
 \end{itemize}
@@
 
-\begin{frame}{notation}
+\begin{frame}{notation for the simplified model}
 
 \begin{itemize}
-%\item but this is a mathematics talk \dots notation! equations!
-\item the data $a(t,x,y)$, $b(x,y)$ is defined on fixed time and space domains:
+\item the \alert{data} $a(t,x,y)$, $b(x,y)$ \alert{is defined on a fixed domain}:
 	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
@@
 \vspace{-2mm}
-    \begin{itemize}
-    \item[$\circ$] $\Gamma_s(t),\Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
-    \end{itemize}
 \end{itemize}
@@
 
-\begin{frame}{conservation}
+\begin{frame}{mass and momentum conservation}
 
 \begin{itemize}
-\item ice sheet evolution is a \alert{free-boundary conservation} problem
+\item ice sheet evolution should conserve physical quantities:
     \begin{itemize}
-    \item[$\circ$] mass conservation
-    \item[$\circ$] momentum conservation
-    \item[$\circ$] energy conservation (\emph{ignored here})
+    \item[$\circ$] mass
+    \item[$\circ$] momentum
+    \item[$\circ$] \st{energy} \hfill $\leftarrow$ \emph{ignored in this talk}
     \end{itemize}
-\item key idea: conservation of mass means both conserve mass in interior (incompressible) and at surface (SKE)
+\item a \emph{key idea} is that conservation of mass is important both
+    \begin{itemize}
+    \item[$\circ$] \emph{in} the icy domain $\Lambda(t) \subset \RR^3$:
+\begin{align*}
+\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
+    \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
+\text{\alert{surface kinematic equation}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
+\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
+\end{align*}
+    \end{itemize}
+\item where $\Gamma_s(t)$ and $\Gamma_b(t)$, subsets of $\partial \Lambda(t)$, denote the surface and base of the ice
 \end{itemize}
@@
 
-\begin{frame}{strong form of ice sheet model}
+\begin{frame}{free boundary problem}
 
 \begin{itemize}
-\item nonlinear complementarity problem coupled to Stokes:
+\item but ice sheet evolution is a \alert{free-boundary} problem for conserved quantities in a (thin) fluid layer
+\item specifically, the surface kinematic equation
+  $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
+applies \emph{only} on the ice upper surface $\Gamma_s(t)$
+\item in the remainder of the fixed map-plane domain $\Omega\subset \RR^2$, glacier \alert{complementarity} holds:
+  $$s=b \quad \text{and} \quad a \le 0$$
+
+\bigskip
+\item<2> {\footnotesize for more on this free-boundary perspective, see Jouvet \& Bueler (2012) and Bueler (2021)}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{simplified ice sheet model (strong form)}
+
+\begin{itemize}
+\item nonlinear complementarity problem \onslide<2->{coupled to Stokes}:
 \begin{align*}
@@
 (s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
-- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda(t) \subset \RR^3$} \\
+\onslide<2->{- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
 \nabla \cdot \bu &= 0 && \text{''} \\
-\bu\cdot \bn = 0 \quad \& \quad \btau_b - \bbf(\bu) &= \bzero && \text{on $\Gamma_b(t) \subset \partial \Lambda(t)$} \\
-\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t) \subset \partial \Lambda(t)$}
+\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t) \subset \partial \Lambda(t)$} \\
+\bu|_b \cdot \bn_b &= 0 && \text{''} \\
+\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t) \subset \partial \Lambda(t)$}}
 \end{align*}
@@
     \begin{itemize}
-    \item viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
+    \item $\bu|_s=0$ where no ice
+    \item<2-> viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
     \end{itemize}
@@
 \section{time-stepping and stability}
+
+\begin{frame}{ice thickness and ``mass continuity equation'' view}
+
+FIXME
+\end{frame}
 
@@
 
-\section{Stokes solver scaling}
+\section{on Stokes solver scaling}
 
@@
    \end{itemize}
-\item current-technology ice sheet models mostly use \alert{explicit time stepping}, \alert{non-optimal stress-balance solvers}, and shallow assumptions
-\item \alert{coming soon} from current research:
+\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
+\item<3-> \alert{coming soon} from current research:
    \begin{itemize}
@@
    \end{itemize}
-\item scalable solvers for the implicit-step, coupled problem for ice sheet geometry requires advances in \alert{multilevel solutions of non-local variational inequalities}
+\item<4-> scalable solvers for the implicit-step, coupled problem for ice sheet geometry requires advances in \alert{multilevel solutions of non-local variational inequalities}
 \end{itemize}
@@
 
+{\scriptsize
 \begin{itemize}
-\footnotesize
 \item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
@@
 \emph{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics}, SIAM J.~Sci.~Comput.~37 (6), B804--B833, \href{https://doi.org/10.1137/140974407}{10.1137/140974407}
+\item G.~Jouvet \& E.
+Bueler (2012).
+\emph{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation}, SIAM J.~Appl.~Math.
+72 (4), 1292--1314, \href{https://doi.org/10.1137/110856654}{10.1137/110856654}
 \item G.~Jouvet, G.~Cordonnier, B.~Kim, M.~L\"uthi, A.~Vieli, A.~Aschwanden (2021).
@@
 \end{itemize}
+}
 \end{frame}

### 2023-01-06 talk/slides.tex
a bit more

@@
 \newcommand{\bF}{\mathbf{F}}
+\newcommand{\bU}{\mathbf{U}}
 \newcommand{\bV}{\mathbf{V}}
@@
 \bigskip
-\item<2> {\footnotesize for more on this free-boundary perspective, see Jouvet \& Bueler (2012) and Bueler (2021)}
+\item<2> {\footnotesize for more on this free-boundary perspective see Bueler (2021)}
 \end{itemize}
@@
 
+\begin{itemize}
+\item a helpful viewpoint on this complicated system is the equation
+
+FIXME
+
+where $\bU$ is vertically-averaged horizontal velocity within icy domain $\Lambda(t)$
+\item \emph{question:} is it really an advection?
+
+FIXME
+
+\item[] \emph{answer:} no, because ice flows (mostly) downhill
+
+\item alternative equation, the \emph{shallow diffusion} model of ice sheets:
+
+FIXME
+\bigskip
+\item<2> {\footnotesize for more on the shallow diffusion perspective see Jouvet \& Bueler (2012)}
+\end{itemize}
+\end{frame}
+
+\section{on Stokes solver scaling}
+
+\begin{frame}{regarding PDE solver complexity}
+
 FIXME
@@
 
-\begin{frame}{complexity table}
-
-%\setlength{\tabcolsep}{5pt}
-%\renewcommand{\arraystretch}{1.5}
+\begin{frame}{complexity of ice sheet models}
 
 \begin{tabular}{llll}
-\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{[pessimistic stability]} \\ \hline
-explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
-explicit & Stokes & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
+\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
+\onslide<1->{
+\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
+}
+\onslide<1->{
+explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
+}
+\onslide<2->{
+explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
+}
+\onslide<3->{
 implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$ \\
+}
+\onslide<3->{
 implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}$
+}
 \end{tabular}
 
-\begin{itemize}
-\item Asymptotic estimates of algorithmic scaling, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$
-\end{itemize}
+FIXME asymptotic estimates of algorithmic scaling, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$
 \end{frame}
 
-\section{on Stokes solver scaling}
+\section{scalable implicit steps by multigrid?}
 
-\section{scalable implicit steps by multigrid?}
+\begin{frame}{obstacle problem solver scaling}
+
+FIXME
+\end{frame}
 

### 2023-01-06 talk/slides.tex
progress

@@
 \item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})} for a model which does \emph{not} make these simplifications}
-\item<2> {\footnotesize next, a simple definition of ``ice sheet model''}
+\item<2> {\footnotesize next, a basic definition of ``ice sheet model''}
 \end{itemize}
@@
 
-\begin{frame}{notation for the simplified model}
+\begin{frame}{notation for the basic model}
 
@@
 \end{itemize}
+\end{frame}
+
+\begin{frame}{shallow ice sheet models}
+
+\begin{itemize}
+\item the model on the last slide is \emph{not yet} in common use for ice sheet simulations, because for now it is expensive to compute
+    \begin{itemize}
+    \item goal: make it fast enough to use!
+    \end{itemize}
+\item current-technology models use various shallow simplifications
+\item simplest is a lubrication approximation, the \alert{shallow ice approximation}, in which $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator  
+\end{itemize}
+\end{frame}
+
+\begin{frame}{shallow ice approximation: details}
+
+{\small
+\begin{align*}
+s - b &\ge 0 \\
+\frac{\partial s}{\partial t} + \Phi(s) - a &\ge 0 \\
+(s - b) \left(\frac{\partial s}{\partial t} + \Phi(s) - a\right) &= 0
+\end{align*}
+where
+\begin{align*}
+\Phi(s) &= - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} \\
+        &\qquad\qquad - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)
+\end{align*}
+here $\nabla$ only acts in horizontal variables, and constants $\pp = \nn+1$ and $\gamma > 0$ are related to ice deformation
+
+\begin{itemize}
+\item $\Phi$ is a \alert{doubly-nonlinear differential operator}, simultaneously of porous medium and $\pp$-Laplacian type
+\end{itemize}
+}
 \end{frame}

### 2023-01-06 talk/slides.tex
more progress I hope

--- 2023-01-06 talk/slides.tex
@@
 
-\begin{frame}{simplified ice sheet model (strong form)}
+\begin{frame}{basic ice sheet model (strong form)}
 
@@
 
-\section{time-stepping and stability}
-
-\begin{frame}{ice thickness and ``mass continuity equation'' view}
+\begin{frame}{the mass-continuity equation view}
 
 \begin{itemize}
-\item a helpful viewpoint on this complicated system is the equation
-
-FIXME
-
-where $\bU$ is vertically-averaged horizontal velocity within icy domain $\Lambda(t)$
+\item a thickness-transport form exposes the basic ice sheet model
+\item define:
+\begin{align*}
+H(t,x,y) &=s-b && \text{ice thickness} \\
+\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
+\end{align*}
+\item the \alert{mass continuity equation} follows from the surface kinematical equation and incompressibility:
+   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$
 \item \emph{question:} is it really an advection?
-
-FIXME
-
-\item[] \emph{answer:} no, because ice flows (mostly) downhill
-
-\item alternative equation, the \emph{shallow diffusion} model of ice sheets:
-
-FIXME
-\bigskip
+\item[] \emph{answer:} not really, because ice flows (mostly) downhill
+\item an alternative form is the \emph{shallow diffusion} equation for ice sheets:
+   $$\frac{\partial H}{\partial t} = \nabla \cdot \left(D \grad s\right) + a$$
 \item<2> {\footnotesize for more on the shallow diffusion perspective see Jouvet \& Bueler (2012)}
@@
 \end{frame}
+
+\begin{frame}{shallow ice sheet models}
+
+\begin{itemize}
+\item the model on the last slide is \emph{not yet} in common use for ice sheet simulations
+    \begin{itemize}
+    \item[$\circ$] for now, it is too expensive at scale
+    \item[$\circ$] one goal: \alert{make it fast enough to use!}
+    \end{itemize}
+\item<2-> current-technology models use various shallow simplifications
+\item<2-> simplest is a lubrication approximation, the \alert{shallow ice approximation}, in which $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator  
+\end{itemize}
+\end{frame}
+
+\begin{frame}{shallow ice approximation: details}
+
+{\small
+\begin{align*}
+s - b &\ge 0 \\
+\frac{\partial s}{\partial t} + \Phi(s) - a &\ge 0 \\
+(s - b) \left(\frac{\partial s}{\partial t} + \Phi(s) - a\right) &= 0
+\end{align*}
+where
+\begin{align*}
+\Phi(s) &= - \bu|_s \cdot \bn_s \\
+        &= - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)
+\end{align*}
+
+\begin{itemize}
+\item $\nabla$ only acts in horizontal variables
+\item constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation
+\item $\Phi$ is a \alert{doubly-nonlinear differential operator}, simultaneously of porous medium and $\pp$-Laplacian type
+\end{itemize}
+}
+\end{frame}
+
+\section{time-stepping and stability}
 

### 2023-01-07 talk/slides.tex
progress I hope

@@
 \newcommand{\bF}{\mathbf{F}}
+\newcommand{\bQ}{\mathbf{Q}}
 \newcommand{\bU}{\mathbf{U}}
@@
 \begin{itemize}
-\item the geometry and velocity evolve \emph{in contact with its climate},
+\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
+    \begin{itemize}
+    \item[$\circ$] more on that soon
+    \end{itemize}
+\item glacier geometry and velocity evolve \emph{in contact with its climate}:
     \begin{itemize}
@@
 \item and the \emph{topography} on which it sits
-\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
-    \begin{itemize}
-    \item[$\circ$] more on that soon
-    \end{itemize}
 \end{itemize}
@@
 \medskip
-\item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})} for a model which does \emph{not} make these simplifications}
+\item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
 \item<2> {\footnotesize next, a basic definition of ``ice sheet model''}
@@
 $$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
-\text{balance of precipitation vs} \\
-\text{melt \& runoff from surfaces}
+\text{precipitation minus} \\
+\text{melt \& runoff}
 \end{pmatrix}$$
@@
 
-\bigskip
+\smallskip
 \item<2-> {\small map: \quad $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$}
@@
 \begin{itemize}
-\item nonlinear complementarity problem \onslide<2->{coupled to Stokes}:
+\item nonlinear complementarity problem (NCP) \onslide<2->{coupled to Stokes}:
 \begin{align*}
@@
 \begin{itemize}
-\item a thickness-transport form exposes the basic ice sheet model
+\item a thickness-transport form exposes evolution type
 \item define:
@@
 \end{align*}
+
+    \begin{itemize}
+    \item[$\circ$] note: $s$ and $H$ are equivalent variables for modeling ice geometry
+    \end{itemize}
 \item the \alert{mass continuity equation} follows from the surface kinematical equation and incompressibility:
    $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$
-\item \emph{question:} is it really an advection?
-\item[] \emph{answer:} not really, because ice flows (mostly) downhill
-\item an alternative form is the \emph{shallow diffusion} equation for ice sheets:
-   $$\frac{\partial H}{\partial t} = \nabla \cdot \left(D \grad s\right) + a$$
-\item<2> {\footnotesize for more on the shallow diffusion perspective see Jouvet \& Bueler (2012)}
+
+\smallskip
+\item<2-> \emph{question:} is it really an advection?
+\item<2->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
+  $$\bU \sim - \grad s \sim - \grad H$$
+\end{itemize}
+\end{frame}
+
+\begin{frame}{mass continuity equation: advection or diffusion?}
+
+\begin{align*}
+\text{advective form:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
+\text{diffusion form:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
+\end{align*}
+\begin{itemize}
+\item both forms are highly-nonlinear: \, $\bU=\bU(H,\grad s), \, D = D(H,\grad s)$
 \end{itemize}
@@
 \begin{itemize}
-\item the model on the last slide is \emph{not yet} in common use for ice sheet simulations
+\item the ``basic model'', i.e.~NCP coupled to Stokes, is \emph{not yet} in common use for ice sheet simulations
     \begin{itemize}
     \item[$\circ$] for now, it is too expensive at scale
-    \item[$\circ$] one goal: \alert{make it fast enough to use!}
+    \item[$\circ$] goal: \alert{make it fast enough to use!}
     \end{itemize}
+
+\smallskip
 \item<2-> current-technology models use various shallow simplifications
 \item<2-> simplest is a lubrication approximation, the \alert{shallow ice approximation}, in which $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator
+
+\smallskip
+\item<3-> the diffusion form of mass continuity ($\frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) = a$) is only literally valid in the lubrication approximation, but \emph{the fact that ice flows downhill has time-stepping stability consequences}
 \end{itemize}
@@
 
-\begin{frame}{shallow ice approximation: details}
+\begin{frame}{shallow ice approximation NCP: details}
 
@@
 \section{time-stepping and stability}
+
+\begin{frame}{time-stepping the mass continuity equation}
+
+{\footnotesize
+\begin{align*}
+\text{advective form:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
+\text{diffusion form:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
+\end{align*}
+}
+
+\begin{itemize}
+\item recall some traditional numerical analysis \dots
+\end{itemize}
+\begin{minipage}[t][60cm][t]{\textwidth}
+\begin{itemize}
+\only<1>{
+\item we expect \alert{explicit} time stepping for advections
+\item for example, forward Euler in 1D gives
+	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{\bQ_{j+1/2}^\ell - \bQ_{j-1/2}^\ell}{\Delta x} = a_j^\ell$$
+
+    \begin{itemize}
+    \item[$\circ$] where $\bQ = \bU H$
+    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
+    \end{itemize}
+}
+\only<2>{
+\item we expect \alert{implicit} time stepping for diffusions
+\item for example, backward Euler in 1D gives
+	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (H_{j+1}^{\ell+1} + H_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (H_{j}^{\ell+1} + H_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$
+
+    \begin{itemize}
+    \item[$\circ$] or use: Crank-Nicolson, BDF, \dots
+    \end{itemize}
+}
+\end{itemize}
+\end{minipage}
+\end{frame}
+
+\begin{frame}{time-stepping in ice sheet models}
+
+\begin{itemize}
+\item current-technology models mostly ``believe'' that the problem is an advection and use explicit time stepping
+\item FIXME but this has performance consequences
+\item FIXME the problem is a DAE!
+\item FIXME implicit alternative requires solving step equations: simultaneous NCP for SKE and Stokes
+\end{itemize}
+\end{frame}
 

### 2023-01-21 talk/slides.tex
better slides

@@
     \end{itemize}
-\item glacier geometry and velocity evolve \emph{in contact with its climate}:
+\item glaciers lie on \emph{topography}, and sometimes they float  (\emph{ice shelf})
+\item glacier geometry and velocity evolve \emph{in contact with climate}:
     \begin{itemize}
@@
     \item[$\circ$] subglacial melt
-    \item[$\circ$] melt when floating (\emph{ice shelf})
+    \item[$\circ$] melt when floating
     \item[$\circ$] calving into the ocean
     \end{itemize}
-\item and the \emph{topography} on which it sits
 \end{itemize}
@@
 \item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
-\item<2> {\footnotesize next, a basic definition of ``ice sheet model''}
 \end{itemize}
@@
 \end{definition} 
-\item two inputs:
+\item at least two inputs:
     \begin{itemize}
@@
     \end{itemize}
-\item two outputs:
+\item at least two outputs:
     \begin{itemize}
@@
 
-\begin{frame}{notation for the basic model}
+\begin{frame}{basic ice sheet model: notation}
 
+\begin{columns}
+\begin{column}{0.6\textwidth}
 \begin{itemize}
-\item the \alert{data} $a(t,x,y)$, $b(x,y)$ \alert{is defined on a fixed domain}:
+\item the \alert{data} $a(t,x,y)$, $b(x,y)$  is defined on a fixed domain:
 	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
-
-\begin{center}
-\includegraphics<1>[width=0.55\textheight]{images/domain-data.png}
-\includegraphics<2>[width=0.6\textheight]{images/domain-velocity.png}
-\end{center}
-
-\medskip
-\item<2> the solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
+\item<2> the solution \alert{surface elevation} $s(t,x,y)$ is defined on $[0,T]\times \Omega$
     \begin{itemize}
@@
 
-\item<2> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \Omega$ on which the velocity $\bu(t,x,y,z)$ is defined:
+\medskip
+\item<2> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$ on which the solution \alert{velocity} $\bu(t,x,y,z)$ is defined:
     $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$
@@
 \end{itemize}
+\end{column}
+\begin{column}{0.4\textwidth}
+\includegraphics<1>[width=\textwidth]{images/domain-data.png}
+
+\includegraphics<2>[width=\textwidth]{images/domain-velocity.png}
+\end{column}
+\end{columns}
 \end{frame}
@@
     \item[$\circ$] momentum
-    \item[$\circ$] \st{energy} \hfill $\leftarrow$ \emph{ignored in this talk}
+    \item[$\circ$] \st{energy} \hfill $\leftarrow$ \emph{ignored for simplicity in this talk}
     \end{itemize}
-\item a \emph{key idea} is that conservation of mass is important both
+\item a key idea is that conservation of mass is important both
     \begin{itemize}
@@
 \end{align*}
+
+    \item[$\circ$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
     \end{itemize}
-\item where $\Gamma_s(t)$ and $\Gamma_b(t)$, subsets of $\partial \Lambda(t)$, denote the surface and base of the ice
+
 \end{itemize}
@@
 \begin{itemize}
-\item but ice sheet evolution is a \alert{free-boundary} problem for conserved quantities in a (thin) fluid layer
+\item ice sheet evolution is a \alert{free-boundary} problem for conserved mass and momentum quantities
 \item specifically, the surface kinematic equation
@@
 applies \emph{only} on the ice upper surface $\Gamma_s(t)$
-\item in the remainder of the fixed map-plane domain $\Omega\subset \RR^2$, glacier \alert{complementarity} holds:
+\item in the remainder of the (fixed, map-plane) domain $\Omega\subset \RR^2$, \alert{complementarity} holds:
   $$s=b \quad \text{and} \quad a \le 0$$
@@
 \bigskip
-\item<2> {\footnotesize for more on this free-boundary perspective see Bueler (2021)}
+\item<2> {\footnotesize for more on this perspective see Bueler (2021)}
 \end{itemize}
@@
 
-\begin{frame}{basic ice sheet model (strong form)}
+\begin{frame}{basic ice sheet model: strong form}
 
@@
 
+\begin{frame}{basic ice sheet model: its a DAE system anyway}
+
+\begin{itemize}
+\item for clarity and simplicity, declutter by forgetting complementarity and all the other boundary conditions, too \dots
+\item get surface kinematic equation coupled to Stokes:
+\begin{align*}
+\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 && \text{on } \Omega \subset \RR^2 \\
+- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
+\nabla \cdot \bu &= 0 && \text{''}
+\end{align*}
+\item observe that $s(t,x,y)$ and $\bu(t,x,y,z)$ both evolve in time
+\item \emph{however}, only the first of the 5 equations has a time derivative
+\item the problem is a \alert{differential algebraic equation} system, e.g.
+\begin{align*}
+\dot x &= f(x,y) \\
+     0 &= g(x,y)
+\end{align*}
+
+    \begin{itemize}
+    \item[$\circ$] \dots actually in infinite dimensions (\emph{PDAE}?), and subject to complementarity (\emph{free boundary problem for PDAE})
+    \end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{important context (current research)}
+
+\begin{itemize}
+\item to the best of my knowledge, \emph{no} current research groups are studying the well-posedness or regularity of this ``NCP coupled to Stokes'' glacier evolution model
+    \begin{itemize}
+    \item[$\circ$] however, I think most researchers would agree this is the intended model!
+    \item[$\circ$] well-posedness of shallow approximations has been considered\footnote{e.g.~Jouvet \& Bueler (2012)}
+    \end{itemize}
+
+\medskip
+\item<2> numerical modelers tend think of the Stokes problem separately from surface evolution (\emph{time-split} or \emph{explicit time-stepping}),
+\item<2> and glacier evolution is addressed with minimal awareness of complementarity
+\end{itemize}
+\end{frame}
+
 \begin{frame}{the mass-continuity equation view}
@@
 \begin{itemize}
-\item a thickness-transport form exposes evolution type
+\item thickness-transport form helps expose evolution character
 \item define:
@@
     \end{itemize}
-\item the \alert{mass continuity equation} follows from the surface kinematical equation and incompressibility:
+\item<2-> the \alert{mass continuity equation} follows from the surface kinematical equation and incompressibility:
    $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$
@@
 \smallskip
-\item<2-> \emph{question:} is it really an advection?
-\item<2->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
+\item<3-> \emph{question:} is it really an advection?
+\item<3->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
   $$\bU \sim - \grad s \sim - \grad H$$
@@
 \item both forms are highly-nonlinear: \, $\bU=\bU(H,\grad s), \, D = D(H,\grad s)$
+\item the diffusion form is only \emph{literally} valid in the lubrication approximation (more below)
+\item the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences} regardless
 \end{itemize}
@@
 
-\begin{frame}{shallow ice sheet models}
+\begin{frame}{numerical shallow ice sheet models}
 
 \begin{itemize}
-\item the ``basic model'', i.e.~NCP coupled to Stokes, is \emph{not yet} in common use for ice sheet simulations
+\item the ``basic ice sheet model'' (NCP coupled to Stokes) is \emph{not yet} in common use for ice sheet simulations
     \begin{itemize}
-    \item[$\circ$] for now, it is too expensive at scale
+    \item[$\circ$] usual justification for falling back to shallow approximations: ``Stokes is too expensive at scale''
     \item[$\circ$] goal: \alert{make it fast enough to use!}
@@
 \item<2-> current-technology models use various shallow simplifications
-\item<2-> simplest is a lubrication approximation, the \alert{shallow ice approximation}, in which $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator
+    \begin{itemize}
+    \item<2->[$\circ$] usual justification for falling back to shallow approximations: ``Stokes is too expensive at scale''
+    \end{itemize}
+\item<2-> the simplest shallow approximation is a lubrication approximation, the \alert{shallow ice approximation} (SIA)
+\item<2-> in the SIA, $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator
 
 \smallskip
-\item<3-> the diffusion form of mass continuity ($\frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) = a$) is only literally valid in the lubrication approximation, but \emph{the fact that ice flows downhill has time-stepping stability consequences}
 \end{itemize}
@@
 \end{align*}
-where
-\begin{align*}
-\Phi(s) &= - \bu|_s \cdot \bn_s \\
-        &= - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)
-\end{align*}
+\begin{equation*}
+\Phi(s) = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)
+\end{equation*}
+}
 
 \begin{itemize}
-\item $\nabla$ only acts in horizontal variables
 \item constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation
-\item $\Phi$ is a \alert{doubly-nonlinear differential operator}, simultaneously of porous medium and $\pp$-Laplacian type
+
+\medskip
+\item<2> thus: $\Phi(s) = - \bu|_s \cdot \bn_s$ resolves to a \alert{doubly-nonlinear} differential operator, of porous medium and $\pp$-Laplacian type simultaneously
+\item<2> Jouvet \& Bueler (2012): existence for this problem
 \end{itemize}
+\end{frame}
+
+\AtBeginSection[]
+{
+  \begin{frame}{Outline}
+    \tableofcontents[currentsection]
+  \end{frame}
 }
-\end{frame}
 
@@
 \begin{align*}
-\text{advective form:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
-\text{diffusion form:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
+\text{advective ice sheet model schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
+\text{diffusion ice sheet model schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
 \end{align*}
@@
 
+\bigskip
 \begin{itemize}
-\item recall some traditional numerical analysis \dots
+\item recall: traditional numerical analysis applied to these equations
 \end{itemize}
+
 \begin{minipage}[t][60cm][t]{\textwidth}
 \begin{itemize}
-\only<1>{
+\only<1>{\item[]}
+\only<2>{
 \item we expect \alert{explicit} time stepping for advections
-\item for example, forward Euler in 1D gives
-	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{\bQ_{j+1/2}^\ell - \bQ_{j-1/2}^\ell}{\Delta x} = a_j^\ell$$
+\item for example, forward Euler:
+	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{(\bU H)_{j+1/2}^\ell - (\bU H)_{j-1/2}^\ell}{\Delta x} = a_j^\ell$$
 
     \begin{itemize}
-    \item[$\circ$] where $\bQ = \bU H$
     \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
+    \item[$\circ$] maximum time step from CFL using $\max |\bU|$
     \end{itemize}
 }
-\only<2>{
+\only<3>{
 \item we expect \alert{implicit} time stepping for diffusions
-\item for example, backward Euler in 1D gives
+\item for example, backward Euler:
 	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (H_{j+1}^{\ell+1} + H_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (H_{j}^{\ell+1} + H_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$
@@
 
-\begin{frame}{time-stepping in ice sheet models}
+\begin{frame}{time-stepping in current ice sheet models}
 
 \begin{itemize}
-\item current-technology models mostly ``believe'' that the problem is an advection and use explicit time stepping
-\item FIXME but this has performance consequences
-\item FIXME the problem is a DAE!
-\item FIXME implicit alternative requires solving step equations: simultaneous NCP for SKE and Stokes
+\item current-technology models mostly ``believe'' the advection schema
+\item current-technology models mostly use explicit time stepping
+    \begin{itemize}
+    \item[$\circ$] an embarrassing situation \dots the problem is a DAE!
+    \item[$\circ$] performance consequences: stability is determined by the coupled solution velocity $\bU$ not a pre-determined velocity
+    \end{itemize}
+
+\medskip
+\item<2> an alternative should be obvious to any numerical analyst \dots
+\item<2> \alert{implicit time-stepping} is expected for DAE problems
+\item<2> implicit steps imply a sequence of problems (``equations'') to solve at each time step
+\item<2> here: implicit $\implies$ NCP coupled to Stokes at each time step
 \end{itemize}
@@
 
-\section{on Stokes solver scaling}
+\section{Stokes solver scaling}
 
-\begin{frame}{regarding PDE solver complexity}
+\begin{frame}{PDE solver complexity}
 
-FIXME
+FIXME explain in terms of 2D Poisson
 \end{frame}
 
-\section{a simplified performance analysis}
+\begin{frame}{a scalable Stokes solver for ice sheets}
 
-\begin{frame}{complexity of ice sheet models}
+FIXME Isaac et al.~2015
+\end{frame}
 
-\begin{tabular}{llll}
-\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
-\onslide<1->{
-\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
-}
-\onslide<1->{
-explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
-}
-\onslide<2->{
-explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
-}
-\onslide<3->{
-implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$ \\
-}
-\onslide<3->{
-implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}$
-}
-\end{tabular}
+\begin{frame}{a scalable Stokes solver for ice sheets 2}
+
+FIXME table from Isaac, and estimate power $\alpha$
+\end{frame}
+
+\section{simplified performance analysis}
+
+\begin{frame}{analyzing ice sheet model performance}
 
@@
 
-\section{scalable implicit steps by multigrid?}
+\begin{frame}{ice sheet model performance: table}
 
-\begin{frame}{obstacle problem solver scaling}
+\begin{tabular}{llll}
+\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
+\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
+explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
+explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$
+\end{tabular}
 
-FIXME
+FIXME table from Bueler (2022)
 \end{frame}
+
+\begin{frame}{complexity for implicit solves}
+
+FIXME $\beta$, $\gamma$
+\end{frame}
+
+\begin{frame}{ice sheet model performance: table cont.}
+
+\begin{tabular}{llll}
+\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
+\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
+explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
+explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
+implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$ \\
+implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}$
+\end{tabular}
+
+FIXME table from Bueler (2022)
+\end{frame}
+
+\section{scalable approaches: candidates}
+
+\begin{frame}{semi-coupled time-stepping}
+
+FIXME L{\"o}fgren et al.~2022
+\end{frame}
+
+\begin{frame}{machine learning}
+
+FIXME Jouvet et al.~2021
+\end{frame}
+
+\begin{frame}{scalable obstacle problem solvers}
+
+FIXME chapter 9 from my book
+\end{frame}
+
+\begin{frame}{scalable obstacle problem solvers}
+
+FIXME multilevel multigrid approach
+\end{frame}
+
+\section{conclusion}
 

### 2023-01-21 talk/slides.tex
a bit more talk

@@
 \usepackage[T1]{fontenc}
-% Or whatever. Note that the encoding and the font should match. If T1
-% does not look nice, try deleting the line with the fontenc.
-
 \usepackage{empheq,bm,xspace,fancyvrb,soul}
-
 \usepackage{tikz}
 \usetikzlibrary{shapes,arrows.meta,decorations.markings,decorations.pathreplacing,fadings,positioning}
-
 \usepackage[kw]{pseudo}
 \pseudoset{left-margin=15mm,topsep=5mm,idfont=\texttt,st-left=,st-right=}
-
-% If you wish to uncover everything in a step-wise fashion, uncomment
-% the following command:
-%\beamerdefaultoverlayspecification{<+->}
 
@@
 \begin{itemize}
-\item before numerical modeling \dots \emph{what} are we modeling?
 \item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width
@@
     \item[$\circ$] subglacial melt
-    \item[$\circ$] melt when floating
-    \item[$\circ$] calving into the ocean
+    \item[$\circ$] sub-shelf melt (when floating)
+    \item[$\circ$] calving (into ocean)
     \end{itemize}
@@
 \begin{itemize}
-\item for simplicity/clarity of the upcoming model performance analysis, I will ignore quite a bit of physics
+\item for simplicity/clarity of the upcoming model performance analysis, I will ignore much of glacier physics
 \item \alert{ignoring}:
@@
 \smallskip
-\item<2-> {\small map: \quad $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$}
+\item {\small \alert{map: \, $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$} }
 \end{itemize}
@@
 \begin{columns}
-\begin{column}{0.6\textwidth}
+\begin{column}{0.55\textwidth}
 \begin{itemize}
-\item the \alert{data} $a(t,x,y)$, $b(x,y)$  is defined on a fixed domain:
+\item data $a(t,x,y)$, $b(x,y)$  is defined on a \alert{fixed domain}:
 	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
-\item<2> the solution \alert{surface elevation} $s(t,x,y)$ is defined on $[0,T]\times \Omega$
+\item<2-> solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
     \begin{itemize}
-    \item[$\circ$] $s=b$ where there is no ice
+    \item[$\circ$] also a fixed domain,
+    \item[$\circ$] but $s=b$ where there is no ice!
     \end{itemize}
-
-\medskip
-\item<2> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$ on which the solution \alert{velocity} $\bu(t,x,y,z)$ is defined:
+\item<3> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$, on which (solution) velocity $\bu(t,x,y,z)$ is defined:
     $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$
@@
 \end{column}
-\begin{column}{0.4\textwidth}
+\begin{column}{0.46\textwidth}
 \includegraphics<1>[width=\textwidth]{images/domain-data.png}
 
-\includegraphics<2>[width=\textwidth]{images/domain-velocity.png}
+\includegraphics<2>[width=\textwidth]{images/domain-surface.png}
+
+\includegraphics<3>[width=\textwidth]{images/domain-velocity.png}
 \end{column}
@@
 
-\begin{frame}{mass and momentum conservation}
+\begin{frame}{basic ice sheet model: conservation}
 
@@
     \end{itemize}
-\item a key idea is that conservation of mass is important both
+
+\medskip
+\item conservation of mass is important both
     \begin{itemize}
@@
 
-    \item[$\circ$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
+        \begin{itemize}
+        \item[$\vartriangleright$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
+        \end{itemize}
     \end{itemize}
@@
 \begin{itemize}
-\item ice sheet evolution is a \alert{free-boundary} problem for conserved mass and momentum quantities
+\item ice sheet evolution is a \alert{free-boundary} problem for conserved quantities
 \item specifically, the surface kinematic equation
@@
 applies \emph{only} on the ice upper surface $\Gamma_s(t)$
-\item in the remainder of the (fixed, map-plane) domain $\Omega\subset \RR^2$, \alert{complementarity} holds:
+\item in the remainder of the (fixed) domain $\Omega\subset \RR^2$, \alert{complementarity} holds:
   $$s=b \quad \text{and} \quad a \le 0$$
@@
 
-\begin{frame}{basic ice sheet model: strong form}
+\begin{frame}{basic ice sheet model: strong form = NCP coupled to Stokes}
 
 \begin{itemize}
-\item nonlinear complementarity problem (NCP) \onslide<2->{coupled to Stokes}:
+\item \only<1>{\alert{nonlinear complementarity problem} (NCP)} \only<2>{nonlinear complementarity problem (NCP) coupled to Stokes}:
 \begin{align*}
@@
 \nabla \cdot \bu &= 0 && \text{''} \\
-\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t) \subset \partial \Lambda(t)$} \\
+\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t)$} \\
 \bu|_b \cdot \bn_b &= 0 && \text{''} \\
-\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t) \subset \partial \Lambda(t)$}}
+\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t)$}}
 \end{align*}
@@
     \begin{itemize}
-    \item $\bu|_s=0$ where no ice
-    \item<2-> viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
+    \item[$\circ$] $\bu|_s=0$ where no ice
+    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
     \end{itemize}
@@
 
-\begin{frame}{basic ice sheet model: its a DAE system anyway}
+\begin{frame}{basic ice sheet model: is a DAE system}
 
 \begin{itemize}
-\item for clarity and simplicity, declutter by forgetting complementarity and all the other boundary conditions, too \dots
-\item get surface kinematic equation coupled to Stokes:
+\item for this slide, forget complementarity and all boundary conditions
+\item result is surface kinematic equation coupled to Stokes:
 \begin{align*}
-\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 && \text{on } \Omega \subset \RR^2 \\
-- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
-\nabla \cdot \bu &= 0 && \text{''}
+\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
+- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
+\nabla \cdot \bu &= 0
 \end{align*}
-\item observe that $s(t,x,y)$ and $\bu(t,x,y,z)$ both evolve in time
-\item \emph{however}, only the first of the 5 equations has a time derivative
-\item the problem is a \alert{differential algebraic equation} system, e.g.
+
+\vspace{-2mm}
+    \begin{itemize}
+    \item[$\circ$] $s(t,x,y)$ and $\bu(t,x,y,z)$ both evolve in time
+    \end{itemize}
+\item<2-> only the first of these 5 equations has a time derivative
+\item<3> the problem is a \alert{differential algebraic equation} (DAE) system:
 \begin{align*}
@@
     \begin{itemize}
-    \item[$\circ$] \dots actually in infinite dimensions (\emph{PDAE}?), and subject to complementarity (\emph{free boundary problem for PDAE})
+    \item<3>[$\circ$] in infinite dimensions (PDAE?), and subject to complementarity
     \end{itemize}
@@
 
-\begin{frame}{important context (current research)}
+\begin{frame}{basic ice sheet model: current research and numerics}
 
 \begin{itemize}
-\item to the best of my knowledge, \emph{no} current research groups are studying the well-posedness or regularity of this ``NCP coupled to Stokes'' glacier evolution model
+\item to the best of my knowledge, \emph{no} current research groups are studying the well-posedness or regularity of this basic ice sheet model
     \begin{itemize}
-    \item[$\circ$] however, I think most researchers would agree this is the intended model!
-    \item[$\circ$] well-posedness of shallow approximations has been considered\footnote{e.g.~Jouvet \& Bueler (2012)}
+    \item[$\circ$] but most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model?
+    \item[$\circ$] well-posedness of shallow approximations \emph{has} been considered
     \end{itemize}
@@
 \medskip
-\item<2> numerical modelers tend think of the Stokes problem separately from surface evolution (\emph{time-split} or \emph{explicit time-stepping}),
-\item<2> and glacier evolution is addressed with minimal awareness of complementarity
+\item<2> numerical modelers tend think of the Stokes problem separately from surface evolution
+    \begin{itemize}
+    \item[$\circ$] \emph{time-splitting} or \emph{explicit time-stepping} is taken for granted
+    \end{itemize}
+\item<2> glacier evolution is addressed with minimal awareness of complementarity
 \end{itemize}
@@
 
-\begin{frame}{the mass-continuity equation view}
+\AtBeginSection[]
+{
+  \begin{frame}{Outline}
+    \tableofcontents[currentsection]
+  \end{frame}
+}
+
+\section{time-stepping and stability}
+
+\begin{frame}{ice sheet models: the mass-continuity equation view}
 
 \begin{itemize}
-\item thickness-transport form helps expose evolution character
-\item define:
+\item thickness transport form helps expose evolution character
+\item define
 \begin{align*}
@@
 \item<3->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
-  $$\bU \sim - \grad s \sim - \grad H$$
+  $$\bU \sim - \grad s \sim - \grad H,$$
+and ice is slow (DAE) so there are no characteristics
 \end{itemize}
@@
 \begin{align*}
-\text{advective form:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
-\text{diffusion form:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
+\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
+\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
 \end{align*}
 \begin{itemize}
-\item both forms are highly-nonlinear: \, $\bU=\bU(H,\grad s), \, D = D(H,\grad s)$
-\item the diffusion form is only \emph{literally} valid in the lubrication approximation (more below)
-\item the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences} regardless
+\item both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
+\item the diffusion schema is only literal in the lubrication approximation
+    \begin{itemize}
+    \item[$\circ$] more on this momentarily
+    \end{itemize}
+\item the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences}, regardless of your advective schema preference
 \end{itemize}
@@
 
-\begin{frame}{numerical shallow ice sheet models}
+\begin{frame}{shallow ice sheet models}
 
 \begin{itemize}
-\item the ``basic ice sheet model'' (NCP coupled to Stokes) is \emph{not yet} in common use for ice sheet simulations
+\item the basic ice sheet model (NCP coupled to Stokes) is \emph{not yet} in common use for ice sheet simulations
     \begin{itemize}
-    \item[$\circ$] usual justification for falling back to shallow approximations: ``Stokes is too expensive at scale''
-    \item[$\circ$] goal: \alert{make it fast enough to use!}
+    \item[$\circ$] \alert{can we make it fast enough to use?}
     \end{itemize}
@@
 \smallskip
-\item<2-> current-technology models use various shallow simplifications
+\item current-technology models use various shallow simplifications
     \begin{itemize}
-    \item<2->[$\circ$] usual justification for falling back to shallow approximations: ``Stokes is too expensive at scale''
+    \item[$\circ$] usual justification: ``Stokes is too expensive at scale''
     \end{itemize}
-\item<2-> the simplest shallow approximation is a lubrication approximation, the \alert{shallow ice approximation} (SIA)
-\item<2-> in the SIA, $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator
+\item the simplest of several shallow approximations is the ``lubrication'' approximation, the \alert{shallow ice approximation} (SIA)
 
@@
 
-\begin{frame}{shallow ice approximation NCP: details}
+\begin{frame}{shallow ice approximation NCP}
 
+\begin{itemize}
+\item in the shallow ice approximation, the surface motion contribution $\Phi(s) = - \bu|_s \cdot \bn_s$ has a formula:
 {\small
-\begin{align*}
-s - b &\ge 0 \\
-\frac{\partial s}{\partial t} + \Phi(s) - a &\ge 0 \\
-(s - b) \left(\frac{\partial s}{\partial t} + \Phi(s) - a\right) &= 0
-\end{align*}
-\begin{equation*}
-\Phi(s) = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)
-\end{equation*}
+\begin{gather*}
+s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0 \\
+{\color{FireBrick} \Phi(s) = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)}
+\end{gather*}
+
+\vspace{-2mm}
+    \begin{itemize}
+    \item[$\circ$] constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation
+    \end{itemize}
 }
 
+\medskip
+\item<2> $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
     \begin{itemize}
-\item constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation
-
-\medskip
-\item<2> thus: $\Phi(s) = - \bu|_s \cdot \bn_s$ resolves to a \alert{doubly-nonlinear} differential operator, of porous medium and $\pp$-Laplacian type simultaneously
-\item<2> Jouvet \& Bueler (2012): existence for this problem
+    \item<2>[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
+    \item<2>[$\circ$] \emph{local} in surface and bed topography
+    \item<2>[$\circ$] existence is known for this problem
+         \begin{itemize}
+         \item[$\vartriangleright$] Jouvet \& Bueler (2012); NCP is written as a \alert{variational inequality} weak form
+         \end{itemize}
+    \end{itemize}
 \end{itemize}
@@
 
-\AtBeginSection[]
-{
-  \begin{frame}{Outline}
-    \tableofcontents[currentsection]
+\begin{frame}{nonlocality}
+
+\begin{itemize}
+\item the basic ice sheet model (NCP coupled to Stokes) problem actually has a \emph{non}-local surface velocity function $\Phi(s) = - \bu|_s \cdot \bn_s$
+{\small
+\begin{equation*}
+s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0
+\end{equation*}
+}
+\item \emph{figure:} the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the shallow ice approximation velocity responds only underneath the surface perturbation}
+\end{itemize}
+
+\begin{center}
+\includegraphics<1>[width=0.6\textwidth]{images/stokes-greens-arndt.png}
+\includegraphics<2>[width=0.6\textwidth]{images/sia-greens-arndt.png}
+\end{center}
 \end{frame}
-}
-
-\section{time-stepping and stability}
 
@@
 \begin{align*}
-\text{advective ice sheet model schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
-\text{diffusion ice sheet model schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
+\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
+\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
 \end{align*}
@@
 \begin{itemize}
-\item recall: traditional numerical analysis applied to these equations
+\item recall traditional numerical analysis applied to these equations
 \end{itemize}
@@
     \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
+    \item[$\circ$] conditional stability
     \item[$\circ$] maximum time step from CFL using $\max |\bU|$
@@
 \item for example, backward Euler:
-	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (H_{j+1}^{\ell+1} + H_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (H_{j}^{\ell+1} + H_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$
+	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell+1} + s_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (s_{j}^{\ell+1} + s_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$
 
     \begin{itemize}
-    \item[$\circ$] or use: Crank-Nicolson, BDF, \dots
+    \item[$\circ$] or: Crank-Nicolson, BDF, \dots
+    \item[$\circ$] unconditional stability
+    \item[$\circ$] must solve equations at each step
     \end{itemize}
@@
 \begin{itemize}
-\item current-technology models mostly ``believe'' the advection schema
+\item current researchers mostly believe the advection schema
 \item current-technology models mostly use explicit time stepping
     \begin{itemize}
-    \item[$\circ$] an embarrassing situation \dots the problem is a DAE!
-    \item[$\circ$] performance consequences: stability is determined by the coupled solution velocity $\bU$ not a pre-determined velocity
+    \item[$\circ$] embarrassing: the problem is a DAE
+    \item[$\circ$] performance and design consequences: stability is determined by the coupled solution velocity $\bU$
     \end{itemize}
@@
 \medskip
-\item<2> an alternative should be obvious to any numerical analyst \dots
 \item<2> \alert{implicit time-stepping} is expected for DAE problems
-\item<2> implicit steps imply a sequence of problems (``equations'') to solve at each time step
-\item<2> here: implicit $\implies$ NCP coupled to Stokes at each time step
+    \begin{itemize}
+    \item[$\circ$] the obvious alternative
+    \end{itemize}
+\item<2> a sequence of problems (``equations'') must be solved at each time step: NCP coupled to Stokes
 \end{itemize}
@@
 
+\begin{frame}{the Stokes problem for ice}
+
+\begin{itemize}
+\item from now on assume our ice sheet model is \emph{not} shallow
+\item whether time steps are explicit or implicit, a Stokes problem is solved at each step
+\item for a given icy domain $\Lambda \subset \RR^3$:
+\begin{align*}
+- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda$} \\
+\nabla \cdot \bu &= 0 && \text{''} \\
+\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
+\bu|_b \cdot \bn_b &= 0 && \text{''} \\
+\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
+\end{align*}
+\item how fast is the numerical solution process?
+\end{itemize}
+\end{frame}
+
 \begin{frame}{PDE solver complexity}
 
-FIXME explain in terms of 2D Poisson
+FIXME explain for Poisson
 \end{frame}
@@
 FIXME Jouvet et al.~2021
+
+\begin{center}
+\includegraphics[width=0.45\textwidth]{images/cnn-jouvet.png}
+
+\includegraphics[width=0.65\textwidth]{images/tiles-jouvet.png}
+\end{center}
 \end{frame}
@@
 \item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
+   \begin{itemize}
+   \item[$\circ$] progress will be made in all of these areas
+   \end{itemize}
 \item<3-> \alert{coming soon} from current research:
-   \begin{itemize}
-   \item[$\circ$] scalable Stokes solvers (Isaac et al.~2015)
-   \item[$\circ$] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
-   \item[$\circ$] ML emulators (Jouvet et al.~2021)
-   \end{itemize}
-\item<4-> scalable solvers for the implicit-step, coupled problem for ice sheet geometry requires advances in \alert{multilevel solutions of non-local variational inequalities}
+   \begin{enumerate}
+   \item[1.] machine learning emulators (Jouvet et al.~2021)
+   \item[2.] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
+   \item[3.] scalable Stokes solvers (Isaac et al.~2015)   
+   \end{enumerate}
+\item<4-> scalable solvers for the implicit-step problem require advances in \alert{multilevel solutions of non-local variational inequalities}
 \end{itemize}

### 2023-01-22 talk/slides.tex
last section still needs to be written

@@
 
-\section{Stokes solver scaling}
+\section{stress balance solver scaling}
 
@@
 \begin{itemize}
-\item from now on assume our ice sheet model is \emph{not} shallow
-\item whether time steps are explicit or implicit, a Stokes problem is solved at each step
-\item for a given icy domain $\Lambda \subset \RR^3$:
+\item assume an ice sheet model which is \emph{not} shallow
+\item whether explicit or implicit, solves a Stokes problem at each step:
 \begin{align*}
@@
 \end{align*}
+\item this problem is the \alert{stress balance} (conservation of momentum) which determines velocity $\bu$ and pressure $p$
 \item how fast is the numerical solution process?
@@
 
-\begin{frame}{PDE solver complexity}
+\begin{frame}{PDE solver algorithmic scaling}
 
-FIXME explain for Poisson
+\begin{itemize}
+\item as intro to scaling, consider the 3D Poisson equation:
+    $$-\grad^2 u = f \text{ in $\Omega$}, \quad u=0 \text{ on $\partial\Omega$}$$
+\item discretization generates a linear system $A\bu = \bb$
+\item (data size $m$) = \alert{number of unknowns}
+    \begin{itemize}
+    \item[$\circ$] for low-order discretizations, $m$ = (number of nodes in the grid)
+    \item[$\circ$] for cell size $h$, $m \sim h^{-3}$ in 3D
+    \end{itemize}
+\item solver choice determines asymptotic growth of computational expense (complexity)
+\item \alert{algorithmic scaling} of flops as $m\to\infty$:
+    \begin{itemize}
+    \item[$\circ$] $O(m^3)$ for direct linear algebra, ignoring matrix structure
+    \item[$\circ$] $\approx O(m^2)$ for sparsity-exploiting direct linear algebra
+    \item[$\circ$] $O(m^1)$, \alert{optimal}, for \alert{multigrid} solvers (below)
+    \end{itemize}
+\end{itemize}
+
+\medskip
+\begin{center}
+\includegraphics[width=0.5\textwidth]{images/multigrid-grids.png}
+\end{center}
 \end{frame}
 
-\begin{frame}{a scalable Stokes solver for ice sheets}
+\begin{frame}{scalable stress balance solvers for ice sheets}
 
-FIXME Isaac et al.~2015
+\begin{itemize}
+\item ice sheet stress balance solvers can be analyzed similarly
+\item $m$ = (number of velocity and pressure unknowns)
+    \begin{itemize}
+    \item[$\circ$] indicate scaling by ``$O(m^{1+\alpha})$'', with $\alpha=0$ optimal
+    \end{itemize}
+\item \alert{near-optimal solvers} already exist: \hfill \emph{good news!}
+    \begin{itemize}
+    \item[$\circ$] $\alpha=0.08$ for Isaac et al.~(2015) Stokes solver
+        \begin{itemize}
+        \item[$\vartriangleright$] unstructured quadrilateral/tetrahedral mesh, $Q_k\times Q_{k-2}$ stable elements, Schur-preconditioned Newton-Krylov, column-oriented algebraic multigrid (AMG) preconditioner for $(\bu,\bu)$ block
+        \end{itemize}
+
+\begin{center}
+\includegraphics[width=0.6\textwidth]{images/isaac-antarctica.png}
+\end{center}
+    \item[$\circ$] $\alpha=0.05$ for Tuminaro et al~(2016) 1st-order (shallow) AMG solver
+    \item[$\circ$] similar for Brown et al~(2013) GMG solver
+    \end{itemize}
+\end{itemize}
 \end{frame}
 
-\begin{frame}{a scalable Stokes solver for ice sheets 2}
-
-FIXME table from Isaac, and estimate power $\alpha$
-\end{frame}
-
-\section{simplified performance analysis}
+\section{ice sheet model performance analysis}
 
@@
 
-FIXME asymptotic estimates of algorithmic scaling, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$
+\begin{itemize}
+\item for a time-stepping model our metric is \alert{flops per model year}
+\item $m$ = (\# of surface elevation, velocity, and pressure unknowns)
+\item domain $\Omega \subset \RR^2$ (map-plane) has characteristic width $L$
+\item map-plane (2D) mesh cells have characteristic diameter $h$:
+  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}}$$
+\item recall mass-continuity equation schemata:
+{\footnotesize
+\begin{align*}
+\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
+\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
+\end{align*}
+}
+\end{itemize}
 \end{frame}
 
-\begin{frame}{ice sheet model performance: table}
+\begin{frame}{ice sheet model performance}
 
 \begin{tabular}{llll}
-\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
-\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
-explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
-explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$
+\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
+\\
+explicit & SIA    & $\oo{\frac{D\, L^2}{h^4}} = \oo{\frac{D}{L^2} m^2}$ \\
+\\
+explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U L^{2+2\alpha}}{h^{3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
+\\
+\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{4+2\alpha}}} = \oo{\frac{D}{L^2} m^{2+\alpha}}$
 \end{tabular}
-
-FIXME table from Bueler (2022)
 \end{frame}
 
-\begin{frame}{complexity for implicit solves}
+\begin{frame}{implicit time-stepping for ice sheet models}
 
-FIXME $\beta$, $\gamma$
+\begin{itemize}
+\item main point: \alert{explicit time-stepping stability restrictions imply many stress-balance solves}
+    \begin{itemize}
+    \item[$\circ$] $\alpha$ for stress balance solver important
+    \item[$\circ$] however, even optimality ($\alpha=0$) cannot fix low model performance
+    \end{itemize}
+\item switch to \alert{implicit time-stepping} for unconditional stability?
+    \begin{itemize}
+    \item[$\circ$] each step is a NCP-coupled-to-Stokes problem
+        \begin{itemize}
+        \item[$\vartriangleright$] or SIA NCP
+        \item[$\vartriangleright$] in any case, a \alert{free-boundary problem}
+        \item[$\vartriangleright$] next slide parameterizes expense of these coupled, implicit-step solves as $O(m^{1+\beta})$; optimal is $\beta=0$
+        \end{itemize}
+    \item[$\circ$] need $q$ model updates per year to integrate climate influences
+    \end{itemize}
+
+\bigskip
+\item<2> \alert{existing} implicit time-stepping ice sheet models:
+    \begin{itemize}
+    \item[$\circ$] Bueler (2016) SIA NCP solver scales badly ($\beta=0.8$)
+    \item[$\circ$] no other fully-implicit NCP solvers known?
+    \end{itemize}
+\end{itemize}
 \end{frame}
 
-\begin{frame}{ice sheet model performance: table cont.}
+\begin{frame}{ice sheet model performance 2}
 
 \begin{tabular}{llll}
-\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
-\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
-explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
-explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
-implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$ \\
-implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}$
+\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
+\\
+explicit & SIA    & $\oo{\frac{D\, L^2}{h^4}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
+\\
+explicit \, {\footnotesize \emph{advective}} & Stokes & $\oo{\frac{U L^{2+2\alpha}}{h^{3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
+\\
+\phantom{explicit} \, {\footnotesize \emph{diffusive}}  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{4+2\alpha}}} = \oo{\frac{D}{L^2} m^{2+\alpha}}$ \\
+\\
+implicit & \emph{either} & $\oo{\frac{q\, L^{2+2\beta}}{h^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$
 \end{tabular}
-
-FIXME table from Bueler (2022)
 \end{frame}
@@
    \end{enumerate}
-\item<4-> scalable solvers for the implicit-step problem require advances in \alert{multilevel solutions of non-local variational inequalities}
+\item<4-> scalable solvers for implicit-step models require advances in \alert{multilevel solutions of non-local variational inequalities}
 \end{itemize}

### 2023-01-22 paper/data/scaling.m
clean up script used in paper ... realize I probably mis-read Isaac ... actually alpha=0.07?

@@ -1,4 +1,4 @@
-% generate serial scaling for bueler 2016 and isaac et al 2015 and tumarino et al 2016
+% generate complexity data for bueler 2016 and isaac et al 2015 and tumarino et al 2016
 
 % note on iterations: if iterations scale as O(m^alpha) then flops scale as O(m^{1+alpha}); this is under assumption that each iteration costs O(m) flops
 
@@ -6,19 +6,28 @@
 % isaac: AMG-Krylov-Newton Stokes solve for antarctica; see table 8.1 in paper
 % tumarino: MDSC–AMG-Krylov-Newton higher-order solve for greenland; see table 7.5 in paper; area of Greenland taken from Searise data: 300x560 5km grid
 
+% BUELER
 nb = [18 36 72 144 288].^2;
 flopsb = [1.776e7 1.749e8 2.374e9 2.608e10 2.927e11];
-pb = polyfit(log(nb),log(flopsb),1);
+pb = polyfit(log(nb),log(flopsb),1)
 %ans = 1.76186   6.45957   ==>  beta = 0.76186
 
+% ISAAC:  publication uses kitsi.*nitsi (last), but probably 1st is correct
 ni = [7e6 51e6 383e6];
-itsi = [7*66 10*75 11*90]; % total Krylov iterations over Newton steps
-pi = polyfit(log(ni),log(itsi),1);
+kitsi = [66 75 90]; % total ("overall") Krylov iterations; averaged over P vals
+pi = polyfit(log(ni),log(kitsi),1)
+%ans = 0.0775   2.9589   ==>  alpha = 0.0775
+%flopsi = ni .* kitsi;
+%pflopsi = polyfit(log(ni),log(flopsi),1)
+%ans = 1.0775   2.9589  ==>  alpha = 0.0775  (same as above)
+%nitsi = [7 10 11]; % round down average Newton iterations to integer
+%pi = polyfit(log(ni),log(kitsi .* nitsi),1)
 %ans = 0.19030   3.17188   ==>  alpha = 0.19030
 
+% TUMINARO
 nt = ((300*560*5^2) / (8^2)) * [1 4 16 64 256];
 itst = [36*13.3 34*12.7 35*12.3 36*13.3 37*17.4]; % total Krylov iterations over Newton steps
-pt = polyfit(log(nt),log(itst),1);
+pt = polyfit(log(nt),log(itst),1)
[…]

### 2023-01-23 talk/slides.tex
finally close to done with the draft talk ... but not done

@@
 
-\bigskip
-\begin{itemize}
-\item recall traditional numerical analysis applied to these equations
-\end{itemize}
-
 \begin{minipage}[t][60cm][t]{\textwidth}
 \begin{itemize}
-\only<1>{\item[]}
+\only<1>{
+\item \emph{caution}: for ice sheet models, $\bU(t,x,y)$ and $D(t,x,y)>0$ vary in space and time, from nonlinear coupling
+\item nonetheless, let us recall some traditional numerical analysis of these equations \dots
+}
 \only<2>{
-\item we expect \alert{explicit} time stepping for advections
-\item for example, forward Euler:
-	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{(\bU H)_{j+1/2}^\ell - (\bU H)_{j-1/2}^\ell}{\Delta x} = a_j^\ell$$
+\item \alert{explicit} time stepping is common for \alert{advections}
+\item for example, forward Euler using spacing $h$ and time step $\Delta t$:
+	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{(\bU H)_{j+1/2}^\ell - (\bU H)_{j-1/2}^\ell}{h} = a_j^\ell$$
 
     \begin{itemize}
+    \item[$\circ$] conditionally stable, with maximum time step
+        $$\Delta t \le \frac{h}{\max |\bU|} = O(h)$$
     \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
-    \item[$\circ$] conditional stability
-    \item[$\circ$] maximum time step from CFL using $\max |\bU|$
     \end{itemize}
@@
 \only<3>{
-\item we expect \alert{implicit} time stepping for diffusions
-\item for example, backward Euler:
-	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell+1} + s_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (s_{j}^{\ell+1} + s_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$
+\item \alert{explicit} time stepping for \alert{diffusions} is best avoided
+\item for example, forward Euler:
+	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell} + s_{j}^{\ell}) - D_{j-\frac{1}{2}} (s_{j}^{\ell} + s_{j-1}^{\ell})}{h^2} = a_j^\ell$$
 
     \begin{itemize}
-    \item[$\circ$] or: Crank-Nicolson, BDF, \dots
-    \item[$\circ$] unconditional stability
-    \item[$\circ$] must solve equations at each step
+    \item[$\circ$] conditionally stable, with maximum time step
+        $$\Delta t \le \frac{h^2}{\max D} = O(h^2)$$
+    \end{itemize}
+}
+\only<4>{
+\item \alert{implicit} time stepping for \alert{diffusions} is recommended
+\item for example, backward Euler:
+	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell+1} + s_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (s_{j}^{\ell+1} + s_{j-1}^{\ell+1})}{h^2} = a_j^\ell$$
+
+    \begin{itemize}
+    \item[$\circ$] unconditionally stable, but must solve equations at each step
+    \item[$\circ$] alternative implicit schemes: Crank-Nicolson, BDF, \dots
     \end{itemize}
@@
     \begin{itemize}
-    \item[$\circ$] embarrassing: the problem is a DAE
-    \item[$\circ$] performance and design consequences: stability is determined by the coupled solution velocity $\bU$
+    \item[$\circ$] embarrassing: \alert{the mathematical problem is a DAE}
+    \item[$\circ$] time step size is determined by the coupled solution velocity $\bU$,
+    \item[$\circ$] but there are performance and design consequences of hidden DAE \& diffusive character
     \end{itemize}
@@
 \medskip
-\item<2> \alert{implicit time-stepping} is expected for DAE problems
+\item<2> \alert{implicit time-stepping} is expected for DAE problems,
+\item<2> but a sequence of difficult free-boundary problems must be solved at each time step: \alert{NCP coupled to Stokes at each time step}
     \begin{itemize}
-    \item[$\circ$] the obvious alternative
+    \item[$\circ$] not really ``equations'' any more because of $s\ge b$ inequality
     \end{itemize}
-\item<2> a sequence of problems (``equations'') must be solved at each time step: NCP coupled to Stokes
 \end{itemize}
@@
 
-\begin{frame}{analyzing ice sheet model performance}
+\begin{frame}{simplified analysis of ice sheet model performance}
 
@@
 \item for a time-stepping model our metric is \alert{flops per model year}
-\item $m$ = (\# of surface elevation, velocity, and pressure unknowns)
-\item domain $\Omega \subset \RR^2$ (map-plane) has characteristic width $L$
-\item map-plane (2D) mesh cells have characteristic diameter $h$:
-  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}}$$
-\item recall mass-continuity equation schemata:
-{\footnotesize
-\begin{align*}
-\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
-\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
-\end{align*}
-}
+    \begin{itemize}
+    \item[$\circ$] glaciologists want high-resolution simulations of ice sheets over 100k year ice age cycles
+    \end{itemize}
+\item assume map-plane domain $\Omega \subset \RR^2$ of width $L$
+\item assume map-plane mesh cells have characteristic diameter $h$
+\item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
+\item the question addressed by upcoming performance table:
+
+\bigskip
+\begin{center}
+\begin{minipage}{0.82\textwidth}
+how does flops per model year \alert{scale} in the \alert{high spatial resolution limit} $h\to 0$, equivalently $m\to \infty$?
+\end{minipage}
+\end{center}
 \end{itemize}
@@
 
-\begin{frame}{ice sheet model performance}
+\begin{frame}{preparation: observe and recall}
+
+\begin{itemize}
+\item ice sheets are thin layers, thus even non-shallow Stokes models often have $O(1)$ mesh points in the vertical direction
+    \begin{itemize}
+    \item[$\circ$] e.g.~Issac et al (2015) Stokes solver
+    \item[$\circ$] \emph{message}: I am ignoring refinement in the vertical
+    \end{itemize}
+
+\medskip
+\item scale relationships:
+  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}} \hspace{20mm}$$
+
+\vspace{-18mm}
+\hfill \includegraphics[width=0.15\textwidth]{images/one-grid.png} \hspace{5mm}
+
+\medskip
+\item mass-continuity equation time-stepping schemata:
+{\footnotesize
+\begin{align*}
+\text{advective} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a & \Delta t &\le \frac{h}{U} \\
+\text{diffusion} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a & \Delta t &\le \frac{h^2}{D}
+\end{align*}
+}
+
+\item stress-balance solver scaling parameterized as $O(m^{1+\alpha})$
+\end{itemize}
+\end{frame}
+
+\begin{frame}{\emph{explicit} ice sheet model performance}
 
@@
 \\
-explicit & SIA    & $\oo{\frac{D\, L^2}{h^4}} = \oo{\frac{D}{L^2} m^2}$ \\
+explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
 \\
-explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U L^{2+2\alpha}}{h^{3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
+explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
 \\
-\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{4+2\alpha}}} = \oo{\frac{D}{L^2} m^{2+\alpha}}$
+\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$
 \end{tabular}
+
+\bigskip\bigskip
+\begin{itemize}
+\item but what about implicit?
+\end{itemize}
 \end{frame}
@@
 \begin{itemize}
-\item main point: \alert{explicit time-stepping stability restrictions imply many stress-balance solves}
+\item main point so far: \alert{explicit} time-stepping \alert{implies many stress-balance solves}
     \begin{itemize}
-    \item[$\circ$] $\alpha$ for stress balance solver important
-    \item[$\circ$] however, even optimality ($\alpha=0$) cannot fix low model performance
+    \item[$\circ$] yes, stress balance solver scaling ($\alpha$) is important
+    \item[$\circ$] however, even optimality ($\alpha=0$) cannot fix performance
     \end{itemize}
@@
         \begin{itemize}
-        \item[$\vartriangleright$] or SIA NCP
-        \item[$\vartriangleright$] in any case, a \alert{free-boundary problem}
-        \item[$\vartriangleright$] next slide parameterizes expense of these coupled, implicit-step solves as $O(m^{1+\beta})$; optimal is $\beta=0$
+        \item[$\vartriangleright$] a \alert{free-boundary problem}
+        \item[$\vartriangleright$] table on next slide parameterizes cost of coupled implicit-step solves as $O(m^{1+\beta})$; optimal is $\beta=0$
         \end{itemize}
@@
 \bigskip
-\item<2> \alert{existing} implicit time-stepping ice sheet models:
+\item<2> \alert{existing} implicit time-stepping ice sheet models?
     \begin{itemize}
+    \item[$\circ$] no fully-implicit NCP-coupled-to-Stokes models exist (?)
     \item[$\circ$] Bueler (2016) SIA NCP solver scales badly ($\beta=0.8$)
-    \item[$\circ$] no other fully-implicit NCP solvers known?
     \end{itemize}
@@
 
-\begin{frame}{ice sheet model performance 2}
+\begin{frame}{ice sheet model performance table (Bueler, 2022)}
 
@@
 \\
-explicit & SIA    & $\oo{\frac{D\, L^2}{h^4}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
+explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
 \\
-explicit \, {\footnotesize \emph{advective}} & Stokes & $\oo{\frac{U L^{2+2\alpha}}{h^{3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
+explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
 \\
-\phantom{explicit} \, {\footnotesize \emph{diffusive}}  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{4+2\alpha}}} = \oo{\frac{D}{L^2} m^{2+\alpha}}$ \\
+\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$ \\
 \\
-implicit & \emph{either} & $\oo{\frac{q\, L^{2+2\beta}}{h^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$
+implicit & \emph{either} & $\oo{\frac{q\, L^{2+2\beta}}{h^{\,2+2\beta}}} = \oo{q\, m^{1+\beta}}$
 \end{tabular}
@@
 
-\begin{frame}{semi-coupled time-stepping}
+\begin{frame}{scalable approach 1: machine learning}
+
+\begin{itemize}
+\item apply \alert{machine learning}
+\item \emph{concept (Jouvet et al.~2021)}: run existing non-scalable ice sheet models on a range of hypothetical/real ice sheets, and train an ML on the results
+    \begin{itemize}
+    \item[$\circ$] supervised learning of physically-based model yields \alert{emulator}
+    \end{itemize}
+\end{itemize}
+
+\vspace{3mm}
+\begin{minipage}[t][60cm][t]{\textwidth}
+\begin{center}
+\only<1>{\includegraphics[height=0.52\textheight]{images/tiles-jouvet.png}}
+\only<2>{\includegraphics[height=0.6\textheight]{images/cnn-jouvet.png}}
+\end{center}
+\end{minipage}
+\end{frame}
+
+\begin{frame}{scalable approach 2: smarter explicit time-stepping}
+
+semi-coupled
 
@@
 
-\begin{frame}{machine learning}
-
-FIXME Jouvet et al.~2021
-
-\begin{center}
-\includegraphics[width=0.45\textwidth]{images/cnn-jouvet.png}
-
-\includegraphics[width=0.65\textwidth]{images/tiles-jouvet.png}
-\end{center}
-\end{frame}
-
-\begin{frame}{scalable obstacle problem solvers}
+\begin{frame}{scalable approach 3: multigrid implicit solvers}
 
@@
 
-\begin{frame}{scalable obstacle problem solvers}
+\begin{frame}{scalable approach 3: multigrid implicit solvers}
 

### 2023-01-23 talk/slides.tex
fiddle with the ML slide

@@
 \item apply \alert{machine learning}
-\item \emph{concept (Jouvet et al.~2021)}: run existing non-scalable ice sheet models on a range of hypothetical/real ice sheets, and train an ML on the results
+\item \emph{concept}: apply non-scalable ice sheet models on many hypothetical/real ice sheets, and train ML \alert{emulator} on results
     \begin{itemize}
-    \item[$\circ$] supervised learning of physically-based model yields \alert{emulator}
+    \item[$\circ$] supervised learning of physically-based model results
+    \item[$\circ$] compute map on CPUs, then learn \& evaluate map on GPUs
+    \item[$\circ$] convincing demo in Jouvet et al.~(2021)
     \end{itemize}
@@
 \only<1>{\includegraphics[height=0.52\textheight]{images/tiles-jouvet.png}}
-\only<2>{\includegraphics[height=0.6\textheight]{images/cnn-jouvet.png}}
+\only<2>{\includegraphics[height=0.57\textheight]{images/cnn-jouvet.png}}
 \end{center}

### 2023-01-24 talk/slides.tex
complete first draft I guess

@@
 \begin{definition}
-an \alert{ice sheet model} is a \underline{map} which simulates an ice sheet in a climate
+an \alert{ice sheet model} is a map which simulates an ice sheet in a climate
 \end{definition} 
@@
 \smallskip
-\item {\small \alert{map: \, $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$} }
+\item {\small \alert{map}: \, $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$ }
 \end{itemize}
@@
     \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
-\text{\alert{surface kinematic equation}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
+\text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
 \text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
@@
 \item ice sheet evolution is a \alert{free-boundary} problem for conserved quantities
-\item specifically, the surface kinematic equation
+\item specifically, the surface kinematic equation (SKE)
   $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
@@
 \begin{itemize}
-\item for this slide, forget complementarity and all boundary conditions
-\item result is surface kinematic equation coupled to Stokes:
+\item for this slide, forget complementarity and boundary conditions
+\item result is SKE coupled to Stokes:
 \begin{align*}
@@
 
-\vspace{-2mm}
+\item<2-> only the first of these 5 equations has a time derivative
     \begin{itemize}
-    \item[$\circ$] $s(t,x,y)$ and $\bu(t,x,y,z)$ both evolve in time
+    \item[$\circ$] because ice is slow and incompressible
     \end{itemize}
-\item<2-> only the first of these 5 equations has a time derivative
-\item<3> the problem is a \alert{differential algebraic equation} (DAE) system:
+\item<3> the problem is a \alert{differential algebraic equation} (DAE), an extremely \alert{stiff}, system:
 \begin{align*}
@@
     \begin{itemize}
-    \item<3>[$\circ$] in infinite dimensions (PDAE?), and subject to complementarity
+    \item<3>[$\circ$] but in $\infty$ dimensions (PDAE?) and subject to complementarity
     \end{itemize}
@@
 
-\begin{frame}{basic ice sheet model: current research and numerics}
+\begin{frame}{basic ice sheet model: current research and thinking}
 
@@
 \medskip
-\item<2> numerical modelers tend think of the Stokes problem separately from surface evolution
+\item<2> numerical modelers tend to think of the Stokes problem separately from surface evolution
     \begin{itemize}
@@
     \end{itemize}
-\item<2> glacier evolution is addressed with minimal awareness of complementarity
+\item<2> ice sheet geometry evolution is addressed with minimal awareness of complementarity
 \end{itemize}
@@
 
-\section{time-stepping and stability}
+\section{time-stepping}
 
@@
 \item thickness transport form helps expose evolution character
-\item define
+\item define:
 \begin{align*}
@@
     \begin{itemize}
-    \item[$\circ$] note: $s$ and $H$ are equivalent variables for modeling ice geometry
+    \item[$\circ$] note $s$ and $H$ are equivalent variables for modeling ice geometry
     \end{itemize}
-\item<2-> the \alert{mass continuity equation} follows from the surface kinematical equation and incompressibility:
+\item<2-> the \alert{mass continuity equation} follows from the SKE and incompressibility:
    $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$
@@
 \smallskip
-\item<3-> \emph{question:} is it really an advection?
+\item<3-> \emph{question:} is this really an advection equation?
 \item<3->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
@@
 \begin{itemize}
-\item both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
 \item the diffusion schema is only literal in the lubrication approximation
@@
     \end{itemize}
-\item the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences}, regardless of your advective schema preference
+\item but the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences}
+    \begin{itemize}
+    \item[$\circ$] regardless of your advective schema preference
+    \end{itemize}
+\item in any case, both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
 \end{itemize}
@@
 \begin{itemize}
-\item the basic ice sheet model (NCP coupled to Stokes) is \emph{not yet} in common use for ice sheet simulations
+\item the basic ice sheet model (NCP coupled to Stokes) is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
     \begin{itemize}
@@
     \begin{itemize}
-    \item[$\circ$] usual justification: ``Stokes is too expensive at scale''
+    \item[$\circ$] justification: ``Stokes is too expensive at scale''
     \end{itemize}
@@
 \begin{itemize}
-\item in the shallow ice approximation, the surface motion contribution $\Phi(s) = - \bu|_s \cdot \bn_s$ has a formula:
+\item in the shallow ice approximation (SIA) version of the NCP,
 {\small
-\begin{gather*}
-s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0 \\
-{\color{FireBrick} \Phi(s) = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)}
-\end{gather*}
+  $$s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0$$
+}
+the surface motion contribution ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$ has a formula:
+  $${\color{FireBrick} \Phi(s)} = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)$$
 
@@
     \end{itemize}
-}
 
@@
     \item<2>[$\circ$] \emph{local} in surface and bed topography
-    \item<2>[$\circ$] existence is known for this problem
+    \item<2>[$\circ$] existence is known for this NCP problem (Jouvet \& Bueler, 2012)
          \begin{itemize}
-         \item[$\vartriangleright$] Jouvet \& Bueler (2012); NCP is written as a \alert{variational inequality} weak form
+         \item[$\vartriangleright$] once written as a \alert{variational inequality} weak form
          \end{itemize}
@@
 \begin{itemize}
-\item the basic ice sheet model (NCP coupled to Stokes) problem actually has a \emph{non}-local surface velocity function $\Phi(s) = - \bu|_s \cdot \bn_s$
+\item from now on, let's try to avoid shallowness approximations
+\item then the basic ice sheet model (NCP coupled to Stokes) problem has a \alert{non-local} surface velocity function ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$
 {\small
@@
 }
-\item \emph{figure:} the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the shallow ice approximation velocity responds only underneath the surface perturbation}
+\item \emph{figure:} the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the SIA velocity responds only underneath the surface perturbation}
 \end{itemize}
@@
 
-\begin{frame}{time-stepping the mass continuity equation}
+\begin{frame}{traditional PDE time-stepping}
 
@@
 \item \emph{caution}: for ice sheet models, $\bU(t,x,y)$ and $D(t,x,y)>0$ vary in space and time, from nonlinear coupling
-\item nonetheless, let us recall some traditional numerical analysis of these equations \dots
+\item nonetheless, let us recall some traditional numerical analysis \dots
 }
@@
     \begin{itemize}
-    \item[$\circ$] conditionally stable, with maximum time step
+    \item[$\circ$] conditionally stable, with CFL maximum time step
         $$\Delta t \le \frac{h}{\max |\bU|} = O(h)$$
@@
 \begin{itemize}
-\item current researchers mostly believe the advection schema
+\item most researchers believe the advection schema
 \item current-technology models mostly use explicit time stepping
     \begin{itemize}
-    \item[$\circ$] embarrassing: \alert{the mathematical problem is a DAE}
-    \item[$\circ$] time step size is determined by the coupled solution velocity $\bU$,
-    \item[$\circ$] but there are performance and design consequences of hidden DAE \& diffusive character
+    \item[$\circ$] embarrassing: the mathematical problem is a DAE
+    \item[$\circ$] time step size is determined by the coupled solution velocity $\bU$
+         \begin{itemize}
+         \item[$\vartriangleright$] CFL
          \end{itemize}
+    \end{itemize}
+\item however, the performance/usability/accuracy consequences of suppressed free-boundary/DAE/diffusive character are hard to sweep under the rug
+\end{itemize}
+\end{frame}
 
-\medskip
-\item<2> \alert{implicit time-stepping} is expected for DAE problems,
-\item<2> but a sequence of difficult free-boundary problems must be solved at each time step: \alert{NCP coupled to Stokes at each time step}
+\begin{frame}{time-stepping in future ice sheet models}
+
 \begin{itemize}
-    \item[$\circ$] not really ``equations'' any more because of $s\ge b$ inequality
+\item \alert{implicit time-stepping} should be used for DAE problems
+\item a sequence of difficult free-boundary problems must be solved at each time step: \alert{NCP coupled to Stokes}
+    \begin{itemize}
+    \item[$\circ$] not really ``equations'', because of $s\ge b$ inequality
     \end{itemize}
@@
 
-\section{stress balance solver scaling}
+\section{stress-balance solver scaling}
 
@@
 \begin{itemize}
-\item assume an ice sheet model which is \emph{not} shallow
-\item whether explicit or implicit, solves a Stokes problem at each step:
+\item whether explicit or implicit, a non-shallow model solves a Stokes problem at each step:
 \begin{align*}
@@
 \end{align*}
-\item this problem is the \alert{stress balance} (conservation of momentum) which determines velocity $\bu$ and pressure $p$
+\item this is the \alert{stress balance} (conservation of momentum) problem which determines velocity $\bu$ and pressure $p$
 \item how fast is the numerical solution process?
+    \begin{itemize}
+    \item[$\circ$] how do solution algorithms \alert{scale with increasing resolution}?
+    \end{itemize}
 \end{itemize}
@@
 
-\begin{frame}{PDE solver algorithmic scaling}
+\begin{frame}{summary: PDE solver algorithmic scaling}
 
 \begin{itemize}
-\item as intro to scaling, consider the 3D Poisson equation:
+\item consider the 2D Poisson equation:
     $$-\grad^2 u = f \text{ in $\Omega$}, \quad u=0 \text{ on $\partial\Omega$}$$
-\item discretization generates a linear system $A\bu = \bb$
-\item (data size $m$) = \alert{number of unknowns}
+\item discretization generates a linear system $A\,\bu = \bb$
+\item data size $m$ = (number of unknowns)
     \begin{itemize}
     \item[$\circ$] for low-order discretizations, $m$ = (number of nodes in the grid)
-    \item[$\circ$] for cell size $h$, $m \sim h^{-3}$ in 3D
+    \item[$\circ$] $m$ scales with mesh cell diameter $h$: \quad $m \sim h^{-2}$ in 2D
     \end{itemize}
-\item solver choice determines asymptotic growth of computational expense (complexity)
-\item \alert{algorithmic scaling} of flops as $m\to\infty$:
+\item \alert{complexity} or \alert{algorithmic scaling} of flops, as $m\to\infty$, depends on solver algorithm:
     \begin{itemize}
@@
 
-\begin{frame}{scalable stress balance solvers for ice sheets}
+\begin{frame}{ice sheet stress-balance solver complexity}
 
 \begin{itemize}
-\item ice sheet stress balance solvers can be analyzed similarly
 \item $m$ = (number of velocity and pressure unknowns)
-    \begin{itemize}
-    \item[$\circ$] indicate scaling by ``$O(m^{1+\alpha})$'', with $\alpha=0$ optimal
-    \end{itemize}
-\item \alert{near-optimal solvers} already exist: \hfill \emph{good news!}
+\item I'll parameterize scaling as $O(m^{1+\alpha})$, with $\alpha=0$ optimal
+\item \alert{near-optimal solvers} already exist: \hfill $\leftarrow$ \emph{good news!}
     \begin{itemize}
@@
         \begin{itemize}
-        \item[$\vartriangleright$] unstructured quadrilateral/tetrahedral mesh, $Q_k\times Q_{k-2}$ stable elements, Schur-preconditioned Newton-Krylov, column-oriented algebraic multigrid (AMG) preconditioner for $(\bu,\bu)$ block
+        \item[$\vartriangleright$] unstructured quadrilateral/tetrahedral mesh, $Q_k\times Q_{k-2}$ stable elements, Schur-preconditioned Newton-Krylov, ice-column-oriented algebraic multigrid (AMG) preconditioner for $(\bu,\bu)$ block
         \end{itemize}
@@
     \item[$\circ$] $\alpha=0.05$ for Tuminaro et al~(2016) 1st-order (shallow) AMG solver
-    \item[$\circ$] similar for Brown et al~(2013) GMG solver
+    \item[$\circ$] similar for Brown et al~(2013) 1st-order GMG solver
     \end{itemize}
@@
 \section{ice sheet model performance analysis}
+
+\begin{frame}{the analysis set-up: definitions and relationships}
+
+\begin{itemize}
+\item ice sheets are thin layers, thus ice sheet models often have $O(1)$ mesh points in the vertical direction
+    \begin{itemize}
+    \item[$\circ$] e.g.~Issac et al (2015) Stokes solver
+    \item[$\circ$] \emph{message}: I am ignoring refinement in the vertical
+    \end{itemize}
+
+\medskip
+\item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
+\item relationships for map-plane domain $\Omega \subset \RR^2$ of width $L$ and cells of diameter $h$:
+  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}} \hspace{20mm}$$
+
+\vspace{-18mm}
+\hfill \includegraphics[width=0.15\textwidth]{images/one-grid.png} \hspace{5mm}
+
+\item explicit time-stepping schemata:
+{\footnotesize
+\begin{align*}
+\text{advective} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a & \Delta t &\le \frac{h}{U} \\
+\text{diffusion} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a & \Delta t &\le \frac{h^2}{D}
+\end{align*}
+}
+
+\item stress-balance solver scaling parameterized as $O(m^{1+\alpha})$
+\end{itemize}
+\end{frame}
 
@@
     \end{itemize}
-\item assume map-plane domain $\Omega \subset \RR^2$ of width $L$
-\item assume map-plane mesh cells have characteristic diameter $h$
-\item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
-\item the question addressed by upcoming performance table:
+
+\bigskip
+\item the question:
 
@@
 \end{center}
-\end{itemize}
-\end{frame}
-
-\begin{frame}{preparation: observe and recall}
-
-\begin{itemize}
-\item ice sheets are thin layers, thus even non-shallow Stokes models often have $O(1)$ mesh points in the vertical direction
-    \begin{itemize}
-    \item[$\circ$] e.g.~Issac et al (2015) Stokes solver
-    \item[$\circ$] \emph{message}: I am ignoring refinement in the vertical
-    \end{itemize}
-
-\medskip
-\item scale relationships:
-  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}} \hspace{20mm}$$
-
-\vspace{-18mm}
-\hfill \includegraphics[width=0.15\textwidth]{images/one-grid.png} \hspace{5mm}
-
-\medskip
-\item mass-continuity equation time-stepping schemata:
-{\footnotesize
-\begin{align*}
-\text{advective} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a & \Delta t &\le \frac{h}{U} \\
-\text{diffusion} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a & \Delta t &\le \frac{h^2}{D}
-\end{align*}
-}
-
-\item stress-balance solver scaling parameterized as $O(m^{1+\alpha})$
 \end{itemize}
@@
 \end{tabular}
-
-\bigskip\bigskip
-\begin{itemize}
-\item but what about implicit?
-\end{itemize}
 \end{frame}
@@
 \begin{itemize}
-\item main point so far: \alert{explicit} time-stepping \alert{implies many stress-balance solves}
+\item thus \alert{explicit} time-stepping \alert{implies many stress-balance solves}
     \begin{itemize}
-    \item[$\circ$] yes, stress balance solver scaling ($\alpha$) is important
-    \item[$\circ$] however, even optimality ($\alpha=0$) cannot fix performance
+    \item[$\circ$] stress-balance solver scaling $\alpha$ is important, but even optimality ($\alpha=0$) cannot fix performance
     \end{itemize}
@@
     \begin{itemize}
-    \item[$\circ$] each step is a NCP-coupled-to-Stokes problem
-        \begin{itemize}
-        \item[$\vartriangleright$] a \alert{free-boundary problem}
-        \item[$\vartriangleright$] table on next slide parameterizes cost of coupled implicit-step solves as $O(m^{1+\beta})$; optimal is $\beta=0$
-        \end{itemize}
     \item[$\circ$] need $q$ model updates per year to integrate climate influences
-    \end{itemize}
-
-\bigskip
-\item<2> \alert{existing} implicit time-stepping ice sheet models?
-    \begin{itemize}
-    \item[$\circ$] no fully-implicit NCP-coupled-to-Stokes models exist (?)
-    \item[$\circ$] Bueler (2016) SIA NCP solver scales badly ($\beta=0.8$)
+    \item[$\circ$] each step is a \alert{free-boundary} NCP-coupled-to-Stokes problem
+    \item[$\circ$] parameterize cost of these solves as $O(m^{1+\beta})$
     \end{itemize}
@@
 \end{tabular}
+
+\bigskip
+\begin{itemize}
+\item to really reduce power on $m$, use \alert{implicit} time-stepping and build a \alert{near-optimal ($\beta \approx 0$) NCP-coupled-to-Stokes solver}
+\end{itemize}
 \end{frame}
 
-\section{scalable approaches: candidates}
+\begin{frame}{existing implicit models?}
 
-\begin{frame}{scalable approach 1: machine learning}
+\begin{itemize}
+\item no convincing NCP-coupled-to-Stokes (free-boundary) solvers exist yet
+    \begin{itemize}
+    \item[$\circ$] Wirbel \& Jarosch~(2020) is an important attempt \dots
+    \end{itemize}
+\item the Bueler (2016) free-boundary solver uses the SIA
+    \begin{itemize}
+    \item[$\circ$] and it scales badly: $\beta=0.8$
+    \end{itemize}
+\end{itemize}
+\end{frame}
+
+\section{3 approaches to better performance}
+
+\begin{frame}{approach 1: machine learning}
 
@@
 \item apply \alert{machine learning}
-\item \emph{concept}: apply non-scalable ice sheet models on many hypothetical/real ice sheets, and train ML \alert{emulator} on results
+\item run non-scalable ice sheet models on many hypothetical/real ice sheets, and train ML \alert{emulator} on results
     \begin{itemize}
@@
 
-\begin{frame}{scalable approach 2: smarter explicit time-stepping}
+\begin{frame}{approach 2: semi-coupled time-stepping}
 
-semi-coupled
+\begin{itemize}
+\item idea from L{\"o}fgren et al.~(2022)
+    \begin{itemize}
+    \item[$\circ$] earlier use in mantle/crust simulations (Kaus et al.~2010)
+    \end{itemize}
+\item \emph{idea.} keep time-splitting (de-coupled Stokes, SKE solves) but \alert{modify the Stokes problem to see the updated surface}
+\item that is, modify the Stokes weak form \onslide<2>{by adding body force terms corresponding to the updated-surface icy domain}
 
-FIXME L{\"o}fgren et al.~2022
+\vspace{-3mm}
+\begin{minipage}[t][30mm][t]{0.8\textwidth}
+\only<1>{
+{\small
+\begin{align*}
+\int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv &= \int_{\Lambda^\ell} \bbf\cdot \bv \\
+\int_{\Lambda^\ell} q \Div \bu &= 0
+\end{align*}
+}
+}
+\only<2>{
+{\small
+\begin{align*}
+\int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv \qquad\qquad &= \int_{\Lambda^\ell} \bbf\cdot \bv + \Delta t \int_{\Gamma_s^\ell} a\, (\bbf\cdot \bv)\,dx\\
+\hspace{20mm} - \Delta t \int_{\Gamma_s^\ell} (\bu \cdot \bn_s^\ell) (\bbf\cdot \bv)\,dx \qquad
+&\phantom{=} \\
+\int_{\Lambda^\ell} q \Div \bu &= 0
+\end{align*}
+}
+}
+\end{minipage}
+
+\bigskip
+\item<2> early experiments suggest $\sim 10$ times longer stable time steps
+\end{itemize}
+
 \end{frame}
 
-\begin{frame}{scalable approach 3: multigrid implicit solvers}
+\begin{frame}{approach 3: multilevel NCP-coupled-to-Stokes solvers}
 
-FIXME chapter 9 from my book
-\end{frame}
-
-\begin{frame}{scalable approach 3: multigrid implicit solvers}
-
-FIXME multilevel multigrid approach
+\begin{itemize}
+\item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs), but in the \alert{non-local residual case}
+    \begin{itemize}
+    \item[$\circ$] this seems not to exist
+    \item[$\circ$] the \alert{smoother} must compute an implicit-step SKE residual including surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ from a scalable Stokes solver
+    \end{itemize}
+\item near-optimal multilevel VI solvers exist for toy problems
+    \begin{itemize}
+    \item[$\circ$] Poisson equation obstacle problem (Brandt \& Cryer 1983; Gr\"aeser \& Kornhuber 2009)
+    \end{itemize}
+\item<2> \alert{help me}
+\end{itemize}
 \end{frame}
@@
    \item[1.] machine learning emulators (Jouvet et al.~2021)
-   \item[2.] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
+   \item[2.] semi-coupled time stepping (L{\"o}fgren et al.~2022)
    \item[3.] scalable Stokes solvers (Isaac et al.~2015)   
    \end{enumerate}
-\item<4-> scalable solvers for implicit-step models require advances in \alert{multilevel solutions of non-local variational inequalities}
+\item<4-> scalable solvers for implicit-step NCP-coupled-to-Stokes models require advances in \alert{multilevel solutions of non-local variational inequalities}
 \end{itemize}

### 2023-01-24 talk/slides.tex
making small things better

@@
 \item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
-\item relationships for map-plane domain $\Omega \subset \RR^2$ of width $L$ and cells of diameter $h$:
-  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}} \hspace{20mm}$$
+\item for map-plane domain $\Omega \subset \RR^2$ of width $L$ and cells of diameter $h$:
+  $$m \sim \frac{L^2}{h^2} \hspace{20mm}$$
 
-\vspace{-18mm}
-\hfill \includegraphics[width=0.15\textwidth]{images/one-grid.png} \hspace{5mm}
+\vspace{-14mm}
+\hfill
+\begin{tikzpicture}[scale=1.5]
+  \pgfmathsetmacro\fourth{1.0/4.0}
+  \draw[xstep=\fourth,ystep=\fourth,black,thin] (0.0,0.0) grid (1.0,1.0);
+  \draw[<->] (-0.2,0.0) -- (-0.2,1.0);
+  \draw[<->] (1.2,0.76) -- (1.2,0.99);
+  \node at (-0.35,0.5) {$L$};
+  \node at (1.4,0.86) {$h$};
+\end{tikzpicture}
 
+\vspace{-2mm}
 \item explicit time-stepping schemata:
@@
 \end{tabular}
+
+\bigskip
+\begin{itemize}
+\item \alert{explicit} time-stepping \alert{implies many stress-balance solves}
+    \begin{itemize}
+    \item[$\circ$] stress-balance solver scaling $\alpha$ is important, but even optimality ($\alpha=0$) cannot fix performance
+    \end{itemize}
+\end{itemize}
 \end{frame}
@@
 \begin{itemize}
-\item thus \alert{explicit} time-stepping \alert{implies many stress-balance solves}
-    \begin{itemize}
-    \item[$\circ$] stress-balance solver scaling $\alpha$ is important, but even optimality ($\alpha=0$) cannot fix performance
-    \end{itemize}
 \item switch to \alert{implicit time-stepping} for unconditional stability?
     \begin{itemize}
-    \item[$\circ$] need $q$ model updates per year to integrate climate influences
     \item[$\circ$] each step is a \alert{free-boundary} NCP-coupled-to-Stokes problem
@@
     \end{itemize}
+\item nonetheless need $q$ model updates per year to integrate climate influences
 \end{itemize}
@@
     \end{itemize}
-\item<2> \alert{help me}
+\item<2> join me on this quest?
 \end{itemize}

### 2023-02-10 talk/slides.tex
sweep through and simplify/clarify/shorten

@@
 \newcommand{\bn}{\mathbf{n}}
+\newcommand{\bq}{\mathbf{q}}
 \newcommand{\br}{\mathbf{r}}
@@
     \item[$\circ$] also a fixed domain,
-    \item[$\circ$] but $s=b$ where there is no ice!
+    \item[$\circ$] but $s=b$ where there is no ice
     \end{itemize}
@@
     \begin{itemize}
-    \item[$\circ$] \emph{in} the icy domain $\Lambda(t) \subset \RR^3$:
+    \item[$\circ$] in the icy domain $\Lambda(t) \subset \RR^3$:
 \begin{align*}
 \text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
-    \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
+    \intertext{\item[$\circ$] and on the ice surfaces:}
 \text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
@@
 \bigskip
-\item<2> {\footnotesize for more on this perspective see Bueler (2021)}
+\item {\footnotesize for more on this perspective see Bueler (2021)}
 \end{itemize}
@@
 \begin{itemize}
-\item \only<1>{\alert{nonlinear complementarity problem} (NCP)} \only<2>{nonlinear complementarity problem (NCP) coupled to Stokes}:
+\item \only<1>{\alert{nonlinear complementarity problem} (NCP)} \only<2>{nonlinear complementarity problem (NCP) coupled to \alert{Stokes}}:
 \begin{align*}
@@
 \item for this slide, forget complementarity and boundary conditions
-\item result is SKE coupled to Stokes:
+\item result: SKE coupled to Stokes
 \begin{align*}
@@
 
-\item<2-> only the first of these 5 equations has a time derivative
+\item only the first of these 5 equations has a time derivative
     \begin{itemize}
-    \item[$\circ$] because ice is slow and incompressible
+    \item[$\circ$] because ice is very viscous and incompressible
     \end{itemize}
-\item<3> the problem is a \alert{differential algebraic equation} (DAE), an extremely \alert{stiff}, system:
+\item<2> this time-dependent problem is a \alert{differential algebraic equation} (DAE), an extremely stiff system:
 \begin{align*}
@@
     \begin{itemize}
-    \item<3>[$\circ$] but in $\infty$ dimensions (PDAE?) and subject to complementarity
+    \item<2>[$\circ$] but in $\infty$ dimensions (PDAE?), and subject to complementarity
     \end{itemize}
@@
 \begin{itemize}
-\item to the best of my knowledge, \emph{no} current research groups are studying the well-posedness or regularity of this basic ice sheet model
+\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
     \begin{itemize}
-    \item[$\circ$] but most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model?
-    \item[$\circ$] well-posedness of shallow approximations \emph{has} been considered
+    \item[$\circ$] when pressed, most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model?
+    \item[$\circ$] well-posedness of the lubrication approximation of the model has been considered; existence proved in (Jouvet \& Bueler 2012)
     \end{itemize}
@@
 \medskip
-\item<2> numerical modelers tend to think of the Stokes problem separately from surface evolution
+\item numerical modelers tend to think of the Stokes problem separately from surface evolution
     \begin{itemize}
-    \item[$\circ$] \emph{time-splitting} or \emph{explicit time-stepping} is taken for granted
+    \item[$\circ$] \emph{time-splitting} or \emph{explicit time-stepping} is often taken for granted
     \end{itemize}
-\item<2> ice sheet geometry evolution is addressed with minimal awareness of complementarity
+
+\medskip
+\item ice sheet geometry evolution is addressed with minimal awareness of complementarity
+
+\medskip
+\item NCP-coupled-to-Stokes is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
+    \begin{itemize}
+    \item[$\circ$] \alert{can we make it fast enough to use?}
+    \end{itemize}
 \end{itemize}
@@
 \begin{itemize}
-\item thickness transport form helps expose evolution character
+\item thickness transport form helps for evolution or stability questions
 \item define:
@@
     \end{itemize}
-\item<2-> the \alert{mass continuity equation} follows from the SKE and incompressibility:
+\item the \alert{mass continuity equation} for thickness follows from SKE and incompressibility:
    $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$
@@
 \smallskip
-\item<3-> \emph{question:} is this really an advection equation?
-\item<3->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
-  $$\bU \sim - \grad s \sim - \grad H,$$
-and ice is slow (DAE) so there are no characteristics
+\item<2> \emph{question:} is this really an advection equation?
+\item<2>[] \emph{answer:} not really \dots ice flows (mostly) downhill so
+  $$\bU \sim - \grad s \sim - \grad H$$
+\item<2> NCP-coupled-to-Stokes DAE system has no characteristic curves
 \end{itemize}
@@
 \begin{itemize}
-\item the diffusion schema is only literal in the lubrication approximation
+\item the diffusion schema is literal in the lubrication approximation
     \begin{itemize}
@@
     \end{itemize}
-\item but the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences}
+\item but the fact that ice flows downhill has \emph{time-stepping stability} consequences
     \begin{itemize}
-    \item[$\circ$] regardless of your advective schema preference
+    \item[$\circ$] regardless of your preference for the advective schema!
     \end{itemize}
-\item in any case, both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
-\end{itemize}
-\end{frame}
-
-\begin{frame}{shallow ice sheet models}
-
-\begin{itemize}
-\item the basic ice sheet model (NCP coupled to Stokes) is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
-    \begin{itemize}
-    \item[$\circ$] \alert{can we make it fast enough to use?}
-    \end{itemize}
-
-\smallskip
-\item current-technology models use various shallow simplifications
-    \begin{itemize}
-    \item[$\circ$] justification: ``Stokes is too expensive at scale''
-    \end{itemize}
-\item the simplest of several shallow approximations is the ``lubrication'' approximation, the \alert{shallow ice approximation} (SIA)
-
-\smallskip
+\item note both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
 \end{itemize}
@@
 \begin{itemize}
-\item in the shallow ice approximation (SIA) version of the NCP,
+\item the simplest of several shallow approximations is the ``lubrication'' approximation, the \alert{shallow ice approximation} (SIA)
+\item SIA version of the NCP:
 {\small
@@
 \medskip
-\item<2> $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
+\item $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
     \begin{itemize}
-    \item<2>[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
-    \item<2>[$\circ$] \emph{local} in surface and bed topography
-    \item<2>[$\circ$] existence is known for this NCP problem (Jouvet \& Bueler, 2012)
-         \begin{itemize}
-         \item[$\vartriangleright$] once written as a \alert{variational inequality} weak form
-         \end{itemize}
+    \item[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
+    \item[$\circ$] \emph{local} in surface and bed topography
+    \item[$\circ$] existence is known for this NCP problem (Jouvet \& Bueler, 2012), when written as a \alert{variational inequality} weak form
     \end{itemize}
@@
 \begin{itemize}
-\item from now on, let's try to avoid shallowness approximations
+\item from now on, let us avoid shallowness approximations
 \item then the basic ice sheet model (NCP coupled to Stokes) problem has a \alert{non-local} surface velocity function ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$
@@
 \only<1>{
-\item \emph{caution}: for ice sheet models, $\bU(t,x,y)$ and $D(t,x,y)>0$ vary in space and time, from nonlinear coupling
-\item nonetheless, let us recall some traditional numerical analysis \dots
+\item let us recall some traditional numerical analysis
 }
@@
 \item for example, forward Euler using spacing $h$ and time step $\Delta t$:
-	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{(\bU H)_{j+1/2}^\ell - (\bU H)_{j-1/2}^\ell}{h} = a_j^\ell$$
+	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{\bq_{j+1/2}^\ell - \bq_{j-1/2}^\ell}{h} = a_j^\ell$$
 
     \begin{itemize}
+    \item[$\circ$] need good approximations of flux $\bq=\bU H$: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
     \item[$\circ$] conditionally stable, with CFL maximum time step
         $$\Delta t \le \frac{h}{\max |\bU|} = O(h)$$
-    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
     \end{itemize}
@@
 \only<4>{
-\item \alert{implicit} time stepping for \alert{diffusions} is recommended
+\item \alert{implicit} time stepping for \alert{diffusions} is often recommended
 \item for example, backward Euler:
@@
     \item[$\circ$] unconditionally stable, but must solve equations at each step
-    \item[$\circ$] alternative implicit schemes: Crank-Nicolson, BDF, \dots
+    \item[$\circ$] further implicit schemes: Crank-Nicolson, BDF, \dots
     \end{itemize}
@@
 
-\begin{frame}{time-stepping in current ice sheet models}
+\begin{frame}{time-stepping in current and future ice sheet models}
 
 \begin{itemize}
+\item current-technology large-scale models use explicit time stepping
+    \begin{itemize}
+    \item[$\circ$] this is embarrassing: the mathematical problem is a DAE
+    \item[$\circ$] the accuracy/performance/usability consequences of the suppressed free-boundary/DAE/diffusive character are hard to sweep under the rug
+    \end{itemize}
 \item most researchers believe the advection schema
-\item current-technology models mostly use explicit time stepping
     \begin{itemize}
-    \item[$\circ$] embarrassing: the mathematical problem is a DAE
-    \item[$\circ$] time step size is determined by the coupled solution velocity $\bU$
-         \begin{itemize}
-         \item[$\vartriangleright$] CFL
+    \item[$\circ$] time step is determined by CFL using coupled solution velocity $\bU$
     \end{itemize}
-    \end{itemize}
-\item however, the performance/usability/accuracy consequences of suppressed free-boundary/DAE/diffusive character are hard to sweep under the rug
-\end{itemize}
-\end{frame}
 
-\begin{frame}{time-stepping in future ice sheet models}
-
-\begin{itemize}
-\item \alert{implicit time-stepping} should be used for DAE problems
-\item a sequence of difficult free-boundary problems must be solved at each time step: \alert{NCP coupled to Stokes}
-    \begin{itemize}
-    \item[$\circ$] not really ``equations'', because of $s\ge b$ inequality
-    \end{itemize}
+\medskip
+\item \alert{implicit time-stepping} is appropriate for DAE problems
+\item a sequence of NCP-coupled-to-Stokes free-boundary problems must be solved at each time step
 \end{itemize}
@@
 \begin{itemize}
-\item whether explicit or implicit, a non-shallow model solves a Stokes problem at each step:
+\item a non-shallow model solves a Stokes problem at each step:
 \begin{align*}
-- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda$} \\
+- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda \subset \RR^3$} \\
 \nabla \cdot \bu &= 0 && \text{''} \\
@@
     \begin{itemize}
-    \item[$\circ$] how do solution algorithms \alert{scale with increasing resolution}?
+    \item[$\circ$] how do solution algorithms \alert{scale} with increasing spatial resolution?
     \end{itemize}
@@
 \begin{itemize}
-\item consider the 2D Poisson equation:
-    $$-\grad^2 u = f \text{ in $\Omega$}, \quad u=0 \text{ on $\partial\Omega$}$$
-\item discretization generates a linear system $A\,\bu = \bb$
-\item data size $m$ = (number of unknowns)
+\item for example, consider the 2D Poisson equation:
+    $$-\grad^2 u = f \,\text{ in $\Omega$}, \qquad u=0 \,\text{ on $\partial\Omega$}$$
+\item discretization generates a linear system $A\,\bu = \bb$ with $\bu\in\RR^m$
+\item data size $m$ is the number of unknowns
     \begin{itemize}
-    \item[$\circ$] for low-order discretizations, $m$ = (number of nodes in the grid)
+    \item[$\circ$] for low-order discretizations, $m$ = \#(nodes in the grid)
     \item[$\circ$] $m$ scales with mesh cell diameter $h$: \quad $m \sim h^{-2}$ in 2D
@@
     \item[$\circ$] $\approx O(m^2)$ for sparsity-exploiting direct linear algebra
-    \item[$\circ$] $O(m^1)$, \alert{optimal}, for \alert{multigrid} solvers (below)
+    \item[$\circ$] $O(m^1)$, \alert{optimal}, e.g.~for \alert{multigrid} solvers (below)
     \end{itemize}
@@
 \begin{itemize}
-\item $m$ = (number of velocity and pressure unknowns)
-\item I'll parameterize scaling as $O(m^{1+\alpha})$, with $\alpha=0$ optimal
+\item Stokes: \quad $m$ = \#(velocity and pressure unknowns)
+\item model the scaling as $O(m^{1+\alpha})$, with $\alpha=0$ optimal
 \item \alert{near-optimal solvers} already exist: \hfill $\leftarrow$ \emph{good news!}
@@
     \item[$\circ$] $\alpha=0.05$ for Tuminaro et al~(2016) 1st-order (shallow) AMG solver
-    \item[$\circ$] similar for Brown et al~(2013) 1st-order GMG solver
+    \item[$\circ$] similar for Brown et al~(2013) 1st-order (shallow) GMG solver
     \end{itemize}
@@
 
-\begin{frame}{the analysis set-up: definitions and relationships}
+\begin{frame}{the analysis set-up}
 
@@
     \item[$\circ$] e.g.~Issac et al (2015) Stokes solver
-    \item[$\circ$] \emph{message}: I am ignoring refinement in the vertical
+    \item[$\circ$] \emph{simple message}: I am ignoring refinement in the vertical
     \end{itemize}
@@
 \medskip
-\item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
+\item let $m$ = \#(surface elevation \& velocity \& pressure unknowns)
 \item for map-plane domain $\Omega \subset \RR^2$ of width $L$ and cells of diameter $h$:
@@
 
-\begin{frame}{simplified analysis of ice sheet model performance}
+\begin{frame}{the simplified ice sheet model performance question}
 
 \begin{itemize}
-\item for a time-stepping model our metric is \alert{flops per model year}
-    \begin{itemize}
-    \item[$\circ$] glaciologists want high-resolution simulations of ice sheets over 100k year ice age cycles
-    \end{itemize}
+\item glaciologists want to run time-stepping high-resolution simulations of ice sheets over e.g.~$10^5$ year ice age cycles
+
+\bigskip
+\item proposed metric: \quad \alert{flops per model year}
 
@@
 \begin{minipage}{0.82\textwidth}
-how does flops per model year \alert{scale} in the \alert{high spatial resolution limit} $h\to 0$, equivalently $m\to \infty$?
+how does this metric \alert{scale} in the \alert{high spatial resolution limit} $h\to 0$, equivalently $m\to \infty$?
 \end{minipage}
@@
     \begin{itemize}
-    \item[$\circ$] stress-balance solver scaling $\alpha$ is important, but even optimality ($\alpha=0$) cannot fix performance
+    \item[$\circ$] stress-balance solver scaling $\alpha$ is important,
+    \item[$\circ$] but even optimality ($\alpha=0$) cannot rescue performance
     \end{itemize}

### 2023-02-10 talk/slides.tex
I think this is good enough

@@
 \end{center}
+
+\bigskip
+\item the goal: \hspace{20mm} $O(h^{-2}) = O(m^1)$
 \end{itemize}
@@
 
-\begin{frame}{\emph{explicit} ice sheet model performance}
+\begin{frame}{explicit ice sheet model performance}
 
@@
 \\
-explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
+explicit ({\footnotesize \emph{advective}}) & Stokes \phantom{xxxx} & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
 \\
@@
 
-\bigskip
+\vspace{10mm}
 \begin{itemize}
-\item \alert{explicit} time-stepping \alert{implies many stress-balance solves}
+\item explicit time-stepping implies \alert{many stress-balance solves}
     \begin{itemize}
-    \item[$\circ$] stress-balance solver scaling $\alpha$ is important,
-    \item[$\circ$] but even optimality ($\alpha=0$) cannot rescue performance
+    \item[$\circ$] while stress-balance scaling exponent $\alpha$ is important, even optimality ($\alpha=0$) cannot rescue performance
     \end{itemize}
@@
     \end{itemize}
-\item nonetheless need $q$ model updates per year to integrate climate influences
+\item need $q$ model updates per year to integrate climate influences
 \end{itemize}
@@
 \\
-implicit & \emph{either} & $\oo{\frac{q\, L^{2+2\beta}}{h^{\,2+2\beta}}} = \oo{q\, m^{1+\beta}}$
+implicit & & $\oo{\frac{q\, L^{2+2\beta}}{h^{\,2+2\beta}}} = \oo{q\, m^{1+\beta}}$
 \end{tabular}
@@
 \begin{itemize}
-\item to really reduce power on $m$, use \alert{implicit} time-stepping and build a \alert{near-optimal ($\beta \approx 0$) NCP-coupled-to-Stokes solver}
+\item goal for optimists: implicit time-stepping \emph{and} build a $\beta \approx 0$ NCP-coupled-to-Stokes solver for the time step problems
 \end{itemize}
@@
     \end{itemize}
-\item the Bueler (2016) free-boundary solver uses the SIA
-    \begin{itemize}
-    \item[$\circ$] and it scales badly: $\beta=0.8$
-    \end{itemize}
+\item the Bueler (2016) implicit (free-boundary) SIA solver scales badly: $\beta=0.8$
 \end{itemize}
@@
     \end{itemize}
-\item \emph{idea.} keep time-splitting (de-coupled Stokes, SKE solves) but \alert{modify the Stokes problem to see the updated surface}
-\item that is, modify the Stokes weak form \onslide<2>{by adding body force terms corresponding to the updated-surface icy domain}
+\item \emph{idea.} remain explicit, but \alert{modify the Stokes problem to ``see'' the updated (extrapolated) surface}
+\item that is, modify the Stokes problem at time $t^\ell$ \onslide<2>{by adding body force terms corresponding to the updated-surface icy domain}
 
@@
 \int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv &= \int_{\Lambda^\ell} \bbf\cdot \bv \\
+\phantom{x} \\
+\phantom{x} \\
 \int_{\Lambda^\ell} q \Div \bu &= 0
@@
 \int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv \qquad\qquad &= \int_{\Lambda^\ell} \bbf\cdot \bv + \Delta t \int_{\Gamma_s^\ell} a\, (\bbf\cdot \bv)\,dx\\
-\hspace{20mm} - \Delta t \int_{\Gamma_s^\ell} (\bu \cdot \bn_s^\ell) (\bbf\cdot \bv)\,dx \qquad
+\hspace{20mm} - \Delta t \int_{\Gamma_s^\ell} (\bu \cdot \bn_{s^\ell}) (\bbf\cdot \bv)\,dx \qquad
 &\phantom{=} \\
@@
 \begin{itemize}
-\item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs), but in the \alert{non-local residual case}
+\item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs)
+\item but in the \alert{non-local residual case} \hfill $\gets$ \emph{yesterday's seminar}
     \begin{itemize}
     \item[$\circ$] this seems not to exist
-    \item[$\circ$] the \alert{smoother} must compute an implicit-step SKE residual including surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ from a scalable Stokes solver
+    \item[$\circ$] the \alert{smoother} must reduce a residual formed from surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ (from a scalable Stokes solver)
     \end{itemize}
-\item near-optimal multilevel VI solvers exist for toy problems
-    \begin{itemize}
-    \item[$\circ$] Poisson equation obstacle problem (Brandt \& Cryer 1983; Gr\"aeser \& Kornhuber 2009)
-    \end{itemize}
-\item<2> join me on this quest?
+\item near-optimal multilevel solvers exist for simpler VI problems
 \end{itemize}
@@
    \begin{itemize}
-   \item[$\circ$] progress will be made in all of these areas
+   \item[$\circ$] progress is being made in all of these areas
    \end{itemize}
@@
    \end{enumerate}
-\item<4-> scalable solvers for implicit-step NCP-coupled-to-Stokes models require advances in \alert{multilevel solutions of non-local variational inequalities}
+\item<4-> scalable solvers for implicit-step NCP-coupled-to-Stokes models, which would seem to be the recommended numerical design, require \alert{multilevel solvers for non-local variational inequalities}
 \end{itemize}

### 2023-02-10 talk/slides.tex
the talk I will give

@@
 \includegraphics<2>[width=\textwidth]{images/ant-schoofhewitt2013.png}
-\only<2>{\vspace{13mm}}
-\only<2>{\par {\scriptsize note vertical exaggeration and rough bed (Schoof \& Hewitt 2013)}}
+\only<2>{\bigskip \hfill {\tiny (Schoof \& Hewitt 2013)}}
+\only<2>{\vspace{4mm}}
+\only<2>{\begin{center} {\footnotesize note vertical exaggeration, smooth surface, and rough bed} \end{center} }
 \includegraphics<3>[height=0.69\textheight]{images/alps-seguinot2018.png}
-\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum (Seguinot et al 2018)}}
+\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum} \hfill {\tiny (Seguinot et al 2018)}}
 \includegraphics<4>[height=0.69\textheight]{images/british-clark2022.png}
-\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum (Clark et al 2022)}}
+\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum} \hfill {\tiny  (Clark et al 2022)}}
 \includegraphics<5>[height=0.69\textheight]{images/not-sea-ice.png}
-\only<5>{\par {\scriptsize \emph{not} sea ice!}}
+\only<5>{\par {\footnotesize an ice sheet is \emph{not} sea ice!}}
 \end{center}
@@
 \begin{itemize}
-\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
+\item glacier ice is modeled as a \emph{very viscous, incompressible, non-Newtonian fluid}
     \begin{itemize}
@@
     \end{itemize}
-\item glaciers lie on \emph{topography}, and sometimes they float  (\emph{ice shelf})
+\item glaciers lie on \emph{topography}
+    \begin{itemize}
+    \item[$\circ$] sometimes they float  (\emph{ice shelf})
+    \end{itemize}
 \item glacier geometry and velocity evolve \emph{in contact with climate}:
@@
 \begin{itemize}
-\item for simplicity/clarity of the upcoming model performance analysis, I will ignore much of glacier physics
+\item for simplicity/clarity of the upcoming modeling, I will ignore much of glacier physics
 \item \alert{ignoring}:
@@
 \medskip
-\item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
+\item {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
 \end{itemize}
@@
 \begin{itemize}
-\item data $a(t,x,y)$, $b(x,y)$  is defined on a \alert{fixed domain}:
+\item data $a(t,x,y)$, $b(x,y)$ are defined on a \alert{fixed domain}:
 	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
@@
     \end{itemize}
-\item<3> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$, on which (solution) velocity $\bu(t,x,y,z)$ is defined:
+\item<3> $s(t,x,y)$ determines the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$, on which the solution velocity $\bu(t,x,y,z)$ is defined:
     $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$
@@
         \item[$\vartriangleright$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
+        \item[$\vartriangleright$] $\bn_s = \left<-\grad s,1\right>$ is upward surface normal
         \end{itemize}
@@
 (s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
-\onslide<2->{- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
+\onslide<2->{- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
 \nabla \cdot \bu &= 0 && \text{''} \\
@@
 \bu|_b \cdot \bn_b &= 0 && \text{''} \\
-\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t)$}}
+\left(2 \nu(D\bu) D\bu - pI\right) \bn_s &= \bzero && \text{on $\Gamma_s(t)$}}
 \end{align*}
@@
     \item[$\circ$] $\bu|_s=0$ where no ice
-    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
+    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu(D\bu) = \Gamma |D\bu|^{\pp-2}$
     \end{itemize}
@@
 \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
-- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
+- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
 \nabla \cdot \bu &= 0
@@
     \begin{itemize}
-    \item[$\circ$] when pressed, most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model?
+    \item[$\circ$] when pressed, most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model
     \item[$\circ$] well-posedness of the lubrication approximation of the model has been considered; existence proved in (Jouvet \& Bueler 2012)
@@
     \begin{itemize}
+    \item[$\circ$] because it is too slow
     \item[$\circ$] \alert{can we make it fast enough to use?}
@@
 \begin{align*}
-- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda \subset \RR^3$} \\
+- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda \subset \RR^3$} \\
 \nabla \cdot \bu &= 0 && \text{''} \\
@@
 \bu|_b \cdot \bn_b &= 0 && \text{''} \\
-\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
+\left(2 \nu(D\bu) D\bu - pI\right) \bn_s &= \bzero && \text{on $\Gamma_s$}
 \end{align*}

### 2023-05-28 talk-oxford/slides.tex
move images/ to faciliate reuse

--- 2023-02-10 talk/slides.tex
@@
 
-\titlegraphic{\hfill \includegraphics[width=0.15\textwidth]{images/uafbw.png}}
+\titlegraphic{\hfill \includegraphics[width=0.15\textwidth]{../images/uafbw.png}}
 
@@
 {
-  \usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/gray-british-clark2022.png}}
+  \usebackgroundtemplate{\includegraphics[width=\paperwidth]{../images/gray-british-clark2022.png}}
   \begin{frame}
@@
 \begin{center}
-\includegraphics<1>[height=0.69\textheight]{images/ant-pittard2021.png}
+\includegraphics<1>[height=0.69\textheight]{../images/ant-pittard2021.png}
 \only<1>{\par {\scriptsize Antarctic ice sheet}} % (Pittard et al 2021)
 \only<2>{\vspace{17mm}}
-\includegraphics<2>[width=\textwidth]{images/ant-schoofhewitt2013.png}
+\includegraphics<2>[width=\textwidth]{../images/ant-schoofhewitt2013.png}
 \only<2>{\bigskip \hfill {\tiny (Schoof \& Hewitt 2013)}}
@@
 \only<2>{\begin{center} {\footnotesize note vertical exaggeration, smooth surface, and rough bed} \end{center} }
-\includegraphics<3>[height=0.69\textheight]{images/alps-seguinot2018.png}
+\includegraphics<3>[height=0.69\textheight]{../images/alps-seguinot2018.png}
 \only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum} \hfill {\tiny (Seguinot et al 2018)}}
-\includegraphics<4>[height=0.69\textheight]{images/british-clark2022.png}
+\includegraphics<4>[height=0.69\textheight]{../images/british-clark2022.png}
 \only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum} \hfill {\tiny  (Clark et al 2022)}}
-\includegraphics<5>[height=0.69\textheight]{images/not-sea-ice.png}
+\includegraphics<5>[height=0.69\textheight]{../images/not-sea-ice.png}
 \only<5>{\par {\footnotesize an ice sheet is \emph{not} sea ice!}}
@@
 \begin{column}{0.42\textwidth}
-\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
+\hfill \includegraphics[width=0.9\textwidth]{../images/map-glacier-ice.png}
 \end{column}
@@
 \begin{column}{0.42\textwidth}
-\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}
+\hfill \includegraphics[width=0.9\textwidth]{../images/map-velocity.png}
 
@@
 \begin{column}{0.46\textwidth}
-\includegraphics<1>[width=\textwidth]{images/domain-data.png}
+\includegraphics<1>[width=\textwidth]{../images/domain-data.png}
 
-\includegraphics<2>[width=\textwidth]{images/domain-surface.png}
+\includegraphics<2>[width=\textwidth]{../images/domain-surface.png}
 
-\includegraphics<3>[width=\textwidth]{images/domain-velocity.png}
+\includegraphics<3>[width=\textwidth]{../images/domain-velocity.png}
 \end{column}
@@
 \begin{center}
-\includegraphics<1>[width=0.6\textwidth]{images/stokes-greens-arndt.png}
-\includegraphics<2>[width=0.6\textwidth]{images/sia-greens-arndt.png}
+\includegraphics<1>[width=0.6\textwidth]{../images/stokes-greens-arndt.png}
+\includegraphics<2>[width=0.6\textwidth]{../images/sia-greens-arndt.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{images/multigrid-grids.png}
+\includegraphics[width=0.5\textwidth]{../images/multigrid-grids.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{images/isaac-antarctica.png}
+\includegraphics[width=0.6\textwidth]{../images/isaac-antarctica.png}
 \end{center}
@@
 \begin{center}
-\only<1>{\includegraphics[height=0.52\textheight]{images/tiles-jouvet.png}}
-\only<2>{\includegraphics[height=0.57\textheight]{images/cnn-jouvet.png}}
+\only<1>{\includegraphics[height=0.52\textheight]{../images/tiles-jouvet.png}}
+\only<2>{\includegraphics[height=0.57\textheight]{../images/cnn-jouvet.png}}
 \end{center}

### 2023-05-28 talk-iu/slides.tex
serious talk start

--- 2023-05-28 talk-oxford/slides.tex
@@
 
-\title[Making ice sheet models scale properly]{Making ice sheet models \\ scale properly}
+\title{Glacial flows, simulated faster}
 
@@
 
-\date[]{February 2023}
+\date[]{June 2023}
 
@@
 {
-  \usebackgroundtemplate{\includegraphics[width=\paperwidth]{../images/gray-british-clark2022.png}}
+  \usebackgroundtemplate{\includegraphics[height=\paperheight]{../images/fade-moraines-indiana.png}}
   \begin{frame}
@@
 
-\section{what is an ice sheet model?}
+\section{the basics of glaciers and ice sheets}
 
-\begin{frame}{what is an ice sheet?}
-
-\begin{itemize}
-\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width
-\end{itemize}
-
-\bigskip
-\begin{minipage}[t][60cm][t]{\textwidth}
-\begin{center}
-\includegraphics<1>[height=0.69\textheight]{../images/ant-pittard2021.png}
-\only<1>{\par {\scriptsize Antarctic ice sheet}} % (Pittard et al 2021)
-\only<2>{\vspace{17mm}}
-\includegraphics<2>[width=\textwidth]{../images/ant-schoofhewitt2013.png}
-\only<2>{\bigskip \hfill {\tiny (Schoof \& Hewitt 2013)}}
-\only<2>{\vspace{4mm}}
-\only<2>{\begin{center} {\footnotesize note vertical exaggeration, smooth surface, and rough bed} \end{center} }
-\includegraphics<3>[height=0.69\textheight]{../images/alps-seguinot2018.png}
-\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum} \hfill {\tiny (Seguinot et al 2018)}}
-\includegraphics<4>[height=0.69\textheight]{../images/british-clark2022.png}
-\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum} \hfill {\tiny  (Clark et al 2022)}}
-\includegraphics<5>[height=0.69\textheight]{../images/not-sea-ice.png}
-\only<5>{\par {\footnotesize an ice sheet is \emph{not} sea ice!}}
-\end{center}
-\end{minipage}
-\end{frame}
-
-\begin{frame}{basic facts about glaciers}
+\newcommand{\basicglaciers}[1]{
+\begin{frame}{basic facts about glaciers #1}
 
@@
 \begin{itemize}
-\item glacier ice is modeled as a \emph{very viscous, incompressible, non-Newtonian fluid}
+\item glacier ice is a \emph{very viscous, incompressible, non-Newtonian fluid}
     \begin{itemize}
-    \item[$\circ$] more on that soon
+    \item[$\circ$] more soon \dots
     \end{itemize}
@@
     \begin{itemize}
-    \item[$\circ$] sometimes they float  (\emph{ice shelf})
+    \item[$\circ$] except sometimes they float on water (floating tongue or ice shelf)
     \end{itemize}
-\item glacier geometry and velocity evolve \emph{in contact with climate}:
+\item a glacier's geometry (\emph{free surface}), and its velocity, \emph{evolve in contact with the climate}:
     \begin{itemize}
@@
 \end{frame}
+}
+
+\basicglaciers{}
+
+\begin{frame}{glaciers}
+
+\begin{center}
+\only<1>{\includegraphics[height=0.8\textheight]{../images/polaris.png}} \only<2>{\includegraphics[height=0.8\textheight]{../images/taku-truffer2016.jpg}}
+\only<3>{\vspace{-5mm} \mbox{\includegraphics[height=0.8\textheight]{../images/columbiasentinel2.jpg} \includegraphics[height=0.55\textheight]{../images/natgeocolumbia1910.png}}}
+\end{center}
+
+{\scriptsize
+\only<1>{Polaris Glacier \hfill {\tiny (Post and LaChappelle 1971)}}
+\only<2>{Taku Glacier \hfill {\tiny (M.~Truffer 2016)}}
+\only<3>{Columbia Glacier \hfill {\tiny (Sentinel-2B 2018, National Geographic 1910)}}
+}
+\end{frame}
+
+\begin{frame}{what is an ice sheet?}
+
+\begin{itemize}
+\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width
+\end{itemize}
+
+\bigskip
+\begin{minipage}[t][60cm][t]{\textwidth}
+\begin{center}
+\includegraphics<1>[height=0.69\textheight]{../images/ant-pittard2021.png}
+\only<1>{\par {\scriptsize Antarctic ice sheet} \hfill {\tiny (Pittard et al 2021)}}
+\only<2>{\vspace{17mm}}
+\includegraphics<2>[width=\textwidth]{../images/ant-schoofhewitt2013.png}
+\only<2>{\vspace{4mm}\par {\scriptsize note smooth surface and rough bed \dots and vertical exaggeration} \hfill {\tiny (Schoof \& Hewitt 2013)}}
+\includegraphics<3>[height=0.69\textheight]{../images/alps-seguinot2018.png}
+\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum} \hfill {\tiny (Seguinot et al 2018)}}
+\includegraphics<4>[height=0.69\textheight]{../images/moraines-indiana.jpg}
+\only<4>{\par {\scriptsize moraines in Illinois, Indiana, Ohio} \hfill {\tiny (Larsen 1986 and other sources)}}
+\includegraphics<5>[height=0.69\textheight]{../images/laurentide-margold2018.png}
+\only<5>{\par {\scriptsize Laurentide ice sheet} \hfill {\tiny (Margold, Stokes, Clark 2018)}}
+\end{center}
+\end{minipage}
+\end{frame}
+
+\begin{frame}{an ice sheet is \emph{not} sea ice!}
+\begin{center}
+\includegraphics[height=0.8\textheight]{../images/not-sea-ice.png}
+\end{center}
+\end{frame}
+
+\basicglaciers{\dots again}
 
@@
 
-\begin{frame}{what is an ice sheet model?}
+\begin{frame}{what is a glacier model?}
 
@@
 \begin{definition}
-an \alert{ice sheet model} is a map which simulates an ice sheet in a climate
+a \alert{glacier} or \alert{ice sheet model} is a \underline{map}
+
+which evolves a glacier in a climate
 \end{definition} 
@@
 
-\begin{frame}{basic ice sheet model: notation}
+\begin{frame}{basic model: notation}
 
@@
     \end{itemize}
-\item<3> $s(t,x,y)$ determines the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$, on which the solution velocity $\bu(t,x,y,z)$ is defined:
+\item<3-> $s(t,x,y)$ determines the \alert{icy domain} $\Lambda(t) \subset \RR^3$:
     $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$
+\item<4> the solution velocity $\bu(t,x,y,z)$ is defined on $\Lambda(t)$
 
@@
 
-\begin{frame}{basic ice sheet model: conservation}
+\begin{frame}{basic model: conservation}
 
@@
     \item[$\circ$] momentum
-    \item[$\circ$] \st{energy} \hfill $\leftarrow$ \emph{ignored for simplicity in this talk}
+    \item[$\circ$] energy \hfill $\leftarrow$ \emph{ignored for simplicity in this talk}
     \end{itemize}
@@
 \begin{align*}
-\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
+\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t) \\
     \intertext{\item[$\circ$] and on the ice surfaces:}
-\text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
-\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
+\text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t) \\
+\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t)
 \end{align*}
@@
 
-\begin{frame}{basic ice sheet model: strong form = NCP coupled to Stokes}
+\begin{frame}{basic model: strong form = NCP coupled to Stokes}
 
@@
 
-\begin{frame}{basic ice sheet model: is a DAE system}
+\begin{frame}{the basic model is a DAE system}
 
@@
 
-\begin{frame}{basic ice sheet model: current research and thinking}
+\begin{frame}{basic model: current research}
 
@@
     \begin{itemize}
-    \item[$\circ$] when pressed, most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model
-    \item[$\circ$] well-posedness of the lubrication approximation of the model has been considered; existence proved in (Jouvet \& Bueler 2012)
+    \item[$\circ$] though most researchers would agree NCP-coupled-to-Stokes is indeed the intended model \dots
+    \end{itemize}
+\item progress has been made on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
+    \begin{itemize}
+    \item[$\circ$] 1D well-posedness on flat bed (Calvo et al 2003)
+    \item[$\circ$] 2D steady-state existence on general beds (Jouvet \& Bueler 2012)
+    \item[$\circ$] 2D well-posedness on flat bed (Piersanti \& Temam 2022)
+    \end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{basic model: current \emph{thinking}}
+
+\begin{itemize}
+\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
+    \begin{itemize}
+    \item[$\circ$] most researchers would agree NCP-coupled-to-Stokes is indeed the intended model!
+    \item[$\circ$] progress on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
+        \begin{itemize}
+        \item 1D well-posedness on flat bed (Calvo et al 2003)
+        \item 2D steady-state existence on general beds (Jouvet \& Bueler 2012)
+        \item 2D well-posedness on flat bed (Piersanti \& Temam 2022)
+        \end{itemize}
     \end{itemize}

### 2023-05-28 talk-iu/slides.tex
a few edits and simplifications

@@
 \hypersetup{
-    pdftitle = {Making ice sheet models scale properly},
+    pdftitle = {Glacial flows, simulated faster},
     pdfauthor = {Ed Bueler},
@@
 
-\section{the basics of glaciers and ice sheets}
+\section{introduction to glaciers and ice sheets}
 
@@
 \end{frame}
+
+\AtBeginSection[]
+{
+  \begin{frame}{Outline}
+    \tableofcontents[currentsection]
+  \end{frame}
+}
+
+\section{the basic glacier model}
 
@@
 \begin{itemize}
-\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
-    \begin{itemize}
-    \item[$\circ$] most researchers would agree NCP-coupled-to-Stokes is indeed the intended model!
-    \item[$\circ$] progress on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
-        \begin{itemize}
-        \item 1D well-posedness on flat bed (Calvo et al 2003)
-        \item 2D steady-state existence on general beds (Jouvet \& Bueler 2012)
-        \item 2D well-posedness on flat bed (Piersanti \& Temam 2022)
-        \end{itemize}
-    \end{itemize}
-
-\medskip
 \item numerical modelers tend to think of the Stokes problem separately from surface evolution
@@
     \item[$\circ$] because it is too slow
-    \item[$\circ$] \alert{can we make it fast enough to use?}
+    \item<2>[$\circ$] \alert{can we make it fast enough to use?}
     \end{itemize}
@@
 
-\AtBeginSection[]
-{
-  \begin{frame}{Outline}
-    \tableofcontents[currentsection]
-  \end{frame}
-}
-
-\section{time-stepping}
+\section{time-stepping and solver scaling}
 
@@
 \end{frame}
-
-\section{stress-balance solver scaling}
 
@@
 \item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs)
-\item but in the \alert{non-local residual case} \hfill $\gets$ \emph{yesterday's seminar}
+\item but in the non-local residual case this seems not to exist
     \begin{itemize}
-    \item[$\circ$] this seems not to exist
     \item[$\circ$] the \alert{smoother} must reduce a residual formed from surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ (from a scalable Stokes solver)

### 2023-05-29 talk-iu/slides.tex
aligned better to my goals: no deep learning or Ahlkrona trick ...

@@
 \begin{itemize}
-\item for simplicity/clarity of the upcoming modeling, I will ignore much of glacier physics
+\item for simplicity/clarity of the upcoming modeling, I will ignore parts of glacier physics in the remainder of my talk
+
+\medskip
 \item \alert{ignoring}:
@@
     \item[$\circ$] ice temperature
-    \item[$\circ$] fracture processes (calving, crevasses)
+    \item[$\circ$] fracture processes (e.g.~calving)
     \item[$\circ$] solid earth deformation
@@
 \medskip
-\item {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
+\item {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, which includes these and other processes}
 \end{itemize}
@@
     \begin{itemize}
-    \item[$\circ$] $\bu|_s=0$ where no ice
+    \item[$\circ$] note: $\bu|_s=0$ where no ice
     \item<2->[$\circ$] viscosity by Glen law: \, $2\nu(D\bu) = \Gamma |D\bu|^{\pp-2}$
@@
     \begin{itemize}
-    \item<2>[$\circ$] but in $\infty$ dimensions (PDAE?), and subject to complementarity
+    \item<2>[$\circ$] but in $\infty$ dimensions \dots and actually subject to complementarity
     \end{itemize}
@@
     \end{itemize}
-\item progress has been made on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
+
+\medskip
+\item<2> progress has been made on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
     \begin{itemize}
@@
 \begin{itemize}
-\item numerical modelers tend to think of the Stokes problem separately from surface evolution
+\item numerical glacier and ice sheet modelers tend to think of the Stokes problem separately from surface evolution
     \begin{itemize}
@@
 \medskip
-\item ice sheet geometry evolution is addressed with minimal awareness of complementarity
+\item \dots and ice sheet geometry evolution is addressed with minimal awareness of complementarity
 
 \medskip
-\item NCP-coupled-to-Stokes is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
+\item the NCP-coupled-to-Stokes basic model is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
     \begin{itemize}
     \item[$\circ$] because it is too slow
-    \item<2>[$\circ$] \alert{can we make it fast enough to use?}
+    \item[$\circ$] \alert{can we make it fast enough to use?}
     \end{itemize}
@@
 
-\section{time-stepping and solver scaling}
+\section{numerics: time-stepping and solver scaling}
 
@@
   $$\bU \sim - \grad s \sim - \grad H$$
-\item<2> NCP-coupled-to-Stokes DAE system has no characteristic curves
+\item<2> observation: the NCP-coupled-to-Stokes system has no characteristic curves
 \end{itemize}
@@
 \begin{itemize}
-\item the diffusion schema is literal in the lubrication approximation
+\item the diffusion schema is literal in the shallow ice approximation
     \begin{itemize}
@@
     \end{itemize}
-\item but the fact that ice flows downhill has \emph{time-stepping stability} consequences
+\item the fact that ice flows downhill has \emph{time-stepping stability} consequences for the basic model
     \begin{itemize}
-    \item[$\circ$] regardless of your preference for the advective schema!
+    \item[$\circ$] regardless of your above schema preference!
     \end{itemize}
@@
 \begin{itemize}
-\item the simplest of several shallow approximations is the ``lubrication'' approximation, the \alert{shallow ice approximation} (SIA)
+\item as mentioned, the simplest of glacier flow approximations is the ``lubrication'' approximation: \alert{shallow ice approximation} (SIA)
 \item SIA version of the NCP:
@@
     \item[$\circ$] \emph{local} in surface and bed topography
-    \item[$\circ$] existence is known for this NCP problem (Jouvet \& Bueler, 2012), when written as a \alert{variational inequality} weak form
+    \item[$\circ$] well-posedness for this problem as weak form \alert{variational inequality} (Calvo et al 2002, Jouvet \& Bueler 2012, Piersanti \& Temam 2022)
     \end{itemize}
@@
 
-\begin{frame}{time-stepping in current and future ice sheet models}
+\begin{frame}{time-stepping in current ice sheet models}
 
@@
     \end{itemize}
-\item most researchers believe the advection schema
+\item yet most researchers believe the advection schema
     \begin{itemize}
@@
     \end{itemize}
+\end{itemize}
+\end{frame}
 
-\medskip
+\begin{frame}{time-stepping in \emph{future} ice sheet models}
+
+\begin{itemize}
 \item \alert{implicit time-stepping} is appropriate for DAE problems
@@
 
-\section{3 approaches to better performance}
+\section{a multilevel approach to better performance}
 
-\begin{frame}{approach 1: machine learning}
+\begin{frame}{multilevel NCP-coupled-to-Stokes solvers}
 
 \begin{itemize}
-\item apply \alert{machine learning}
-\item run non-scalable ice sheet models on many hypothetical/real ice sheets, and train ML \alert{emulator} on results
-    \begin{itemize}
-    \item[$\circ$] supervised learning of physically-based model results
-    \item[$\circ$] compute map on CPUs, then learn \& evaluate map on GPUs
-    \item[$\circ$] convincing demo in Jouvet et al.~(2021)
-    \end{itemize}
-\end{itemize}
-
-\vspace{3mm}
-\begin{minipage}[t][60cm][t]{\textwidth}
-\begin{center}
-\only<1>{\includegraphics[height=0.52\textheight]{../images/tiles-jouvet.png}}
-\only<2>{\includegraphics[height=0.57\textheight]{../images/cnn-jouvet.png}}
-\end{center}
-\end{minipage}
-\end{frame}
-
-\begin{frame}{approach 2: semi-coupled time-stepping}
-
-\begin{itemize}
-\item idea from L{\"o}fgren et al.~(2022)
-    \begin{itemize}
-    \item[$\circ$] earlier use in mantle/crust simulations (Kaus et al.~2010)
-    \end{itemize}
-\item \emph{idea.} remain explicit, but \alert{modify the Stokes problem to ``see'' the updated (extrapolated) surface}
-\item that is, modify the Stokes problem at time $t^\ell$ \onslide<2>{by adding body force terms corresponding to the updated-surface icy domain}
-
-\vspace{-3mm}
-\begin{minipage}[t][30mm][t]{0.8\textwidth}
-\only<1>{
-{\small
-\begin{align*}
-\int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv &= \int_{\Lambda^\ell} \bbf\cdot \bv \\
-\phantom{x} \\
-\phantom{x} \\
-\int_{\Lambda^\ell} q \Div \bu &= 0
-\end{align*}
-}
-}
-\only<2>{
-{\small
-\begin{align*}
-\int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv \qquad\qquad &= \int_{\Lambda^\ell} \bbf\cdot \bv + \Delta t \int_{\Gamma_s^\ell} a\, (\bbf\cdot \bv)\,dx\\
-\hspace{20mm} - \Delta t \int_{\Gamma_s^\ell} (\bu \cdot \bn_{s^\ell}) (\bbf\cdot \bv)\,dx \qquad
-&\phantom{=} \\
-\int_{\Lambda^\ell} q \Div \bu &= 0
-\end{align*}
-}
-}
-\end{minipage}
-
-\bigskip
-\item<2> early experiments suggest $\sim 10$ times longer stable time steps
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}{approach 3: multilevel NCP-coupled-to-Stokes solvers}
-
-\begin{itemize}
+\item FIXME
 \item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs)
@@
    \begin{itemize}
-   \item[$\circ$] progress is being made in all of these areas
+   \item[$\circ$] progress is being made in all of these areas, e.g.~ scalable Stokes solvers (Isaac et al.~2015)
    \end{itemize}
-\item<3-> \alert{coming soon} from current research:
-   \begin{enumerate}
-   \item[1.] machine learning emulators (Jouvet et al.~2021)
-   \item[2.] semi-coupled time stepping (L{\"o}fgren et al.~2022)
-   \item[3.] scalable Stokes solvers (Isaac et al.~2015)   
-   \end{enumerate}
-\item<4-> scalable solvers for implicit-step NCP-coupled-to-Stokes models, which would seem to be the recommended numerical design, require \alert{multilevel solvers for non-local variational inequalities}
+\item<3> scalable solvers for implicit-step NCP-coupled-to-Stokes models, which would seem to be the recommended numerical design, require \alert{multilevel solvers for non-local variational inequalities}
 \end{itemize}
@@
 \item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
+\item E.~Bueler \& P.~Farrell (in preparation). \emph{A full approximation storage multilevel method for nonlinear variational inequalities}
+% Calvo et al 2002
 \item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015).
 \emph{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics}, SIAM J.~Sci.~Comput.~37 (6), B804--B833, \href{https://doi.org/10.1137/140974407}{10.1137/140974407}
-\item G.~Jouvet \& E.
-Bueler (2012).
-\emph{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation}, SIAM J.~Appl.~Math.
-72 (4), 1292--1314, \href{https://doi.org/10.1137/110856654}{10.1137/110856654}
-\item G.~Jouvet, G.~Cordonnier, B.~Kim, M.~L\"uthi, A.~Vieli, A.~Aschwanden (2021).
-\emph{Deep learning speeds up ice flow modelling by several orders of magnitude}, J.~Glaciol.~\href{https://doi.org/10.1017/jog.2021.120}{10.1017/jog.2021.120}
-\item A.~L{\"o}fgren, J.
-Ahlkrona, \& C.
-Helanow (2022).
-\emph{Increasing stable time-step sizes of the free-surface problem arising in ice-sheet simulations}, J.~Comput.~Phys.~X 16, \href{https://doi.org/10.1016/j.jcpx.2022.100114}{10.1016/j.jcpx.2022.100114}
+\item G.~Jouvet \& E.~Bueler (2012).
+\emph{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation}, SIAM J.~Appl.~Math.~72 (4), 1292--1314, \href{https://doi.org/10.1137/110856654}{10.1137/110856654}
+\item P.~Piersanti \& R.~Temam (2022). \emph{On the dynamics of grounded shallow ice
+sheets: modeling and analysis}, Adv.~Nonlinear Analysis 12, \href{https://doi.org/10.1515/anona-2022-0280}{10.1515/anona-2022-0280}
+% Shapero et al ?
+% Wirbel & Jarosch
 \end{itemize}

### 2023-06-05 talk-iu/slides.tex
good edits ... it is alreaady a goot talk

@@
 \begin{column}{0.42\textwidth}
-\hfill \includegraphics[width=0.9\textwidth]{../images/map-glacier-ice.png}
+\hfill \includegraphics[width=0.9\textwidth]{../images/map-velocity.png}
 \end{column}
@@
 
-\begin{frame}{glaciers}
+\begin{frame}{pictures of glaciers}
 
@@
 \begin{itemize}
-\item \emph{def.} an \alert{ice sheet} is a large glacier with small thickness$/$width
+\item \emph{def.} \, \alert{ice sheet} $=$ a large glacier with small thickness$/$width ratio
 \end{itemize}
@@
 \only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum} \hfill {\tiny (Seguinot et al 2018)}}
-\includegraphics<4>[height=0.69\textheight]{../images/moraines-indiana.jpg}
-\only<4>{\par {\scriptsize moraines in Illinois, Indiana, Ohio} \hfill {\tiny (Larsen 1986 and other sources)}}
-\includegraphics<5>[height=0.69\textheight]{../images/laurentide-margold2018.png}
-\only<5>{\par {\scriptsize Laurentide ice sheet} \hfill {\tiny (Margold, Stokes, Clark 2018)}}
+\includegraphics<4>[height=0.69\textheight]{../images/laurentide-margold2018.png}
+\only<4>{\par {\scriptsize Laurentide ice sheet, $\approx$ 22,000 years ago} \hfill {\tiny (Margold, Stokes, Clark 2018)}}
+\includegraphics<5>[height=0.69\textheight]{../images/moraines-indiana.jpg}
+\only<5>{\par {\scriptsize moraines in Illinois, Indiana, Ohio} \hfill {\tiny (Larsen 1986 and other sources)}}
 \end{center}
@@
 
-\begin{frame}{an ice sheet is \emph{not} sea ice!}
+\begin{frame}{finally, an ice sheet is \emph{not} sea ice!}
 \begin{center}
@@
 
+\AtBeginSection[]
+{
+  \begin{frame}{Outline}
+    \tableofcontents[currentsection]
+  \end{frame}
+}
+
+\section{the basic mathematical model for glaciers}
+
 \basicglaciers{\dots again}
 
-\begin{frame}{simplifications}
+\begin{frame}{modeling simplifications}
 
-\begin{columns}
-\begin{column}{0.6\textwidth}
 \begin{itemize}
-\item for simplicity/clarity of the upcoming modeling, I will ignore parts of glacier physics in the remainder of my talk
-
-\medskip
-\item \alert{ignoring}:
+\item for simplicity/clarity of the upcoming model, I will \alert{ignore} these aspects of glacier physics in my talk:
     \begin{itemize}
@@
 \medskip
-\item {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, which includes these and other processes}
+\item all are important for doing science!
+\item UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, includes these and other processes
 \end{itemize}
-\end{column}
-\begin{column}{0.42\textwidth}
-\hfill \includegraphics[width=0.9\textwidth]{../images/map-glacier-ice.png}
-\end{column}
-\end{columns}
 \end{frame}
-
-\AtBeginSection[]
-{
-  \begin{frame}{Outline}
-    \tableofcontents[currentsection]
-  \end{frame}
-}
-
-\section{the basic glacier model}
 
@@
 \begin{definition}
-a \alert{glacier} or \alert{ice sheet model} is a \underline{map}
+a \alert{glacier model} is a \underline{map}
 
@@
 $$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
-\text{precipitation minus} \\
+\text{snowfall minus} \\
 \text{melt \& runoff}
@@
 \smallskip
-\item {\small \alert{map}: \, $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$ }
+\item<2> {\small \underline{map}: \, $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$ }
 \end{itemize}
@@
 
-\begin{frame}{basic model: notation}
+\begin{frame}{the basic glacier model: notation}
 
@@
 	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
-\item<2-> solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
+\item<2-> solution \alert{surface elevation} $s(t,x,y)$ is defined on $[0,T]\times \Omega$
     \begin{itemize}
@@
     $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$
-\item<4> the solution velocity $\bu(t,x,y,z)$ is defined on $\Lambda(t)$
+\item<3-> the solution \alert{velocity} $\bu(t,x,y,z)$ is defined on $\Lambda(t)$
 
@@
 
-\begin{frame}{basic model: conservation}
+\begin{frame}{the basic glacier model: conservation}
 
 \begin{itemize}
-\item ice sheet evolution should conserve physical quantities:
+\item glacier evolution is merely physics \dots so it \alert{conserves}
     \begin{itemize}
@@
 \medskip
-\item conservation of mass is important both
+\item<2> conservation of mass happens
     \begin{itemize}
-    \item[$\circ$] in the icy domain $\Lambda(t) \subset \RR^3$:
+    \item[$\circ$] within the icy domain $\Lambda(t) \subset \RR^3$:
 \begin{align*}
 \text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t) \\
-    \intertext{\item[$\circ$] and on the ice surfaces:}
+    \intertext{\item[$\circ$] on the surfaces $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$:}
 \text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t) \\
@@
         \begin{itemize}
-        \item[$\vartriangleright$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
+        \item[$\vartriangleright$] $\Gamma_s(t)$ is upper surface of the ice
+        \item[$\vartriangleright$] $\Gamma_b(t)$ is base of the ice
         \item[$\vartriangleright$] $\bn_s = \left<-\grad s,1\right>$ is upward surface normal
@@
 
-\begin{frame}{free boundary problem}
+\begin{frame}{the free boundary problem for a fluid layer}
 
 \begin{itemize}
-\item ice sheet evolution is a \alert{free-boundary} problem for conserved quantities
+\item glacier evolution is a \alert{free-boundary} problem
 \item specifically, the surface kinematic equation (SKE)
@@
 \bigskip
-\item {\footnotesize for more on this perspective see Bueler (2021)}
+\item {\footnotesize for more on this perspective see Bueler (2021), \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math}
 \end{itemize}
@@
 
-\begin{frame}{basic model: strong form = NCP coupled to Stokes}
+\begin{frame}{the basic glacier model strong form: \, NCP coupled to Stokes}
 
@@
     \item[$\circ$] note: $\bu|_s=0$ where no ice
-    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu(D\bu) = \Gamma |D\bu|^{\pp-2}$
+    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu(D\bu) = \Gamma |D\bu|^{\pp-2}$, \,$p \approx 4$
     \end{itemize}
@@
 
-\begin{frame}{the basic model is a DAE system}
+\begin{frame}{the basic glacier model is a DAE system}
 
 \begin{itemize}
-\item for this slide, forget complementarity and boundary conditions
-\item result: SKE coupled to Stokes
+\item for this slide, forget complementarity and boundary conditions to get simplified model ``SKE coupled to Stokes'':
 \begin{align*}
@@
     \begin{itemize}
-    \item[$\circ$] because ice is very viscous and incompressible
+    \item[$\circ$] recall: ice is very viscous and incompressible
     \end{itemize}
@@
     \begin{itemize}
-    \item<2>[$\circ$] but in $\infty$ dimensions \dots and actually subject to complementarity
+    \item<2>[$\circ$] in $\infty$ dimensions, of course,
+    \item<2> and also subject to complementarity
     \end{itemize}
@@
 
-\begin{frame}{basic model: current research}
+\begin{frame}{the basic glacier model: current research}
 
@@
     \begin{itemize}
-    \item[$\circ$] though most researchers would agree NCP-coupled-to-Stokes is indeed the intended model \dots
+    \item[$\circ$] though most researchers would agree NCP-coupled-to-Stokes is indeed the intended model!
     \end{itemize}
@@
 
-\begin{frame}{basic model: current \emph{thinking}}
+\begin{frame}{the basic glacier model: current numerical \emph{thinking}}
 
@@
 \medskip
-\item \dots and ice sheet geometry evolution is addressed with minimal awareness of complementarity
+\item \dots and ice sheet geometry evolution is often addressed with minimal awareness of complementarity
 
 \medskip
-\item the NCP-coupled-to-Stokes basic model is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
+\item<2> the NCP-coupled-to-Stokes basic model is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
     \begin{itemize}
     \item[$\circ$] because it is too slow
-    \item[$\circ$] \alert{can we make it fast enough to use?}
+    \item[$\circ$] \alert{can we make it fast enough to use?} \hfill \begin{minipage}{0.3\textwidth}{\footnotesize \color{FireBrick}{$\leftarrow$ \emph{what I am working on}}} \end{minipage}
     \end{itemize}
@@
 
-\section{numerics: time-stepping and solver scaling}
+\section{numerics: time-stepping}
 
-\begin{frame}{ice sheet models: the mass-continuity equation view}
+\begin{frame}{the mass-continuity equation view}
 
 \begin{itemize}
-\item thickness transport form helps for evolution or stability questions
+\item ``thickness transport form'' helps for evolution or stability questions
 \item define:
@@
     \end{itemize}
-\item the \alert{mass continuity equation} for thickness follows from SKE and incompressibility:
+\item the \alert{mass continuity equation} for thickness, an apparent \alert{advection equation}, follows from the SKE and incompressibility:
    $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$
@@
   $$\bU \sim - \grad s \sim - \grad H$$
-\item<2> observation: the NCP-coupled-to-Stokes system has no characteristic curves
+\item<2> in any case, the NCP-coupled-to-Stokes system \emph{has no characteristic curves}
 \end{itemize}
@@
 \begin{itemize}
+\item both forms are nonlinear: \, $\bU=\bU(H,\grad s), \, D=D(H,\grad s)$
+\item the glacier modeling literature is confusing!
 \item the diffusion schema is literal in the shallow ice approximation
@@
     \end{itemize}
-\item the fact that ice flows downhill has \emph{time-stepping stability} consequences for the basic model
-    \begin{itemize}
-    \item[$\circ$] regardless of your above schema preference!
+\item regardless of your schema preference, the fact that ice flows downhill has \emph{time-stepping stability} consequences!
+\item \dots so let us recall some traditional numerical analysis
 \end{itemize}
-\item note both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
-\end{itemize}
-\end{frame}
-
-\begin{frame}{shallow ice approximation NCP}
-
-\begin{itemize}
-\item as mentioned, the simplest of glacier flow approximations is the ``lubrication'' approximation: \alert{shallow ice approximation} (SIA)
-\item SIA version of the NCP:
-{\small
-  $$s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0$$
-}
-the surface motion contribution ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$ has a formula:
-  $${\color{FireBrick} \Phi(s)} = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)$$
-
-\vspace{-2mm}
-    \begin{itemize}
-    \item[$\circ$] constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation
-    \end{itemize}
-
-\medskip
-\item $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
-    \begin{itemize}
-    \item[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
-    \item[$\circ$] \emph{local} in surface and bed topography
-    \item[$\circ$] well-posedness for this problem as weak form \alert{variational inequality} (Calvo et al 2002, Jouvet \& Bueler 2012, Piersanti \& Temam 2022)
-    \end{itemize}
-\end{itemize}
-\end{frame}
-
-\begin{frame}{nonlocality}
-
-\begin{itemize}
-\item from now on, let us avoid shallowness approximations
-\item then the basic ice sheet model (NCP coupled to Stokes) problem has a \alert{non-local} surface velocity function ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$
-{\small
-\begin{equation*}
-s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0
-\end{equation*}
-}
-\item \emph{figure:} the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the SIA velocity responds only underneath the surface perturbation}
-\end{itemize}
-
-\begin{center}
-\includegraphics<1>[width=0.6\textwidth]{../images/stokes-greens-arndt.png}
-\includegraphics<2>[width=0.6\textwidth]{../images/sia-greens-arndt.png}
-\end{center}
 \end{frame}
@@
 \only<1>{
-\item let us recall some traditional numerical analysis
-}
-\only<2>{
 \item \alert{explicit} time stepping is common for \alert{advections}
@@
 }
-\only<3>{
+\only<2>{
 \item \alert{explicit} time stepping for \alert{diffusions} is best avoided
@@
 }
-\only<4>{
+\only<3>{
 \item \alert{implicit} time stepping for \alert{diffusions} is often recommended
@@
 \begin{itemize}
-\item current-technology large-scale models use explicit time stepping
+\item current-technology, large-scale numerical models, including \href{https://www.pism.io/}{PISM}, use explicit time stepping
     \begin{itemize}
     \item[$\circ$] this is embarrassing: the mathematical problem is a DAE
-    \item[$\circ$] the accuracy/performance/usability consequences of the suppressed free-boundary/DAE/diffusive character are hard to sweep under the rug
     \end{itemize}
-\item yet most researchers believe the advection schema
+\item many researchers ``believe'' the advection schema
     \begin{itemize}
-    \item[$\circ$] time step is determined by CFL using coupled solution velocity $\bU$
+    \item[$\circ$] time step is supposed to be determined by CFL using the coupled solution velocity $\bU$
     \end{itemize}
+\item the accuracy/performance/usability consequences of the suppressed DAE/diffusive character are hard to sweep under the rug
+\item the whole situation is a cry for mathematical clarity!
 \end{itemize}
@@
 \item \alert{implicit time-stepping} is appropriate for DAE problems
-\item a sequence of NCP-coupled-to-Stokes free-boundary problems must be solved at each time step
+\item future models will solve a sequence of NCP-coupled-to-Stokes free-boundary problems at each time step
 \end{itemize}
+\end{frame}
+
+\section{numerics: Stokes models}
+
+\begin{frame}{shallow ice approximation \quad $\leftarrow$ \emph{not Stokes}}
+
+\begin{itemize}
+\item the simplest of glacier flow approximations is the ``lubrication'' approximation: \alert{shallow ice approximation} (SIA)
+\item SIA version of the NCP:
+{\small
+  $$s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0$$
+}
+the surface motion contribution ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$ has a formula:
+  $${\color{FireBrick} \Phi(s)} = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)$$
+
+\vspace{-2mm}
+    \begin{itemize}
+    \item[$\circ$] constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation
+    \end{itemize}
+
+\medskip
+\item ${\color{FireBrick} \Phi(s)}$ is a \alert{doubly-nonlinear differential operator}
+    \begin{itemize}
+    \item[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
+    \item[$\circ$] but \emph{local} in surface and bed topography, which Stokes is not
+    \item[$\circ$] well-posedness holds for the weak form $=$ \alert{variational inequality} (Calvo et al 2002, Jouvet \& Bueler 2012, Piersanti \& Temam 2022)
+    \end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{nonlocality}
+
+\begin{itemize}
+\item however, from now on, let us avoid shallowness approximations
+\item the basic glacier model (NCP coupled to Stokes) problem has a \alert{non-local} surface velocity function ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$
+{\small
+\begin{equation*}
+s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0
+\end{equation*}
+}
+\item the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the SIA velocity responds only underneath the surface perturbation}
+\end{itemize}
+
+\begin{center}
+\includegraphics<1>[width=0.6\textwidth]{../images/stokes-greens-arndt.png}
+\includegraphics<2>[width=0.6\textwidth]{../images/sia-greens-arndt.png}
+\end{center}
 \end{frame}
@@
 \item this is the \alert{stress balance} (conservation of momentum) problem which determines velocity $\bu$ and pressure $p$
-\item how fast is the numerical solution process?
+\item<2> how fast is the numerical solution process?
     \begin{itemize}
@@
 
-\begin{frame}{summary: PDE solver algorithmic scaling}
+\begin{frame}{one slide course: numerical PDE solver algorithmic scaling}
 
 \begin{itemize}
-\item for example, consider the 2D Poisson equation:
+\item consider the Poisson equation:
     $$-\grad^2 u = f \,\text{ in $\Omega$}, \qquad u=0 \,\text{ on $\partial\Omega$}$$
 \item discretization generates a linear system $A\,\bu = \bb$ with $\bu\in\RR^m$
-\item data size $m$ is the number of unknowns
+\item the number of unknowns is the data size $m$:
     \begin{itemize}
-    \item[$\circ$] for low-order discretizations, $m$ = \#(nodes in the grid)
-    \item[$\circ$] $m$ scales with mesh cell diameter $h$: \quad $m \sim h^{-2}$ in 2D
+    \item[$\circ$] $m$ = \#(nodes in the mesh)
+    \item[$\circ$] $m$ scales with mesh cell diameter: \quad $m \sim h^{-2}$ \quad in 2D
     \end{itemize}
@@
     \item[$\circ$] $\approx O(m^2)$ for sparsity-exploiting direct linear algebra
-    \item[$\circ$] $O(m^1)$, \alert{optimal}, e.g.~for \alert{multigrid} solvers (below)
+    \item[$\circ$] $O(m^1)$, \alert{optimal}, for \alert{multigrid} solvers
     \end{itemize}
@@
 
-\begin{frame}{ice sheet stress-balance solver complexity}
+\begin{frame}{ice sheet models: stress-balance solver complexity}
 
@@
     \end{itemize}
+\item \emph{but} this is for Stokes solvers \alert{de-coupled} from the surface-evolution NCP
 \end{itemize}
@@
 
-\section{ice sheet model performance analysis}
+\section{numerics: comparative performance analysis}
 
-\begin{frame}{the analysis set-up}
+\begin{frame}{ice sheet models: the analysis set-up}
 
@@
     \item[$\circ$] e.g.~Issac et al (2015) Stokes solver
-    \item[$\circ$] \emph{simple message}: I am ignoring refinement in the vertical
+    \item[$\circ$] I am ignoring refinement in the vertical
     \end{itemize}
@@
 \medskip
-\item let $m$ = \#(surface elevation \& velocity \& pressure unknowns)
-\item for map-plane domain $\Omega \subset \RR^2$ of width $L$ and cells of diameter $h$:
+\item data size: \quad {\small $m$ = \#(surface elevation \& velocity \& pressure unknowns)}
+\item assume domain $\Omega \subset \RR^2$ with width $L$ and cell diameter $h$:
   $$m \sim \frac{L^2}{h^2} \hspace{20mm}$$
@@
 \vspace{-2mm}
-\item explicit time-stepping schemata:
+\item recall explicit time-stepping stability:
 {\footnotesize
 \begin{align*}
-\text{advective} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a & \Delta t &\le \frac{h}{U} \\
-\text{diffusion} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a & \Delta t &\le \frac{h^2}{D}
+\text{advective} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a & &\implies & \Delta t &\le \frac{h}{U} \\
+\text{diffusion} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a & &\implies & \Delta t &\le \frac{h^2}{D}
 \end{align*}
@@
 
-\item stress-balance solver scaling parameterized as $O(m^{1+\alpha})$
+\item recall stress-balance solver complexity: \quad $O(m^{1+\alpha})$
 \end{itemize}
@@
 
-\begin{frame}{the simplified ice sheet model performance question}
+\begin{frame}{ice sheet models: the performance question}
 
@@
 \bigskip
-\item the goal: \hspace{20mm} $O(h^{-2}) = O(m^1)$
+\item the goal is optimality: \hspace{20mm} $\text{flops} \sim O(h^{-2}) = O(m^1)$
 \end{itemize}

### 2023-06-05 talk-iu/slides.tex
edits to later slides ... I think only a couple of slides on multilevel VI method are needed

@@
     \begin{itemize}
-    \item[$\circ$] 1D well-posedness on flat bed (Calvo et al 2003)
+    \item[$\circ$] 1D well-posedness on flat bed (Calvo et al 2002)
     \item[$\circ$] 2D steady-state existence on general beds (Jouvet \& Bueler 2012)
@@
 
-\begin{frame}{explicit ice sheet model performance}
+\begin{frame}{ice sheet models: explicit time-stepping performance}
 
@@
 \begin{itemize}
-\item explicit time-stepping implies \alert{many stress-balance solves}
+\item we \emph{want} optimality: \quad $O(m^1)$ flops per model year
+\item explicit time-stepping implies \alert{too many stress-balance solves}
     \begin{itemize}
-    \item[$\circ$] while stress-balance scaling exponent $\alpha$ is important, even optimality ($\alpha=0$) cannot rescue performance
+    \item[$\circ$] while the Stokes (stress-balance) scaling exponent $\alpha$ is important, even Stokes solver optimality ($\alpha=0$) cannot yield optimality
     \end{itemize}
@@
 \begin{itemize}
-\item switch to \alert{implicit time-stepping} for unconditional stability?
-    \begin{itemize}
-    \item[$\circ$] each step is a \alert{free-boundary} NCP-coupled-to-Stokes problem
-    \item[$\circ$] parameterize cost of these solves as $O(m^{1+\beta})$
-    \end{itemize}
-\item need $q$ model updates per year to integrate climate influences
+\item let us try \alert{implicit time-stepping}, for its unconditional stability
+\item each step is now a \alert{free-boundary NCP-coupled-to-Stokes problem}
+\item let us parameterize cost of these solves as $O(m^{1+\beta})$
+\item we still need $q$ model updates per year to integrate climate influences, and track evolution for the simulation purpose
 \end{itemize}
@@
 \begin{itemize}
-\item goal for optimists: implicit time-stepping \emph{and} build a $\beta \approx 0$ NCP-coupled-to-Stokes solver for the time step problems
+\item new goal: use implicit time-stepping \emph{and}~build a $\beta \approx 0$ NCP-coupled-to-Stokes solver for problem at each time step
 \end{itemize}
@@
     \begin{itemize}
-    \item[$\circ$] Wirbel \& Jarosch~(2020) is an important attempt \dots
+    \item[$\circ$] however, Wirbel \& Jarosch~(2020) is an important beginning
     \end{itemize}
-\item the Bueler (2016) implicit (free-boundary) SIA solver scales badly: $\beta=0.8$
+\item the Bueler (2016) implicit and NCP SIA solver scales badly: $\beta=0.8$
 \end{itemize}
@@
 
-\section{a multilevel approach to better performance}
+\section{a multilevel approach}
 
@@
 \begin{itemize}
-\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
+\item glacier simulations are both \alert{important to humanity} and a rich \alert{source of interesting mathematics}
+   \begin{itemize}
+   \item[$\circ$] predict sea level rise!
+   \end{itemize}
+\item<2-> ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
    \begin{itemize}
@@
    \end{itemize}
-\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
+\item<3-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
    \begin{itemize}
-   \item[$\circ$] progress is being made in all of these areas, e.g.~ scalable Stokes solvers (Isaac et al.~2015)
+   \item[$\circ$] progress is being made in all of these areas, e.g.~scalable Stokes solvers (Isaac et al.~2015)
    \end{itemize}
-\item<3> scalable solvers for implicit-step NCP-coupled-to-Stokes models, which would seem to be the recommended numerical design, require \alert{multilevel solvers for non-local variational inequalities}
+\item<4> scalable solvers for implicit-step, NCP-coupled-to-Stokes models require \alert{multilevel solvers for non-local variational inequalities}
+   \begin{itemize}
+   \item[$\circ$] is this the preferred numerical design for the basic glacier model?
+   \end{itemize}
 \end{itemize}
@@
 
-\begin{frame}{references}
+\begin{frame}{references (partial list)}
 
@@
 \item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
-\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
+%\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
 \item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
 \item E.~Bueler \& P.~Farrell (in preparation). \emph{A full approximation storage multilevel method for nonlinear variational inequalities}
-% Calvo et al 2002
+\item N.~Calvo, J.~D\'iaz, J.~Durany, E.~Schiavi, \& C.~V\'azquez (2002).
+\emph{On a doubly-nonlinear parabolic obstacle problem modelling ice sheet dynamics}, SIAM J.~Appl.~Math., 63(2), 683--707 \href{https://doi.org/10.1137/S0036139901385345}{10.1137/S0036139901385345}
 \item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015).
@@
 sheets: modeling and analysis}, Adv.~Nonlinear Analysis 12, \href{https://doi.org/10.1515/anona-2022-0280}{10.1515/anona-2022-0280}
-% Shapero et al ?
-% Wirbel & Jarosch
+\item A.~Wirbel \& A.~Jarosch (2020).
+\emph{Inequality-constrained free-surface evolution in a full Stokes ice flow model (evolve\_glacier v1.1)}, Geosci.~Model Dev.~13 (12), 6425--6445, \href{https://doi.org/10.5194/gmd-13-6425-2020}{10.5194/gmd-13-6425-2020}
 \end{itemize}

### 2023-06-06 talk-iu/slides.tex
I think this is a complete talk ... but I need to practice

@@
 
-%\subtitle{\emph{x}}
+%\subtitle{}
 
@@
 
-%\titlegraphic{\begin{picture}(0,0)
-%    \put(0,180){\makebox(0,0)[rt]{\includegraphics[width=4cm]{figs/software.png}}}
-%  \end{picture}
-%}
-
 \titlegraphic{\hfill \includegraphics[width=0.15\textwidth]{../images/uafbw.png}}
-
-%% to start section counter at 0 see
-%% https://tex.stackexchange.com/questions/170222/change-the-numbering-in-beamers-table-of-content
 
@@
 \beamertemplatenavigationsymbolsempty
-
-%\begin{frame}
-%  \maketitle
-%\end{frame}
 
@@
 \begin{itemize}
-\item FIXME
-\item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs)
-\item but in the non-local residual case this seems not to exist
+\item direct attack on the NCP-coupled-to-Stokes problem, to get an optimal ($\beta=0$) solver, seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs)
+\item but in the \alert{non-local residual} case
     \begin{itemize}
-    \item[$\circ$] the \alert{smoother} must reduce a residual formed from surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ (from a scalable Stokes solver)
+    \item[$\circ$] application of the smoother needs to reduce the NCP residual from the surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$, where $\bu|_s$ is evaluated from a scalable Stokes solver
     \end{itemize}
-\item near-optimal multilevel solvers exist for simpler VI problems
+\item this seems not to exist, but we are starting to work on it \dots
 \end{itemize}
+\end{frame}
+
+\begin{frame}{a \emph{new} multilevel SIA solver (joint with P.~Farrell)}
+
+\begin{block}{FASCD}
+\emph{full approximation storage constraint decomposition}
+
+a multilevel method for box constrained NCPs and VIs
+\end{block}
+
+\begin{itemize}
+\item in preparation, but here are fresh preliminary results \dots
+\end{itemize}
+
+\medskip
+\begin{center}
+\includegraphics[height=0.55\textheight]{../images/siabumps.png}
+\end{center}
+\end{frame}
+
+\begin{frame}{a \emph{new} multilevel SIA solver (joint with P.~Farrell)}
+
+\begin{columns}
+\begin{column}{0.6\textwidth}
+\begin{itemize}
+\item results below show FASCD F-cycles give optimal ($\beta=0$) performance for the SIA problem
+    \begin{itemize}
+    \item[$\circ$] \emph{iterations} counts number of V-cycles after F-cycle ``ramp''
+    \item[$\circ$] \emph{time} is for 4 core runs on my laptop
+    \end{itemize}
+\end{itemize}
+\end{column}
+\begin{column}{0.4\textwidth}
+\hfill \input{fullcycle.tex}
+\end{column}
+\end{columns}
+
+\bigskip
+\begin{center}
+\begin{tabular}{c|r|c|r}
+\emph{levels} & $N$ & \emph{iterations} & \emph{time} (s) \\ \hline
+\input{siaresults.txt} \hspace{-2.2mm}
+\end{tabular}
+\end{center}
 \end{frame}

### 2023-06-07 talk-iu/slides.tex
I am going with this version

@@
     \end{itemize}
-\item this seems not to exist, but we are starting to work on it \dots
+\item this seems not to exist, but we are making progress \dots
 \end{itemize}
@@
 
-a multilevel method for box constrained NCPs and VIs
+a multilevel method for box-constrained NCPs and VIs
 \end{block}
@@
 \begin{itemize}
-\item results below show FASCD F-cycles give optimal ($\beta=0$) performance for the SIA problem
+\item results below show FASCD F-cycles give optimal ($\beta=0$) performance for the SIA NCP problem
     \begin{itemize}
-    \item[$\circ$] \emph{iterations} counts number of V-cycles after F-cycle ``ramp''
-    \item[$\circ$] \emph{time} is for 4 core runs on my laptop
+    \item[$\circ$] \emph{iterations} $=$ number of V-cycles after F-cycle ``ramp''
+    \item[$\circ$] \emph{time} is for 4-core runs on my laptop
     \end{itemize}
@@
 \begin{tabular}{c|r|c|r}
-\emph{levels} & $N$ & \emph{iterations} & \emph{time} (s) \\ \hline
+\emph{levels} & $m$ & \emph{iterations} & \emph{time} (s) \\ \hline
 \input{siaresults.txt} \hspace{-2.2mm}
